### 2017-05-02 paper.tex
init commit to git

%\documentclass[a4paper,conference]{IEEEtran}
%\documentclass[a4paper,10pt,conference]{ieeeconf}
%\documentclass[usletter, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
%\IEEEoverridecommandlockouts                              % This command is only needed if 
%\overrideIEEEmargins                                      % Needed to meet printer requirements.

\documentclass{llncs}

\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage[ruled,lined]{algorithm2e}
\usepackage{multirow}
\usepackage[bookmarks=false]{hyperref}
\pdfminorversion=4

\newcommand{\red}[1]{\textcolor{red}{#1}}

\hypersetup{colorlinks=true,linkcolor=black,citecolor=black}

\SetKwInput{KwData}{Global params.}

\title{
    Analysis of protein tunnel traversability for flexible ligands
}

\author{Vojt\v ech Von\' asek\inst{1} \and Barbora Kozl\'\i kov\'a\inst{2} \and Adam Jur\v{c}\'\i k\inst{2} \and Martin Saska\inst{1}}
\institute{
Faculty of Electrical Engineering,  
Czech Technical University in Prague, 
Technick\'a 2, 166 27, Prague 6, Czech Republic
\email{vonasek@labe.felk.cvut.cz}
\and
Faculty of Informatics,  
Masaryk University, 
Botanick\'a 68a, 602 00 Brno,
}

%This work was supported by Grant Agency of the Czech Technical University in Prague, grant No. SGS15/157/OHK3/2T/13.
%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
%vonasek@labe.felk.cvut.cz.

\def\qrand{q_{rand}}
\def\qstart{q_{start}}
\def\qinit{\qstart}
\def\qnear{q_{near}}
\def\qnew{q_{new}}
\def\C{\mathcal{C}}
\def\T{\mathcal{T}}
\def\CF{\mathcal{C}_{free}}
\def\Imax{I_{max}} %max number of iterations of RRT-based planners

\def\dist{\mathrm{dist}}
\def\dists{\mathrm{dist}_{\mathrm{s}}}

\SetKw{return}{return}

\def\VV{V_{vor}}
\def\VVA{V_{vor}^{act}}
\def\da{d_{high}}
\def\db{d_{low}}

\def\TA{$A_1$}
\def\TB{$A_2$}
\def\TC{$A_3$}

\def\probe{r_{\mathrm{probe}}}
\def\Sprobe{S_{\mathrm{probe}}}

\def\gprobe{r_{\mathrm{out}}}
\def\Sgprobe{S_{\mathrm{out}}}

\def\CG{\mathcal{C}_{goal}}
\def\SB{\mathbf{S}_{blocking}}
\def\SS{\mathbf{S}}

\def\RRTTD{RRT$_{\mathrm{td}}$}
\def\RRTN{RRT$_{\mathrm{n}}$}
\def\ths{d_\mathrm{surf}}

%spacing for algorithm environment. 1.0 mean normal spacing
\def\gb{p_{c}}

\begin{document}

\maketitle
%\frontmatter          % for the preliminaries

\begin{abstract}
Proteins are involved in many biochemical processes.
%Behavior and properties of proteins as well as other bio-macromolecules are influenced by internal void space such as tunnels or cavities.
Behavior of proteins is influenced by internal void space such as tunnels or cavities.
Tunnels are paths leading from an active site inside the protein to its surface.
Knowledge about tunnels and their evolution in time provides an insight into protein properties
(e.g., stability).
Sampling-based motion planning methods like Rapidly Exploring Random Tree (RRT) can be used to find the tunnels by
sampling the corresponding configuration space.
The main advantage of these methods is that they can be easily adapted to consider protein dynamics.
%i.e., they can find tunnels in proteins moving in time.
However, the inner void space of proteins is  very limited, 
    which may decrease ability of RRT to find a solution due to the narrow passage problem.
In this paper, we propose to generate the random samples using Voronoi Diagram of the atoms.
A subset of Voronoi vertices is dynamically maintained to support generation of samples in promising regions inside the protein.

%VD provides initial information about protein void space, which helps to sample the corresponding configuration space.

%While VD of the protein atoms can be easily computed, the RRT is used to cope with the protein dynamics<F12>
%While the VD provides
%Random samples are generated around a subset of vertices of VD

%and modifications to support searching of multiple tunnels in each frame.
%In the comparison to existing literature on sampling using Voronoi diagrams, the proposed technique supports finding of multiple
%tunnels even in dynamic proteins.
%The proposed approach is compared to CAVER 3.0, one of the widely used freely available tools for protein analysis.
%In this paper, we propose a novel sampling technique for the purpose of tunnel detection in protein structures.
%The protein void space can be search using sampling-based
%The classical approach to find tunnels in protein structures is based on Voronoi diagrams (VD), which is suitable
%for tunnel detection in a single snapshot of protein's dynamics.
%To consider protein dynamics, that is represented by a sequence of protein snapshots, correspondences between
%VD in these snapshots need to be found, which is time and memory consuming.
%An alternative approach is to employ motion planning methods to detect the tunnels.
%However, VD-based computation can be time consuming when protein dynamics needs to be considered, as it
%requires to find correspondences between VD in a sequence of frames of protein dynamics.
%The performance of VD-based tunnel detection is decreased in the case of dynamic proteins, which requires
%to find correspondences between VD in frames of molecular dynamics.
%In order to apply the VD-based approach for dynamic proteins,
%Available methods for tunnel detection assume static proteins, which allows us to detect tunnels using Voronoi Diagrams.
%Tunnels in static proteins however  provide only a  rough idea about protein behavior.
%Tunnels in real proteins are however influenced by dynamics of the protein.
%In this paper, we describe tunnel detection using Rapidly Exploring Random Tree (RRT).
%The method builds a single configuration tree describing free space of the protein. 
%A crucial part of this novel approach is the sampling of the internal void space of the protein.
%The nodes of the tree are pruned according to protein dynamics.
\end{abstract}

\section{Introduction}
%{\color{red}Co bys rikala na to dat na 1. stranku nejaky obrazek tunelu?}

Protein structures are essential components of all living organisms.
Proper understanding of their structure and function is important in many fields, including drug design, agriculture, cosmetics, etc.
Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. % which led to their better understanding. 
Such investigation can incorporate computational methods which help to analyze the geometry of protein structure,  its shape, surface area, and inner void space. 
%These parameters then suggest and influence protein behavior, namely its reactivity with other molecules.
%consists of the analysis of protein structure, its shape, surface area, and inner void space. 
%These parameters influence namely the protein reactivity. % which is one of the most often studied behavior of the protein.

%Proteins interact with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.) which can lead to changes in properties of the protein or the interacting molecule.
%Thanks to protein interaction with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.), 
% we are able to change the properties of the protein or the interacting molecule. 
 For example, one task in protein engineering is to change selected properties of a protein, e.g. its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
%This is reached by studying so called tunnels in proteins which can serve as the transportation paths for the ligand from the outside environment to the active site or vice versa.
%For example, typical task in protein engineering is to change selected properties of a protein, e.g. its stability under different outer conditions [11] or activity of the protein towards other molecules [16]. 
This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
ligand from the outside environment to the active site or vice versa. 
The active site is a specific site, usually deeply buried inside the protein, 
    where the chemical reaction between the protein and ligand can undergo. 
The importance of tunnels can be demonstrated on the study where mutations of amino acids located directly around the tunnel substantially improved the structural and kinetic stability of the studied protein, while
surface mutations almost did not contribute to the stabilization~\cite{Koudelakova2013}.

%For example, one task in protein engineering is to change selected properties of a protein, e.g. its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
%This is reached by studying so called tunnels in proteins which can serve as the transportation paths for the ligand from the outside environment to the active site or vice versa.
%The active site is a specific site, usually deeply buried inside the protein, where the chemical reaction between the protein and ligand can undergo.
%Since the importance of the presence of tunnels in proteins was revealed, the researchers started to study them intensively.
%Koudel\'{a}kov\'{a} et al.~\cite{Koudelakova2013} show that mutations of amino acids located directly around the tunnel improved the structural and kinetic stability of the studied protein substantially, while the surface mutations almost did not contribute to protein stabilization.
%The importance of tunnel can be demonstrated on an example, 
% where mutations of amino acids located directly around the tunnel improved the structural 
%and kinetic stability of the studied protein substantially, while surface mutations almost did not contribute to
%the protein stabilization~\cite{Koudelakova2013}.

%On the other hand, surface mutations almost did not contribute to protein stabilization~\cite{Koudelakova2013}.

%The task of tunnel detection is to find a collision-free path leading from an active site inside the protein to its surface.
%As not all paths are biochemically relevant, only paths with a given minimal bottleneck have to be reported (the bottleneck is
%the radius of the smallest sphere that can traverse the whole tunnel)

Generally, the task of tunnel detection is to find a collision-free path leading from the active site to the protein's surface. 
The path is usually searched for a spherical probe of a given radius, which serves as a proxy geometry for a ligand. 
An example is depicted in Fig.~\ref{fig::motiv}.
Early solutions focused on static molecules. 
However, from a static snapshot it is hard to assess the biochemical relevance of the tunnel as it does not
provide information about its temporal stability.

Recently, researchers started to focus on molecular dynamics simulations and study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
%The path is searched for a spherical probe of a given radius.
%Early algorithms focused on static molecules, % and the tools were able to detect dozens of putative tunnels.
%but it was hard to assess their biochemical relevance because from a static molecule it is impossible to derive the temporal stability
%of the tunnels.
%Recently, researchers started to focus on molecular dynamics simulations and studying the behavior of individual 
%tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
Molecular dynamics is represented as a sequence of frames (molecule snapshots), and the tunnels need to be detected through these frames.
%With the increasing size of simulations the correctness of predicting the most biochemically relevant tunnel increases as well.
Existing solutions based exclusively on Voronoi diagrams and clustering methods are time and memory consuming
which also gives the limitation for the maximum number of frames which they are able to analyze.
In such cases, the biochemists have to select a subset of the whole simulation and perform the analysis only for this selection.
Such analysis however provides only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted.
%This opens new possibilities for alternative solutions enabling to explore the tunnels in each frame of the molecular dynamics.
%In such cases the biochemists have to skip each n-th snapshot or select a subset of the whole simulation and perform the analysis only for this selection.
%Such analysis however provides only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted. 
%Recently, we have proposed a novel method for tunnel detection using sampling-based motion planning [23]. In this method, RRT is used to ?nd tunnels in each frame and to consider the molecular dynamics, the tree is continuously pruned in each frame. In comparison to existing approaches [17], [18], this approach [23] handles the dynamics without need to cluster and match Voronoi diagrams from consecutive frames.

Recently, we have proposed a novel method for tunnel detection using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT) method~\cite{vonasek2016application}.
RRT builds a tree of collision-free configurations of a spherical probe moving in the protein. 
%In this method, the corresponding configuration space is searched using RRT (Rapidly Exploring Random Tree)~\cite{lavalleRRT}.
To consider the molecular dynamics, the tree is continuously pruned in each frame.
%The proposed solution is based on RRT technique, which constructs a configuration tree separately in each frame of protein dynamics.
In comparison to existing approaches~\cite{Petrek20071357,citeulike:6257975}, this RRT-based approach~\cite{vonasek2016application} 
handles the dynamics without need to cluster and match Voronoi diagrams from consecutive frames.

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} \\
%\includegraphics[width=0.16\textwidth]{fig/motiv3} \\
Protein 1CQW & Active site & Detected tunnels \\
             &            & (orange)
\end{tabular}
}
\caption{\label{fig::motiv}
    Example of tunnel detection in haloalkane dehalogenase.
}
\end{figure}

Due to low volume of protein void space, sampling-based tunnel detection may suffer from the narrow passage problem.
%Building a configuration tree of a spherical probe moving in a protein is however very challenging due to low volume of the protein void space.
To cope with narrow passages, we propose to generate the samples along a subset of Voronoi vertices of the atoms.
%a novel approach that combines VD-based techniques with sampling-based algorithms.
%Instead of generating samples randomly in the whole configuration space, they are generated around a changing subset of vertices of Weighted Voronoi Diagram.
The subset is automatically adapted so it always attracts the tree towards unexplored regions of the configuration space.
The configuration tree is expanded towards the random samples and also by the Voronoi vertices.
Construction of the configuration tree only from Voronoi vertices is not suitable, as it would result in a sparse configuration tree, that would be discontinued when switching to the next frame of protein dynamics.
To enable tunnel detection among changing (moving) atoms, it is necessary to fill the void space by a dense configuration tree, which
is achieved by the randomized sampling.

%vertices of Voronoi diagram.
%The set is maintained in such a way, so it attracts the tree towards unexplored parts of the Voronoi diagram.

%To prevent the tree growth outside the protein, which could lead to detection of invalid (outer) tunnels, we propose
%to block its growth using a set of blocking spheres.
%This results in a tree representing collision-free configurations inside the protein.
%Before the tree is used in the next frame, its nodes that collide with new positions of atoms are removed.
%It is therefore ensured, that for each node in each frame, a path leading to the starting node in the first frame can be found in the tree.
%a path between each node and the starting configuration can be found in the sequence of
%trees.
%It is therefore ensured, that all configurations in the tree are connected to the starting configuration in the first frame.

%The tunnels have low volume compared to the volume of the whole protein, which may slow down sampling-based planning.
%We propose several modifications to RRT to cope with this problem.

%Moreover, RRT allows us to easily cope with additional constraints
%As the biomolecules are densely filled with atoms, their tunnels behave like narrow passages.
%To cope with this problem, we propose two novel modifications of RRT.

\section{Related work}

The analysis of protein structure aiming to reveal the tunnels has been supported by different computational software tools which take the geometry of the protein as an input and explore the inner void space (e.g., CAVER 1.0~\cite{citeulike:6257975} or MOLE~\cite{Petrek20071357}). 
Early methods for tunnel detection utilized a discretized 3D grid, where each cell is considered as occupied or free depending
on the presence of atoms of the protein.
Tunnels can be then searched using standard graph-search methods like Dijkstra's algorithm.
Besides, the grid can be used to identify other relevant properties like 
pockets, cavities, or channels~\cite{sehnal2013mole,citeulike:6257975}.
%One of the first grid-based approaches to the detection of tunnels in protein is the CAVER 1.0 algorithm by Pet\v{r}ek et al.~\cite{citeulike:6257975}.
The obvious disadvantage of the grid-based methods is the high memory demand and their dependency on the grid resolution.
Due to the high memory consumption, these methods are not suitable for tunnel detection in dynamic proteins and therefore they
are used primarily for analysis of static molecules or individual snapshots of dynamics.

%hese approaches are not suitable for tunnel detection in dynamic proteins, a
%, which limits the usage of these methods for spherical ligands.
%To increase precision by 10, the number of cells increases by $10^3$.
%Another disadvantage is that the 3D grid does not consider orientation of the ligand and therefore, the computations 
%are useful only for spherical ligands.

Currently the most widely used approach to tunnel detection is based on Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD).
The ordinary VD is computed on points representing centers of all atoms, without considering radii of the atoms.
This may lead to detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
To consider atoms with different radii, weights of individual points are determined by the van der Walls radii of the atoms in WVD.
An alternative solution is to compute a non-weighted VD on an extended point set, where 
each atom is approximated by several spheres with small radius~\cite{yaffe2008,caver3}.
%In this case, each atom is approximated by several spheres with small radius~\cite{yaffe2008,caver3}.
%The VD-based methods remove one of the main limitations of the grid-based approach which is the dependency on the grid resolution.
VD-based methods are memory less demanding, and also faster than grid-based methods.
%They can be easily used for static tunnel detection, as they are based on Voronoi diagrams of static frames.
%The grid-based and VD-based methods are usually employed on static proteins.
%Finding tunnels through time-changing proteins increases the dimension of the space to be searched, which prevents the grid-based methods
%to be used for this task.
The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
finding correspondences between them.
The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
%The main limitation of this approach is that this solution builds a complex and large hierarchical tree which has to be stored in memory and subsequently traversed which is computationally demanding. %~\cite{lindow2012dynamic}.

The tunnel detection in dynamic proteins can also be formulated as path planning in 4D (i.e., 3D $\times$ time) 
configuration space assuming a spherical probe.
%Tunnel detection in dynamic proteins leads to path planning 
%in a high-dimensional configuration space (4D for the probe and 7D for the ligand).
This space can be searched using sampling-based motion planning methods.
%Instead of a systematic search, which would be time and memory consuming, randomized sampling can be used
%to obtain necessary information about the configuration space. %~\cite{Lav06}. 
%Path planning approaches studied in robotics can be used to solve tunnel detection~\cite{guieysse2008structure} as well.
%Sampling-based path planning methods like Probabilistic Roadmaps (PRM)~\cite{kavrakiPRM} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT} are able to find feasible paths for complex objects with arbitrary geometries.
The idea of sampling-based motion planning is to randomly sample the configuration space $\C$ of the robot. % (i.e., the spherical probe in this case).
%The configuration space is a space of all possible configurations of the robot.
The random samples are classified as free or non-free using collision detection and  the free ones are stored into a roadmap.
A path in the roadmap then represents a motion in the workspace.
%The usage of collision detection for classification of samples allows the application of sampling-based methods for path planning for robots (objects) of arbitrary shapes.
%As the sampling-based methods work in the configuration space, they can cope with objects (robots) of many degrees of freedom
% and arbitrary shapes.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query motion planning method.
RRT incrementally builds a configuration tree $\T$ rooted at the initial configuration.
In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node in the tree $\qnear \in \T$ is found.
A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$.
If $\qnew$ is collision-free, it is added to the tree.
The algorithm terminates if the tree approaches the goal configuration close enough.

Beyond robotics, sampling-based motion planners have been successfully applied in computational biology in studies
related to 
protein folding~\cite{al2012motion,gipson2012computational,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro}, or analysis of loop motions~\cite{cortes2004geometric}.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
The most relevant paper is~\cite{guieysse2008structure}, where RRT is used to compute pathways for flexible ligands towards 
the active site.
This method however does not compute explicitly all tunnels, rather a single pathway.

In our recent paper~\cite{vonasek2016application}, we have proposed to detect tunnels using RRT.
In this approach, RRT samples the configuration space of a small spherical probe moving in a single frame of protein dynamics.
After the tunnels are found in one frame, the built  configuration tree is transferred to the next frame of protein dynamics and pruned to remove
nodes colliding with the new positions of atoms.
This results in a sequence of connected trees, that can be searched for path representing the tunnels in the protein dynamics.

%This allows us to build a sequence of trees, where each tree corresponds to each frame of protein dynamics.
%The nodes of the trees are connected to their parents in the same frame, or to their parents in the previous frame.
%This allows us to find paths in this sequence of tree between two given frames, that correspond to the dynamic tunnels.

In comparison to~\cite{guieysse2008structure}, the approach~\cite{vonasek2016application} assumes spherical probes, which is motivated
by practical reasons.
%The tunnels can be computed for a spherical probe or even for non-spherical ligands --- small molecules.
%Computing tunnels considering only spherical probes is more practical.
%Despite the ability of sampling-based planners to find path for robots (objects) of arbitrary shapes, which is achieved by the collision detection, the above described approach~\cite{vonasek2016application} detect tunnels only for spherical ligands.
%The reasons are practical.
Flexible ligands have many degrees of freedom, which increases dimension of the configuration space to be searched and makes
the collision-detection more complex.
Moreover, it is required to define suitable metric and possibly also an efficient data structure for nearest-neighbor search.
%Dimension of the configuration space of flexible ligands is higher and it is determined by the Degrees Of Freedom of the ligand.
%It is required to define proper metric to efficiently search this space using RRT.
%Planning patways for flexible ligands leads to a search in a high-di
%Taking into account flexible ligands increases the dimension of the configuration space.
Despite the flexibility of ligands, their movements in proteins are strongly limited, as they are typically larger than spherical
probes.
Traversability of the ligand through a protein may depend on its initial orientation, 
 which can further limit the growth of the configuration tree.
Contrary, configuration space of  spherical probes is only 3D with the possibility to employ Euclidean metric and KD-trees for the
nearest neighbor search.
It is more practical to first detect tunnels for a spherical probe and then analyze traversability of the actual shape of ligand.
In this case, the traversability analysis is limited only to a tunnel and its vicinity, which decreases volume of the configuration space and
decreases the number of atoms involved in collision detection.
%which is faster than searching path for the ligand in the full protein.

%The ability of RRT to quickly explore the configuration space is caused by the combination
%of uniform sampling and the nearest-neighbor rule used to select nodes for the expansion.
The growth of the configuration tree into some regions of the configuration space can however be slowed down due to the narrow passages.
A narrow passage is a small region of the configuration space containing part of the solution.
As the samples are generated from uniform distribution in RRT, probability of placing the samples into the narrow passages is low and therefore,
   the probability of expanding the tree through a narrow passage is small~\cite{hannaWIS}.
%Due to its low volume, the probability of expanding the nodes towards the passage is small~\cite{hannaWIS}.
%    many iterations are needed before a tree can be built through the passage~\cite{hannaWIS}.
A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
in the difficult areas.
For Probabilistic Roadmaps~\cite{kavrakiPRM}, this can be achieved simply by generating more samples in the difficult areas, e.g. along
medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
%This requires to precompute the medial axis, which can be time consuming.
%Another approach is to shift the random samples towards the medial axis~\cite{amatoOBPRM} or into a direction of estimated
%medial axis~\cite{hollemanMAPRM}.
Generating random samples in difficult regions however does not ensure construction of a better tree in the case of RRT-based methods. 
In RRT, the configuration tree is expanded in the direction of random samples, but the obstacles may prevent to reach them closely~\cite{vonasekphd}.
Increased probability of sampling in a given region, e.g., in a narrow passage, 
brings advantage only if the tree is close to the region and if it can reach it.

In this paper, we propose how to combine RRT-based search with Voronoi-based methods in the task of tunnel detection. 
To guide the configuration tree using Voronoi diagram, a subset of Voronoi vertices is determined and
the samples are generated around vertices in this subset.
The subset is maintained according the distance from the tree, so the random samples
are generated only if they are in a suitable distance from the tree.
This allows us to sample simultaneously along all promissing parts of VD that are accessible from the tree, which is 
suitable for detection of multiple tunnels.

%In this paper, we propose how to combine the VD-based methods and sampling-based approaches for tunnel detection.
%Instead of uninformed sampling of the configuration space, which has been proposed in~\cite{vonasek2016application}, we propose
%to generate the random samples using Voronoi diagrams.

%~\cite{amatoOBPRM,hollemanMAPRM} shift random samples towards medial axis of the environment, which helps to sample
%narrow passages more densely. 
%For example, the random samples generated in the configuration space can be shifted
%towards the medial axis of the environment~\cite{amatoOBPRM,hollemanMAPRM} or even generated around the medial 
%axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
%The paper~\cite{amatoOBPRM} suggests to sample uniformly in the configuration space and shift the samples towards the medial axis.
%Another approach is to generate the samples randomly in \CS\ and shift them closer to the medial axis~\cite{amatoOBPRM}.
%The medial axis can be computed exactly using GVD.
%While the medial axis can be easily computed for 2D or 3D workspaces using GVD, its computation become complicated in many dimensions.
%In such case, the samples can be generated uniformly from the whole \CS\ and moved in an estimated direction towards the medial axis~\cite{hollemanMAPRM}.

%sampling-based for 
%protein-ligand access and docking \cite{bayazit2001ligand,singhLig,apaydin2004stochastic,cortes2010simulating}
%protein and RNA folding~\cite{amatoPF1,ApaBru03}
%protein loop motions~\cite{cortes2004geometric},
%domain motions~\cite{kirillova2008an}, 
%and motions of pairs of alpha-helices in transmembrane proteins~\cite{enosh2007prediction}

%The tree is extended from a node, that is selected as a nearest node to a randomly generated configuration in the configuration space.
%The highest probability of expansion have nodes with large Voronoi cells. 
%Due to this Voronoi-bias, the tree grows towards unexplored regions of the configuration space.
%However, the Voronoi-bias brings disadvantages in narrow passages.
%nodes in the narrow passages.

%The behavior of RRT in the narrow passages was analyzed using Voronoi diagrams in~\cite{yershovaDDRRT}.
%The nodes in the tree can be divided into two groups: frontier nodes, whose Voronoi cells grow together with growth
%of the environment and boundary nodes, which are close to the obstacles. %~\cite{yershovaDDRRT}.
%The tree cannot be expanded from the boundary nodes.
%In a narrow passage, the nodes are both boundary and frontier.
%These nodes are frequently selected for the expansion, because they are the frontier nodes, however the tree cannot be expanded
%from them, because they are also the boundary nodes. 
%To suppress selection of the boundary nodes, authors of~\cite{yershovaDDRRT} suggest to define an action radius around each node.
%The node is selected for the expansion only if its radius is larger than the distance to the random sample.
%The radius of new nodes is set to $\infty$ and it is decreased to a predefined value $\rdd$ if the node cannot be expanded.
%A Dynamic-Domain strategy for RRT (RRT--DD) is proposed in~\cite{yershovaDDRRT}:
%each node holds an action radius defining how far can be a random sample $\qrand$ that activates the node for the expansion.
%The RRT--DD algorithm generates random samples $\qrand$  and finds its nearest neighbor in the tree
%$\qnear$ until $\rho(\qrand,\qnear) < \qnear.radius$. 
%Although the algorithm is efficient in the narrow passages, it is strongly influenced by the parameter $\rdd$.
%To decrease the sensitivity of the method to the parameter $\rdd$, it can be automatically adjusted, which
%was proposed in RRT--ADD (RRT with Adaptive Dynamic Domain)~\cite{jailletATDDRRT}.
%Another schema to automatically adjust parameters of the RRT-based planners was proposed in~\cite{schneider2015completely}.

%Retraction-RRT~\cite{zhangRetraction} generates random samples uniformly as in the original RRT, but it
%attempts to shift them into the narrow passages.
%A~random configuration $\qrand \in \CF$ is generated and a close non-free configuration $q \in \CO$ is found. 
%A~contact configuration $q_c$ on a segment $(\qrand,q)$ is found and its neighborhood is searched for
%$q_c'$ minimizing the distance between $q$ and $q_c'$. 
%The configuration $q_c'$ is then added to the tree.
%It was shown that this approach can deal with narrow passages efficiently,
%because the generated contact configurations penetrate into the narrow passages.
%However, to find the contact configurations, the collision detection algorithm is called frequently, which can decrease
%the performance of the algorithm.

%The goal-bias principle can be further extended by sampling the configuration space along a path computed in the workspace~\cite{vonasek2009rrt,amatoOBRRT}.
%%Geometric path in the workspace~\cite{vonasek2009rrt} or medial axis of the 3D space~\cite{amatoOBRRT} can be used to guide the tree.
%Sampling along geometric paths constructed in the workspace is suitable for low-dimensional configuration spaces, e.g. for
%path planning of mobile robots.
%However, the geometric paths computed in workspace are less effective for sampling in high-dimensional configuration spaces~\cite{hannaWIS}.
%

%of the narrow passages is related to narrow passaged of the workspace~\cite{hannaWIS}.
%
%is to manipulate the random samples, e.g. to change their rotation, rotation or both parts or to shift them closer to the medial
%axis of the workspace~\cite{amatoOBPRM}.

%In~\cite{amatoOBPRM} several strategies for validating non-free random samples have been proposed.
%Authors propose different manipulation procedures for random configurations generated during the learning phase.
%For example, the random configuration can be placed into the roadmap with changed rotation, translation or both, or it can be
%shifted away from an obstacle allong a line connecting the sample and medial axis of the obstacles.
%In a configuration is invalid, it is pushed randomly into various directions to gen free samples around boundaries of $\CO$.

% ===============================================================================

%25\% of VDW radii, T-RRT \cite{jaillet10costmap}

%\red{The tunnel detection task differs from classic motion planning it two main aspects: a) the goal configuration
%is not explicitly defined, and b) multiple pathways (tunnels) should be detected.}

%Molecular structures can be represented as articulated bodies and sampling-based methods can be used
%Path planning methods have been used to expore the conformational space of 
%proteins~\cite{novinskaya2015improving,songPFintro,mollProt,proteinRRT},
%Other applications of sampling-based approaches is in detection of 
%folding pathways~\cite{amato2002using}, analyzing protein loops~\cite{cortes2004geometric}, 
%or modeling large-scale transitions in a protein structure~\cite{raveh2009rapid}
%The geometric-based tunnel detection approaches provide fast computation of tunels in large protein structures and they
%have been integrated in several tools like Caver~\cite{bara2014caver}, Chexvis~\cite{masood2015chexvis} and other.
%Most of the proposed methods consider only spherical models of ligand~\cite{benkaidali2014computing}.

%dxTuber: Detecting protein cavities, tunnels and clefts based on protein and solvent dynamics:
%furtng protein cavities, tunnels and clefts based on protein and
%solvent dynamics insights into the protein in question. For example, empty
%space in a protein structure can provide valuable insight into protein properties such as internal hydration, structure stabilization,
%substrate translocation, storage compartments or substrate binding sites [1,2]. This information can be visualized by means of cav-
%ity analysis. Over the years numerous cavity detection tools have been developed including [3–17] that depict cavities either directly
%[3–6,8,10–17] or indirectly by identifying lining residues [9] or filling a cavity with water molecules [7]. The main strategies used
%in these geometry-based algorithms [1] can be grouped into four categories plus combinations of these.
%n the motion planning problem, that is widely studied in robotics, the task is to find a feasible trajectory for a robot between two

%given positions in an enironment.
%To utilize the motion planning approaches, the ligand is considere as the robot and the protein's atoms as the obstacles.
%Computing feasible (traversable) tunnels for non-spheric ligands requires to consider also rotations of the ligands, which 
%can be solved using sampling-based approaches like Probabilistic Roadmaps (PRM)~\cite{kavrakiPRM} or Rapidly Exploring
%Random Trees (RRT)~\cite{lavalleRRT}.
%Sampling-based methods randomly samples configuration space of the robot (ligand) and the samples are classified
%as free or non-free using collision detection.
%The free samples are stored in a graph structure.
%A path in the grap then corresponds to a motion in the workspace.

%In~\cite{guieysse2008structure}, RRT method is used to compute pathways for flexible ligands.

%\cite{lindow2012dynamic}
%Analysis of protein dynamics suggests that internal cavities and channels can be rather dynamic structures. 
%Voronoi-based algorithm to extract the geometry and the dynamics of cavities and channels from a molecular dynamics trajectory.
%The algorithm requires a pre-processing step in which the Voronoi diagram of the van der Waals spheres is used to calculate the cav-
%ity structure for each coordinate set of the trajectory. 
%In the next step, we interactively compute dynamic channels by analyzing the time evolution of components of the cavity structure. Tracing of the
%cavity dynamics is supported by timeline visualization tools that allow the user to select specific components of the cavity structures
%for detailed exploration. All visualization methods are interactive and enable the user to animate the time-dependent molecular struc-
%ture together with its cavity structure. To facilitate a comprehensive overview of the dynamics of a channel, we have also developed a
%visualization technique that renders a dynamic channel in a single image and color-codes time on its extension surface. 
%
%
%Sampling-based motion planning algorithms from the field of robotics have been very successful in exploring the
%conformational space of proteins
%\cite{novinskaya2015improving}
%Sampling-based algorithms explore the conformational space of a protein by randomly sampling it (usually using a special
%heuristic) and constructing a graph where each node represents a feasible low-energy protein conformation (or state), and each
%edge represents a possible low-energy local transition between two states. 
%The computed graph describes the topology of a protein’s energy landscape and the connectivity of its lowenergy areas. 
%This graph can be used to find possible large-scale transitions between two given protein conformations.
%

%Sampling-based methods have been very effective for the fast computation of representative motions of molecular systems \cite{al2012motion, gipson2012computational}

%A broad range of approaches exploit sampling-based techniques to address various biological problems, such
%as exploring energy landscapes \cite{devaur2014sampling} ,
%modeling protein folding pathways \cite{amato2002using}, analyzing protein loops \cite{cortes2004geometric}, 
%or modeling large-scale transitions in a protein structure \cite{raveh2009rapid}

%In computational biology, molecules and proteins are modeled as articulated bodies and sampling based planners are used to simulate
%protein folding and protein-ligand interactions  \cite{al2012motion} 
%asi spis necitovat \cite{gipson2012computational}

%Ms have been applied to model molecular motions by modeling the molecule as an articulated linkage and 
%replacing the typical collision detection validity check with some measure of physical viability (e.g., potential energy).
%Protein motions, ranging from molecular flexibility to large-scale conformational change, play an essential role in many biochemical processes. 
%!!to map transitinos between specific conmfotrmations!!!!
%ample conformation space more effectively, and we describe extensions of our framework to automate
%the process and to map transitions between specified conformations.
%\cite{thomas2006simulating}
%

%In subsequent work, our group used another PRM variant on this problem \cite{bayazit2001ligand}
%Our group was the first to adapt PRMs to model protein folding pathways 
%\cite{amato2003using, amato2002using, song2003apath}

%Using a standard modeling assumption for proteins that bond angles and bond lengths are fixed \cite{sternberg1996protein}, 
%the only dof in our model are the backbone’s phi and psi torsional angles which are modeled as revolute joints with values 0 .. 2pi

%
%The roadmap produced by our technique is an approximation of the protein’s energy landscape. 
%Roadmap quality is measured both by how realistic (as compared to experimental data) its pathways are and by how many samples are required
%to achieve the desired accuracy. 
%The latter is important because it determines what size molecules can be analyzed.
%Hence, sampling is the key to producing a good approximation of the land-scape. 
%Note that only a relatively small portion of the conformation space ‘near’ the target conformation(s) is of interest in modeling motions. 
%This implies that we should use biased sampling to cover the regions of interest efficiently.
%
%genrovani lepsich vzorku: using rigidity analysis~\cite{thomas2006simulating}
%or perturbing:
%iterative sampling process where we apply small Gaussian perturbations to existing conformations~\cite{amato2003using, amato2002using, song2003apath}
%
%rrt for tunnel (pathway) detection
%\cite{cortes2010simulating}
%\cite{corset2005path}
%\cite{guieysse2008structure}
%

%Many modifications have been proposed for RRT to speed up the growth of the tree~\cite{kuffnerRRTC}, 
%     to consider an optimality criteria such as path length~\cite{karaman2011sampling}, and to improve behavior in narrow passages~\cite{amatoOBRRT,vonasek2009rrt,zhangRetraction,schneider2015completely}. 
%For example, RRT--Retraction~\cite{zhangRetraction} analyzes the line segment from $\qnear$ to $\qrand$.
%If the line segment is collision-free, the tree is expanded normally using the straight-line planner.
%Otherwise, the contact configuration on the line segment is constructed and its neighborhood is searched
%for other contact configurations in order to minimize the distance to $\qrand$.
%This retraction step allows RRT--Retraction to place the nodes of the tree into narrow passages.

%The behavior of RRT in the narrow passages was analyzed using Voronoi diagrams in~\cite{yershovaDDRRT}.
%The nodes in the tree can be divided into two groups: frontier nodes, whose Voronoi cells grow together with growth
%of the environment and boundary nodes, which are close to the obstacles. %~\cite{yershovaDDRRT}.
%The tree cannot be expanded from the boundary nodes.
%In a narrow passage, the nodes are both boundary and frontier.
%These nodes are frequently selected for the expansion, because they are the frontier nodes, however the tree cannot be expanded
%from them, because they are also the boundary nodes. 
%To suppress selection of the boundary nodes, authors of~\cite{yershovaDDRRT} suggest to define an action radius around each node.
%The node is selected for the expansion only if its radius is larger than the distance to the random sample.
%The radius of new nodes is set to $\infty$ and it is decreased to a predefined value $\rdd$ if the node cannot be expanded.
%A Dynamic-Domain strategy for RRT (RRT--DD) is proposed in~\cite{yershovaDDRRT}:
%each node holds an action radius defining how far can be a random sample $\qrand$ that activates the node for the expansion.
%The RRT--DD algorithm generates random samples $\qrand$  and finds its nearest neighbor in the tree
%$\qnear$ until $\rho(\qrand,\qnear) < \qnear.radius$. 
%Although the algorithm is efficient in the narrow passages, it is strongly influenced by the parameter $\rdd$.
%To decrease the sensitivity of the method to the parameter $\rdd$, it can be automatically adjusted, which
%was proposed in RRT--ADD (RRT with Adaptive Dynamic Domain)~\cite{jailletATDDRRT}.
%Another schema to automatically adjust parameters of the RRT-based planners was proposed in~\cite{schneider2015completely}.

%Retraction-RRT~\cite{zhangRetraction} generates random samples uniformly as in the original RRT, but it
%attempts to shift them into the narrow passages.
%A~random configuration $\qrand \in \CF$ is generated and a close non-free configuration $q \in \CO$ is found. 
%A~contact configuration $q_c$ on a segment $(\qrand,q)$ is found and its neighborhood is searched for
%$q_c'$ minimizing the distance between $q$ and $q_c'$. 
%The configuration $q_c'$ is then added to the tree.
%It was shown that this approach can deal with narrow passages efficiently,
%because the generated contact configurations penetrate into the narrow passages.
%However, to find the contact configurations, the collision detection algorithm is called frequently, which can decrease
%the performance of the algorithm.

%Another approach to biased sampling was presented in~\cite{bruceERRT}, where the trajectories found in previous
%iterations are used to sample the configuration space. 
%This modification is suitable for dynamic environments.

%The growth of the tree can be attracted into a region $\R \in \C$ by increasing probability of sampling in that region.
%This is used in the goal-bias principle~\cite{lavalleRRTPP}, where $\qgoal$ is used instead of $\qrand$ with the probability $\gb$.
%The goal-bias suppresses the exploration of $\C$ and the tree is expanded more preferably towards the goal.
%An extension of the goal-bias was presented in~\cite{kardossRRTKK}, where a predefined key-configuration is 
%used to focus sampling to difficult areas of the configuration space.
%%Authors suggest to place the key-configuration in a way that connects easy and difficult areas of the configuration space.
%%For example, the key-configuration can be placed at the enter to a narrow passage.
%%Although the goal-bias speeds up the growth towards the goal, it may cause a congestion problem of the tree growth due to existence of an obstacle in the pathway.
%
%
%The goal-bias principle can be further extended by sampling the configuration space along a path computed in the workspace~\cite{vonasek2009rrt,amatoOBRRT}.
%%Geometric path in the workspace~\cite{vonasek2009rrt} or medial axis of the 3D space~\cite{amatoOBRRT} can be used to guide the tree.
%Sampling along geometric paths constructed in the workspace is suitable for low-dimensional configuration spaces, e.g. for
%path planning of mobile robots.
%However, the geometric paths computed in workspace are less effective for sampling in high-dimensional configuration spaces~\cite{hannaWIS}.
%

\section{Preliminaries}

Proteins are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
%   can be detected as atoms located at $\alpha$-shape of the molecule.
The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
%We further consider a spherical probe $\Sprobe$  of radius $\probe$ encapsulating the ligand.
Let $q=(x,y,z)\in\C$ denote the configuration (position) of the probe, where
the configuration space $\C$ consists of all possible configurations.
For each configuration $q\in\C$ we assume, that its largest collision-free radius $r(q) \in \mathbf{R}, r(q)\ge 0$ can be computed, e.g.
using collision detection.
%using collision detection.
%Let $L(q) \subseteq \mathbf{R}^3$ denote the geometry of the ligand at position $q \in \C$.
%Similarly to geometry of molecule, the geometry of ligand is union of hard sphere model.
%A probe is represented by one sphere of radius $\probe$.
The collision-free region $\CF \subseteq \C$ is formed by configurations, where $\Sprobe$ can be placed without any collision, i.e., 
$\CF = \{q \in \C | \Sprobe(q) \cap \SS = \emptyset\}$, where $\Sprobe(q)$ is the sphere with radius $\probe$ at configuration $q$.
%We assume, that atoms located on the surface of the molecule are known, these are reffered to as {\sl surface atoms}.
The distance $\dist(q_1,q_2)$  between two configurations $q_1,q_2\in\C$ is measured using 3D Euclidean metric.

%This radius can be detected using collision detection method.
The configurations outside the protein $\CG \subseteq \C$ are collision-free with sphere $\Sgprobe$ of radius $\gprobe$, $\gprobe> \probe$,
$\CG=\{q\in \C| \Sgprobe(q) \cap \SS = \emptyset \}$.
%A configuration $q \in \CG$ is referred to as {\sl goal configuration} in the rest of the paper.
By computing $\alpha$-shape of the molecule, we can identify atoms located at the surface. 
These atoms are referred to as {\sl surface atoms} in the rest of the paper.
The distance between a configuration $q \in \C$ and its nearest surface atom is referred to as $\dists(q)$.
%We assume, that for each configuration $q$, the largest collision-free radius $r(q)$ can be computed.
%Let $r(q)$ denote the radius of the largest collision-free sphere at configuration $q$.

Dynamic proteins are represented by a sequence of $n$ snapshots (frames).
The tunnels change with the protein dynamics as well: their width and position change, they can merge with other tunnels
or even disappear.
The free region $\CF$ is therefore different in each frame, as positions of atoms change.
For the sake of simplicity, a single notation $\C$ resp. $\CF$ is used in this paper and it is valid for the frame being processed.

Approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing ordinary Voronoi diagram.
Computing WVD this way is more efficient and numerically stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
Vertices of this WVD are filtered out if their distance to an atom is less than probe radius $\probe$.
Let $\VV$ contain the remaining Voronoi vertices, and let $r_v \in \VV$ denote the distance to the nearest atom.
%The remaining Voronoi vertices $\VV$ of the WVD and their distances $r_v, v\in \VV$ to the nearest atom are then used in the following method.

%\section{Tunnel detection}

%The tunnel detection using motion planning leads to search in 4-dimensional $\C\!\times\!time$ space, which can be searched using randomized sampling-based methods.
%The task of tunnel detection is to find a collision-free path from a given starting point located inside a dynamic protein 
%to the outer environment. %: for a probe $\Sprobe$.
%The idea of the proposed approach is to detect tunnels in each frame separately using a novel modification of RRT algorithm.
%The configuration tree built in a given frame is transferred to the next frame, where the search continues.

%We propose to detect the tunnels in this 4D space
%(the space consists of pairs $(q,t) \in \C \times [1,n]$, where $n$ is the number of frames) by randomized sampling.
%where $\C$ is three-dimensional configuration space
%of the spherical probe and $n$ is the number of frames.
%It is not computationally feasible to discretize the space to a grid, in which a path is found using standard space-search methods.
%Moreover, finding a collision-free path in the space does not solve the problem, because the probe cannot move back in time.
%Tunnel detection in dynamic proteins leads to search in $\C \times time$ space. %~\cite{Kindel00kinodynamic}.
%sometimes called state-time space~\cite{berg2005roadmap}.
%The proposed solution samples the configuration space of each frame separately in order to detect possible tunnels in the frame.
%The resulting tree is then pruned to remove collision with atoms of the next frame.

\section{Tunnel detection}

The tunnel detection problem is defined by specifying the coordinates of the active site $\qinit \in \CF$, 
    the radius $\probe$ of the probe, and the radius 
$\gprobe$ that determines outer space $\CG$ of the protein.
The dynamic proteins are searched for tunnels in a two-stage process: first, the tunnels are searched in a single frame, then the
built configuration tree is transferred to the next frame and pruned.
Due to space limit, this paper focuses mainly on the first stage. 
We refer to~\cite{vonasek2016application} for detailed description of the second stage.

\subsection{Sampling of configuration space of a single frame}

%The configuration space is sampled using the RRT method. 
%The tunnels detection requires to find tunnels only inside the protein, not on its surface.
%However, proteins have jagged surface with protrusions and small cavities. 
%The growth of the tree outside the protein needs to be prohibited.
%Otherwise, many invalid tunnels hidden in this bumpy surface would be detected.
%These cavities are smaller than $\Sgprobe$ but larger than $\Sprobe$, which prevents us to determine
%configurations at the surface using simple collision detection between $\Sgprobe$ and the protein surface.
%Therefore, simple collision detection between $\Sgprobe$ and the protein may fail to determine configurations located at the surface
%of the protein.

The configuration space is searched using the RRT principle~\cite{lavalleRRT}, which is extended by three main modifications.
First, the tree expanded by several new nodes in each iteration, in order to boost growth in the narrow passages.
The second modification is designed to block growth of the tree outside the protein, which is necessary to prevent detection of false
tunnels. 
The last modification is the novel VD-based sampling of configuration space.
The proposed method is listed in Alg.~\ref{alg::tunnel}.

Instead of adding only a single node in each iteration, which is done in original RRT, we expand the tree by $m$ nodes.
This improves the growth of the tree inside the narrow passages, as these are filled by the tree in less number of iterations.
After a node $\qnear$ for the expansion is determined, its neighborhood in the distance $\varepsilon$ is sampled.
The collision-free samples (i.e., $r(q) \ge \probe$)  are stored in the set $N$.
%These configurations are randomly sampled around $\qnear$ in the distance $\varepsilon$ and these are stored in the set $N$.
%After $\qnear$ is found in the tree, its neighborhood is sampled by at most $m$ random samples in the distance $\varepsilon$. 
%These samples are stored in the set $N$ and added to the tree if they are not too close to the protein surface.
%Otherwise, blocking spheres are added at the corresponding configuration.
%a sample $q$ with the maximal collision-free radius $r(q)$. % (Fig.~\ref{fig::expansion}b,c).
The configuration $q \in N$ is added to the tree if it does not regress to the tree.
% $r(q) \ge \probe$ and 
A sample $q$ is regressing if its nearest neighbor in the tree is closer than $\qnear$. %parent node. % (Alg.~\ref{alg::lexp}, line 7).
This regression test prevents the expansion to large sites, that are already filled with other nodes. 
Without this test, the tree would densely occupy large areas, but less densely the small ones.
Tree expansion by multiple nodes result in more dense trees. 
%If $\qnew$ is located in a large area, i.e., $r(\qnew) > \probe$, additional $m$ samples located around $\qnew$ are added to
%the tree. % (Alg~\ref{alg::lexp}, lines 15--22), which is depicted in Fig.~\ref{fig::expansion}d.
%This procedure expands the tree by multiple nodes (at most $m+1$), but it guarantees that these nodes
%are collision-free at radius at least $\probe$ and that these are not regressing towards the tree.

%The active sites are typically connected with the protein's surface using multiple tunnels.
%It is therefore desired to detect as many tunnels as possible.
%On the other hand, not all path within the protein are considered as good tunnels, e.g. the tunnels partly located in the protrusions and 
%small cavities of the protein surface.
%Such tunnels, despite they are connected with the active site, are not relevant for chemists.

The active sites are typically connected with the protein’s surface by multiple tunnels.
It is therefore desired to detect as many tunnels as possible. 
On the other hand, not all paths within the protein are biochemically relevant, e.g., long thin tunnels with highly curved trajectory or
tunnels partly located on the surface of the protein.
Such tunnels, despite they are connected with the active site, are not relevant because there is a 
high probability that they will be closed in molecular dynamics.

To prevent the growth of the tree on the bumpy surface and to boost searching of tunnels inside the protein,
 algorithm searches the configuration space in two phases.
 In the first phase, { blocking spheres} $\SB$ of radius $\gprobe$ are placed
to configurations that are believed to be close to the surface.
The distance of each configuration $q \in N$ from the nearest surface atom is determined and $q$ is
added to the tree only if this distance is larger than $\ths$. 
Otherwise, $q$ is added to $\SB$ (Alg.~\ref{alg::tunnel}, line 21), because a tunnel can exit around $q$ from the protein.

%Each node $q \in N$ is therefore tested for being close to surface atoms before it is added to the tree (Alg.~\ref{alg::tunnel}, line 20).
%If this distance is less than the threshold $\ths$, the configuration $q$ is considered as being close to the surface, which
%indicates that a tunnel can exit from the protein around this point.
%Therefore, new blocking sphere at this position is introduced into the set $\SB$. % (Fig.~\ref{fig::timexp}d).

After a predefined fraction of iterations (parameter $\alpha$ in Alg.~\ref{alg::tunnel}, lines 32--35), the algorithm
switches to the second phase, and the blocking atoms $\SB$ are deleted and removed from collision detection data structures.
In this second phase, the configuration tree is allowed to exit the protein and reach the outer space.
%This allows the tree to finally expand out of  the protein.
%In such a case (lines 24--26 in Alg.~\ref{alg::tunnel}), $q$ is added to the set of tunnel exit-points.
An exit from the tunnel is detected if $q \in N$ is collision-free with sphere $\Sgprobe$ (or $r(q) > \gprobe$). 
The exit points are stored in the set $G$ (Alg.~\ref{alg::tunnel}, line 28--30).
%The algorithm can be initialized with a non-empty set $\SB$ (line 1 in Alg.~\ref{alg::tunnel}), which allows to block the growth of the tree according to exit-points detected in the previous frame.

%The second modification aims to improve the expansion step.
%Tunnel are narrow passages and the straight-line expansion of the tree, that is often used in RRT-based planners,
%to collision (Fig.~\ref{fig::expansion}a).
%In RRT literature, the tree is often expanded using the straight-line expansion, i.e., $\qnew$ is constructed on a line
%connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
%The protein void space is small and this type of expansion would lead to collisions. % (Fig.~\ref{fig::expansion}a).
%To cope with this issue, it is necessary to find $\qnew$ around $\qnear$ and not only on the line 
%from $\qnear$ to $\qrand$. % (Alg.~\ref{alg::lexp}).

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
{\small
\caption{\label{alg::tunnel}static tunnel detection}
\KwIn{Initial configuration tree $\T$,
    initial set of blocking spheres $\SB'$
}
\KwData{goal-probe radius $\gprobe$,
    probe radius $\probe$, 
    max. number of iterations $\Imax$, 
    phase switching threshold $\alpha$,
    threshold $\ths$,
    num. of new nodes $m$ in each iteration,
    prob. $\gb$ of sampling randomly from $\C$
}
\KwOut{Configuration tree $\T$,
    list of tunnel exit-points $G$
}
\hrule
$\SB = \SB'$; //blocking atoms added during phase~1 \\
$G = \emptyset$; // list of tunnel exit-points \\
$phase=1$\;
\For{$iteration = 1:\Imax$}{
    \If(// Voronoi-based sample --- Section~\ref{sec::vbg}){$phase=1$}{
        $\VVA =$ update active Voronoi points from $\VV$\;
        $v =$ random point from $\VVA$\;
        $\qrand =$ random point around $v$ in the distance $r_v$\;
%        $a_1$ = random atom on surface\;
%        $a_2$ = random atom on surface\;
%        $\qrand$ = random point on line $|a_1, a_2|$\;
    }
    \If{$rand() < \gb$}{
        $\qrand$ = random sample from $\C$\;
        $v = \emptyset$; //no Voronoi vertex used
    }

    $\qnear$ = nearest node in $\T$ towards $\qrand$\;

    \If{$v \ne \emptyset$ {\bf and} $\dist(\qnear, v) < \probe$} {
        add $v$ to $\T$, set its parent to $\qnear$\;
    }

    $N\!=\!m$ collision-free nodes around $\!\qnear$ in the distance $\!\varepsilon$\;
%    $(\qnew,r(\qnew))$ = expand($\T$,$\qnear$, $\SS \cup \SB$)\;
    \For{$q \in N $}{
        \eIf{$\dists(q) < \ths $ {\bf and} $phase=1$}{
            $\SB = \SB \cup \{\Sgprobe(q)\}$ \;
        }{
            $q'$ = nearest node in $\T$ towards $q$\;
            \If{$\dist(q,q') > \dist(q,\qnear)$}{
                add $q$ to $\T$, set its parent to $\qnear$\;
            }
        }
        \If{ $r(q) > \gprobe$ }{
            $G = G \cup \{ q \} $\;
        }
    }
    \If{$phase = 1$ {\bf and} $iteration > \alpha \Imax$}{
        $phase = 2$\;
        $\SB = \emptyset$; // block. atoms are not used in phase 2 \\
    }
}
\return $(\T, G)$\; 
}
\end{algorithm}

%\begin{algorithm}[h]
%%\setstretch{\straa}
%{\small
%\caption{\label{alg::lexp}expand()}
%\KwIn{Configuration tree $\T$, configuration $\qnear$ to be expanded, 
%    collision-geometry $\SS$
%}
%\KwData{num. of random samples $n$ around $\qnear$, 
%    max. number of additional nodes $m$,
%    resolution $\varepsilon$
%}
%\KwOut{Return new node $\qnew$ and its collision-free radius $r(\qnew)$ if it is added to tree, or NULL otherwise
%}
%\hrule
%    $\qnew = \emptyset$; // new node to be added to the tree \\
%    $r' = 0$; // largest collision-free radius of node $\qnew$\\
%    \For{$i=1:n$}{
%        $q$ = random sample around $\qnear$ in distance $\varepsilon$\;
%        $q'$ = nearest node in $\T$ towards $q$\;
%        $r(q)$ = collision-free radius of sample $q$ w.r.t. $\SS$\;
%        \If{$dist(q,q') < dist(q,\qnear$) {\bf and} $r(q) > r'$}{
%               $r' = r(q)$\; 
%               $\qnew = q$\;
%        }
%    }
%    \If{$\qnew \ne \emptyset$}{
%        $\T$.addNode($\qnew$)\;
%        $\T$.addEdge($\qnear, \qnew)$\;
%        \If{$r' > \probe$}{
%            $r_{s} = r' - \probe$\;
%            \For{$i=1:m$}{
%                $q$ = random sample in distance $r_s$ from $\qnew$\;
%                $\T.$addNode($q$)\;
%                $\T.$addEdge($\qnew, q$)\;
%            } 
%        }
%        \return $(\qnew,r')$\;
%    }
%    \return NULL;
%}
%\end{algorithm}
%

\subsection{Tunnel detection in dynamic proteins}

A sequence of frames of protein dynamics is searched by repeated construction of a tree and its transfer to the next frame.
The collisions between the nodes of the actual tree and atoms of the next frame are checked and the colliding nodes are removed.
The parents of the remaining nodes are changed to corresponding nodes from the previous tree.
The search then continues by expanding the existing tree using Alg.~\ref{alg::tunnel}.
This results in a sequence of trees.
Each node in the tree points to its parent, which is either a node of the same tree, or a node of the tree in the previous frame.
Also, the result is a set of exit points (the set $G$ in Alg.~\ref{alg::tunnel}), which contains ending points of tunnels in each frame. 
The tunnels are then reconstructed by searching paths from each $g \in G$ to the first frame (or to other frame of interest).

%The algorithm starts with the first frame at configuration $\qinit \in \CF$.
%We assume that this position is collision-free at frame $1$, otherwise it would not be possible to construct the tree around it.
%The result of the static tunnel detection (Alg.~\ref{alg::tunnel}) is a new configuration tree of collision-free nodes for frame $i$ and
%tunnel exit-points $G$.
%Further expansions from the tunnel exit-nodes have to be prohibited, otherwise the tree would fill the outer region of the molecule.
%Similarly to the Alg.~\ref{alg::tunnel}, this is achieved using the blocking spheres $\SB$.
%The blocking spheres have radius $\gprobe$ and they are placed at tunnel exit-points $G$ (Fig.~\ref{fig::timexp}d).
%The result of the Alg.~\ref{alg::tunnel} is a configuration tree $\T$ representing the protein void space, together with
%a set of tunnel exit-points $G$. 
%The result of Alg.~\ref{alg::main} is represented by a sequence of trees $\T_0,\ldots,\T_n$, and the tunnel exit-points stored in the set $G$.
%$t$ is represented by the sequence 
%$t=\left((q_1,f_1), \ldots, (q_{k},f_{k})\right)$ of $k$ points $q_i$ and their corresponding frames $f_i$.

%\begin{algorithm}
%%\setstretch{\straa}
%{\small
%\caption{tree pruning\label{alg::copyTree}}
%\KwIn{Geometry $\SS_{i+1}$ for frame $i+1$,
%    configuration tree $\T_i$ constructed at frame $i$
%}
%\KwData{probe radius $\probe$, goal radius $\gprobe$}
%\KwOut{Configuration tree $\T_{i+1}$ valid for frame $i+1$}
%\hrule
%$\SB = \emptyset $\;
%$\T_{i+1} = $empty tree\;
%\For{$q \in \T_i$}{
%    $r(q)$ = largest collision-free radius at $q$ w.r.t. $\SS_{i+1}$\;
%    \If{$r(q) \ge \probe$ {\bf and} $r(q) < \gprobe$ }{
%        $T_{i+1}.$addNode($q$)\;
%        set parent of $\T_{i+1}$.node($q$) to $\T_i.$node($q$)\;
%    }
%    \If(// $q$ is tunnel exit-point){$r(q) \ge \gprobe$}{
%        $\SB = \SB \cup \{\Sgprobe(q)\}$\;
%    }
%}
%\return $(\T_{i+1}, \SB)$\;
%}
%\end{algorithm}
%
%

\section{Voronoi-based guided sampling}
\label{sec::vbg}

Due to limited void space of the protein, tunnel detection suffers from the narrow passage problem.
To cope with this problem, we propose to guide the sampling in the configuration space using vertices $\VV$ of WVD of the atoms.
To prevent stucking of the tree due to obstacles (i.e., atoms), it is necessary to generate random samples around $\VV$ only if the tree has
a chance to grow towards the random samples.
To achieve this, we maintain a set of {\sl active} Voronoi vertices called $\VVA$. 
A Voronoi vertex $v \in \VV$ is considered as active, if the distance to its nearest node in the tree is less than $\da$, but larger than
$\db$ (Fig.~\ref{fig::av}).
The upper value $\da$ ensures, that the active Voronoi vertices are close to the boundary of the configuration tree, whilst
the lower value $\db$ prevents to activate Voronoi vertices that have already been approached by the tree.
This sampling process therefore automatically adapts according to the progress of the configuration tree:
after the tree approaches a given $v \in \VVA$, this vertex is removed from $\VVA$ so its surrounding is not sampled anymore.
Contrary, distant vertices $v \in \VV$ are not used for sampling.

To generate a random sample $\qrand$, $v \in \VVA$ is randomly selected and $\qrand$ is generated around $v$ using Gaussian distribution with 
$\sigma = r_v$ (Alg.~\ref{alg::tunnel}, lines 5--9).
%This Guassian distribution is used because it allows to generate $\qrand$ also out of the sphere ce
%so $\qrand$ can be placed also out of the sphere of the Voronoi vertex $v$.
%which has been shown to be useful.
The Voronoi-based sampling is used only in the first phase of the method, where the tree growth outside the protein is blocked by $\SB$.  
In the second phase, it is desired to grow outside the protein and the VD structure is not needed, so
 $\qrand$ is randomly taken from the whole $\C$ with probability $\gb$ (Alg.~\ref{alg::tunnel}, line 10--13).

%It should be noted, that Voronoi vertices alone does not necessary represent the tunnels, 
%as they can be located too close to the atoms (closer than the radius of the probe).

Besides generating $\qrand$ around $v \in \VVA$, the Voronoi vertex $v$ is added to the tree if its distance to $\qnear$ is less than
 $\probe$ (Alg.~\ref{alg::tunnel}, lines 15--17).
Such addition is valid as all vertices in $\VV$ (and consequently in $\VVA$) have collision-free radius at least $\probe$.

The set $\VVA$ does not need to be updated in each iteration, because the tree is expanded only by the distance $\varepsilon$ in each
iteration. 
The update can be realized e.g. after every $k$ nodes added to the tree.

%The configuration tree is expanded by several nodes in each iteration, but in the maximal distance $\varepsilon$, so several
%terations is needed to approach a given $v \in \VVA$. 
%Therefore, it is not necessary to update $\VVA$ in each iteration, but only after $k$ nodes is added to the tree.

%As the configuration tree is extended by several spheres in each step that are at most in distance $\varepsilon$ from the expanded node $\qnear$.
%It is therefore needed several iterations to extend the tree close enough to the nodes in $\VVA$.
%Consequently, it is not necessary to update $\VVA$ in each iteration, but rather after every $k$ nodes added to the tree (Alg.

\begin{figure}
\centering
%\includegraphics[width=0.23\textwidth]{fig/activevor}
%\includegraphics[width=0.23\textwidth]{fig/activevor2}
\includegraphics[width=0.23\textwidth]{fig/av1}
\includegraphics[width=0.23\textwidth]{fig/av2}
\caption{\label{fig::av}
    Simplified example of a progress of generating random samples using WVD of a set of 2D atoms.
    Atoms are depicted in blue, WVD edges in green.
    Dashed circles are centered at Voronoi vertices, and the radius is given by the distance to the nearest atom.
    Small red circles represent a partially built configuration tree.
    Active Voronoi vertices, that are close to the tree are depicted in light yellow.
    Random samples are generated around these active vertices.
    The tree is attracted by samples in the yellow regions (left). 
    Voronoi vertices, that have been already approached, are depicted in light pink (right).
    Random samples are not generated around these vertices, so the tree is still attracted towards unexplored regions.
}
\end{figure}

\section{Discussion}
%+proc nepridavat rovnou koule z WVD
%+multiple tunnels

%The advantage of the proposed approach is that the sequence of protein snapshots is searched frame by frame.
%By transferring the constructed tree to the next frame it is ensured that tunnels can be search not only within frame, but also
%between two given frames.

As each frame of protein dynamics is analyzed separately, only the actual configuration tree and frame-related data structures 
(e.g., WVD, collision-detection structures) need to be held in the memory.
After the frame is switched, the full tree from the previous frame can be stored.
According to our experiments, 50--70\% of nodes are kept to the next frame.
The amount of required memory therefore does not depend on the number of frames to be processed, but mainly on the size of the protein.
The sampling in the next frame does not start from scratch, but it continues by adding new nodes to the tree.
Therefore, the number of iterations $\Imax$ has to be large for the first frame in order to maximize the number of detected tunnels, but it can 
be decreased for the subsequent frames.

Detecting tunnels for spherical probes brings several advantages.
The collision detection can be computed very fast between a spherical probe and the hard sphere model of the protein.
Besides, simple 3D Euclidean metric can be used to measure the distance in the configuration space.
Consequently, the nearest neighbor search can be realized using KD-trees in $\mathcal{O}(\log n)$, where $n$ is the number of 
nodes in the tree.
KD-trees can also be used to compute the distance $\dists$ to surface atoms 
(line 20 in Alg.~\ref{alg::tunnel}), and for the update of set $\VVA$ (Alg.~\ref{alg::tunnel}, line 6), 
which makes these procedures fast.

%The sampling in a given frame is terminated after $\Imax$ iterations.
%The tree is then pruned to remove the nodes colliding with atoms of the next frame.
%The amount of memory required to detect tunnels in a sequence of protein dynamics is given only by the number of nodes in each frame.

The number of atoms in proteins, their arrangement as well as roughness of the surface may significantly vary, 
       and the behavior of the algorithms have to be adjusted.
The ability to detect tunnels is influenced by the parameter $\alpha$.
The first phase should be long enough in order to build the tree inside the protein.
According to experiments on proteins listed in Tab.~\ref{tab::results}, 
we suggest to set $\alpha=0.7$, i.e., 70\% of iterations is performed in the first phase.
The distance to surface atoms $\ths$ indicating tunnel exit points depends on the roughness of the protein surface and we recommend
to set it less than $\gprobe$, e.g. $\ths=\gprobe/5$ in our case.
The parameters of Voronoi-based sampling, $\da$ and $\db$ depend on the amount of void space and also on the probe radius $\probe$.
We recommend to use $\db=0.5\probe$ and $\da=4\probe$.

%For the purpose of visualization and further analysis, it is convenient to fill the spheres, i.e.,  to find $r(q_i)$ for each point $q_i \in t$.

\section{Experimental verification}

{\bf Static tunnel detection: }
The proposed method (Alg.~\ref{alg::tunnel}), referred to as \TA, is compared to~\cite{vonasek2016application} (referred to as \TB) and
to basic search of 3D path for sphere using the basic RRT~\cite{lavalleRRT}, denoted as \TC.
In \TB, the Voronoi-based sampling is not used.
%Algorithms are run with the following parameters.

%Collision detection was realized using OZCollide library~\cite{ozcollide}.
Voronoi vertices $\VV$ used in \TA\ are computed using WVD with the weight set to the radii of the atoms.
Only Voronoi vertices located inside the alpha-shape are considered to be in the set $\VV$.
The set $\VVA$ is updated after 1000 new nodes in the tree.
\TA\ and \TB\ strategies are run with $\alpha=0.7$ and $\Imax=50\cdot 10^3$ iterations.
The thresholds for Voronoi sampling in \TA\ are $\da=4r_v$ ($r_v$ is radius of a Voronoi vertex $v \in \VV$), $\db=0.5\probe$.
\TA\ and \TB\ expand the tree by up to $m=4$ new nodes in each iteration, \TC\ only by one.
\TC\ therefore needs more iterations, so for a fair comparison, \TC\ is run with 10 times more iterations ($\Imax=500\cdot10^3$).
All other settings are the same for all three tested strategies:  $\varepsilon=0.1$~\AA, $\gprobe=5$~\AA, $\ths=1$~\AA.

%(Alg.~\ref{alg::tunnel}), referred to as \RRTTD\ in the rest of the paper, 
%was launched with $\Imax=50\cdot10^3$ iterations, $\alpha=0.7$, and for probes $\probe=0.8, 0.9$, and $1.1$~\AA, resolution
%$\varepsilon=0.1$~\AA, and $\gprobe=5$~\AA, expansion parameters $n=50$, $m=3$, and $\ths=1$~\AA.
%The resulting tunnels were compared with CAVER 3.0~\cite{caver3} analytical tool that computes tunnels using approximated Voronoi diagrams.

%The proposed \RRTTD\ method was also compared with a na\"{i}ve RRT, referred to as \RRTN\ in the following text.
%\RRTN\ was implemented according~\cite{lavalleRRT} and it generates random samples uniformly in the configuration space and expands
%the tree using the straight-line expansion with resolution $\varepsilon=0.1$~\AA.
%d as RRT-na\"{i}ve in the rest of the text, gienerates random samples uniformly from the whole configuration space, and it does not use the two-phase blocking mechanisms to prevent growing of the tree outside molecules. 
%\RRTN\ builds the configuration tree slower than \RRTTD\, therefore $\Imax=500\cdot10^3$ iterations were used in \RRTN\ in order to build
%trees with similar number of nodes~as~\RRTTD.

Tunnels are searched for several proteins from the Protein Data Bank (PDB) (\url{www.pdb.org}) and compared
to results of CAVER~3.0~\cite{caver3} tool.
The surface atoms are detected using alpha-shapes with radius 5~\AA. 
The resolution of all methods is set to $\varepsilon=0.1$~\AA.
The default setting is used for CAVER, except for the shell radius which is set to $5$~\AA~and the resolution of provided tunnels is $0.1$~\AA.

The performance of RRT-based methods is evaluated as the success rate of finding tunnels provided by CAVER.
For each protein and probe radius, methods \TA, \TB\ and \TC\ are run 50 times.
The tunnels are matched with the tunnels provided by CAVER for the same molecule and probe.
The tunnels are represented as a sequence of 3D spheres, $t=(q_1,\ldots,q_k), q\in\CF$.
The one-way distance between two tunnels $d'(t_1,t_2)$ is defined as the average distance between points $q_i \in t_1$ and 
their closest points $q'_2 \in t_2$ in the tunnel $t_2$, $d'(t_1,t_2)={1\over n_1}\sum_{i=1}^{n_1} \dist(q_i, q_2')$, where $n_1$ is the
number of points in $t_1$.
To match tunnels of CAVER and RRT methods, the distance
%The distance between two tunnels is computed as %that was used to match tunnels of CAVER and RRT is computed as 
$D(t_1,t_2) = \max( d'(t_1, t_2), d'(t_2, t_1))$ is used.
Two tunnels can be considered as the same if their distance is less or equal 
$2.5$~\AA~(examples of same/different tunnels according this metric are depicted in Fig.~\ref{fig::pic2}c,d).
The success rate for a given CAVER's tunnel is the percentage of RRT trials (out of 50) where this tunnel is detected.
CAVER, depending on the size of the probe, can provide more tunnels, that are sorted according 
to a cost function which is derived from the tunnel length, its bottleneck, and the curvature of the tunnel.
The default sorting criteria, that prefer shorter and thicker tunnels, are used.
We are therefore interested in the success rate of the first three tunnels, because they are 
supposed to be the most important for further biochemical analysis.
The examples of detected tunnels are depicted in Fig.~\ref{fig::pic2} and~\ref{fig::pic1}.

%This distance can be interpreted as the worst mean distance between tunnel points.
%The RRT-based tunnels end by spheres with radius $\gprobe$, while CAVER tunnels end much closer to the surface of the molecule.
%Therefore, we compare the tunnels only using first $n'_1$ points, where $n'$ is the index of point of the first tunnel, whose distance
%from the last point on the tunnel is $\Sprobe$.

The results are summarized in Tab.~\ref{tab::results}.
The first column (denoted {\sl T}) is the index of the CAVER's tunnel.
The column {\sl Length} shows the length of the tunnels in~\AA.
Columns \TA,\TB\ and \TC\ show the success rate (in percents) of finding these tunnels by the tested RRT-based strategies.
The first three rows (starting with $1$, $2$, and $3$) in each section show success rates of finding the first three tunnels provided by CAVER.
%The next two rows show the tunnels that were found by \RRTTD\ with high probability.
%For example, first three tunnels of radius 0.7~\AA~in 1BL8 were found with success rates 63\%, 60\%, and 68\% by \RRTTD.
%\RRTTD\ also found tunnel \#10 in 87\% of cases and tunnel \#18 in 84\% of cases.
Not all three tunnels can be found in all proteins, especially with larger probes, which is indicated by the blank space 
(e.g., protein 1CQW contains only one tunnel for probe 1.1~\AA).
%It is not possible to find at least 3 tunnels in each molecule and probe,
% therefore less than 5 tunnels can be reported (e.g. in 1CQW, probe 1.1~\AA).
%On the contrary, many tunnels exist for smaller probes (0.7~\AA). % because there is more space to move amongst the atoms.
%This gives RRT a chance to expand the configuration tree through these tunnels, but due to randomization, not all tunnels are explored in each trial and the success rate is smaller than with a larger probe.

From the biochemical point of view, the most important is the first tunnel (denoted as $1$) detected by CAVER.
Generally, the first tunnels are always found with high success rate (sometimes even 100~\%), but the success rate is lower
for other tunnels.
Lengths of the tunnels show that the first tunnel is typically shorter than other tunnels.
These tunnels are more tortuous and they are not going directly to the surface.

By comparing the sampling strategies, the proposed method \TA\ outperforms other strategies in most of the cases, except for the 1BL8 protein.
The possible reason is that 1BL8 has larger internal void space (many tunnels for large spheres can be found in it), but
the parameters $\da,\db$ have been used the same for all methods.
It is worth to emphasize that \TC\ strategy (naive RRT sampling) is run with 10 times more iterations $\Imax$ and still it finds tunnels
with significantly lower success rate.
This shows that the two-phase approach for configuration space sampling used in \TA\ and \TB\ brings advantage.
\TA\ and \TB\ perform similarly, but \TA\ has higher success rate for the tunnels of higher order, which
is achieved by the Voronoi guided sampling.

%Detection of such tunnels can be difficult for RRT-based methods and it can be increased by tuning parameters related
%to protein properties,  such as $\da$ and $\db$. 
%In this comparison, all settings are same for all proteins.

%For the probe 0.9 and 1.1~\AA, \RRTTD\ finds the tunnel \#1 with high probability (90~\% or more) for most of the tested proteins.
%The exception is the protein 1BL8, where the first three tunnels are found with 50--60~\% success rate with probes 0.9 and 1.1~\AA.
%The reason is caused by the initial position, that is on the tip of the protein (Fig.~\ref{fig::pic1}c).
%During the first phase of \RRTTD, most samples are generated inside the protein.
%The tree is therefore attracted towards the protein, which
%decreases the probability to expand the tree in the opposite direction, where the short tunnel can be found.
%As \RRTTD\ generates more samples inside the molecule during the first phase, the tree is attracted towa
%probability to find the first CAVER's tunnel. % (Fig.~\ref{fig::pic}b).
%Therefore, other tunnels, e.g. \#8 are found by \RRTTD\ more often (Fig.~\ref{fig::pic1}d)  as these go through center of the protein.
%Another starting position was tested (1BL8$^c$ in Tab.~\ref{tab::results}, Fig.~\ref{fig::pic1}c), where the success rates of 
%finding the first three tunnels are significantly higher.

%Experiments have also shown superior performance of \RRTTD\ over \RRTN, as \RRTTD\ provides tunnels with significantly higher probability.
%It is worth to emphasize that \RRTN\ was run with 10 times more iterations $\Imax$ to enable the construction of trees with similar number of nodes.
%This shows that the introduced modifications of RRT for the tunnel detection improve the planning process.

\begin{table}
\centering
\caption{\label{tab::results}
    Comparison of CAVER~3.0 and \TA, \TB\ and \TC\ methods. 
        Bold font denotes the best success rate for each tunnel and probe.
}
{
\footnotesize
\renewcommand{\arraystretch}{1.1}
\renewcommand{\tabcolsep}{2.3pt}
%\input com-all.tex
}
\end{table}

\begin{figure*}
\centering
{
\renewcommand{\arraystretch}{0.2}
%\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccccc}
\includegraphics[width=0.22\textwidth]{fig/1akdprobe09-1} &
\includegraphics[width=0.18\textwidth]{fig/1akdprobe09-0-label} &
\includegraphics[width=0.12\textwidth]{fig/1akd-good} &
\includegraphics[width=0.105\textwidth]{fig/1akd-bad}  \\
(a) 1AKD tunnels      & (b) 1AKD tunnels only & (c) Correct tunnels (A,B)  & (d) Incorrect tunnels (C)
\end{tabular}
}
\caption{\label{fig::pic2}
    Tunnels detected in the protein 1AKD for $\probe=0.9$~\AA. 
Tunnels detected by \TA\ are depicted in orange, green spheres show tunnels computed by CAVER 3.0.
The protein with all tunnels detected by \TA\ method (a). 
The comparison between tunnels detected by \TA\ (orange) and CAVER 3.0 (green) (b) with highlighted three tunnels.
Examples of correctly detected tunnels --- green and orange spheres overlap in the most length of the tunnels (c).
Example of a CAVER tunnel (green) that has not been detected by \TA, as its nearest runnel (orange) does not overlap (d).
}
\end{figure*}

\begin{figure*}
\centering
{
\renewcommand{\arraystretch}{0.2}
%\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccccc}
\includegraphics[width=0.22\textwidth]{fig/1cqw-main} &
%\includegraphics[width=0.3\textwidth]{fig/1cqw-caver} &
%\includegraphics[width=0.3\textwidth]{fig/1cqw-rrt} &
\includegraphics[width=0.22\textwidth]{fig/1cqw-tunnels} &
%\includegraphics[width=0.3\textwidth]{fig/1cqw-detailt2} \\
\includegraphics[width=0.23\textwidth]{fig/1bl8-center} &
\includegraphics[width=0.23\textwidth]{fig/result1bl8-comp-09-1}  \\
%(a) Tunnels in 1CQW, probe $0.7$\AA & (b) 1CQW tunnels only & (c) Start positions in 1BL8 and 1BL8$^c$ & (d) Tunnel \#8 in 1BL8, probe 0.9~\AA
(a) Tunnels in 1CQW,      & (b) 1CQW tunnels only & (c) Start positions in  & (d) Tunnel \#8 in 1BL8, \\
probe 0.7~\AA            &                       & 1BL8        &     probe 0.9~\AA \\                                 
\end{tabular}
}
\caption{\label{fig::pic1}
Tunnels detected by \TA\ are depicted in orange, green spheres show tunnels computed by CAVER 3.0.
}
\end{figure*}

{\bf Tunnel detection in dynamic protein: }
A sequence of 1,000 frames of protein Haloalkane dehalogenase (DhaA) was analyzed with $\probe=0.9$.
Due to protein dynamics, the tunnels change rapidly and they often disappear or merge with others.
Each frame was analyzed with Alg.~\ref{alg::tunnel} with $\Imax=1000$ iterations.
Fig.~\ref{fig::dyn} shows visualization of selected frames.
In comparison to CAVER, that clusters the first tunnels found in each frame to a single 3D tunnel, the proposed
method can compute behavior or each detected tunnel separately.
Due to the CAVER's implicit clustering of the first tunnels, we do show any comparison of tunnels.
The total time to process the whole sequence was $\sim\!\!2$ minutes (@ 2.9 GHz), while CAVER took $\sim\!70$ minutes.
Due to space limits, only few images are shown in this section. 
We refer to \url{http://mrs.felk.cvut.cz/romoco2017} for more information.

\begin{figure*}
\centering
\begin{tabular}{cccc}
\includegraphics[width=0.2\textwidth]{fig/dcpmld2-6-label} &
\includegraphics[width=0.2\textwidth]{fig/dcpmld2-12} &
\includegraphics[width=0.2\textwidth]{fig/dcpmld2-19} &
\includegraphics[width=0.2\textwidth]{fig/dcpmld2-20} \\
Frame 6 & Frame 12 & Frame 19 & Frame 20 \\                       
\includegraphics[width=0.2\textwidth]{fig/dcpmld2-48} &
\includegraphics[width=0.2\textwidth]{fig/dcpmld2-54} &
\includegraphics[width=0.2\textwidth]{fig/dcpmld2-55} &
\includegraphics[width=0.2\textwidth]{fig/dcpmld2-56} \\
Frame 48 & Frame 54 & Frame 55 & Frame 56 \\                       
\end{tabular}
\caption{\label{fig::dyn}
Evolution of void space in molecular dynamics of the DHA protein. 
The active site is highlighted by the red circle in the top left image.
In the frame 20, the single tunnel is separated into three smaller tunnels.
Two of them temporarily connect in the frame 55, but then they remain unconnected.
}
\end{figure*}

\section{Conclusion }

The problem of tunnel detection in protein structures can be formulated as a path planning problem which can be solved by RRT.
%The paper shows how to utilize the RRT planner for this task.
%In comparison to existing techniques for tunnel detection, which are based on Voronoi Diagrams, the proposed method can find tunnels in a sequence of protein dynamics.
The tunnels have properties of narrow passages and it is therefore challenging to find path inside the protein.
We propose to generate random samples around a subset of Voronoi Diagram of the protein's atoms.
The subset is maintained automatically according to the distance to the configuration tree, so only the regions not reached yet
by the tree are sampled.
While the Voronoi diagram provides valuable information about suitable void regions in the protein, the
proposed RRT-based technique can cope with the protein dynamics.
The proposed technique can therefore be considered as an extension of the existing VD-based tools for tunnel detection.

\section{Acknowledgments}

The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.

\bibliographystyle{plain}
\bibliography{paper}

\end{document}

### 2017-05-03 paper.tex
deleted old stuff

@@
 }
-
-%This work was supported by Grant Agency of the Czech Technical University in Prague, grant No. SGS15/157/OHK3/2T/13.
-%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
-%vonasek@labe.felk.cvut.cz.
 
@@
 
-%VD provides initial information about protein void space, which helps to sample the corresponding configuration space.
-
-%While VD of the protein atoms can be easily computed, the RRT is used to cope with the protein dynamics<F12>
-%While the VD provides
-%Random samples are generated around a subset of vertices of VD
-
-%and modifications to support searching of multiple tunnels in each frame.
-%In the comparison to existing literature on sampling using Voronoi diagrams, the proposed technique supports finding of multiple
-%tunnels even in dynamic proteins.
-%The proposed approach is compared to CAVER 3.0, one of the widely used freely available tools for protein analysis.
-%In this paper, we propose a novel sampling technique for the purpose of tunnel detection in protein structures.
-%The protein void space can be search using sampling-based
-%The classical approach to find tunnels in protein structures is based on Voronoi diagrams (VD), which is suitable
-%for tunnel detection in a single snapshot of protein's dynamics.
-%To consider protein dynamics, that is represented by a sequence of protein snapshots, correspondences between
-%VD in these snapshots need to be found, which is time and memory consuming.
-%An alternative approach is to employ motion planning methods to detect the tunnels.
-%However, VD-based computation can be time consuming when protein dynamics needs to be considered, as it
-%requires to find correspondences between VD in a sequence of frames of protein dynamics.
-%The performance of VD-based tunnel detection is decreased in the case of dynamic proteins, which requires
-%to find correspondences between VD in frames of molecular dynamics.
-%In order to apply the VD-based approach for dynamic proteins,
-%Available methods for tunnel detection assume static proteins, which allows us to detect tunnels using Voronoi Diagrams.
-%Tunnels in static proteins however  provide only a  rough idea about protein behavior.
-%Tunnels in real proteins are however influenced by dynamics of the protein.
-%In this paper, we describe tunnel detection using Rapidly Exploring Random Tree (RRT).
-%The method builds a single configuration tree describing free space of the protein. 
-%A crucial part of this novel approach is the sampling of the internal void space of the protein.
-%The nodes of the tree are pruned according to protein dynamics.
 \end{abstract}
@@
 \section{Introduction}
-%{\color{red}Co bys rikala na to dat na 1. stranku nejaky obrazek tunelu?}
 
@@
 
-Recently, we have proposed a novel method for tunnel detection using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT) method~\cite{vonasek2016application}.
-RRT builds a tree of collision-free configurations of a spherical probe moving in the protein. 
-%In this method, the corresponding configuration space is searched using RRT (Rapidly Exploring Random Tree)~\cite{lavalleRRT}.
-To consider the molecular dynamics, the tree is continuously pruned in each frame.
-%The proposed solution is based on RRT technique, which constructs a configuration tree separately in each frame of protein dynamics.
-In comparison to existing approaches~\cite{Petrek20071357,citeulike:6257975}, this RRT-based approach~\cite{vonasek2016application} 
-handles the dynamics without need to cluster and match Voronoi diagrams from consecutive frames.
+Curently available tools for protein analysis support only the tunnel detection both on single protein snapshots and sequences
+of molecular dynamics.
+The properties of tunnels like bottleneck radius, their length and the surrounding residues can be used
+to estimate suitability of the tunnel for a ligand.
+However, such estimation still involve lot of experience.
+
+\red{TODO}
+something about screeing: what is it, what's is the input and output, in which applications (research/industry) it is useful.
 
@@
 
-Due to low volume of protein void space, sampling-based tunnel detection may suffer from the narrow passage problem.
-%Building a configuration tree of a spherical probe moving in a protein is however very challenging due to low volume of the protein void space.
-To cope with narrow passages, we propose to generate the samples along a subset of Voronoi vertices of the atoms.
-%a novel approach that combines VD-based techniques with sampling-based algorithms.
-%Instead of generating samples randomly in the whole configuration space, they are generated around a changing subset of vertices of Weighted Voronoi Diagram.
-The subset is automatically adapted so it always attracts the tree towards unexplored regions of the configuration space.
-The configuration tree is expanded towards the random samples and also by the Voronoi vertices.
-Construction of the configuration tree only from Voronoi vertices is not suitable, as it would result in a sparse configuration tree, that would be discontinued when switching to the next frame of protein dynamics.
-To enable tunnel detection among changing (moving) atoms, it is necessary to fill the void space by a dense configuration tree, which
-is achieved by the randomized sampling.
+\red{TODO}
+what is traversability analysis.
 
-%vertices of Voronoi diagram.
-%The set is maintained in such a way, so it attracts the tree towards unexplored parts of the Voronoi diagram.
-
-%To prevent the tree growth outside the protein, which could lead to detection of invalid (outer) tunnels, we propose
-%to block its growth using a set of blocking spheres.
-%This results in a tree representing collision-free configurations inside the protein.
-%Before the tree is used in the next frame, its nodes that collide with new positions of atoms are removed.
-%It is therefore ensured, that for each node in each frame, a path leading to the starting node in the first frame can be found in the tree.
-%a path between each node and the starting configuration can be found in the sequence of
-%trees.
-%It is therefore ensured, that all configurations in the tree are connected to the starting configuration in the first frame.
-
-%The tunnels have low volume compared to the volume of the whole protein, which may slow down sampling-based planning.
-%We propose several modifications to RRT to cope with this problem.
-
-%Moreover, RRT allows us to easily cope with additional constraints
-%As the biomolecules are densely filled with atoms, their tunnels behave like narrow passages.
-%To cope with this problem, we propose two novel modifications of RRT.
+The goal is the traversability analysis is to estimate traversability of a tunnel with a given ligand from geometrical point of view.
 
 \section{Related work}
+
+\red{TODO include}
+Recently, we have proposed a novel method for tunnel detection using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT) method~\cite{vonasek2016application}.
+In comparison to existing approaches~\cite{Petrek20071357,citeulike:6257975}, this RRT-based approach~\cite{vonasek2016application} 
+handles the dynamics without need to cluster and match Voronoi diagrams from consecutive frames.
 
@@
 The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
+
+A tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, where $n$ denotes the number of spheres and
+$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius for a spherical probe centered
+at $t_i$. 
+The tunnels can be found by tunnel-detection tools like CAVER 3.0~\cite{caver3}.
+
+Ligands are small molecules interacting with the protein. 
+Ligands are not rigid structures, they rather change due to interactions with the protein's atoms.
+Flexible ligands can be modeled as a kinematic chain with predefied degrees of freedom.
+An alternative approach, that is used in this work, is to consider a predefined set of possible conformations.
+The conformations are prepared using Rosetta software TODO. and their are denoted $c$.
+
 %We further consider a spherical probe $\Sprobe$  of radius $\probe$ encapsulating the ligand.
@@
 
-\section{Tunnel detection}
+\section{Ligand traversability}
 
-The tunnel detection problem is defined by specifying the coordinates of the active site $\qinit \in \CF$, 
-    the radius $\probe$ of the probe, and the radius 
-$\gprobe$ that determines outer space $\CG$ of the protein.
-The dynamic proteins are searched for tunnels in a two-stage process: first, the tunnels are searched in a single frame, then the
-built configuration tree is transferred to the next frame and pruned.
-Due to space limit, this paper focuses mainly on the first stage. 
-We refer to~\cite{vonasek2016application} for detailed description of the second stage.
-
-\subsection{Sampling of configuration space of a single frame}
-
-%The configuration space is sampled using the RRT method. 
-%The tunnels detection requires to find tunnels only inside the protein, not on its surface.
-%However, proteins have jagged surface with protrusions and small cavities. 
-%The growth of the tree outside the protein needs to be prohibited.
-%Otherwise, many invalid tunnels hidden in this bumpy surface would be detected.
-%These cavities are smaller than $\Sgprobe$ but larger than $\Sprobe$, which prevents us to determine
-%configurations at the surface using simple collision detection between $\Sgprobe$ and the protein surface.
-%Therefore, simple collision detection between $\Sgprobe$ and the protein may fail to determine configurations located at the surface
-%of the protein.
+The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
+The tunnels are usually computed on proteins without ligands, which implies narrow void space.
 
@@
 The last modification is the novel VD-based sampling of configuration space.
-The proposed method is listed in Alg.~\ref{alg::tunnel}.
-
-Instead of adding only a single node in each iteration, which is done in original RRT, we expand the tree by $m$ nodes.
-This improves the growth of the tree inside the narrow passages, as these are filled by the tree in less number of iterations.
-After a node $\qnear$ for the expansion is determined, its neighborhood in the distance $\varepsilon$ is sampled.
-The collision-free samples (i.e., $r(q) \ge \probe$)  are stored in the set $N$.
-%These configurations are randomly sampled around $\qnear$ in the distance $\varepsilon$ and these are stored in the set $N$.
-%After $\qnear$ is found in the tree, its neighborhood is sampled by at most $m$ random samples in the distance $\varepsilon$. 
-%These samples are stored in the set $N$ and added to the tree if they are not too close to the protein surface.
-%Otherwise, blocking spheres are added at the corresponding configuration.
-%a sample $q$ with the maximal collision-free radius $r(q)$. % (Fig.~\ref{fig::expansion}b,c).
-The configuration $q \in N$ is added to the tree if it does not regress to the tree.
-% $r(q) \ge \probe$ and 
-A sample $q$ is regressing if its nearest neighbor in the tree is closer than $\qnear$. %parent node. % (Alg.~\ref{alg::lexp}, line 7).
-This regression test prevents the expansion to large sites, that are already filled with other nodes. 
-Without this test, the tree would densely occupy large areas, but less densely the small ones.
-Tree expansion by multiple nodes result in more dense trees. 
-%If $\qnew$ is located in a large area, i.e., $r(\qnew) > \probe$, additional $m$ samples located around $\qnew$ are added to
-%the tree. % (Alg~\ref{alg::lexp}, lines 15--22), which is depicted in Fig.~\ref{fig::expansion}d.
-%This procedure expands the tree by multiple nodes (at most $m+1$), but it guarantees that these nodes
-%are collision-free at radius at least $\probe$ and that these are not regressing towards the tree.
-
-%The active sites are typically connected with the protein's surface using multiple tunnels.
-%It is therefore desired to detect as many tunnels as possible.
-%On the other hand, not all path within the protein are considered as good tunnels, e.g. the tunnels partly located in the protrusions and 
-%small cavities of the protein surface.
-%Such tunnels, despite they are connected with the active site, are not relevant for chemists.
-
-The active sites are typically connected with the protein’s surface by multiple tunnels.
-It is therefore desired to detect as many tunnels as possible. 
-On the other hand, not all paths within the protein are biochemically relevant, e.g., long thin tunnels with highly curved trajectory or
-tunnels partly located on the surface of the protein.
-Such tunnels, despite they are connected with the active site, are not relevant because there is a 
-high probability that they will be closed in molecular dynamics.
-
-To prevent the growth of the tree on the bumpy surface and to boost searching of tunnels inside the protein,
- algorithm searches the configuration space in two phases.
- In the first phase, { blocking spheres} $\SB$ of radius $\gprobe$ are placed
-to configurations that are believed to be close to the surface.
-The distance of each configuration $q \in N$ from the nearest surface atom is determined and $q$ is
-added to the tree only if this distance is larger than $\ths$. 
-Otherwise, $q$ is added to $\SB$ (Alg.~\ref{alg::tunnel}, line 21), because a tunnel can exit around $q$ from the protein.
-
-%Each node $q \in N$ is therefore tested for being close to surface atoms before it is added to the tree (Alg.~\ref{alg::tunnel}, line 20).
-%If this distance is less than the threshold $\ths$, the configuration $q$ is considered as being close to the surface, which
-%indicates that a tunnel can exit from the protein around this point.
-%Therefore, new blocking sphere at this position is introduced into the set $\SB$. % (Fig.~\ref{fig::timexp}d).
-
-After a predefined fraction of iterations (parameter $\alpha$ in Alg.~\ref{alg::tunnel}, lines 32--35), the algorithm
-switches to the second phase, and the blocking atoms $\SB$ are deleted and removed from collision detection data structures.
-In this second phase, the configuration tree is allowed to exit the protein and reach the outer space.
-%This allows the tree to finally expand out of  the protein.
-%In such a case (lines 24--26 in Alg.~\ref{alg::tunnel}), $q$ is added to the set of tunnel exit-points.
-An exit from the tunnel is detected if $q \in N$ is collision-free with sphere $\Sgprobe$ (or $r(q) > \gprobe$). 
-The exit points are stored in the set $G$ (Alg.~\ref{alg::tunnel}, line 28--30).
-%The algorithm can be initialized with a non-empty set $\SB$ (line 1 in Alg.~\ref{alg::tunnel}), which allows to block the growth of the tree according to exit-points detected in the previous frame.
-
-%The second modification aims to improve the expansion step.
-%Tunnel are narrow passages and the straight-line expansion of the tree, that is often used in RRT-based planners,
-%to collision (Fig.~\ref{fig::expansion}a).
-%In RRT literature, the tree is often expanded using the straight-line expansion, i.e., $\qnew$ is constructed on a line
-%connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
-%The protein void space is small and this type of expansion would lead to collisions. % (Fig.~\ref{fig::expansion}a).
-%To cope with this issue, it is necessary to find $\qnew$ around $\qnear$ and not only on the line 
-%from $\qnear$ to $\qrand$. % (Alg.~\ref{alg::lexp}).
 
@@
 
-%\begin{algorithm}[h]
-%%\setstretch{\straa}
-%{\small
-%\caption{\label{alg::lexp}expand()}
-%\KwIn{Configuration tree $\T$, configuration $\qnear$ to be expanded, 
-%    collision-geometry $\SS$
-%}
-%\KwData{num. of random samples $n$ around $\qnear$, 
-%    max. number of additional nodes $m$,
-%    resolution $\varepsilon$
-%}
-%\KwOut{Return new node $\qnew$ and its collision-free radius $r(\qnew)$ if it is added to tree, or NULL otherwise
-%}
-%\hrule
-%    $\qnew = \emptyset$; // new node to be added to the tree \\
-%    $r' = 0$; // largest collision-free radius of node $\qnew$\\
-%    \For{$i=1:n$}{
-%        $q$ = random sample around $\qnear$ in distance $\varepsilon$\;
-%        $q'$ = nearest node in $\T$ towards $q$\;
-%        $r(q)$ = collision-free radius of sample $q$ w.r.t. $\SS$\;
-%        \If{$dist(q,q') < dist(q,\qnear$) {\bf and} $r(q) > r'$}{
-%               $r' = r(q)$\; 
-%               $\qnew = q$\;
-%        }
-%    }
-%    \If{$\qnew \ne \emptyset$}{
-%        $\T$.addNode($\qnew$)\;
-%        $\T$.addEdge($\qnear, \qnew)$\;
-%        \If{$r' > \probe$}{
-%            $r_{s} = r' - \probe$\;
-%            \For{$i=1:m$}{
-%                $q$ = random sample in distance $r_s$ from $\qnew$\;
-%                $\T.$addNode($q$)\;
-%                $\T.$addEdge($\qnew, q$)\;
-%            } 
-%        }
-%        \return $(\qnew,r')$\;
-%    }
-%    \return NULL;
-%}
-%\end{algorithm}
-%
-
-\subsection{Tunnel detection in dynamic proteins}
-
-A sequence of frames of protein dynamics is searched by repeated construction of a tree and its transfer to the next frame.
-The collisions between the nodes of the actual tree and atoms of the next frame are checked and the colliding nodes are removed.
-The parents of the remaining nodes are changed to corresponding nodes from the previous tree.
-The search then continues by expanding the existing tree using Alg.~\ref{alg::tunnel}.
-This results in a sequence of trees.
-Each node in the tree points to its parent, which is either a node of the same tree, or a node of the tree in the previous frame.
-Also, the result is a set of exit points (the set $G$ in Alg.~\ref{alg::tunnel}), which contains ending points of tunnels in each frame. 
-The tunnels are then reconstructed by searching paths from each $g \in G$ to the first frame (or to other frame of interest).
-
-%The algorithm starts with the first frame at configuration $\qinit \in \CF$.
-%We assume that this position is collision-free at frame $1$, otherwise it would not be possible to construct the tree around it.
-%The result of the static tunnel detection (Alg.~\ref{alg::tunnel}) is a new configuration tree of collision-free nodes for frame $i$ and
-%tunnel exit-points $G$.
-%Further expansions from the tunnel exit-nodes have to be prohibited, otherwise the tree would fill the outer region of the molecule.
-%Similarly to the Alg.~\ref{alg::tunnel}, this is achieved using the blocking spheres $\SB$.
-%The blocking spheres have radius $\gprobe$ and they are placed at tunnel exit-points $G$ (Fig.~\ref{fig::timexp}d).
-%The result of the Alg.~\ref{alg::tunnel} is a configuration tree $\T$ representing the protein void space, together with
-%a set of tunnel exit-points $G$. 
-%The result of Alg.~\ref{alg::main} is represented by a sequence of trees $\T_0,\ldots,\T_n$, and the tunnel exit-points stored in the set $G$.
-%$t$ is represented by the sequence 
-%$t=\left((q_1,f_1), \ldots, (q_{k},f_{k})\right)$ of $k$ points $q_i$ and their corresponding frames $f_i$.
-
-%\begin{algorithm}
-%%\setstretch{\straa}
-%{\small
-%\caption{tree pruning\label{alg::copyTree}}
-%\KwIn{Geometry $\SS_{i+1}$ for frame $i+1$,
-%    configuration tree $\T_i$ constructed at frame $i$
-%}
-%\KwData{probe radius $\probe$, goal radius $\gprobe$}
-%\KwOut{Configuration tree $\T_{i+1}$ valid for frame $i+1$}
-%\hrule
-%$\SB = \emptyset $\;
-%$\T_{i+1} = $empty tree\;
-%\For{$q \in \T_i$}{
-%    $r(q)$ = largest collision-free radius at $q$ w.r.t. $\SS_{i+1}$\;
-%    \If{$r(q) \ge \probe$ {\bf and} $r(q) < \gprobe$ }{
-%        $T_{i+1}.$addNode($q$)\;
-%        set parent of $\T_{i+1}$.node($q$) to $\T_i.$node($q$)\;
-%    }
-%    \If(// $q$ is tunnel exit-point){$r(q) \ge \gprobe$}{
-%        $\SB = \SB \cup \{\Sgprobe(q)\}$\;
-%    }
-%}
-%\return $(\T_{i+1}, \SB)$\;
-%}
-%\end{algorithm}
-%
-%
-
-\section{Voronoi-based guided sampling}
-\label{sec::vbg}
-
-Due to limited void space of the protein, tunnel detection suffers from the narrow passage problem.
-To cope with this problem, we propose to guide the sampling in the configuration space using vertices $\VV$ of WVD of the atoms.
-To prevent stucking of the tree due to obstacles (i.e., atoms), it is necessary to generate random samples around $\VV$ only if the tree has
-a chance to grow towards the random samples.
-To achieve this, we maintain a set of {\sl active} Voronoi vertices called $\VVA$. 
-A Voronoi vertex $v \in \VV$ is considered as active, if the distance to its nearest node in the tree is less than $\da$, but larger than
-$\db$ (Fig.~\ref{fig::av}).
-The upper value $\da$ ensures, that the active Voronoi vertices are close to the boundary of the configuration tree, whilst
-the lower value $\db$ prevents to activate Voronoi vertices that have already been approached by the tree.
-This sampling process therefore automatically adapts according to the progress of the configuration tree:
-after the tree approaches a given $v \in \VVA$, this vertex is removed from $\VVA$ so its surrounding is not sampled anymore.
-Contrary, distant vertices $v \in \VV$ are not used for sampling.
-
-To generate a random sample $\qrand$, $v \in \VVA$ is randomly selected and $\qrand$ is generated around $v$ using Gaussian distribution with 
-$\sigma = r_v$ (Alg.~\ref{alg::tunnel}, lines 5--9).
-%This Guassian distribution is used because it allows to generate $\qrand$ also out of the sphere ce
-%so $\qrand$ can be placed also out of the sphere of the Voronoi vertex $v$.
-%which has been shown to be useful.
-The Voronoi-based sampling is used only in the first phase of the method, where the tree growth outside the protein is blocked by $\SB$.  
-In the second phase, it is desired to grow outside the protein and the VD structure is not needed, so
- $\qrand$ is randomly taken from the whole $\C$ with probability $\gb$ (Alg.~\ref{alg::tunnel}, line 10--13).
-
-%It should be noted, that Voronoi vertices alone does not necessary represent the tunnels, 
-%as they can be located too close to the atoms (closer than the radius of the probe).
-
-Besides generating $\qrand$ around $v \in \VVA$, the Voronoi vertex $v$ is added to the tree if its distance to $\qnear$ is less than
- $\probe$ (Alg.~\ref{alg::tunnel}, lines 15--17).
-Such addition is valid as all vertices in $\VV$ (and consequently in $\VVA$) have collision-free radius at least $\probe$.
-
-The set $\VVA$ does not need to be updated in each iteration, because the tree is expanded only by the distance $\varepsilon$ in each
-iteration. 
-The update can be realized e.g. after every $k$ nodes added to the tree.
-
-%The configuration tree is expanded by several nodes in each iteration, but in the maximal distance $\varepsilon$, so several
-%terations is needed to approach a given $v \in \VVA$. 
-%Therefore, it is not necessary to update $\VVA$ in each iteration, but only after $k$ nodes is added to the tree.
-
-%As the configuration tree is extended by several spheres in each step that are at most in distance $\varepsilon$ from the expanded node $\qnear$.
-%It is therefore needed several iterations to extend the tree close enough to the nodes in $\VVA$.
-%Consequently, it is not necessary to update $\VVA$ in each iteration, but rather after every $k$ nodes added to the tree (Alg.
-
-\begin{figure}
-\centering
-%\includegraphics[width=0.23\textwidth]{fig/activevor}
-%\includegraphics[width=0.23\textwidth]{fig/activevor2}
-\includegraphics[width=0.23\textwidth]{fig/av1}
-\includegraphics[width=0.23\textwidth]{fig/av2}
-\caption{\label{fig::av}
-    Simplified example of a progress of generating random samples using WVD of a set of 2D atoms.
-    Atoms are depicted in blue, WVD edges in green.
-    Dashed circles are centered at Voronoi vertices, and the radius is given by the distance to the nearest atom.
-    Small red circles represent a partially built configuration tree.
-    Active Voronoi vertices, that are close to the tree are depicted in light yellow.
-    Random samples are generated around these active vertices.
-    The tree is attracted by samples in the yellow regions (left). 
-    Voronoi vertices, that have been already approached, are depicted in light pink (right).
-    Random samples are not generated around these vertices, so the tree is still attracted towards unexplored regions.
-}
-\end{figure}
-
-\section{Discussion}
-%+proc nepridavat rovnou koule z WVD
-%+multiple tunnels
-
-%The advantage of the proposed approach is that the sequence of protein snapshots is searched frame by frame.
-%By transferring the constructed tree to the next frame it is ensured that tunnels can be search not only within frame, but also
-%between two given frames.
-
-As each frame of protein dynamics is analyzed separately, only the actual configuration tree and frame-related data structures 
-(e.g., WVD, collision-detection structures) need to be held in the memory.
-After the frame is switched, the full tree from the previous frame can be stored.
-According to our experiments, 50--70\% of nodes are kept to the next frame.
-The amount of required memory therefore does not depend on the number of frames to be processed, but mainly on the size of the protein.
-The sampling in the next frame does not start from scratch, but it continues by adding new nodes to the tree.
-Therefore, the number of iterations $\Imax$ has to be large for the first frame in order to maximize the number of detected tunnels, but it can 
-be decreased for the subsequent frames.
-
-Detecting tunnels for spherical probes brings several advantages.
-The collision detection can be computed very fast between a spherical probe and the hard sphere model of the protein.
-Besides, simple 3D Euclidean metric can be used to measure the distance in the configuration space.
-Consequently, the nearest neighbor search can be realized using KD-trees in $\mathcal{O}(\log n)$, where $n$ is the number of 
-nodes in the tree.
-KD-trees can also be used to compute the distance $\dists$ to surface atoms 
-(line 20 in Alg.~\ref{alg::tunnel}), and for the update of set $\VVA$ (Alg.~\ref{alg::tunnel}, line 6), 
-which makes these procedures fast.
-
-%The sampling in a given frame is terminated after $\Imax$ iterations.
-%The tree is then pruned to remove the nodes colliding with atoms of the next frame.
-%The amount of memory required to detect tunnels in a sequence of protein dynamics is given only by the number of nodes in each frame.
-
-The number of atoms in proteins, their arrangement as well as roughness of the surface may significantly vary, 
-       and the behavior of the algorithms have to be adjusted.
-The ability to detect tunnels is influenced by the parameter $\alpha$.
-The first phase should be long enough in order to build the tree inside the protein.
-According to experiments on proteins listed in Tab.~\ref{tab::results}, 
-we suggest to set $\alpha=0.7$, i.e., 70\% of iterations is performed in the first phase.
-The distance to surface atoms $\ths$ indicating tunnel exit points depends on the roughness of the protein surface and we recommend
-to set it less than $\gprobe$, e.g. $\ths=\gprobe/5$ in our case.
-The parameters of Voronoi-based sampling, $\da$ and $\db$ depend on the amount of void space and also on the probe radius $\probe$.
-We recommend to use $\db=0.5\probe$ and $\da=4\probe$.
-
-%For the purpose of visualization and further analysis, it is convenient to fill the spheres, i.e.,  to find $r(q_i)$ for each point $q_i \in t$.
-
 \section{Experimental verification}
-
-{\bf Static tunnel detection: }
-The proposed method (Alg.~\ref{alg::tunnel}), referred to as \TA, is compared to~\cite{vonasek2016application} (referred to as \TB) and
-to basic search of 3D path for sphere using the basic RRT~\cite{lavalleRRT}, denoted as \TC.
-In \TB, the Voronoi-based sampling is not used.
-%Algorithms are run with the following parameters.
-
-%Collision detection was realized using OZCollide library~\cite{ozcollide}.
-Voronoi vertices $\VV$ used in \TA\ are computed using WVD with the weight set to the radii of the atoms.
-Only Voronoi vertices located inside the alpha-shape are considered to be in the set $\VV$.
-The set $\VVA$ is updated after 1000 new nodes in the tree.
-\TA\ and \TB\ strategies are run with $\alpha=0.7$ and $\Imax=50\cdot 10^3$ iterations.
-The thresholds for Voronoi sampling in \TA\ are $\da=4r_v$ ($r_v$ is radius of a Voronoi vertex $v \in \VV$), $\db=0.5\probe$.
-\TA\ and \TB\ expand the tree by up to $m=4$ new nodes in each iteration, \TC\ only by one.
-\TC\ therefore needs more iterations, so for a fair comparison, \TC\ is run with 10 times more iterations ($\Imax=500\cdot10^3$).
-All other settings are the same for all three tested strategies:  $\varepsilon=0.1$~\AA, $\gprobe=5$~\AA, $\ths=1$~\AA.
-
-%(Alg.~\ref{alg::tunnel}), referred to as \RRTTD\ in the rest of the paper, 
-%was launched with $\Imax=50\cdot10^3$ iterations, $\alpha=0.7$, and for probes $\probe=0.8, 0.9$, and $1.1$~\AA, resolution
-%$\varepsilon=0.1$~\AA, and $\gprobe=5$~\AA, expansion parameters $n=50$, $m=3$, and $\ths=1$~\AA.
-%The resulting tunnels were compared with CAVER 3.0~\cite{caver3} analytical tool that computes tunnels using approximated Voronoi diagrams.
-
-%The proposed \RRTTD\ method was also compared with a na\"{i}ve RRT, referred to as \RRTN\ in the following text.
-%\RRTN\ was implemented according~\cite{lavalleRRT} and it generates random samples uniformly in the configuration space and expands
-%the tree using the straight-line expansion with resolution $\varepsilon=0.1$~\AA.
-%d as RRT-na\"{i}ve in the rest of the text, gienerates random samples uniformly from the whole configuration space, and it does not use the two-phase blocking mechanisms to prevent growing of the tree outside molecules.
-%\RRTN\ builds the configuration tree slower than \RRTTD\, therefore $\Imax=500\cdot10^3$ iterations were used in \RRTN\ in order to build
-%trees with similar number of nodes~as~\RRTTD.
-
-Tunnels are searched for several proteins from the Protein Data Bank (PDB) (\url{www.pdb.org}) and compared
-to results of CAVER~3.0~\cite{caver3} tool.
-The surface atoms are detected using alpha-shapes with radius 5~\AA. 
-The resolution of all methods is set to $\varepsilon=0.1$~\AA.
-The default setting is used for CAVER, except for the shell radius which is set to $5$~\AA~and the resolution of provided tunnels is $0.1$~\AA.
-
-The performance of RRT-based methods is evaluated as the success rate of finding tunnels provided by CAVER.
-For each protein and probe radius, methods \TA, \TB\ and \TC\ are run 50 times.
-The tunnels are matched with the tunnels provided by CAVER for the same molecule and probe.
-The tunnels are represented as a sequence of 3D spheres, $t=(q_1,\ldots,q_k), q\in\CF$.
-The one-way distance between two tunnels $d'(t_1,t_2)$ is defined as the average distance between points $q_i \in t_1$ and 
-their closest points $q'_2 \in t_2$ in the tunnel $t_2$, $d'(t_1,t_2)={1\over n_1}\sum_{i=1}^{n_1} \dist(q_i, q_2')$, where $n_1$ is the
-number of points in $t_1$.
-To match tunnels of CAVER and RRT methods, the distance
-%The distance between two tunnels is computed as %that was used to match tunnels of CAVER and RRT is computed as 
-$D(t_1,t_2) = \max( d'(t_1, t_2), d'(t_2, t_1))$ is used.
-Two tunnels can be considered as the same if their distance is less or equal 
-$2.5$~\AA~(examples of same/different tunnels according this metric are depicted in Fig.~\ref{fig::pic2}c,d).
-The success rate for a given CAVER's tunnel is the percentage of RRT trials (out of 50) where this tunnel is detected.
-CAVER, depending on the size of the probe, can provide more tunnels, that are sorted according 
-to a cost function which is derived from the tunnel length, its bottleneck, and the curvature of the tunnel.
-The default sorting criteria, that prefer shorter and thicker tunnels, are used.
-We are therefore interested in the success rate of the first three tunnels, because they are 
-supposed to be the most important for further biochemical analysis.
-The examples of detected tunnels are depicted in Fig.~\ref{fig::pic2} and~\ref{fig::pic1}.
-
-%This distance can be interpreted as the worst mean distance between tunnel points.
-%The RRT-based tunnels end by spheres with radius $\gprobe$, while CAVER tunnels end much closer to the surface of the molecule.
-%Therefore, we compare the tunnels only using first $n'_1$ points, where $n'$ is the index of point of the first tunnel, whose distance
-%from the last point on the tunnel is $\Sprobe$.
-
-The results are summarized in Tab.~\ref{tab::results}.
-The first column (denoted {\sl T}) is the index of the CAVER's tunnel.
-The column {\sl Length} shows the length of the tunnels in~\AA.
-Columns \TA,\TB\ and \TC\ show the success rate (in percents) of finding these tunnels by the tested RRT-based strategies.
-The first three rows (starting with $1$, $2$, and $3$) in each section show success rates of finding the first three tunnels provided by CAVER.
-%The next two rows show the tunnels that were found by \RRTTD\ with high probability.
-%For example, first three tunnels of radius 0.7~\AA~in 1BL8 were found with success rates 63\%, 60\%, and 68\% by \RRTTD.
-%\RRTTD\ also found tunnel \#10 in 87\% of cases and tunnel \#18 in 84\% of cases.
-Not all three tunnels can be found in all proteins, especially with larger probes, which is indicated by the blank space 
-(e.g., protein 1CQW contains only one tunnel for probe 1.1~\AA).
-%It is not possible to find at least 3 tunnels in each molecule and probe,
-% therefore less than 5 tunnels can be reported (e.g. in 1CQW, probe 1.1~\AA).
-%On the contrary, many tunnels exist for smaller probes (0.7~\AA). % because there is more space to move amongst the atoms.
-%This gives RRT a chance to expand the configuration tree through these tunnels, but due to randomization, not all tunnels are explored in each trial and the success rate is smaller than with a larger probe.
-
-From the biochemical point of view, the most important is the first tunnel (denoted as $1$) detected by CAVER.
-Generally, the first tunnels are always found with high success rate (sometimes even 100~\%), but the success rate is lower
-for other tunnels.
-Lengths of the tunnels show that the first tunnel is typically shorter than other tunnels.
-These tunnels are more tortuous and they are not going directly to the surface.
-
-By comparing the sampling strategies, the proposed method \TA\ outperforms other strategies in most of the cases, except for the 1BL8 protein.
-The possible reason is that 1BL8 has larger internal void space (many tunnels for large spheres can be found in it), but
-the parameters $\da,\db$ have been used the same for all methods.
-It is worth to emphasize that \TC\ strategy (naive RRT sampling) is run with 10 times more iterations $\Imax$ and still it finds tunnels
-with significantly lower success rate.
-This shows that the two-phase approach for configuration space sampling used in \TA\ and \TB\ brings advantage.
-\TA\ and \TB\ perform similarly, but \TA\ has higher success rate for the tunnels of higher order, which
-is achieved by the Voronoi guided sampling.
-
-%Detection of such tunnels can be difficult for RRT-based methods and it can be increased by tuning parameters related
-%to protein properties,  such as $\da$ and $\db$. 
-%In this comparison, all settings are same for all proteins.
-
-%For the probe 0.9 and 1.1~\AA, \RRTTD\ finds the tunnel \#1 with high probability (90~\% or more) for most of the tested proteins.
-%The exception is the protein 1BL8, where the first three tunnels are found with 50--60~\% success rate with probes 0.9 and 1.1~\AA.
-%The reason is caused by the initial position, that is on the tip of the protein (Fig.~\ref{fig::pic1}c).
-%During the first phase of \RRTTD, most samples are generated inside the protein.
-%The tree is therefore attracted towards the protein, which
-%decreases the probability to expand the tree in the opposite direction, where the short tunnel can be found.
-%As \RRTTD\ generates more samples inside the molecule during the first phase, the tree is attracted towa
-%probability to find the first CAVER's tunnel. % (Fig.~\ref{fig::pic}b).
-%Therefore, other tunnels, e.g. \#8 are found by \RRTTD\ more often (Fig.~\ref{fig::pic1}d)  as these go through center of the protein.
-%Another starting position was tested (1BL8$^c$ in Tab.~\ref{tab::results}, Fig.~\ref{fig::pic1}c), where the success rates of 
-%finding the first three tunnels are significantly higher.
-
-%Experiments have also shown superior performance of \RRTTD\ over \RRTN, as \RRTTD\ provides tunnels with significantly higher probability.
-%It is worth to emphasize that \RRTN\ was run with 10 times more iterations $\Imax$ to enable the construction of trees with similar number of nodes.
-%This shows that the introduced modifications of RRT for the tunnel detection improve the planning process.
-
-\begin{table}
-\centering
-\caption{\label{tab::results}
-    Comparison of CAVER~3.0 and \TA, \TB\ and \TC\ methods. 
-        Bold font denotes the best success rate for each tunnel and probe.
-}
-{
-\footnotesize
-\renewcommand{\arraystretch}{1.1}
-\renewcommand{\tabcolsep}{2.3pt}
-%\input com-all.tex
-}
-\end{table}
-
-\begin{figure*}
-\centering
-{
-\renewcommand{\arraystretch}{0.2}
-%\renewcommand{\tabcolsep}{0pt}
-\begin{tabular}{ccccc}
-\includegraphics[width=0.22\textwidth]{fig/1akdprobe09-1} &
-\includegraphics[width=0.18\textwidth]{fig/1akdprobe09-0-label} &
-\includegraphics[width=0.12\textwidth]{fig/1akd-good} &
-\includegraphics[width=0.105\textwidth]{fig/1akd-bad}  \\
-(a) 1AKD tunnels      & (b) 1AKD tunnels only & (c) Correct tunnels (A,B)  & (d) Incorrect tunnels (C)
-\end{tabular}
-}
-\caption{\label{fig::pic2}
-    Tunnels detected in the protein 1AKD for $\probe=0.9$~\AA. 
-Tunnels detected by \TA\ are depicted in orange, green spheres show tunnels computed by CAVER 3.0.
-The protein with all tunnels detected by \TA\ method (a). 
-The comparison between tunnels detected by \TA\ (orange) and CAVER 3.0 (green) (b) with highlighted three tunnels.
-Examples of correctly detected tunnels --- green and orange spheres overlap in the most length of the tunnels (c).
-Example of a CAVER tunnel (green) that has not been detected by \TA, as its nearest runnel (orange) does not overlap (d).
-}
-\end{figure*}
-
-\begin{figure*}
-\centering
-{
-\renewcommand{\arraystretch}{0.2}
-%\renewcommand{\tabcolsep}{0pt}
-\begin{tabular}{ccccc}
-\includegraphics[width=0.22\textwidth]{fig/1cqw-main} &
-%\includegraphics[width=0.3\textwidth]{fig/1cqw-caver} &
-%\includegraphics[width=0.3\textwidth]{fig/1cqw-rrt} &
-\includegraphics[width=0.22\textwidth]{fig/1cqw-tunnels} &
-%\includegraphics[width=0.3\textwidth]{fig/1cqw-detailt2} \\
-\includegraphics[width=0.23\textwidth]{fig/1bl8-center} &
-\includegraphics[width=0.23\textwidth]{fig/result1bl8-comp-09-1}  \\
-%(a) Tunnels in 1CQW, probe $0.7$\AA & (b) 1CQW tunnels only & (c) Start positions in 1BL8 and 1BL8$^c$ & (d) Tunnel \#8 in 1BL8, probe 0.9~\AA
-(a) Tunnels in 1CQW,      & (b) 1CQW tunnels only & (c) Start positions in  & (d) Tunnel \#8 in 1BL8, \\
-probe 0.7~\AA            &                       & 1BL8        &     probe 0.9~\AA \\                                 
-\end{tabular}
-}
-\caption{\label{fig::pic1}
-Tunnels detected by \TA\ are depicted in orange, green spheres show tunnels computed by CAVER 3.0.
-}
-\end{figure*}
-
-{\bf Tunnel detection in dynamic protein: }
-A sequence of 1,000 frames of protein Haloalkane dehalogenase (DhaA) was analyzed with $\probe=0.9$.
-Due to protein dynamics, the tunnels change rapidly and they often disappear or merge with others.
-Each frame was analyzed with Alg.~\ref{alg::tunnel} with $\Imax=1000$ iterations.
-Fig.~\ref{fig::dyn} shows visualization of selected frames.
-In comparison to CAVER, that clusters the first tunnels found in each frame to a single 3D tunnel, the proposed
-method can compute behavior or each detected tunnel separately.
-Due to the CAVER's implicit clustering of the first tunnels, we do show any comparison of tunnels.
-The total time to process the whole sequence was $\sim\!\!2$ minutes (@ 2.9 GHz), while CAVER took $\sim\!70$ minutes.
-Due to space limits, only few images are shown in this section. 
-We refer to \url{http://mrs.felk.cvut.cz/romoco2017} for more information.
-
-\begin{figure*}
-\centering
-\begin{tabular}{cccc}
-\includegraphics[width=0.2\textwidth]{fig/dcpmld2-6-label} &
-\includegraphics[width=0.2\textwidth]{fig/dcpmld2-12} &
-\includegraphics[width=0.2\textwidth]{fig/dcpmld2-19} &
-\includegraphics[width=0.2\textwidth]{fig/dcpmld2-20} \\
-Frame 6 & Frame 12 & Frame 19 & Frame 20 \\                       
-\includegraphics[width=0.2\textwidth]{fig/dcpmld2-48} &
-\includegraphics[width=0.2\textwidth]{fig/dcpmld2-54} &
-\includegraphics[width=0.2\textwidth]{fig/dcpmld2-55} &
-\includegraphics[width=0.2\textwidth]{fig/dcpmld2-56} \\
-Frame 48 & Frame 54 & Frame 55 & Frame 56 \\                       
-\end{tabular}
-\caption{\label{fig::dyn}
-Evolution of void space in molecular dynamics of the DHA protein. 
-The active site is highlighted by the red circle in the top left image.
-In the frame 20, the single tunnel is separated into three smaller tunnels.
-Two of them temporarily connect in the frame 55, but then they remain unconnected.
-}
-\end{figure*}
 
@@
 The problem of tunnel detection in protein structures can be formulated as a path planning problem which can be solved by RRT.
-%The paper shows how to utilize the RRT planner for this task.
-%In comparison to existing techniques for tunnel detection, which are based on Voronoi Diagrams, the proposed method can find tunnels in a sequence of protein dynamics.
 The tunnels have properties of narrow passages and it is therefore challenging to find path inside the protein.

### 2017-05-04 paper.tex
added algorithm

@@
 Ligands are small molecules interacting with the protein. 
-Ligands are not rigid structures, they rather change due to interactions with the protein's atoms.
-Flexible ligands can be modeled as a kinematic chain with predefied degrees of freedom.
-An alternative approach, that is used in this work, is to consider a predefined set of possible conformations.
-The conformations are prepared using Rosetta software TODO. and their are denoted $c$.
+Due to the interaction, the relative positions of the ligand's atoms can change.
+the ligands  can be modeled as a multi-link kinematic chain where the torional angles phi/psi can change in a given range~\cite{songPFpath}.
+After the torsional angles are set, 3D position of each atom of the ligand is computed.
+By changing the angles, various conformations can be generated.
+However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
+
+To decrease the computational burgen, a set of feasible conformations $C$ is prepared.
+The conformations can be prepared using Molecular Dynamics software such as Rosetta.
+For each conformation $c\in C$ an energy potential is also calculated.
+The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
+can be switched to the confomration $c_j \in C$.
+
+The proteins are dense structures with narrow tunnels.
+Depending on the protein, the tunnel bottlenecks can be even less then $1$~\AA, which is too narrow for ligands with more than 2 atoms.
+For the purpose of traversability study, the radii of the ligands are scaled down by a factor $s$.
+
+A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,c,s)$ is described
+by the 3D position $(x,y,z)$ of the reference point of the ligand (usually it's geometric center), rotation around $x$, $y$ and $z$ axes,
+index of the conformation $c$ and scale of atom radii $s$.
+All possible configurations form the configuration space $\C$. 
+Let $\CF \in \C$ denote the set of collision-free configurations. 
+A configuration is collision-free if there is no intersection between the ligand's atoms scaled by the actual scale factor and the protein atoms.
 
 %We further consider a spherical probe $\Sprobe$  of radius $\probe$ encapsulating the ligand.
-Let $q=(x,y,z)\in\C$ denote the configuration (position) of the probe, where
-the configuration space $\C$ consists of all possible configurations.
-For each configuration $q\in\C$ we assume, that its largest collision-free radius $r(q) \in \mathbf{R}, r(q)\ge 0$ can be computed, e.g.
-using collision detection.
+%Let $q=(x,y,z)\in\C$ denote the configuration (position) of the probe, where
+%the configuration space $\C$ consists of all possible configurations.
+%For each configuration $q\in\C$ we assume, that its largest collision-free radius $r(q) \in \mathbf{R}, r(q)\ge 0$ can be computed, e.g.
+%using collision detection.
 %using collision detection.
@@
 %A probe is represented by one sphere of radius $\probe$.
-The collision-free region $\CF \subseteq \C$ is formed by configurations, where $\Sprobe$ can be placed without any collision, i.e., 
-$\CF = \{q \in \C | \Sprobe(q) \cap \SS = \emptyset\}$, where $\Sprobe(q)$ is the sphere with radius $\probe$ at configuration $q$.
-%We assume, that atoms located on the surface of the molecule are known, these are reffered to as {\sl surface atoms}.
-The distance $\dist(q_1,q_2)$  between two configurations $q_1,q_2\in\C$ is measured using 3D Euclidean metric.
+%The collision-free region $\CF \subseteq \C$ is formed by configurations, where $\Sprobe$ can be placed without any collision, i.e., 
+%$\CF = \{q \in \C | \Sprobe(q) \cap \SS = \emptyset\}$, where $\Sprobe(q)$ is the sphere with radius $\probe$ at configuration $q$.
+%The distance $\dist(q_1,q_2)$  between two configurations $q_1,q_2\in\C$ is measured using 3D Euclidean metric.
 
 %This radius can be detected using collision detection method.
-The configurations outside the protein $\CG \subseteq \C$ are collision-free with sphere $\Sgprobe$ of radius $\gprobe$, $\gprobe> \probe$,
-$\CG=\{q\in \C| \Sgprobe(q) \cap \SS = \emptyset \}$.
+%The configurations outside the protein $\CG \subseteq \C$ are collision-free with sphere $\Sgprobe$ of radius $\gprobe$, $\gprobe> \probe$,
+%$\CG=\{q\in \C| \Sgprobe(q) \cap \SS = \emptyset \}$.
 %A configuration $q \in \CG$ is referred to as {\sl goal configuration} in the rest of the paper.
-By computing $\alpha$-shape of the molecule, we can identify atoms located at the surface. 
-These atoms are referred to as {\sl surface atoms} in the rest of the paper.
-The distance between a configuration $q \in \C$ and its nearest surface atom is referred to as $\dists(q)$.
+%By computing $\alpha$-shape of the molecule, we can identify atoms located at the surface. 
+%These atoms are referred to as {\sl surface atoms} in the rest of the paper.
+%The distance between a configuration $q \in \C$ and its nearest surface atom is referred to as $\dists(q)$.
 %We assume, that for each configuration $q$, the largest collision-free radius $r(q)$ can be computed.
@@
 
-Dynamic proteins are represented by a sequence of $n$ snapshots (frames).
-The tunnels change with the protein dynamics as well: their width and position change, they can merge with other tunnels
-or even disappear.
-The free region $\CF$ is therefore different in each frame, as positions of atoms change.
-For the sake of simplicity, a single notation $\C$ resp. $\CF$ is used in this paper and it is valid for the frame being processed.
+%Dynamic proteins are represented by a sequence of $n$ snapshots (frames).
+%The tunnels change with the protein dynamics as well: their width and position change, they can merge with other tunnels or even disappear.
+%The free region $\CF$ is therefore different in each frame, as positions of atoms change.
+%For the sake of simplicity, a single notation $\C$ resp. $\CF$ is used in this paper and it is valid for the frame being processed.
 
-Approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing ordinary Voronoi diagram.
-Computing WVD this way is more efficient and numerically stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
-Vertices of this WVD are filtered out if their distance to an atom is less than probe radius $\probe$.
-Let $\VV$ contain the remaining Voronoi vertices, and let $r_v \in \VV$ denote the distance to the nearest atom.
+%Approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing ordinary Voronoi diagram.
+%Computing WVD this way is more efficient and numerically stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
+%Vertices of this WVD are filtered out if their distance to an atom is less than probe radius $\probe$.
+%Let $\VV$ contain the remaining Voronoi vertices, and let $r_v \in \VV$ denote the distance to the nearest atom.
 %The remaining Voronoi vertices $\VV$ of the WVD and their distances $r_v, v\in \VV$ to the nearest atom are then used in the following method.
-
-%\section{Tunnel detection}
-
-%The tunnel detection using motion planning leads to search in 4-dimensional $\C\!\times\!time$ space, which can be searched using randomized sampling-based methods.
-%The task of tunnel detection is to find a collision-free path from a given starting point located inside a dynamic protein 
-%to the outer environment. %: for a probe $\Sprobe$.
-%The idea of the proposed approach is to detect tunnels in each frame separately using a novel modification of RRT algorithm.
-%The configuration tree built in a given frame is transferred to the next frame, where the search continues.
-
-%We propose to detect the tunnels in this 4D space
-%(the space consists of pairs $(q,t) \in \C \times [1,n]$, where $n$ is the number of frames) by randomized sampling.
-%where $\C$ is three-dimensional configuration space
-%of the spherical probe and $n$ is the number of frames.
-%It is not computationally feasible to discretize the space to a grid, in which a path is found using standard space-search methods.
-%Moreover, finding a collision-free path in the space does not solve the problem, because the probe cannot move back in time.
-%Tunnel detection in dynamic proteins leads to search in $\C \times time$ space. %~\cite{Kindel00kinodynamic}.
-%sometimes called state-time space~\cite{berg2005roadmap}.
-%The proposed solution samples the configuration space of each frame separately in order to detect possible tunnels in the frame.
-%The resulting tree is then pruned to remove collision with atoms of the next frame.
 
@@
 %\setstretch{\straa}
-{\small
-\caption{\label{alg::tunnel}static tunnel detection}
-\KwIn{Initial configuration tree $\T$,
-    initial set of blocking spheres $\SB'$
+\caption{\label{alg::tunnel}traversability planning}
+\KwIn{
+    TODO
 }
-\KwData{goal-probe radius $\gprobe$,
-    probe radius $\probe$, 
-    max. number of iterations $\Imax$, 
-    phase switching threshold $\alpha$,
-    threshold $\ths$,
-    num. of new nodes $m$ in each iteration,
-    prob. $\gb$ of sampling randomly from $\C$
+\KwData{
+    todo
 }
-\KwOut{Configuration tree $\T$,
-    list of tunnel exit-points $G$
+\KwOut{
+    todo
 }
 \hrule
-$\SB = \SB'$; //blocking atoms added during phase~1 \\
-$G = \emptyset$; // list of tunnel exit-points \\
-$phase=1$\;
-\For{$iteration = 1:\Imax$}{
-    \If(// Voronoi-based sample --- Section~\ref{sec::vbg}){$phase=1$}{
-        $\VVA =$ update active Voronoi points from $\VV$\;
-        $v =$ random point from $\VVA$\;
-        $\qrand =$ random point around $v$ in the distance $r_v$\;
-%        $a_1$ = random atom on surface\;
-%        $a_2$ = random atom on surface\;
-%        $\qrand$ = random point on line $|a_1, a_2|$\;
+$v = 0$; // index of the virtual goal\\
+$iteration = 0$\;
+\While{$iteration < \Imax$ and $v < N$}{
+    \eIf{$rand() < \gb$}{
+        $\qrand$ = random sample around $P[v]$\;
+    }{
+        $\qrand$ = random sample from $\C$\;
     }
-    \If{$rand() < \gb$}{
-        $\qrand$ = random sample from $\C$\;
-        $v = \emptyset$; //no Voronoi vertex used
+    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
+
+    $S = $expandConfigurations($\qnear$)\;
+
+    \For{$q \in S$}{
+        $\T$.addNode($q$)\;
+        $\T$.addEdge($\qnear, q$)\;
     }
 
-    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
-
-    \If{$v \ne \emptyset$ {\bf and} $\dist(\qnear, v) < \probe$} {
-        add $v$ to $\T$, set its parent to $\qnear$\;
-    }
-
-    $N\!=\!m$ collision-free nodes around $\!\qnear$ in the distance $\!\varepsilon$\;
-%    $(\qnew,r(\qnew))$ = expand($\T$,$\qnear$, $\SS \cup \SB$)\;
-    \For{$q \in N $}{
-        \eIf{$\dists(q) < \ths $ {\bf and} $phase=1$}{
-            $\SB = \SB \cup \{\Sgprobe(q)\}$ \;
-        }{
-            $q'$ = nearest node in $\T$ towards $q$\;
-            \If{$\dist(q,q') > \dist(q,\qnear)$}{
-                add $q$ to $\T$, set its parent to $\qnear$\;
+    \For{$i=N : v$}{
+        $d$ = nearest node in the tree to $P[i]$\;
+        \If{$d < DT$}{
+            $v = i$; // new virtual goal found\;
+            break\;
         }
     }
-        \If{ $r(q) > \gprobe$ }{
-            $G = G \cup \{ q \} $\;
 }
-    }
-    \If{$phase = 1$ {\bf and} $iteration > \alpha \Imax$}{
-        $phase = 2$\;
-        $\SB = \emptyset$; // block. atoms are not used in phase 2 \\
-    }
-}
-\return $(\T, G)$\; 
-}
+trajectory = extract from tree\;
 \end{algorithm}

### 2017-05-05 paper.tex
intro

@@
 \title{
-    Analysis of protein tunnel traversability for flexible ligands
+    Analysis of protein tunnel traversability for flexible ligands using motion planning
 }
@@
 \begin{abstract}
-Proteins are involved in many biochemical processes.
-%Behavior and properties of proteins as well as other bio-macromolecules are influenced by internal void space such as tunnels or cavities.
-Behavior of proteins is influenced by internal void space such as tunnels or cavities.
-Tunnels are paths leading from an active site inside the protein to its surface.
-Knowledge about tunnels and their evolution in time provides an insight into protein properties
-(e.g., stability).
-Sampling-based motion planning methods like Rapidly Exploring Random Tree (RRT) can be used to find the tunnels by
-sampling the corresponding configuration space.
-The main advantage of these methods is that they can be easily adapted to consider protein dynamics.
-%i.e., they can find tunnels in proteins moving in time.
-However, the inner void space of proteins is  very limited, 
-    which may decrease ability of RRT to find a solution due to the narrow passage problem.
-In this paper, we propose to generate the random samples using Voronoi Diagram of the atoms.
-A subset of Voronoi vertices is dynamically maintained to support generation of samples in promising regions inside the protein.
-
+%Proteins are involved in many biochemical processes.
+Behavior and properties of proteins as well as other bio-macromolecules are influenced by internal void space such as tunnels or cavities.
+%Behavior of proteins is influenced by internal void space such as tunnels or cavities.
+Tunnels are paths leading from an active site inside the protein to its surface and they act as transporting pathways
+to the active site.
+To achieve a desired propery of the proteins, the active sites has to be reached by small molecules (ligands).
+The widely used software tools for tunnel detection are based on Voronoi diagrams, which is suitable
+for detecting tunnel of spherical probes.
+It is however not easy to estimate if a given non-spherical ligand can pass a spherical tunnel.
+Moreover, ligands are not rigid structures, but their conformation changes due to interactions with the protein's atoms.
+In this paper, we propose to employ sampling-based motion planning techniques to analyze traversability of ligands in the spherical
+tunnels.
+To speed up the computation, the flexibility of ligands is modeled using a predefined set of conformations suitable (typical) for 
+the given ligand.
+We introduce a novel modification of RRT method, whose search in the corresponding configuration space is guided by
+the tunnel pathway.
 \end{abstract}
@@
 
-%For example, one task in protein engineering is to change selected properties of a protein, e.g.
+For example, one task in protein engineering is to change selected properties of a protein, e.g.
 its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
-%This is reached by studying so called tunnels in proteins which can serve as the transportation paths for the ligand from the outside environment to the active site or vice versa.
-%The active site is a specific site, usually deeply buried inside the protein, where the chemical reaction between the protein and ligand can undergo.
-%Since the importance of the presence of tunnels in proteins was revealed, the researchers started to study them intensively.
+This is reached by studying so called tunnels in proteins which can serve as the transportation paths for the ligand from the outside environment to the active site or vice versa.
+The active site is a specific site, usually deeply buried inside the protein, where the chemical reaction between the protein and ligand can undergo.
+Since the importance of the presence of tunnels in proteins was revealed, the researchers started to study them intensively.
 %Koudel\'{a}kov\'{a} et al.~\cite{Koudelakova2013} show that mutations of amino acids located directly around the tunnel improved the structural and kinetic stability of the studied protein substantially, while the surface mutations almost did not contribute to protein stabilization.
@@
 
-%On the other hand, surface mutations almost did not contribute to protein stabilization~\cite{Koudelakova2013}.
+The task of tunnel detection is to find a collision-free path leading from an active site inside the protein to its surface.
+An example is depicted in Fig.~\ref{fig::motiv}.
+As not all paths are biochemically relevant, only paths with a given minimal bottleneck have to be reported (the bottleneck is
+the radius of the smallest sphere that can traverse the whole tunnel)
 
-%The task of tunnel detection is to find a collision-free path leading from an active site inside the protein to its surface.
-%As not all paths are biochemically relevant, only paths with a given minimal bottleneck have to be reported (the bottleneck is
-%the radius of the smallest sphere that can traverse the whole tunnel)
+Available tools for tunnel detection compute the tunnels assuming spherical ligands (probes). 
+Bio-chemicists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel
+lengths and its bottlenect.
+Such decision is however not precise and require previous expertise with the domain.
+
+To provide more information about the traversability of the tunnels, it is necessary to consider a non-spherical model of 
+the ligand.
+Such task can be then solved using motion planning techniques.
+It is necessary to consider translation, rotation and possibly also other degrees of freedom of the ligands, that are determined
+by the torsional angles, the related configuration space, that need to be searched, is high-dimensional. 
+Sampling-based motion planning techniques can be used to find a trajectory of the ligand.
 
@@
 The path is usually searched for a spherical probe of a given radius, which serves as a proxy geometry for a ligand. 
-An example is depicted in Fig.~\ref{fig::motiv}.
 Early solutions focused on static molecules. 
@@
 %    many iterations are needed before a tree can be built through the passage~\cite{hannaWIS}.
+
 A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
@@
 brings advantage only if the tree is close to the region and if it can reach it.
+To utilize a knowledge about the environment in the sampling process, represented e.g. using a mexial axis or a 2D or 3D path,
+it is necessary to maintain the sampling regions according the progress of the tree.
+For example, random samples can be generated along a given workspace path~\cite{vonasek2009rrt}.
+After the tree reaches a path point, the sampling around the point is disabled and moved to the next point of 
+the path.
+Similarly in
+\cite{jorry14marrt,neddy2016dynnamic}
+ob-rrt
+rrt-retraction
 
@@
 
+\begin{algorithm}[h]
+%\setstretch{\straa}
+\caption{\label{alg::tunnel}expand()}
+\KwIn{
+   configuration $q$ to be expanded
+}
+%\KwData{
+%}
+\KwOut{
+    set $S$ of collision-free configurations reachable from $q$
+}
+\hrule
+$c = $ conformation of $q$\;
+\ForEach{$c_j \in C$}{
+    \If{$canSwtich(c,c_j)$}{
+        \ForEach{$s \in scales$}{
+            \For{i = 1:m}{
+                $q' = q$\;
+                translate by random around $q'$\;
+                rotate randomly \;
+                $q'.c = c_j$\;
+                $q'.s = s$\;
+                \If{$isCollisionFree(q')$}{
+                    $S = S \cup \{ q \}$\;
+                }
+            }
+        }
+    }
+}
+\end{algorithm}
+
 \section{Experimental verification}

### 2017-06-05 paper.tex
new latex template, new refs to planning for deformable objects

@@
-%\documentclass[a4paper,conference]{IEEEtran}
-%\documentclass[a4paper,10pt,conference]{ieeeconf}
-%\documentclass[usletter, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
-%\IEEEoverridecommandlockouts                              % This command is only needed if 
-%\overrideIEEEmargins                                      % Needed to meet printer requirements.
-
-\documentclass{llncs}
+%\documentclass{llncs}
+\documentclass{svmult}
 
@@
 
-\title{
-    Analysis of protein tunnel traversability for flexible ligands using motion planning
+\title*{
+    Testing traversability of protein tunnels with flexible ligands using motion planning
 }
@@
 }
+
+\makeindex
 
@@
 
+% ==================================================================================
+
 \begin{document}
@@
 
-\begin{abstract}
-%Proteins are involved in many biochemical processes.
-Behavior and properties of proteins as well as other bio-macromolecules are influenced by internal void space such as tunnels or cavities.
-%Behavior of proteins is influenced by internal void space such as tunnels or cavities.
-Tunnels are paths leading from an active site inside the protein to its surface and they act as transporting pathways
-to the active site.
-To achieve a desired propery of the proteins, the active sites has to be reached by small molecules (ligands).
-The widely used software tools for tunnel detection are based on Voronoi diagrams, which is suitable
-for detecting tunnel of spherical probes.
-It is however not easy to estimate if a given non-spherical ligand can pass a spherical tunnel.
-Moreover, ligands are not rigid structures, but their conformation changes due to interactions with the protein's atoms.
-In this paper, we propose to employ sampling-based motion planning techniques to analyze traversability of ligands in the spherical
-tunnels.
-To speed up the computation, the flexibility of ligands is modeled using a predefined set of conformations suitable (typical) for 
-the given ligand.
-We introduce a novel modification of RRT method, whose search in the corresponding configuration space is guided by
-the tunnel pathway.
-\end{abstract}
+\abstract{
+%\begin{abstract}
+\red{TODO: rewrite}
+Proteins are involved in many biochemical processes and they function is achieved by interactions with other molecules such as ligands.
+The interaction takes place in an active site inside the protein, which is accesible from the surface by so called tunnels.
+The knowledge about tunnels can help researches for example to better select target ligand in order to achieve a desired
+behavior of the protein.
+Widely used tools for tunnel detection are based on Voronoi diagrams.
+This however requires to approximate the ligand by a sphere, which is not always realistic.
+In this paper, we propose a motion planning technique for analysis of traversability of such tunnels by flexible ligands.
+To decrease the dimension of the related configuration space, the flexibility of ligands is modeled by a set of expected conformations.
+The proposed method is based on the Rapidly Exploring Random Tree motion planning.
+The spheres representing the tunnel are used to guide the growth of the RRT tree in the configuration space.
+Something about visualization ...
+%\end{abstract}
+}
 
@@
 
+\red{prepsat, todle uz jsme meli vickrat}
 Protein structures are essential components of all living organisms.
 Proper understanding of their structure and function is important in many fields, including drug design, agriculture, cosmetics, etc.
-Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. % which led to their better understanding. 
+Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. 
 Such investigation can incorporate computational methods which help to analyze the geometry of protein structure,  its shape, surface area, and inner void space. 
-%These parameters then suggest and influence protein behavior, namely its reactivity with other molecules.
-%consists of the analysis of protein structure, its shape, surface area, and inner void space. 
-%These parameters influence namely the protein reactivity. % which is one of the most often studied behavior of the protein.
+These parameters then suggest and influence protein behavior, namely its reactivity with other molecules.
 
-%Proteins interact with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.) which can lead to changes in properties of the protein or the interacting molecule.
-%Thanks to protein interaction with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.), 
-% we are able to change the properties of the protein or the interacting molecule. 
+Proteins interact with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.) which can lead to changes in properties of the protein or the interacting molecule.
+The chemical reaction between the protein and ligand takes places in so called active site, that is accessible 
+from the protein surface by a tunnel.
+
+The knowledge about the tunnels is important in many fields .. todo.
 For example, one task in protein engineering is to change selected properties of a protein, e.g.
 its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
-%This is reached by studying so called tunnels in proteins which can serve as the transportation paths for the ligand from the outside environment to the active site or vice versa.
-%For example, typical task in protein engineering is to change selected properties of a protein, e.g.
-its stability under different outer conditions [11] or activity of the protein towards other molecules [16].
 This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
 ligand from the outside environment to the active site or vice versa.
-The active site is a specific site, usually deeply buried inside the protein, 
-    where the chemical reaction between the protein and ligand can undergo. 
-The importance of tunnels can be demonstrated on the study where mutations of amino acids located directly around the tunnel substantially improved the structural and kinetic stability of the studied protein, while
-surface mutations almost did not contribute to the stabilization~\cite{Koudelakova2013}.
 
-For example, one task in protein engineering is to change selected properties of a protein, e.g.
-its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
-This is reached by studying so called tunnels in proteins which can serve as the transportation paths for the ligand from the outside environment to the active site or vice versa.
-The active site is a specific site, usually deeply buried inside the protein, where the chemical reaction between the protein and ligand can undergo.
-Since the importance of the presence of tunnels in proteins was revealed, the researchers started to study them intensively.
-%Koudel\'{a}kov\'{a} et al.~\cite{Koudelakova2013} show that mutations of amino acids located directly around the tunnel improved the structural and kinetic stability of the studied protein substantially, while the surface mutations almost did not contribute to protein stabilization.
-%The importance of tunnel can be demonstrated on an example, 
-% where mutations of amino acids located directly around the tunnel improved the structural 
-%and kinetic stability of the studied protein substantially, while surface mutations almost did not contribute to
-%the protein stabilization~\cite{Koudelakova2013}.
-
-The task of tunnel detection is to find a collision-free path leading from an active site inside the protein to its surface.
-An example is depicted in Fig.~\ref{fig::motiv}.
+Generally, the task of tunnel detection is to find a collision-free path leading from the active site to the protein's surface. 
+The path is usually searched for a spherical probe of a given radius, which serves as a proxy geometry for a ligand. 
 As not all paths are biochemically relevant, only paths with a given minimal bottleneck have to be reported (the bottleneck is
 the radius of the smallest sphere that can traverse the whole tunnel) 
-
 Available tools for tunnel detection compute the tunnels assuming spherical ligands (probes). 
-Bio-chemicists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel
-lengths and its bottlenect.
+Bio-chemicists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottlenect.
 Such decision is however not precise and require previous expertise with the domain.
 
-To provide more information about the traversability of the tunnels, it is necessary to consider a non-spherical model of 
-the ligand.
+To provide more information about the traversability of the tunnels, it is necessary to consider a non-spherical model of the ligand.
 Such task can be then solved using motion planning techniques.
@@
 
-Generally, the task of tunnel detection is to find a collision-free path leading from the active site to the protein's surface. 
-The path is usually searched for a spherical probe of a given radius, which serves as a proxy geometry for a ligand. 
-Early solutions focused on static molecules. 
-However, from a static snapshot it is hard to assess the biochemical relevance of the tunnel as it does not
-provide information about its temporal stability.
+%Early solutions focused on static molecules. 
+%However, from a static snapshot it is hard to assess the biochemical relevance of the tunnel as it does not provide information about its temporal stability.
 
 Recently, researchers started to focus on molecular dynamics simulations and study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
-%The path is searched for a spherical probe of a given radius.
-%Early algorithms focused on static molecules, % and the tools were able to detect dozens of putative tunnels.
-%but it was hard to assess their biochemical relevance because from a static molecule it is impossible to derive the temporal stability
-%of the tunnels.
 %Recently, researchers started to focus on molecular dynamics simulations and studying the behavior of individual 
@@
 %In such cases the biochemists have to skip each n-th snapshot or select a subset of the whole simulation and perform the analysis only for this selection.
-%Such analysis however provides only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted. 
-%Recently, we have proposed a novel method for tunnel detection using sampling-based motion planning [23].
-In this method, RRT is used to ?nd tunnels in each frame and to consider the molecular dynamics, the tree is continuously pruned in each frame.
-In comparison to existing approaches [17], [18], this approach [23] handles the dynamics without need to cluster and match Voronoi diagrams from consecutive frames.
 
-Curently available tools for protein analysis support only the tunnel detection both on single protein snapshots and sequences
-of molecular dynamics.
 The properties of tunnels like bottleneck radius, their length and the surrounding residues can be used
@@
    the probability of expanding the tree through a narrow passage is small~\cite{hannaWIS}.
+
+RRT--Retraction~\cite{zhangRetraction} improves sampling in the narrow passages by retracting the newly added configuration
+closer to obstacles.
+%for motion planning of 3D objects.
+After $\qrand$ is generated and its nearest neighbor $\qnear\in \T$ is found, the segment from $\qrand$ to $\qnear$ is checked for collision.
+The tree is extended normally if the segment is free, otherwise the retraction step is performed.
+The task of the retraction step is to find a contact configuration  that minimizes the distance to $\qrand$.
+The contact configuration $q$ is found on the line and its neighborhood is searched to find another contact configuration $q'$ such that 
+$\varrho(q',\qrand) < \varrho(q,\qrand)$.
+The retraction is terminated after predefined number of steps or if no new contact configuration can be found.
+%The configuration trees built by RRT--Retraction can therefore contain more nodes than is the number of planning iterations,
+%as the tree can be expanded by upto $\Nrret$ contact configurations in each iteration.
+%An example of the retraction is depicted in Fig.~\ref{fig::rrtretraction}.
+%The RRT-Retraction was designed for path planning of 3D objects, where the straight-line local planner is used to connect two configurations, which allows us to find the contact configurations by random sampling.
+The RRT--Retraction can quickly penetrate to narrow passages, but its computational burden is significantly
+increased due to necessity to find the contact configurations.
+%The method was designed mainly for 3D path planning, but it can be combined with other planners
+%for motion planning of many-DOF robots~\cite{pan2010retraction}.
+%    e.g. for articulated robots~\cite{pan2010retraction}.
+%RRT--Retraction combined with decomposition-based path planners for articulated robots with many DOF was proposed in~\cite{pan2010retraction}.
+
+%The narrow passage problem has been studied intensively 
 %Due to its low volume, the probability of expanding the nodes towards the passage is small~\cite{hannaWIS}.
@@
 brings advantage only if the tree is close to the region and if it can reach it.
-To utilize a knowledge about the environment in the sampling process, represented e.g. using a mexial axis or a 2D or 3D path,
+To utilize a knowledge about the environment in the sampling process, 
+   %represented e.g. using a mexial axis or a 2D or 3D path,
 it is necessary to maintain the sampling regions according the progress of the tree.
@@
 the path.
-Similarly in
-\cite{jorry14marrt,neddy2016dynnamic}
-ob-rrt
-rrt-retraction
+Similar approach is used in~\cite{denny2014marrt}, where the sampling regions are maintained with a grapth describing topology 
+of the workspace.
+OB-RRT~\cite{amatoOBRRT} defines several types of tree expansions
+steps based on information about obstacles (e.g. using vectors of obstacles, tangential vectors and information about medial axis).
 
-In this paper, we propose how to combine RRT-based search with Voronoi-based methods in the task of tunnel detection. 
-To guide the configuration tree using Voronoi diagram, a subset of Voronoi vertices is determined and
-the samples are generated around vertices in this subset.
-The subset is maintained according the distance from the tree, so the random samples
-are generated only if they are in a suitable distance from the tree.
-This allows us to sample simultaneously along all promissing parts of VD that are accessible from the tree, which is 
-suitable for detection of multiple tunnels.
+deformable objects in many fields like computer graphics, robotics and medicine~\cite{alterovitz2008motion}
+clothing, rope, human tissue
 
-%In this paper, we propose how to combine the VD-based methods and sampling-based approaches for tunnel detection.
-%Instead of uninformed sampling of the configuration space, which has been proposed in~\cite{vonasek2016application}, we propose
-%to generate the random samples using Voronoi diagrams.
+a two-stage approach was presented in~\cite{bayazit2002deformable}. 
+First a path is found with reduced-scale of the robot, or by allowing the robot to penetrate obstacles.
+The roadmap can therefore contain configurations that are infeasible for the rigid version of the robot
+After finding a path, the infeasible configuraions are determined and the robot is deformed there to avoid the collisions.
+the path is refined to remove the collisions
+geometric deformation -- only part of the object that is colliding is deformed,
+and boundix box (first bounding box is deformed and the the model is deformed according to the deformation compute for the bbox)
+
+path planning with non-rigid obstacles~\cite{frank2008efficient}.
+the majority of path planning approached focus on rigid robots and rigid obstacles.
+
+\cite{gayle2005path},planning for a flexible robot in complex environments. 
+algorithm  computes a collision free path by taking into account geometric and physical constraints, including obstacle avoidance, non-
+penetration constraint, volume preservation, surface tension, and energy minimization.
+
+new algorithm for collision detection between a deformable robot and fixed obstacles using  graphics processors. 
+We also present techniques to efficiently handle complex deformable models composed of tens of thousands of polygons and obtain significant performance improvement    over previous approaches. 
+practical application of our algorithm in performing path planning of catheters in liver chemoembolization.
+
+planning for flexible objects~\cite{lamiraux2001flexible},
+
+ f-PRM \cite{kavraki1998towards} framework for flexible robots of simple geometrix shapes\cite{anshelevich2000deformable} or surface patches~\cite{holleman1998planning}.
+
+other planning for defomrable: \cite{anshelevich2000deformable,bayazit2002deformable,gayle2005path}
+deformable envinorments: \cite{rodriguez2006planning,frank2008efficient,phillips2014representation}
+
+representation of deformation~\cite{phillips2014representation}
+
+Robots can be deformed either considering their physical propersites or not.
+
+geometrical or physical-based deformation approaches
+The non-physical deformations include basic deformations  like scale, bend or twist~\cite{barr1984global},
+    free-form deformation~\cite{gibson19973d,sederberg1986free},
+    Physical-based deformations can be achiebed using mass-spring and finite-element methods.
+
+f-PRM~\cite{anshelevich2000deformable}, physically coreect deformations, time consuming
 
@@
 %\setstretch{\straa}
-\caption{\label{alg::tunnel}expand()}
+\caption{\label{alg::tunnel2}expand()}
 \KwIn{

### 2017-06-05 paper.tex
sota

@@
 \def\qinit{\qstart}
+\def\qgoal{\qstart}
 \def\qnear{q_{near}}
@@
 %\begin{abstract}
-\red{TODO: rewrite}
 Proteins are involved in many biochemical processes and they function is achieved by interactions with other molecules such as ligands.
@@
 
-Generally, the task of tunnel detection is to find a collision-free path leading from the active site to the protein's surface. 
-The path is usually searched for a spherical probe of a given radius, which serves as a proxy geometry for a ligand. 
-As not all paths are biochemically relevant, only paths with a given minimal bottleneck have to be reported (the bottleneck is
+The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or 
+a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
+These methods compute the tunnels assuming a spherical ligand (probe).
+The resulting tunnels are represented as a sequence of spheres.
+The active sites are usually connected with the protein surface with multiple pathways.
+However, not all paths are biochemically relevant, so only paths with a given minimal bottleneck have to be reported (the bottleneck is
 the radius of the smallest sphere that can traverse the whole tunnel) 
-Available tools for tunnel detection compute the tunnels assuming spherical ligands (probes). 
-Bio-chemicists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottlenect.
-Such decision is however not precise and require previous expertise with the domain.
-
-To provide more information about the traversability of the tunnels, it is necessary to consider a non-spherical model of the ligand.
-Such task can be then solved using motion planning techniques.
-It is necessary to consider translation, rotation and possibly also other degrees of freedom of the ligands, that are determined
-by the torsional angles, the related configuration space, that need to be searched, is high-dimensional. 
-Sampling-based motion planning techniques can be used to find a trajectory of the ligand.
-
-%Early solutions focused on static molecules. 
-%However, from a static snapshot it is hard to assess the biochemical relevance of the tunnel as it does not provide information about its temporal stability.
-
-Recently, researchers started to focus on molecular dynamics simulations and study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
-%Recently, researchers started to focus on molecular dynamics simulations and studying the behavior of individual 
-%tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
-Molecular dynamics is represented as a sequence of frames (molecule snapshots), and the tunnels need to be detected through these frames.
-%With the increasing size of simulations the correctness of predicting the most biochemically relevant tunnel increases as well.
-Existing solutions based exclusively on Voronoi diagrams and clustering methods are time and memory consuming
-which also gives the limitation for the maximum number of frames which they are able to analyze.
-In such cases, the biochemists have to select a subset of the whole simulation and perform the analysis only for this selection.
-Such analysis however provides only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted.
-%This opens new possibilities for alternative solutions enabling to explore the tunnels in each frame of the molecular dynamics.
-%In such cases the biochemists have to skip each n-th snapshot or select a subset of the whole simulation and perform the analysis only for this selection.
-
-The properties of tunnels like bottleneck radius, their length and the surrounding residues can be used
-to estimate suitability of the tunnel for a ligand.
-However, such estimation still involve lot of experience.
+Bio-chemicists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck.
 
@@
 something about screeing: what is it, what's is the input and output, in which applications (research/industry) it is useful.
+
+Such decision is however not precise and it requires previous expertise with the domain.
+Of the reasons is that typical ligands are non-spherical and therefore it is difficult to estimate their traversability in tunnels
+computed for a spherical probe.
+For better decision, it is necessary to consider also the shape of the ligand being investigaed, and consequently, to also
+consider rotations of the whole ligand.
+Moreover, ligands are rather flexible molecules, so the relative positions of their atoms can change as well.
+
+Sampling-based motion planning methods such as Rapidly Exploring Random Tree~\cite{lavalleRRT} can be
+used to test the traversability of a given ligand in the tunnels.
+These methods can cope with objects of many degrees of freedom and arbitrary geometry.
+
+%Recently, researchers started to focus on molecular dynamics simulations and study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
+%Recently, researchers started to focus on molecular dynamics simulations and studying the behavior of individual 
+%tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
+%Molecular dynamics is represented as a sequence of frames (molecule snapshots), and the tunnels need to be detected through these frames.
+%With the increasing size of simulations the correctness of predicting the most biochemically relevant tunnel increases as well.
+%Existing solutions based exclusively on Voronoi diagrams and clustering methods are time and memory consuming
+%which also gives the limitation for the maximum number of frames which they are able to analyze.
+%In such cases, the biochemists have to select a subset of the whole simulation and perform the analysis only for this selection.
+%Such analysis however provides only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted.
+%This opens new possibilities for alternative solutions enabling to explore the tunnels in each frame of the molecular dynamics.
+%In such cases the biochemists have to skip each n-th snapshot or select a subset of the whole simulation and perform the analysis only for this selection.
+%The properties of tunnels like bottleneck radius, their length and the surrounding residues can be used
+%to estimate suitability of the tunnel for a ligand.
+%However, such estimation still involve lot of experience.
 
@@
 
-\red{TODO include}
-Recently, we have proposed a novel method for tunnel detection using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT) method~\cite{vonasek2016application}.
-In comparison to existing approaches~\cite{Petrek20071357,citeulike:6257975}, this RRT-based approach~\cite{vonasek2016application} 
-handles the dynamics without need to cluster and match Voronoi diagrams from consecutive frames.
-
-The analysis of protein structure aiming to reveal the tunnels has been supported by different computational software tools which take the geometry of the protein as an input and explore the inner void space (e.g., CAVER 1.0~\cite{citeulike:6257975} or MOLE~\cite{Petrek20071357}).
+The analysis of protein structure aiming to reveal the tunnels has been supported by different computational software tools which take the geometry of the protein as an input and explore the inner void space (e.g., CAVER 1.0~\cite{petrek2006caver} or MOLE~\cite{Petrek20071357}).
 Early methods for tunnel detection utilized a discretized 3D grid, where each cell is considered as occupied or free depending
@@
 Besides, the grid can be used to identify other relevant properties like 
-pockets, cavities, or channels~\cite{sehnal2013mole,citeulike:6257975}.
+pockets, cavities, or channels~\cite{sehnal2013mole,petrek2006caver}.
 %One of the first grid-based approaches to the detection of tunnels in protein is the CAVER 1.0 algorithm by Pet\v{r}ek et al.~\cite{citeulike:6257975}.
@@
 each atom is approximated by several spheres with small radius~\cite{yaffe2008,caver3}.
-%In this case, each atom is approximated by several spheres with small radius~\cite{yaffe2008,caver3}.
 %The VD-based methods remove one of the main limitations of the grid-based approach which is the dependency on the grid resolution.
-VD-based methods are memory less demanding, and also faster than grid-based methods.
+VD-based methods are memory less demanding, and also faster than the grid-based methods.
 %They can be easily used for static tunnel detection, as they are based on Voronoi diagrams of static frames.
@@
 
-The tunnel detection in dynamic proteins can also be formulated as path planning in 4D (i.e., 3D $\times$ time) 
-configuration space assuming a spherical probe.
+Finding tunnels for a general, non-spherical and possibly a flexible ligand can be solved by searching
+a feasible path in the corresponding configuration space $\C$.
+The dimension of the space is given by the degrees of freedom (6D for a solid ligand and 6D+$n$ for a flexible ligand with
+$n$ rotational bonds).
+The feasible configurations, where the ligand (robot) does not touch and obstacle, form the free region $\CF$.
+The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
+to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the begining and the end of the
+tunnel, respectively.
+Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
+%Sampling-based techniques like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a high-dimensional space.
+%The tunnel detection in dynamic proteins can also be formulated as path planning~\cite{vonasek2016application}. in 4D (i.e., 3D $\times$ time) 
+%configuration space assuming a spherical probe.
 %Tunnel detection in dynamic proteins leads to path planning 
 %in a high-dimensional configuration space (4D for the probe and 7D for the ligand).
-This space can be searched using sampling-based motion planning methods.
+%This space can be searched using sampling-based motion planning methods.
 %Instead of a systematic search, which would be time and memory consuming, randomized sampling can be used
@@
 %Sampling-based path planning methods like Probabilistic Roadmaps (PRM)~\cite{kavrakiPRM} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT} are able to find feasible paths for complex objects with arbitrary geometries.
-The idea of sampling-based motion planning is to randomly sample the configuration space $\C$ of the robot. % (i.e., the spherical probe in this case).
-%The configuration space is a space of all possible configurations of the robot.
+The idea of sampling-based motion planning is to randomly sample the configuration space.
 The random samples are classified as free or non-free using collision detection and  the free ones are stored into a roadmap.
@@
 %or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
+
 Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query motion planning method.
@@
 related to 
-protein folding~\cite{al2012motion,gipson2012computational,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro}, or analysis of loop motions~\cite{cortes2004geometric}.
+protein folding~\cite{al2012motion,gipson2012computational,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},analysis of loop motions~\cite{cortes2004geometric}, and even for tunnel detection~\cite{vonasek2016application,vonasek2017tunnel}.
 %rrt for molecular: \cite{al2012motion}
 %survey \cite{gipson2012computational}
-Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
-The most relevant paper is~\cite{guieysse2008structure}, where RRT is used to compute pathways for flexible ligands towards 
-the active site.
-This method however does not compute explicitly all tunnels, rather a single pathway.
+%In our previous works, we proposed to employ RRT also for the purp
+%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
 
-In our recent paper~\cite{vonasek2016application}, we have proposed to detect tunnels using RRT.
-In this approach, RRT samples the configuration space of a small spherical probe moving in a single frame of protein dynamics.
-After the tunnels are found in one frame, the built  configuration tree is transferred to the next frame of protein dynamics and pruned to remove
-nodes colliding with the new positions of atoms.
-This results in a sequence of connected trees, that can be searched for path representing the tunnels in the protein dynamics.
-
+The most relevant paper is~\cite{guieysse2008structure}, where RRT is used to compute pathways for flexible ligands towards the active site, but
+without the possibility to select a preferable path/tunnel.
+%This method however does not compute explicitly all tunnels, rather a single pathway.
 %This allows us to build a sequence of trees, where each tree corresponds to each frame of protein dynamics.
@@
 %This allows us to find paths in this sequence of tree between two given frames, that correspond to the dynamic tunnels.
+%In comparison to~\cite{guieysse2008structure}, the approach~\cite{vonasek2016application} assumes spherical probes, which is motivated
+%by practical reasons.
+%Flexible ligands have many degrees of freedom, which increases dimension of the configuration space to be searched and makes
+%the collision-detection more complex.
+%Moreover, it is required to define suitable metric and possibly also an efficient data structure for nearest-neighbor search.
+%Despite the flexibility of ligands, their movements in proteins are strongly limited, as they are typically larger than spherical probes.
+%Traversability of the ligand through a protein may depend on its initial orientation, 
+% which can further limit the growth of the configuration tree.
+%Contrary, configuration space of  spherical probes is only 3D with the possibility to employ Euclidean metric and KD-trees for the
+%nearest neighbor search.
+%It is more practical to first detect tunnels for a spherical probe and then analyze traversability of the actual shape of ligand.
+%In this case, the traversability analysis is limited only to a tunnel and its vicinity, which decreases volume of the configuration space and
+%decreases the number of atoms involved in collision detection.
 
-In comparison to~\cite{guieysse2008structure}, the approach~\cite{vonasek2016application} assumes spherical probes, which is motivated
-by practical reasons.
-%The tunnels can be computed for a spherical probe or even for non-spherical ligands --- small molecules.
-%Computing tunnels considering only spherical probes is more practical.
-%Despite the ability of sampling-based planners to find path for robots (objects) of arbitrary shapes, which is achieved by the collision detection, the above described approach~\cite{vonasek2016application} detect tunnels only for spherical ligands.
-%The reasons are practical.
-Flexible ligands have many degrees of freedom, which increases dimension of the configuration space to be searched and makes
-the collision-detection more complex.
-Moreover, it is required to define suitable metric and possibly also an efficient data structure for nearest-neighbor search.
-%Dimension of the configuration space of flexible ligands is higher and it is determined by the Degrees Of Freedom of the ligand.
-%It is required to define proper metric to efficiently search this space using RRT.
-%Planning patways for flexible ligands leads to a search in a high-di
-%Taking into account flexible ligands increases the dimension of the configuration space.
-Despite the flexibility of ligands, their movements in proteins are strongly limited, as they are typically larger than spherical
-probes.
-Traversability of the ligand through a protein may depend on its initial orientation, 
- which can further limit the growth of the configuration tree.
-Contrary, configuration space of  spherical probes is only 3D with the possibility to employ Euclidean metric and KD-trees for the
-nearest neighbor search.
-It is more practical to first detect tunnels for a spherical probe and then analyze traversability of the actual shape of ligand.
-In this case, the traversability analysis is limited only to a tunnel and its vicinity, which decreases volume of the configuration space and
-decreases the number of atoms involved in collision detection.
-%which is faster than searching path for the ligand in the full protein.
+In this paper, we propose to employ motion planning to compute the trajectory for a given ligand in a given tunnel.
+This brings several advantages.
+From motion planning point of view, the tunnels have properties of narrow passages.
+By focusing the motion planning only to the surrounding of a tunnel, the total volume of the configuration space decreases which increases
+the relative volume of the free space, which increases probability of placing random sampling in the perspective regions of the configruation 
+space.
 
-%The ability of RRT to quickly explore the configuration space is caused by the combination
-%of uniform sampling and the nearest-neighbor rule used to select nodes for the expansion.
+\red{comparison to momaligpath~\cite{cortes2005path} + \cite{guieysse2008structure} }
+
 The growth of the configuration tree into some regions of the configuration space can however be slowed down due to the narrow passages.
@@
 
-RRT--Retraction~\cite{zhangRetraction} improves sampling in the narrow passages by retracting the newly added configuration
-closer to obstacles.
-%for motion planning of 3D objects.
-After $\qrand$ is generated and its nearest neighbor $\qnear\in \T$ is found, the segment from $\qrand$ to $\qnear$ is checked for collision.
-The tree is extended normally if the segment is free, otherwise the retraction step is performed.
-The task of the retraction step is to find a contact configuration  that minimizes the distance to $\qrand$.
-The contact configuration $q$ is found on the line and its neighborhood is searched to find another contact configuration $q'$ such that 
-$\varrho(q',\qrand) < \varrho(q,\qrand)$.
-The retraction is terminated after predefined number of steps or if no new contact configuration can be found.
-%The configuration trees built by RRT--Retraction can therefore contain more nodes than is the number of planning iterations,
+%RRT--Retraction~\cite{zhangRetraction} improves sampling in the narrow passages by retracting the newly added configuration
+%closer to obstacles.
+%%for motion planning of 3D objects.
+%After $\qrand$ is generated and its nearest neighbor $\qnear\in \T$ is found, the segment from $\qrand$ to $\qnear$ is checked for collision.
+%The tree is extended normally if the segment is free, otherwise the retraction step is performed.
+%The task of the retraction step is to find a contact configuration  that minimizes the distance to $\qrand$.
+%The contact configuration $q$ is found on the line and its neighborhood is searched to find another contact configuration $q'$ such that 
+%$\varrho(q',\qrand) < \varrho(q,\qrand)$.
+%The retraction is terminated after predefined number of steps or if no new contact configuration can be found.
+%%The configuration trees built by RRT--Retraction can therefore contain more nodes than is the number of planning iterations,
 %as the tree can be expanded by upto $\Nrret$ contact configurations in each iteration.
@@
 %The RRT-Retraction was designed for path planning of 3D objects, where the straight-line local planner is used to connect two configurations, which allows us to find the contact configurations by random sampling.
-The RRT--Retraction can quickly penetrate to narrow passages, but its computational burden is significantly
-increased due to necessity to find the contact configurations.
+%The RRT--Retraction can quickly penetrate to narrow passages, but its computational burden is significantly
+%increased due to necessity to find the contact configurations.
 %The method was designed mainly for 3D path planning, but it can be combined with other planners
@@
 %    many iterations are needed before a tree can be built through the passage~\cite{hannaWIS}.
+%Recently, we have proposed a novel method for tunnel detection using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT) method~\cite{vonasek2016application}.
+%In comparison to existing approaches~\cite{Petrek20071357,petrek2006caver}, this RRT-based approach~\cite{vonasek2016application} 
+%handles the dynamics without need to cluster and match Voronoi diagrams from consecutive frames.
 
-A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
-in the difficult areas.
-For Probabilistic Roadmaps~\cite{kavrakiPRM}, this can be achieved simply by generating more samples in the difficult areas, e.g. along
-medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
+This technique is similar to generation of samples near the medial exist of the environment
+medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting,amatoOBRRT}.
+or by guiding the growth of the tree around a general path in the workspace~\cite{vonasek2009rrt,denny2014marrt}.
+
+%A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples in the difficult areas.
+%For Probabilistic Roadmaps~\cite{kavrakiPRM}, this can be achieved simply by generating more samples in the difficult areas, e.g. along
+%medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
 %This requires to precompute the medial axis, which can be time consuming.
@@
 %medial axis~\cite{hollemanMAPRM}.
-Generating random samples in difficult regions however does not ensure construction of a better tree in the case of RRT-based methods. 
-In RRT, the configuration tree is expanded in the direction of random samples, but the obstacles may prevent to reach them closely~\cite{vonasekphd}.
-Increased probability of sampling in a given region, e.g., in a narrow passage, 
-brings advantage only if the tree is close to the region and if it can reach it.
-To utilize a knowledge about the environment in the sampling process, 
-   %represented e.g. using a mexial axis or a 2D or 3D path,
-it is necessary to maintain the sampling regions according the progress of the tree.
-For example, random samples can be generated along a given workspace path~\cite{vonasek2009rrt}.
-After the tree reaches a path point, the sampling around the point is disabled and moved to the next point of 
-the path.
-Similar approach is used in~\cite{denny2014marrt}, where the sampling regions are maintained with a grapth describing topology 
-of the workspace.
-OB-RRT~\cite{amatoOBRRT} defines several types of tree expansions
-steps based on information about obstacles (e.g. using vectors of obstacles, tangential vectors and information about medial axis).
+%Generating random samples in difficult regions however does not ensure construction of a better tree in the case of RRT-based methods. 
+%In RRT, the configuration tree is expanded in the direction of random samples, but the obstacles may prevent to reach them closely~\cite{vonasekphd}.
+%Increased probability of sampling in a given region, e.g., in a narrow passage, 
+%brings advantage only if the tree is close to the region and if it can reach it.
+%it is necessary to maintain the sampling regions according the progress of the tree.
+%For example, random samples can be generated along a given workspace path~\cite{vonasek2009rrt}.
+%After the tree reaches a path point, the sampling around the point is disabled and moved to the next point of 
+%the path.
+%Similar approach is used in~\cite{denny2014marrt}, where the sampling regions are maintained with a grapth describing topology 
+%of the workspace.
+%OB-RRT~\cite{amatoOBRRT} defines several types of tree expansions
+%steps based on information about obstacles (e.g. using vectors of obstacles, tangential vectors and information about medial axis).
 
-deformable objects in many fields like computer graphics, robotics and medicine~\cite{alterovitz2008motion}
-clothing, rope, human tissue
+By reducing the scale of the ligand's atomic radii, the configuration tree can explore more areas of the configuration space.
+Such techniques has been used for motion planning of deformable objects, 
+e.g.,~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path} 
+and for motion planning of rigin objects in amongst 
+flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.
 
-a two-stage approach was presented in~\cite{bayazit2002deformable}. 
-First a path is found with reduced-scale of the robot, or by allowing the robot to penetrate obstacles.
-The roadmap can therefore contain configurations that are infeasible for the rigid version of the robot
-After finding a path, the infeasible configuraions are determined and the robot is deformed there to avoid the collisions.
-the path is refined to remove the collisions
-geometric deformation -- only part of the object that is colliding is deformed,
-and boundix box (first bounding box is deformed and the the model is deformed according to the deformation compute for the bbox)
+%deformable objects in many fields like computer graphics, robotics and medicine~\cite{alterovitz2008motion} clothing, rope, human tissue
+%a two-stage approach was presented in~\cite{bayazit2002deformable}. 
+%First a path is found with reduced-scale of the robot, or by allowing the robot to penetrate obstacles.
+%The roadmap can therefore contain configurations that are infeasible for the rigid version of the robot
+%After finding a path, the infeasible configuraions are determined and the robot is deformed there to avoid the collisions.
+%the path is refined to remove the collisions
+%geometric deformation -- only part of the object that is colliding is deformed,
+%and bounding box (first bounding box is deformed and the the model is deformed according to the deformation compute for the bbox)
+%path planning with non-rigid obstacles~\cite{frank2008efficient}.
 
-path planning with non-rigid obstacles~\cite{frank2008efficient}.
-the majority of path planning approached focus on rigid robots and rigid obstacles.
+%the majority of path planning approached focus on rigid robots and rigid obstacles.
+%\cite{gayle2005path},planning for a flexible robot in complex environments. 
+%algorithm  computes a collision free path by taking into account geometric and physical constraints, including obstacle avoidance, non-
+%penetration constraint, volume preservation, surface tension, and energy minimization.
 
-\cite{gayle2005path},planning for a flexible robot in complex environments. 
-algorithm  computes a collision free path by taking into account geometric and physical constraints, including obstacle avoidance, non-
-penetration constraint, volume preservation, surface tension, and energy minimization.
+%new algorithm for collision detection between a deformable robot and fixed obstacles using  graphics processors. 
+%We also present techniques to efficiently handle complex deformable models composed of tens of thousands of polygons and obtain significant performance improvement    over previous approaches. 
+%practical application of our algorithm in performing path planning of catheters in liver chemoembolization.
+%planning for flexible objects~\cite{lamiraux2001flexible},
 
-new algorithm for collision detection between a deformable robot and fixed obstacles using  graphics processors. 
-We also present techniques to efficiently handle complex deformable models composed of tens of thousands of polygons and obtain significant performance improvement    over previous approaches. 
-practical application of our algorithm in performing path planning of catheters in liver chemoembolization.
+% f-PRM \cite{kavraki1998towards} framework for flexible robots of simple geometrix shapes\cite{anshelevich2000deformable} or surface patches~\cite{holleman1998planning}.
 
-planning for flexible objects~\cite{lamiraux2001flexible},
+%other planning for defomrable: \cite{anshelevich2000deformable,bayazit2002deformable,gayle2005path}
+%deformable envinorments: \cite{rodriguez2006planning,frank2008efficient,phillips2014representation}
 
- f-PRM \cite{kavraki1998towards} framework for flexible robots of simple geometrix shapes\cite{anshelevich2000deformable} or surface patches~\cite{holleman1998planning}.
+%representation of deformation~\cite{phillips2014representation}
 
-other planning for defomrable: \cite{anshelevich2000deformable,bayazit2002deformable,gayle2005path}
-deformable envinorments: \cite{rodriguez2006planning,frank2008efficient,phillips2014representation}
-
-representation of deformation~\cite{phillips2014representation}
-
-Robots can be deformed either considering their physical propersites or not.
-
-geometrical or physical-based deformation approaches
-The non-physical deformations include basic deformations  like scale, bend or twist~\cite{barr1984global},
-    free-form deformation~\cite{gibson19973d,sederberg1986free},
-    Physical-based deformations can be achiebed using mass-spring and finite-element methods.
-
-f-PRM~\cite{anshelevich2000deformable}, physically coreect deformations, time consuming
+%Robots can be deformed either considering their physical propersites or not.
+%geometrical or physical-based deformation approaches
+%The non-physical deformations include basic deformations  like scale, bend or twist~\cite{barr1984global},
+%   free-form deformation~\cite{gibson19973d,sederberg1986free},
+%   Physical-based deformations can be achiebed using mass-spring and finite-element methods.
+%f-PRM~\cite{anshelevich2000deformable}, physically coreect deformations, time consuming
 
@@
 Due to the interaction, the relative positions of the ligand's atoms can change.
-the ligands  can be modeled as a multi-link kinematic chain where the torional angles phi/psi can change in a given range~\cite{songPFpath}.
+The ligands can be modeled as a multi-link kinematic chain where the torional angles phi/psi can change in a given range~\cite{songPFpath}.
 After the torsional angles are set, 3D position of each atom of the ligand is computed.
@@
 
+%from wiki:
+%Multiple static structures experimentally determined for the same protein in different conformations are often used to emulate receptor flexibility.[19] Alternatively rotamer libraries of amino acid side chains that surround the binding cavity may be searched to generate alternate but energetically reasonable protein conformations.[20][21]
+% 19 = \cite{totrov2008flexible}
+% 20 = \cite{hartmann2009docking}
+% 21 = \cite{taylor2003fds}
+
+Another approach is to model the ligand flexibility using a predefined set of feasible conformations $C$.
 To decrease the computational burgen, a set of feasible conformations $C$ is prepared.

### 2017-06-06 paper.tex
algorithms

@@
 \def\qnew{q_{new}}
+\def\T{\mathcal{T}}
+
 \def\C{\mathcal{C}}
-\def\T{\mathcal{T}}
 \def\CF{\mathcal{C}_{free}}
+\def\CFD{{\mathcal{C}^s_{free}}}
+\def\dt{d_{goal}}
+
 \def\Imax{I_{max}} %max number of iterations of RRT-based planners
@@
 
-\def\TA{$A_1$}
-\def\TB{$A_2$}
-\def\TC{$A_3$}
+\def\smin{s_{min}}
+\def\smax{s_{max}}
+\def\sdelta{{\Delta}_s}
 
@@
 \def\gb{p_{c}}
+
+\def\L{L}
 
@@
 %   can be detected as atoms located at $\alpha$-shape of the molecule.
-The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
-
-A tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, where $n$ denotes the number of spheres and
-$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius for a spherical probe centered
-at $t_i$. 
-The tunnels can be found by tunnel-detection tools like CAVER 3.0~\cite{caver3}.
+%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
+A tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, where $n$ denotes the number of spheres,
+$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
+The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.
 
@@
 
-Another approach is to model the ligand flexibility using a predefined set of feasible conformations $C$.
-To decrease the computational burgen, a set of feasible conformations $C$ is prepared.
-The conformations can be prepared using Molecular Dynamics software such as Rosetta.
-For each conformation $c\in C$ an energy potential is also calculated.
-The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
-can be switched to the confomration $c_j \in C$.
+%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
+%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
+%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
+%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
+%knowledge-based methods have also been used.[17]
+%13 = https://link.springer.com/article/10.1007/BF00123666
+
+Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
+To decrease the computational burgen, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
+%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
+%For each conformation $c\in \L$ an energy potential is also calculated.
+%Let $\L_i$ denote the set of conformations that 
+We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
+%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
+%can be switched to the confomration $c_j \in C$.
 
@@
 Depending on the protein, the tunnel bottlenecks can be even less then $1$~\AA, which is too narrow for ligands with more than 2 atoms.
-For the purpose of traversability study, the radii of the ligands are scaled down by a factor $s$.
+For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
 
-A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,c,s)$  is described
-by the 3D position $(x,y,z)$ of the reference point of the ligand (usually it's geometric center), rotation around $x$, $y$ and $z$ axes,
-index of the conformation $c$ and scale of atom radii $s$.
+A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
+by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., it's geometric center), rotation around $x$, $y$ and $z$ axes,
+index of the conformation $l\in \L$ and the actual scale $s$.
+A discrete set of scales is used, i.e., $s \in {\smin, \smin+\sdelta, \ldots, \smax}$, where 
+$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
 All possible configurations form the configuration space $\C$. 
-Let $\CF \in \C$ denote the set of collision-free configurations. 
-A configuration is collision-free if there is no intersection between the ligand's atoms scaled by the actual scale factor and the protein atoms.
+A configuration is collision-free if there is no intersection between the protein atoms and ligand atoms scaled by $s$.
+The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.
 
@@
 
-The configuration space is searched using the RRT principle~\cite{lavalleRRT}, which is extended by three main modifications.
-First, the tree expanded by several new nodes in each iteration, in order to boost growth in the narrow passages.
-The second modification is designed to block growth of the tree outside the protein, which is necessary to prevent detection of false
-tunnels. 
-The last modification is the novel VD-based sampling of configuration space.
+The configuration space is searched using the RRT principle~\cite{lavalleRRT}.
+The main loop of the method is described in Alg.~\ref{alg::main}.
+In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found
+and the node is expanded.
+The random samples are not generatd only from whole $\C$, as in the original RRT, but the sampling
+is guided using moving virtual goal~\cite{vonasek2009rrt}.
+The virtual goal $v$ is the index of the actual sphere of the tunnel, around which the samples
+are generated with probability $\gb$. 
+After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
+less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
+that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
+The algorithm termines after a predefined planning trials $\Imax$ or if the tree reaches
+the last sphere in the tunnel, i.e., when $v = N$.
 
@@
 %\setstretch{\straa}
-\caption{\label{alg::tunnel}traversability planning}
+\caption{\label{alg::main}main loop of the RRT planner}
 \KwIn{
-    TODO
+    tunnel $T=(t_i, r_i), n=1,\ldots N$ represented by $N$ spheres,
+    initial configuration $\qinit$,
+    goal configuration $\qgoal$,
 }
 \KwData{
-    todo
+   ligand conformations $\L$,
+   scale limits $\smin, \smax$ and $\sdelta$
 }
 \KwOut{
-    todo
+    configuration tree $\T$\;
 }
@@
     \eIf{$rand() < \gb$}{
-        $\qrand$ = random sample around $P[v]$\;
+        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
     }{
@@
 
-    $S = $expandConfigurations($\qnear$)\;
+    $S = $expandConfigurations($\qnear,\L$)\;
 
@@
 
-    \For{$i=N : v$}{
+    \For{$i=N : v$}{ \nllabel{alg::main:a}
         $d$ = nearest node in the tree to $P[i]$\;
@@
         }
+    } \nllabel{alg::main:b}
+    $iteration = iteration+1$\;
 }
-}
-trajectory = extract from tree\;
+\return $\T$\;
 \end{algorithm}
+
+The core of the method is the expansion procedure, which generates new collision-free nodes around $\qnear\in\T$.
+The algorithm is listed in Alg.~\ref{alg::expand}.
+For each conformation $l \in \L$, the expansion generates up to $m$ samples $q$ with maximal radii around $\qnear$.
+The 3D position of $q$ is generated using Normal distribution $N(0,\sigma^2)$ around $\qnear$, where $\sigma^2$ is the predefined
+parameter.
+The rotation part of $q$ is generated using $U(0,2\pi)$.
+The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated
+with $s$, the next lower $s$ is used.
+The expansion step is repeated for each conformation $l \in L$.
 
@@
 %\setstretch{\straa}
-\caption{\label{alg::tunnel2}expand()}
+\caption{\label{alg::expand}expand}
 \KwIn{
-   configuration $q$ to be expanded
+   configuration $\qnear$ to be expanded,
 }
-%\KwData{
-%}
+\KwData{
+   ligand conformations $\L$,
+   scale limits $\smin, \smax$ and $\sdelta$
+}
 \KwOut{
@@
 \hrule
-$c = $ conformation of $q$\;
-\ForEach{$c_j \in C$}{
-    \If{$canSwtich(c,c_j)$}{
-        \ForEach{$s \in scales$}{
-            \For{i = 1:m}{
+\ForEach{$l \in \L$}{
+    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
+        \For{$i = 1,\ldots,m$}{
             $q' = q$\;
-                translate by random around $q'$\;
+            translate $q$ by random around $q'$\;
             rotate randomly \;
-                $q'.c = c_j$\;
+            $q'.l = l$\;
             $q'.s = s$\;
@@
 }
-}
 \end{algorithm}
+
+\section{Visualization of results}
+
+The result of the proposed planning method is a tree of collision-free configurations, from which a path leading from $\qinit$
+to the nearest node towards $\qgoal$ can be found.
+Due to stochastic nature of RRT, it is not guaranteed that the tree always reached the goal configuration $\qgoal$.
+Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted fromm the tree.
+
+To present the results to chemicists ...
 
@@
 \section{Conclusion }
-
-The problem of tunnel detection in protein structures can be formulated as a path planning problem which can be solved by RRT.
-The tunnels have properties of narrow passages and it is therefore challenging to find path inside the protein.
-We propose to generate random samples around a subset of Voronoi Diagram of the protein's atoms.
-The subset is maintained automatically according to the distance to the configuration tree, so only the regions not reached yet
-by the tree are sampled.
-While the Voronoi diagram provides valuable information about suitable void regions in the protein, the
-proposed RRT-based technique can cope with the protein dynamics.
-The proposed technique can therefore be considered as an extension of the existing VD-based tools for tunnel detection.
 

### 2017-06-06 paper.tex
algorithms

@@
 
-\author{Vojt\v ech Von\' asek\inst{1} \and Barbora Kozl\'\i kov\'a\inst{2} \and Adam Jur\v{c}\'\i k\inst{2} \and Martin Saska\inst{1}}
+\author{
+    Vojt\v ech Von\' asek \and 
+    Barbora Kozl\'\i kov\'a  \and 
+    Adam Jur\v{c}\'\i k  \and 
+    David Bend\'{a}\v{r}  \and
+    Martin Saska }
+
 \institute{
+    V. Von\'{a}sek, Martin Saska \at 
     Faculty of Electrical Engineering,  
@@
 \and
+B. Kozl\'{\i}kov\'{a}, A. Jur\v{c}\'{\i}k \at
 Faculty of Informatics,  
 Masaryk University, 
-Botanick\'a 68a, 602 00 Brno,
+Botanick\'a 68a, 602 00 Brno, \\
+\and 
+\inst{3}
+TODO DAVID \at TODO DAVID
 }
@@
 \abstract{
-%\begin{abstract}
-Proteins are involved in many biochemical processes and they function is achieved by interactions with other molecules such as ligands.
-The interaction takes place in an active site inside the protein, which is accesible from the surface by so called tunnels.
+\red{needs rewrite:
+Function and behavior of proteins are achieved by interactions with other molecules such as ligands.
+The interaction takes place in an active site inside the protein, which is accessible from the surface by so called tunnels.
 The knowledge about tunnels can help researches for example to better select target ligand in order to achieve a desired
 behavior of the protein.
-Widely used tools for tunnel detection are based on Voronoi diagrams.
-This however requires to approximate the ligand by a sphere, which is not always realistic.
-In this paper, we propose a motion planning technique for analysis of traversability of such tunnels by flexible ligands.
-To decrease the dimension of the related configuration space, the flexibility of ligands is modeled by a set of expected conformations.
-The proposed method is based on the Rapidly Exploring Random Tree motion planning.
-The spheres representing the tunnel are used to guide the growth of the RRT tree in the configuration space.
-Something about visualization ...
-%\end{abstract}
+Widely used tools for tunnel detection are based on Voronoi diagrams and they compute the tunnels for a virtual spherical probe.
+The tunnels are then scored based on their bottleneck, length and also curvature. 
+It is however not easy to determine if a given non-spherical ligand can pass such a tunnel.
+To provide the chemists better information about the tunnels, the full shape of ligands should be considered.
+}
+Sampling-based motion planning techniques can be used to compute trajectories for the non-spherical ligands in the protein tunnels.
+In this paper, we propose a modification of Rapidly Exploring Random Tree (RRT) to find the trajectories.
+To enable the ligands to move in the tunnels, the planning is realized with a reduced-scale of the ligands. 
+The flexibility of ligands is achieved by using a predefined set of ligand conformations.
+For the purpose of further analysis, we computed trajectories are visualized ..
 }
 
-\section{Introduction}
+\section{Introduction \& motivation}
 
-\red{prepsat, todle uz jsme meli vickrat}
-Protein structures are essential components of all living organisms.
-Proper understanding of their structure and function is important in many fields, including drug design, agriculture, cosmetics, etc.
-Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. 
-Such investigation can incorporate computational methods which help to analyze the geometry of protein structure,  its shape, surface area, and inner void space. 
-These parameters then suggest and influence protein behavior, namely its reactivity with other molecules.
+\red{this is fast copy and paste from previous paper, please rewrite, cut to at ca. 1 page}
 
+%Protein structures are essential components of all living organisms.
+Proper understanding of structure and function of proteins is important in many fields, including drug design, agriculture, cosmetics, etc.
+%Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. 
+Computational methods are often used to analyze the geometry of protein structure,  its shape, surface area, and inner void space. 
 Proteins interact with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.) which can lead to changes in properties of the protein or the interacting molecule.
@@
 
-The knowledge about the tunnels is important in many fields .. todo.
+The knowledge about the tunnels is important in many fields..
 For example, one task in protein engineering is to change selected properties of a protein, e.g.
@@
 the radius of the smallest sphere that can traverse the whole tunnel) 
-Bio-chemicists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck.
+Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck.
 
 \red{TODO}
-something about screeing: what is it, what's is the input and output, in which applications (research/industry) it is useful.
+something about screening: what is it, what's is the input and output, in which applications (research/industry) it is useful.
 
@@
 computed for a spherical probe.
-For better decision, it is necessary to consider also the shape of the ligand being investigaed, and consequently, to also
+For better decision, it is necessary to consider also the shape of the ligand being investigated, and consequently, to also
 consider rotations of the whole ligand.
@@
 
-%hese approaches are not suitable for tunnel detection in dynamic proteins, a
-%, which limits the usage of these methods for spherical ligands.
-%To increase precision by 10, the number of cells increases by $10^3$.
-%Another disadvantage is that the 3D grid does not consider orientation of the ligand and therefore, the computations 
-%are useful only for spherical ligands.
-
 Currently the most widely used approach to tunnel detection is based on Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD).
@@
 The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
-to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the begining and the end of the
+to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the beginning and the end of the
 tunnel, respectively.
 Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
-%Sampling-based techniques like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a high-dimensional space.
-%The tunnel detection in dynamic proteins can also be formulated as path planning~\cite{vonasek2016application}. in 4D (i.e., 3D $\times$ time) 
-%configuration space assuming a spherical probe.
-%Tunnel detection in dynamic proteins leads to path planning 
-%in a high-dimensional configuration space (4D for the probe and 7D for the ligand).
-%This space can be searched using sampling-based motion planning methods.
-%Instead of a systematic search, which would be time and memory consuming, randomized sampling can be used
-%to obtain necessary information about the configuration space. %~\cite{Lav06}. 
-%Path planning approaches studied in robotics can be used to solve tunnel detection~\cite{guieysse2008structure} as well.
-%Sampling-based path planning methods like Probabilistic Roadmaps (PRM)~\cite{kavrakiPRM} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT} are able to find feasible paths for complex objects with arbitrary geometries.
 The idea of sampling-based motion planning is to randomly sample the configuration space.
@@
 A path in the roadmap then represents a motion in the workspace.
-%The usage of collision detection for classification of samples allows the application of sampling-based methods for path planning for robots (objects) of arbitrary shapes.
-%As the sampling-based methods work in the configuration space, they can cope with objects (robots) of many degrees of freedom
-% and arbitrary shapes.
-%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
-% to study proteins, e.g. in  
-%loop motions~\cite{cortes2004geometric},
-%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
-%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
-%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
-
 Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query motion planning method.
@@
 %Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
+%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
+% to study proteins, e.g. in  
+%loop motions~\cite{cortes2004geometric},
+%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
+%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
+%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
 
-The most relevant paper is~\cite{guieysse2008structure}, where RRT is used to compute pathways for flexible ligands towards the active site, but
+The most relevant paper is~\cite{cortes2005path}, where RRT is used to compute pathways for flexible ligands towards the active site, but
 without the possibility to select a preferable path/tunnel.
-%This method however does not compute explicitly all tunnels, rather a single pathway.
-%This allows us to build a sequence of trees, where each tree corresponds to each frame of protein dynamics.
-%The nodes of the trees are connected to their parents in the same frame, or to their parents in the previous frame.
-%This allows us to find paths in this sequence of tree between two given frames, that correspond to the dynamic tunnels.
-%In comparison to~\cite{guieysse2008structure}, the approach~\cite{vonasek2016application} assumes spherical probes, which is motivated
-%by practical reasons.
-%Flexible ligands have many degrees of freedom, which increases dimension of the configuration space to be searched and makes
-%the collision-detection more complex.
-%Moreover, it is required to define suitable metric and possibly also an efficient data structure for nearest-neighbor search.
-%Despite the flexibility of ligands, their movements in proteins are strongly limited, as they are typically larger than spherical probes.
-%Traversability of the ligand through a protein may depend on its initial orientation, 
-% which can further limit the growth of the configuration tree.
-%Contrary, configuration space of  spherical probes is only 3D with the possibility to employ Euclidean metric and KD-trees for the
-%nearest neighbor search.
-%It is more practical to first detect tunnels for a spherical probe and then analyze traversability of the actual shape of ligand.
-%In this case, the traversability analysis is limited only to a tunnel and its vicinity, which decreases volume of the configuration space and
-%decreases the number of atoms involved in collision detection.
 
@@
 By focusing the motion planning only to the surrounding of a tunnel, the total volume of the configuration space decreases which increases
-the relative volume of the free space, which increases probability of placing random sampling in the perspective regions of the configruation 
+the relative volume of the free space, which increases probability of placing random sampling in the perspective regions of the configuration 
 space.
@@
 \red{comparison to momaligpath~\cite{cortes2005path} + \cite{guieysse2008structure} }
+
+% in cortes2005path:
+%In this paper two kinds of large-amplitude motion are treated: protein loop conformational changes (involving pro-
+%tein backbone flexibility) and ligand trajectories to deep active sites in proteins (involving ligand and protein side-chain flex-
+%ibility). First studies performed using our two-stage approach (geometric search followed by energy refinements) show that,
+%    compared to classical molecular modeling methods, quite   similar results can be obtained with a performance gain of
+% several orders of magnitude. Furthermore, our results also  indicate that the geometric stage can provide highly valuable information to biologists.
 
@@
 
-%RRT--Retraction~\cite{zhangRetraction} improves sampling in the narrow passages by retracting the newly added configuration
-%closer to obstacles.
-%%for motion planning of 3D objects.
-%After $\qrand$ is generated and its nearest neighbor $\qnear\in \T$ is found, the segment from $\qrand$ to $\qnear$ is checked for collision.
-%The tree is extended normally if the segment is free, otherwise the retraction step is performed.
-%The task of the retraction step is to find a contact configuration  that minimizes the distance to $\qrand$.
-%The contact configuration $q$ is found on the line and its neighborhood is searched to find another contact configuration $q'$ such that 
-%$\varrho(q',\qrand) < \varrho(q,\qrand)$.
-%The retraction is terminated after predefined number of steps or if no new contact configuration can be found.
-%%The configuration trees built by RRT--Retraction can therefore contain more nodes than is the number of planning iterations,
-%as the tree can be expanded by upto $\Nrret$ contact configurations in each iteration.
-%An example of the retraction is depicted in Fig.~\ref{fig::rrtretraction}.
-%The RRT-Retraction was designed for path planning of 3D objects, where the straight-line local planner is used to connect two configurations, which allows us to find the contact configurations by random sampling.
-%The RRT--Retraction can quickly penetrate to narrow passages, but its computational burden is significantly
-%increased due to necessity to find the contact configurations.
-%The method was designed mainly for 3D path planning, but it can be combined with other planners
-%for motion planning of many-DOF robots~\cite{pan2010retraction}.
-%    e.g. for articulated robots~\cite{pan2010retraction}.
-%RRT--Retraction combined with decomposition-based path planners for articulated robots with many DOF was proposed in~\cite{pan2010retraction}.
-
-%The narrow passage problem has been studied intensively 
-%Due to its low volume, the probability of expanding the nodes towards the passage is small~\cite{hannaWIS}.
-%    many iterations are needed before a tree can be built through the passage~\cite{hannaWIS}.
-%Recently, we have proposed a novel method for tunnel detection using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT) method~\cite{vonasek2016application}.
-%In comparison to existing approaches~\cite{Petrek20071357,petrek2006caver}, this RRT-based approach~\cite{vonasek2016application} 
-%handles the dynamics without need to cluster and match Voronoi diagrams from consecutive frames.
-
 This technique is similar to generation of samples near the medial exist of the environment
@@
 
-%A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples in the difficult areas.
-%For Probabilistic Roadmaps~\cite{kavrakiPRM}, this can be achieved simply by generating more samples in the difficult areas, e.g. along
-%medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
-%This requires to precompute the medial axis, which can be time consuming.
-%Another approach is to shift the random samples towards the medial axis~\cite{amatoOBPRM} or into a direction of estimated
-%medial axis~\cite{hollemanMAPRM}.
-%Generating random samples in difficult regions however does not ensure construction of a better tree in the case of RRT-based methods. 
-%In RRT, the configuration tree is expanded in the direction of random samples, but the obstacles may prevent to reach them closely~\cite{vonasekphd}.
-%Increased probability of sampling in a given region, e.g., in a narrow passage, 
-%brings advantage only if the tree is close to the region and if it can reach it.
-%it is necessary to maintain the sampling regions according the progress of the tree.
-%For example, random samples can be generated along a given workspace path~\cite{vonasek2009rrt}.
-%After the tree reaches a path point, the sampling around the point is disabled and moved to the next point of 
-%the path.
-%Similar approach is used in~\cite{denny2014marrt}, where the sampling regions are maintained with a grapth describing topology 
-%of the workspace.
-%OB-RRT~\cite{amatoOBRRT} defines several types of tree expansions
-%steps based on information about obstacles (e.g. using vectors of obstacles, tangential vectors and information about medial axis).
-
 By reducing the scale of the ligand's atomic radii, the configuration tree can explore more areas of the configuration space.
@@
 e.g.,~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path} 
-and for motion planning of rigin objects in amongst 
+and for motion planning of rigid objects in among 
 flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.
@@
 %f-PRM~\cite{anshelevich2000deformable}, physically coreect deformations, time consuming
+
+\section{Preliminaries}
+
+Proteins are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
+Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
+%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
+%   can be detected as atoms located at $\alpha$-shape of the molecule.
+%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
+A tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, where $n$ denotes the number of spheres,
+$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
+The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.
+
+Ligands are small molecules interacting with the protein. 
+Due to the interaction, the relative positions of the ligand's atoms can change.
+The ligands can be modeled as a multi-link kinematic chain where the torsional angles phi/psi can change in a given range~\cite{songPFpath}.
+After the torsional angles are set, 3D position of each atom of the ligand is computed.
+By changing the angles, various conformations can be generated.
+However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
+
+%from wiki:
+%Multiple static structures experimentally determined for the same protein in different conformations are often used to emulate receptor flexibility.[19] Alternatively rotamer libraries of amino acid side chains that surround the binding cavity may be searched to generate alternate but energetically reasonable protein conformations.[20][21]
+% 19 = \cite{totrov2008flexible}
+% 20 = \cite{hartmann2009docking}
+% 21 = \cite{taylor2003fds}
+
+%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
+%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
+%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
+%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
+%knowledge-based methods have also been used.[17]
+%13 = https://link.springer.com/article/10.1007/BF00123666
+
+Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
+To decrease the computational burden, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
+%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
+%For each conformation $c\in \L$ an energy potential is also calculated.
+%Let $\L_i$ denote the set of conformations that 
+We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
+%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
+%can be switched to the confomration $c_j \in C$.
+
+The proteins are dense structures with narrow tunnels.
+Depending on the protein, the tunnel bottlenecks can be even less then $1$~\AA, which is too narrow for ligands with more than 2 atoms.
+For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
+
+A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
+by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., it's geometric center), rotation around $x$, $y$ and $z$ axes,
+index of the conformation $l\in \L$ and the actual scale $s$.
+A discrete set of scales is used, i.e., $s \in {\smin, \smin+\sdelta, \ldots, \smax}$, where 
+$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
+All possible configurations form the configuration space $\C$. 
+A configuration is collision-free if there is no intersection between the protein atoms and ligand atoms scaled by $s$.
+The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.
+
+\section{Ligand traversability}
+
+The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
+The tunnels are usually computed on proteins without ligands, which implies narrow void space.
+
+The configuration space is searched using the RRT principle~\cite{lavalleRRT}.
+The main loop of the method is described in Alg.~\ref{alg::main}.
+In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found
+and the node is expanded.
+The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
+is guided using moving virtual goal~\cite{vonasek2009rrt}.
+The virtual goal $v$ is the index of the actual sphere of the tunnel, around which the samples
+are generated with probability $\gb$. 
+After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
+less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
+that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
+The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
+the last sphere in the tunnel, i.e., when $v = N$.
+
+\linesnumbered
+\begin{algorithm}[h]
+%\setstretch{\straa}
+\caption{\label{alg::main}main loop of the RRT planner}
+\KwIn{
+    tunnel $T=(t_i, r_i), n=1,\ldots N$ represented by $N$ spheres,
+    initial configuration $\qinit$,
+    goal configuration $\qgoal$,
+}
+\KwData{
+   ligand conformations $\L$,
+   scale limits $\smin, \smax$ and $\sdelta$
+}
+\KwOut{
+    configuration tree $\T$\;
+}
+\hrule
+$v = 0$; // index of the virtual goal\\
+$iteration = 0$\;
+\While{$iteration < \Imax$ and $v < N$}{
+    \eIf{$rand() < \gb$}{
+        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
+    }{
+        $\qrand$ = random sample from $\C$\;
+    }
+    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
+
+    expand($\qnear$)\;
+
+   % \For{$q \in S$}{
+    %    $\T$.addNode($q$)\;
+    %    $\T$.addEdge($\qnear, q$)\;
+    %}
+
+    \For{$i=N : v$}{ \nllabel{alg::main:a}
+        $d$ = nearest node in the tree to $P[i]$\;
+        \If{$d < DT$}{
+            $v = i$; // new virtual goal found\;
+            break\;
+        }
+    } \nllabel{alg::main:b}
+    $iteration = iteration+1$\;
+}
+\return $\T$\;
+\end{algorithm}
+
+The core of the method is the expansion procedure, which generates new collision-free nodes around $\qnear\in\T$.
+The algorithm is listed in Alg.~\ref{alg::expand}.
+For each conformation $l \in \L$, the expansion generates up to $m$ samples $q$ with maximal radii around $\qnear$.
+The 3D position of $q$ is generated using Normal distribution $N(0,\sigma^2)$ around $\qnear$, where $\sigma^2$ is the predefined
+parameter.
+The rotation part of $q$ is generated using $U(0,2\pi)$.
+The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated
+with $s$, the next lower $s$ is used.
+The expansion step is repeated for each conformation $l \in L$.
+
+\begin{algorithm}[h]
+%\setstretch{\straa}
+\caption{\label{alg::expand}expand}
+\KwIn{
+   configuration $\qnear$ to be expanded,
+}
+\KwData{
+   ligand conformations $\L$,
+   scale limits $\smin, \smax$ and $\sdelta$
+}
+%\KwOut{
+%    set $S$ of collision-free configurations reachable from $\qnear$\;
+%}
+\hrule
+\ForEach{$l \in \L$}{
+%    $newnodes = $ false\;
+    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
+        \For{$i = 1,\ldots,m$}{
+            $q' = q$\;
+            translate $q$ by random around $q'$\;
+            rotate randomly \;
+            $q'.l = l$\;
+            $q'.s = s$\;
+            $qn$ = nearest node in $\T$ towards $q'$\;
+            \If{$isCollisionFree(q')$ {\bf and} $d(q',\qnear) < d(q',qn)$}{
+                $\T$.addNode($q$)\;
+                $\T$.addEdge($\qnear,q$)\;
+%                $S = S \cup \{ q \}$\;
+%                $newnodes = $ true\;
+                {\bf break}\;
+            }
+        }
+    }
+}
+%\return $S$\;
+\end{algorithm}
+
+\section{Visualization of results}
+
+The result of the proposed planning method is a tree of collision-free configurations, from which a path leading from $\qinit$
+to the nearest node towards $\qgoal$ can be found.
+Due to stochastic nature of RRT, it is not guaranteed that the tree always reached the goal configuration $\qgoal$.
+Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted from the tree.
+
+To present the results to chemicists ...
+
+\section{Experimental verification}
+
+\section{Conclusion }
+
+\section{Acknowledgments}
+
+The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
+Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
+
+\bibliographystyle{plain}
+\bibliography{paper}
+
+\end{document}
+
+% ======================================================================
+%
+%    UNUSED TEXT      UNUSED TEXT      UNUSED TEXT     UNUSED TEXT
+% ======================================================================
+
+%RRT--Retraction~\cite{zhangRetraction} improves sampling in the narrow passages by retracting the newly added configuration
+%closer to obstacles.
+%%for motion planning of 3D objects.
+%After $\qrand$ is generated and its nearest neighbor $\qnear\in \T$ is found, the segment from $\qrand$ to $\qnear$ is checked for collision.
+%The tree is extended normally if the segment is free, otherwise the retraction step is performed.
+%The task of the retraction step is to find a contact configuration  that minimizes the distance to $\qrand$.
+%The contact configuration $q$ is found on the line and its neighborhood is searched to find another contact configuration $q'$ such that 
+%$\varrho(q',\qrand) < \varrho(q,\qrand)$.
+%The retraction is terminated after predefined number of steps or if no new contact configuration can be found.
+%%The configuration trees built by RRT--Retraction can therefore contain more nodes than is the number of planning iterations,
+%as the tree can be expanded by upto $\Nrret$ contact configurations in each iteration.
+%An example of the retraction is depicted in Fig.~\ref{fig::rrtretraction}.
+%The RRT-Retraction was designed for path planning of 3D objects, where the straight-line local planner is used to connect two configurations, which allows us to find the contact configurations by random sampling.
+%The RRT--Retraction can quickly penetrate to narrow passages, but its computational burden is significantly
+%increased due to necessity to find the contact configurations.
+%The method was designed mainly for 3D path planning, but it can be combined with other planners
+%for motion planning of many-DOF robots~\cite{pan2010retraction}.
+%    e.g. for articulated robots~\cite{pan2010retraction}.
+%RRT--Retraction combined with decomposition-based path planners for articulated robots with many DOF was proposed in~\cite{pan2010retraction}.
+
+%The narrow passage problem has been studied intensively 
+%Due to its low volume, the probability of expanding the nodes towards the passage is small~\cite{hannaWIS}.
+%    many iterations are needed before a tree can be built through the passage~\cite{hannaWIS}.
+%Recently, we have proposed a novel method for tunnel detection using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT) method~\cite{vonasek2016application}.
+%In comparison to existing approaches~\cite{Petrek20071357,petrek2006caver}, this RRT-based approach~\cite{vonasek2016application} 
+%handles the dynamics without need to cluster and match Voronoi diagrams from consecutive frames.
+
+%We further consider a spherical probe $\Sprobe$  of radius $\probe$ encapsulating the ligand.
+%Let $q=(x,y,z)\in\C$ denote the configuration (position) of the probe, where
+%the configuration space $\C$ consists of all possible configurations.
+%For each configuration $q\in\C$ we assume, that its largest collision-free radius $r(q) \in \mathbf{R}, r(q)\ge 0$ can be computed, e.g.
+%using collision detection.
+%using collision detection.
+%Let $L(q) \subseteq \mathbf{R}^3$ denote the geometry of the ligand at position $q \in \C$.
+%Similarly to geometry of molecule, the geometry of ligand is union of hard sphere model.
+%A probe is represented by one sphere of radius $\probe$.
+%The collision-free region $\CF \subseteq \C$ is formed by configurations, where $\Sprobe$ can be placed without any collision, i.e., 
+%$\CF = \{q \in \C | \Sprobe(q) \cap \SS = \emptyset\}$, where $\Sprobe(q)$ is the sphere with radius $\probe$ at configuration $q$.
+%The distance $\dist(q_1,q_2)$  between two configurations $q_1,q_2\in\C$ is measured using 3D Euclidean metric.
+
+%This radius can be detected using collision detection method.
+%The configurations outside the protein $\CG \subseteq \C$ are collision-free with sphere $\Sgprobe$ of radius $\gprobe$, $\gprobe> \probe$,
+%$\CG=\{q\in \C| \Sgprobe(q) \cap \SS = \emptyset \}$.
+%A configuration $q \in \CG$ is referred to as {\sl goal configuration} in the rest of the paper.
+%By computing $\alpha$-shape of the molecule, we can identify atoms located at the surface. 
+%These atoms are referred to as {\sl surface atoms} in the rest of the paper.
+%The distance between a configuration $q \in \C$ and its nearest surface atom is referred to as $\dists(q)$.
+%We assume, that for each configuration $q$, the largest collision-free radius $r(q)$ can be computed.
+%Let $r(q)$ denote the radius of the largest collision-free sphere at configuration $q$.
+
+%Dynamic proteins are represented by a sequence of $n$ snapshots (frames).
+%The tunnels change with the protein dynamics as well: their width and position change, they can merge with other tunnels or even disappear.
+%The free region $\CF$ is therefore different in each frame, as positions of atoms change.
+%For the sake of simplicity, a single notation $\C$ resp. $\CF$ is used in this paper and it is valid for the frame being processed.
+
+%Approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing ordinary Voronoi diagram.
+%Computing WVD this way is more efficient and numerically stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
+%Vertices of this WVD are filtered out if their distance to an atom is less than probe radius $\probe$.
+%Let $\VV$ contain the remaining Voronoi vertices, and let $r_v \in \VV$ denote the distance to the nearest atom.
+%The remaining Voronoi vertices $\VV$ of the WVD and their distances $r_v, v\in \VV$ to the nearest atom are then used in the following method.
+
+%A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples in the difficult areas.
+%For Probabilistic Roadmaps~\cite{kavrakiPRM}, this can be achieved simply by generating more samples in the difficult areas, e.g. along
+%medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
+%This requires to precompute the medial axis, which can be time consuming.
+%Another approach is to shift the random samples towards the medial axis~\cite{amatoOBPRM} or into a direction of estimated
+%medial axis~\cite{hollemanMAPRM}.
+%Generating random samples in difficult regions however does not ensure construction of a better tree in the case of RRT-based methods. 
+%In RRT, the configuration tree is expanded in the direction of random samples, but the obstacles may prevent to reach them closely~\cite{vonasekphd}.
+%Increased probability of sampling in a given region, e.g., in a narrow passage, 
+%brings advantage only if the tree is close to the region and if it can reach it.
+%it is necessary to maintain the sampling regions according the progress of the tree.
+%For example, random samples can be generated along a given workspace path~\cite{vonasek2009rrt}.
+%After the tree reaches a path point, the sampling around the point is disabled and moved to the next point of 
+%the path.
+%Similar approach is used in~\cite{denny2014marrt}, where the sampling regions are maintained with a grapth describing topology 
+%of the workspace.
+%OB-RRT~\cite{amatoOBRRT} defines several types of tree expansions
+%steps based on information about obstacles (e.g. using vectors of obstacles, tangential vectors and information about medial axis).
 
@@
 %
-
-\section{Preliminaries}
-
-Proteins are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
-Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
-%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
-%   can be detected as atoms located at $\alpha$-shape of the molecule.
-%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
-A tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, where $n$ denotes the number of spheres,
-$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
-The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.
-
-Ligands are small molecules interacting with the protein. 
-Due to the interaction, the relative positions of the ligand's atoms can change.
-The ligands can be modeled as a multi-link kinematic chain where the torional angles phi/psi can change in a given range~\cite{songPFpath}.
-After the torsional angles are set, 3D position of each atom of the ligand is computed.
-By changing the angles, various conformations can be generated.
-However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
-
-%from wiki:
-%Multiple static structures experimentally determined for the same protein in different conformations are often used to emulate receptor flexibility.[19] Alternatively rotamer libraries of amino acid side chains that surround the binding cavity may be searched to generate alternate but energetically reasonable protein conformations.[20][21]
-% 19 = \cite{totrov2008flexible}
-% 20 = \cite{hartmann2009docking}
-% 21 = \cite{taylor2003fds}
-
-%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
-%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
-%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
-%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
-%knowledge-based methods have also been used.[17]
-%13 = https://link.springer.com/article/10.1007/BF00123666
-
-Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
-To decrease the computational burgen, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
-%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
-%For each conformation $c\in \L$ an energy potential is also calculated.
-%Let $\L_i$ denote the set of conformations that 
-We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
-%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
-%can be switched to the confomration $c_j \in C$.
-
-The proteins are dense structures with narrow tunnels.
-Depending on the protein, the tunnel bottlenecks can be even less then $1$~\AA, which is too narrow for ligands with more than 2 atoms.
-For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
-
-A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
-by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., it's geometric center), rotation around $x$, $y$ and $z$ axes,
-index of the conformation $l\in \L$ and the actual scale $s$.
-A discrete set of scales is used, i.e., $s \in {\smin, \smin+\sdelta, \ldots, \smax}$, where 
-$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
-All possible configurations form the configuration space $\C$. 
-A configuration is collision-free if there is no intersection between the protein atoms and ligand atoms scaled by $s$.
-The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.
-
-%We further consider a spherical probe $\Sprobe$  of radius $\probe$ encapsulating the ligand.
-%Let $q=(x,y,z)\in\C$ denote the configuration (position) of the probe, where
-%the configuration space $\C$ consists of all possible configurations.
-%For each configuration $q\in\C$ we assume, that its largest collision-free radius $r(q) \in \mathbf{R}, r(q)\ge 0$ can be computed, e.g.
-%using collision detection.
-%using collision detection.
-%Let $L(q) \subseteq \mathbf{R}^3$ denote the geometry of the ligand at position $q \in \C$.
-%Similarly to geometry of molecule, the geometry of ligand is union of hard sphere model.
-%A probe is represented by one sphere of radius $\probe$.
-%The collision-free region $\CF \subseteq \C$ is formed by configurations, where $\Sprobe$ can be placed without any collision, i.e., 
-%$\CF = \{q \in \C | \Sprobe(q) \cap \SS = \emptyset\}$, where $\Sprobe(q)$ is the sphere with radius $\probe$ at configuration $q$.
-%The distance $\dist(q_1,q_2)$  between two configurations $q_1,q_2\in\C$ is measured using 3D Euclidean metric.
-
-%This radius can be detected using collision detection method.
-%The configurations outside the protein $\CG \subseteq \C$ are collision-free with sphere $\Sgprobe$ of radius $\gprobe$, $\gprobe> \probe$,
-%$\CG=\{q\in \C| \Sgprobe(q) \cap \SS = \emptyset \}$.
-%A configuration $q \in \CG$ is referred to as {\sl goal configuration} in the rest of the paper.
-%By computing $\alpha$-shape of the molecule, we can identify atoms located at the surface. 
-%These atoms are referred to as {\sl surface atoms} in the rest of the paper.
-%The distance between a configuration $q \in \C$ and its nearest surface atom is referred to as $\dists(q)$.
-%We assume, that for each configuration $q$, the largest collision-free radius $r(q)$ can be computed.
-%Let $r(q)$ denote the radius of the largest collision-free sphere at configuration $q$.
-
-%Dynamic proteins are represented by a sequence of $n$ snapshots (frames).
-%The tunnels change with the protein dynamics as well: their width and position change, they can merge with other tunnels or even disappear.
-%The free region $\CF$ is therefore different in each frame, as positions of atoms change.
-%For the sake of simplicity, a single notation $\C$ resp. $\CF$ is used in this paper and it is valid for the frame being processed.
-
-%Approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing ordinary Voronoi diagram.
-%Computing WVD this way is more efficient and numerically stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
-%Vertices of this WVD are filtered out if their distance to an atom is less than probe radius $\probe$.
-%Let $\VV$ contain the remaining Voronoi vertices, and let $r_v \in \VV$ denote the distance to the nearest atom.
-%The remaining Voronoi vertices $\VV$ of the WVD and their distances $r_v, v\in \VV$ to the nearest atom are then used in the following method.
-
-\section{Ligand traversability}
-
-The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
-The tunnels are usually computed on proteins without ligands, which implies narrow void space.
-
-The configuration space is searched using the RRT principle~\cite{lavalleRRT}.
-The main loop of the method is described in Alg.~\ref{alg::main}.
-In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found
-and the node is expanded.
-The random samples are not generatd only from whole $\C$, as in the original RRT, but the sampling
-is guided using moving virtual goal~\cite{vonasek2009rrt}.
-The virtual goal $v$ is the index of the actual sphere of the tunnel, around which the samples
-are generated with probability $\gb$. 
-After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
-less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
-that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
-The algorithm termines after a predefined planning trials $\Imax$ or if the tree reaches
-the last sphere in the tunnel, i.e., when $v = N$.
-
-\linesnumbered
-\begin{algorithm}[h]
-%\setstretch{\straa}
-\caption{\label{alg::main}main loop of the RRT planner}
-\KwIn{
-    tunnel $T=(t_i, r_i), n=1,\ldots N$ represented by $N$ spheres,
-    initial configuration $\qinit$,
-    goal configuration $\qgoal$,
-}
-\KwData{
-   ligand conformations $\L$,
-   scale limits $\smin, \smax$ and $\sdelta$
-}
-\KwOut{
-    configuration tree $\T$\;
-}
-\hrule
-$v = 0$; // index of the virtual goal\\
-$iteration = 0$\;
-\While{$iteration < \Imax$ and $v < N$}{
-    \eIf{$rand() < \gb$}{
-        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
-    }{
-        $\qrand$ = random sample from $\C$\;
-    }
-    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
-
-    $S = $expandConfigurations($\qnear,\L$)\;
-
-    \For{$q \in S$}{
-        $\T$.addNode($q$)\;
-        $\T$.addEdge($\qnear, q$)\;
-    }
-
-    \For{$i=N : v$}{ \nllabel{alg::main:a}
-        $d$ = nearest node in the tree to $P[i]$\;
-        \If{$d < DT$}{
-            $v = i$; // new virtual goal found\;
-            break\;
-        }
-    } \nllabel{alg::main:b}
-    $iteration = iteration+1$\;
-}
-\return $\T$\;
-\end{algorithm}
-
-The core of the method is the expansion procedure, which generates new collision-free nodes around $\qnear\in\T$.
-The algorithm is listed in Alg.~\ref{alg::expand}.
-For each conformation $l \in \L$, the expansion generates up to $m$ samples $q$ with maximal radii around $\qnear$.
-The 3D position of $q$ is generated using Normal distribution $N(0,\sigma^2)$ around $\qnear$, where $\sigma^2$ is the predefined
-parameter.
-The rotation part of $q$ is generated using $U(0,2\pi)$.
-The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated
-with $s$, the next lower $s$ is used.
-The expansion step is repeated for each conformation $l \in L$.
-
-\begin{algorithm}[h]
-%\setstretch{\straa}
-\caption{\label{alg::expand}expand}
-\KwIn{
-   configuration $\qnear$ to be expanded,
-}
-\KwData{
-   ligand conformations $\L$,
-   scale limits $\smin, \smax$ and $\sdelta$
-}
-\KwOut{
-    set $S$ of collision-free configurations reachable from $q$
-}
-\hrule
-\ForEach{$l \in \L$}{
-    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
-        \For{$i = 1,\ldots,m$}{
-            $q' = q$\;
-            translate $q$ by random around $q'$\;
-            rotate randomly \;
-            $q'.l = l$\;
-            $q'.s = s$\;
-            \If{$isCollisionFree(q')$}{
-                $S = S \cup \{ q \}$\;
-            }
-        }
-    }
-}
-\end{algorithm}
-
-\section{Visualization of results}
-
-The result of the proposed planning method is a tree of collision-free configurations, from which a path leading from $\qinit$
-to the nearest node towards $\qgoal$ can be found.
-Due to stochastic nature of RRT, it is not guaranteed that the tree always reached the goal configuration $\qgoal$.
-Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted fromm the tree.
-
-To present the results to chemicists ...
-
-\section{Experimental verification}
-
-\section{Conclusion }
-
-\section{Acknowledgments}
-
-The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
-Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
-
-\bibliographystyle{plain}
-\bibliography{paper}
-
-\end{document}

### 2017-06-07 paper.tex
intro + added images

@@
 a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
-These methods compute the tunnels assuming a spherical ligand (probe).
-The resulting tunnels are represented as a sequence of spheres.
-The active sites are usually connected with the protein surface with multiple pathways.
-However, not all paths are biochemically relevant, so only paths with a given minimal bottleneck have to be reported (the bottleneck is
-the radius of the smallest sphere that can traverse the whole tunnel) 
-Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck.
+These methods compute the tunnels assuming a spherical ligand (probe), and the resulting tunnels are represented as a sequence of spheres.
+Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck (i.e.,
+        the radius of the smallest sphere that can traverse the whole tunnel).
+%The active sites are usually connected with the protein surface with multiple pathways.
+%However, not all paths are biochemically relevant, so only paths with a given minimal bottleneck have to be reported (the bottleneck is
+%the radius of the smallest sphere that can traverse the whole tunnel).
+Such decision is however not precise and it requires previous expertise in the domain.
+
+Of the reasons is that typical ligands are non-spherical and therefore it is difficult to estimate their traversability in tunnels
+computed for a spherical probe.
+For a better decision, it is necessary to consider also the shape of the ligand and possibly also its flexibility.
+Due to the increase degrees of freedom (DOF) that need to be considered, the classical Voronoi-based methods cannot be used 
+to compute trajectories of the ligands.
 
@@
 
-Such decision is however not precise and it requires previous expertise with the domain.
-Of the reasons is that typical ligands are non-spherical and therefore it is difficult to estimate their traversability in tunnels
-computed for a spherical probe.
-For better decision, it is necessary to consider also the shape of the ligand being investigated, and consequently, to also
-consider rotations of the whole ligand.
-Moreover, ligands are rather flexible molecules, so the relative positions of their atoms can change as well.
-
-Sampling-based motion planning methods such as Rapidly Exploring Random Tree~\cite{lavalleRRT} can be
-used to test the traversability of a given ligand in the tunnels.
-These methods can cope with objects of many degrees of freedom and arbitrary geometry.
+In this paper, we propose a novel method for computing the trajectories of the flexible ligands in a tunnel represented by spheres.
+The proposed work is based on the sampling-based Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner, which
+can cope with many-DOF objects of arbitrary geometry.
+The random samples for RRT are generated using the spheres of the tunnel.
+To make the transportation easier, a scaled-down version of the ligand is used, similarly e.g. to~\cite{cortes2005path}.
+The flexiblity of the ligand is modeled using a predefined set of typical conformations.
+We propose novel visualization techniques for presenting the results to bio-chemists.
+The proposed method can be seen as an extension for the classic tunnel detection tools.
 
@@
 \includegraphics[width=0.15\textwidth]{fig/motiv1} &
-\includegraphics[width=0.17\textwidth]{fig/motiv2lab} \\
-%\includegraphics[width=0.16\textwidth]{fig/motiv3} \\
+\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
+\includegraphics[width=0.16\textwidth]{fig/motiv3} \\
 Protein 1CQW & Active site & Detected tunnels \\
@@
 \end{figure}
-
-\red{TODO}
-what is traversability analysis.
-
-The goal is the traversability analysis is to estimate traversability of a tunnel with a given ligand from geometrical point of view.
 

### 2017-06-07 paper.tex
text + images

@@
 
+\begin{figure}[t]
+\centering
+{\footnotesize
+\renewcommand{\arraystretch}{0.1}
+\renewcommand{\tabcolsep}{0pt}
+\begin{tabular}{ccccc}
+\includegraphics[width=0.15\textwidth]{fig/motiv1} &
+\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
+\includegraphics[width=0.16\textwidth]{fig/motiv3} &
+\hbox{
+\vbox{
+\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
+\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
+} 
+}
+\\
+Protein 1CQW & Active site & Detected tunnels & Example of a  \\
+             &            & (orange)          & trajectory
+\end{tabular}
+}
+\caption{\label{fig::motiv}
+    Example of tunnel detection in haloalkane dehalogenase.
+}
+\end{figure}
+
 \red{this is fast copy and paste from previous paper, please rewrite, cut to at ca. 1 page}
@@
 These methods compute the tunnels assuming a spherical ligand (probe), and the resulting tunnels are represented as a sequence of spheres.
-Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck (i.e.,
+Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and bottleneck (i.e.,
         the radius of the smallest sphere that can traverse the whole tunnel).
@@
 %the radius of the smallest sphere that can traverse the whole tunnel).
-Such decision is however not precise and it requires previous expertise in the domain.
 
-Of the reasons is that typical ligands are non-spherical and therefore it is difficult to estimate their traversability in tunnels
+%Such decision is however not precise and it requires previous expertise in the domain.
+Typical ligands are however non-spherical and therefore it is difficult to estimate their traversability in tunnels
 computed for a spherical probe.
-For a better decision, it is necessary to consider also the shape of the ligand and possibly also its flexibility.
+It is therefore necessary to consider also the shape of the ligand and possibly also its flexibility.
 Due to the increase degrees of freedom (DOF) that need to be considered, the classical Voronoi-based methods cannot be used 
@@
 To make the transportation easier, a scaled-down version of the ligand is used, similarly e.g. to~\cite{cortes2005path}.
-The flexiblity of the ligand is modeled using a predefined set of typical conformations.
+The flexibility of the ligand is modeled using a predefined set of typical conformations.
 We propose novel visualization techniques for presenting the results to bio-chemists.
@@
 %However, such estimation still involve lot of experience.
-
-\begin{figure}[t]
-\centering
-{\footnotesize
-\renewcommand{\arraystretch}{0.1}
-\renewcommand{\tabcolsep}{0pt}
-\begin{tabular}{ccc}
-\includegraphics[width=0.15\textwidth]{fig/motiv1} &
-\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
-\includegraphics[width=0.16\textwidth]{fig/motiv3} \\
-Protein 1CQW & Active site & Detected tunnels \\
-             &            & (orange)
-\end{tabular}
-}
-\caption{\label{fig::motiv}
-    Example of tunnel detection in haloalkane dehalogenase.
-}
-\end{figure}
 
@@
 each atom is approximated by several spheres with small radius~\cite{yaffe2008,caver3}.
-%The VD-based methods remove one of the main limitations of the grid-based approach which is the dependency on the grid resolution.
 VD-based methods are memory less demanding, and also faster than the grid-based methods.
-%They can be easily used for static tunnel detection, as they are based on Voronoi diagrams of static frames.
-%The grid-based and VD-based methods are usually employed on static proteins.
-%Finding tunnels through time-changing proteins increases the dimension of the space to be searched, which prevents the grid-based methods
-%to be used for this task.
 The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
@@
 The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
-%The main limitation of this approach is that this solution builds a complex and large hierarchical tree which has to be stored in memory and subsequently traversed which is computationally demanding.
-%~\cite{lindow2012dynamic}.
 
-Finding tunnels for a general, non-spherical and possibly a flexible ligand can be solved by searching
-a feasible path in the corresponding configuration space $\C$.
-The dimension of the space is given by the degrees of freedom (6D for a solid ligand and 6D+$n$ for a flexible ligand with
-$n$ rotational bonds).
-The feasible configurations, where the ligand (robot) does not touch and obstacle, form the free region $\CF$.
-The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
-to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the beginning and the end of the
-tunnel, respectively.
-Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
+The detected tunnels are evaluated using basic characteristics like bottleneck, length, curvature and a list of surrounding
+residues, which are later used to estimate possibility of the interaction.
+The biggest disadvantage is that the shape of the ligand is not considered during the tunnel detection and it is therefore
+not easy to estimate if (and how) a ligand might traverse the tunnel.
+
+The task of trajectory computation for the non-spherical ligand can be formulated as a path search 
+in the related configuration space $\C$.
+The dimension of the configuration space is given my the DOF of the ligand, i.e., 6D for a rigid and 6D + $n$ for a flexible
+ligand with $n$ DOFs.
+%We propose to employ motion planning technique to determine possible trjaectories for a given ligand in the detected tunnels.
+%This can be solved by search a high-dimensional configuration space $\C$. % as both 3D translation, rotation possibly with additional DOFs caused
+%by the ligand's flexiblity need to be considered.
+%The dimension of the configuration space $\C$ is given by the degrees of freedom (6D for a solid ligand and 6D+$n$ for a flexible ligand with
+%$n$ rotational bonds).
+%The feasible configurations, where the ligand (robot) does not touch and obstacle, form the free region $\CF$.
+%The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
+%to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the beginning and the end of the
+%tunnel, respectively.
+%Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
 The idea of sampling-based motion planning is to randomly sample the configuration space.
@@
 A path in the roadmap then represents a motion in the workspace.
+Beyond robotics, sampling-based motion planners have been successfully applied in computational biology in studies
+related to 
+protein folding~\cite{al2012motion,gipson2012computational,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},analysis of loop motions~\cite{cortes2004geometric}, and even for tunnel detection~\cite{vonasek2016application,vonasek2017tunnel}.
+%none of the existing work focuses on the trajectory generation in tunnels.
+
 Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query motion planning method.
@@
 
-Beyond robotics, sampling-based motion planners have been successfully applied in computational biology in studies
-related to 
-protein folding~\cite{al2012motion,gipson2012computational,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},analysis of loop motions~\cite{cortes2004geometric}, and even for tunnel detection~\cite{vonasek2016application,vonasek2017tunnel}.
 %rrt for molecular: \cite{al2012motion}
@@
 
-The most relevant paper is~\cite{cortes2005path}, where RRT is used to compute pathways for flexible ligands towards the active site, but
-without the possibility to select a preferable path/tunnel.
+In this paper, we propose to employ RRT to compute the trajectory for a given ligand in a given tunnel.
+To guide the configuration tree through the configuration space, the random samples are generated gradually using the spheres of the tunnels.
+As the tunnels are usually computed on protein models without ligands, they are usually very narrow and a ligand can barely traverse them.
+It can be assume that both the protein and the ligand adapt to each other and therefore.
+To enable the planner find a solution, the radii of the ligand, and possibly also the radii of the protein atoms, are down-scaled
+by a given factor.
 
-In this paper, we propose to employ motion planning to compute the trajectory for a given ligand in a given tunnel.
-This brings several advantages.
-From motion planning point of view, the tunnels have properties of narrow passages.
-By focusing the motion planning only to the surrounding of a tunnel, the total volume of the configuration space decreases which increases
-the relative volume of the free space, which increases probability of placing random sampling in the perspective regions of the configuration 
-space.
+The most related work to this paper is the MoMa-LigPath~\cite{cortes2005path}, that is based on the ML-RRT method~\cite{cortes2010simulating}.
+In comparison to our solution, the method~\cite{cortes2005path} aims for find any path leading towards the active site without
+possibility to select a tunnel of interest.
 
@@
 
-The growth of the configuration tree into some regions of the configuration space can however be slowed down due to the narrow passages.
-A narrow passage is a small region of the configuration space containing part of the solution.
-As the samples are generated from uniform distribution in RRT, probability of placing the samples into the narrow passages is low and therefore,
-   the probability of expanding the tree through a narrow passage is small~\cite{hannaWIS}.
-
-This technique is similar to generation of samples near the medial exist of the environment
-medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting,amatoOBRRT}.
-or by guiding the growth of the tree around a general path in the workspace~\cite{vonasek2009rrt,denny2014marrt}.
+%The growth of the configuration tree into some regions of the configuration space can however be slowed down due to the narrow passages.
+%A narrow passage is a small region of the configuration space containing part of the solution.
+%As the samples are generated from uniform distribution in RRT, probability of placing the samples into the narrow passages is low and therefore,
+%   the probability of expanding the tree through a narrow passage is small~\cite{hannaWIS}.
+%
+%This technique is similar to generation of samples near the medial exist of the environment
+%medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting,amatoOBRRT}.
+%or by guiding the growth of the tree around a general path in the workspace~\cite{vonasek2009rrt,denny2014marrt}.
 
@@
 flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.
-
-%deformable objects in many fields like computer graphics, robotics and medicine~\cite{alterovitz2008motion} clothing, rope, human tissue
-%a two-stage approach was presented in~\cite{bayazit2002deformable}. 
-%First a path is found with reduced-scale of the robot, or by allowing the robot to penetrate obstacles.
-%The roadmap can therefore contain configurations that are infeasible for the rigid version of the robot
-%After finding a path, the infeasible configuraions are determined and the robot is deformed there to avoid the collisions.
-%the path is refined to remove the collisions
-%geometric deformation -- only part of the object that is colliding is deformed,
-%and bounding box (first bounding box is deformed and the the model is deformed according to the deformation compute for the bbox)
-%path planning with non-rigid obstacles~\cite{frank2008efficient}.
-
-%the majority of path planning approached focus on rigid robots and rigid obstacles.
-%\cite{gayle2005path},planning for a flexible robot in complex environments. 
-%algorithm  computes a collision free path by taking into account geometric and physical constraints, including obstacle avoidance, non-
-%penetration constraint, volume preservation, surface tension, and energy minimization.
-
-%new algorithm for collision detection between a deformable robot and fixed obstacles using  graphics processors. 
-%We also present techniques to efficiently handle complex deformable models composed of tens of thousands of polygons and obtain significant performance improvement    over previous approaches. 
-%practical application of our algorithm in performing path planning of catheters in liver chemoembolization.
-%planning for flexible objects~\cite{lamiraux2001flexible},
-
-% f-PRM \cite{kavraki1998towards} framework for flexible robots of simple geometrix shapes\cite{anshelevich2000deformable} or surface patches~\cite{holleman1998planning}.
-
-%other planning for defomrable: \cite{anshelevich2000deformable,bayazit2002deformable,gayle2005path}
-%deformable envinorments: \cite{rodriguez2006planning,frank2008efficient,phillips2014representation}
-
-%representation of deformation~\cite{phillips2014representation}
-
-%Robots can be deformed either considering their physical propersites or not.
-%geometrical or physical-based deformation approaches
-%The non-physical deformations include basic deformations  like scale, bend or twist~\cite{barr1984global},
-%   free-form deformation~\cite{gibson19973d,sederberg1986free},
-%   Physical-based deformations can be achiebed using mass-spring and finite-element methods.
-%f-PRM~\cite{anshelevich2000deformable}, physically coreect deformations, time consuming
 
@@
 %
+
+%deformable objects in many fields like computer graphics, robotics and medicine~\cite{alterovitz2008motion} clothing, rope, human tissue
+%a two-stage approach was presented in~\cite{bayazit2002deformable}. 
+%First a path is found with reduced-scale of the robot, or by allowing the robot to penetrate obstacles.
+%The roadmap can therefore contain configurations that are infeasible for the rigid version of the robot
+%After finding a path, the infeasible configuraions are determined and the robot is deformed there to avoid the collisions.
+%the path is refined to remove the collisions
+%geometric deformation -- only part of the object that is colliding is deformed,
+%and bounding box (first bounding box is deformed and the the model is deformed according to the deformation compute for the bbox)
+%path planning with non-rigid obstacles~\cite{frank2008efficient}.
+
+%the majority of path planning approached focus on rigid robots and rigid obstacles.
+%\cite{gayle2005path},planning for a flexible robot in complex environments. 
+%algorithm  computes a collision free path by taking into account geometric and physical constraints, including obstacle avoidance, non-
+%penetration constraint, volume preservation, surface tension, and energy minimization.
+
+%new algorithm for collision detection between a deformable robot and fixed obstacles using  graphics processors. 
+%We also present techniques to efficiently handle complex deformable models composed of tens of thousands of polygons and obtain significant performance improvement    over previous approaches. 
+%practical application of our algorithm in performing path planning of catheters in liver chemoembolization.
+%planning for flexible objects~\cite{lamiraux2001flexible},
+
+% f-PRM \cite{kavraki1998towards} framework for flexible robots of simple geometrix shapes\cite{anshelevich2000deformable} or surface patches~\cite{holleman1998planning}.
+
+%other planning for defomrable: \cite{anshelevich2000deformable,bayazit2002deformable,gayle2005path}
+%deformable envinorments: \cite{rodriguez2006planning,frank2008efficient,phillips2014representation}
+
+%representation of deformation~\cite{phillips2014representation}
+
+%Robots can be deformed either considering their physical propersites or not.
+%geometrical or physical-based deformation approaches
+%The non-physical deformations include basic deformations  like scale, bend or twist~\cite{barr1984global},
+%   free-form deformation~\cite{gibson19973d,sederberg1986free},
+%   Physical-based deformations can be achiebed using mass-spring and finite-element methods.
+%f-PRM~\cite{anshelevich2000deformable}, physically coreect deformations, time consuming

### 2017-06-08 paper.tex
ml-rrt refs + atom metric

@@
 \def\dt{d_{goal}}
+\def\da{d_{atom}}
 
@@
 \def\VVA{V_{vor}^{act}}
-\def\da{d_{high}}
-\def\db{d_{low}}
 
@@
 
-The most related work to this paper is the MoMa-LigPath~\cite{cortes2005path}, that is based on the ML-RRT method~\cite{cortes2010simulating}.
+The most related work to this paper is the MoMa-LigPath~\cite{cortes2005path}, that is based on the ML-RRT method~\cite{cortes2010simulating,cortes2007mlrrt}.
 In comparison to our solution, the method~\cite{cortes2005path} aims for find any path leading towards the active site without
@@
 
+todo atom metric
+
 \begin{algorithm}[h]
@@
     \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
+        $\qnew = \emptyset$; // empty configuration\\
         \For{$i = 1,\ldots,m$}{
-            $q' = q$\;
-            translate $q$ by random around $q'$\;
-            rotate randomly \;
-            $q'.l = l$\;
-            $q'.s = s$\;
-            $qn$ = nearest node in $\T$ towards $q'$\;
-            \If{$isCollisionFree(q')$ {\bf and} $d(q',\qnear) < d(q',qn)$}{
+            $q=\qnear$\;
+            $q.position$ = random 3D position around $\qnear$\;
+            $q.rotation$ = random 3D rotation\;
+            $q.l = l$\; 
+            $q.s = s$\;
+            \If{isCollisionFree($q$) {\bf and} \\ $(\qnew = \emptyset$ {\bf or} $\da(q, \qnear) < \da(\qnew,\qnear))$} {
+                $\qnew = q$\;
+            } 
+        }
         $\T$.addNode($q$)\;
         $\T$.addEdge($\qnear,q$)\;
-%                $S = S \cup \{ q \}$\;
-%                $newnodes = $ true\;
-                {\bf break}\;
-            }
-        }
     }

### 2017-06-08 paper.tex
alg updated

@@
 The main loop of the method is described in Alg.~\ref{alg::main}.
-In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found
-and the node is expanded.
+In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
+The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
+both 3D rotation and 3D translation.
+
 The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
@@
 
-The core of the method is the expansion procedure, which generates new collision-free nodes around $\qnear\in\T$.
-The algorithm is listed in Alg.~\ref{alg::expand}.
-For each conformation $l \in \L$, the expansion generates up to $m$ samples $q$ with maximal radii around $\qnear$.
+The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
+For each ligand conformation, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with maximal
+scale.
+For a given conformation $l \in \L$, first the maximal scale $\smax$ is considered and
+$m$ samples are generated around $\qrand$ and tested for collision.
+The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
+If none of the samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
+until a free collision-free samples is found or until the mimimal reduced-scale $\smin$ is reached.
+
+The sampling-based methods are sensitive to the employed metric, especially if the objects are not symmetrical, which is the
+case of the non-spherical ligands.
+To consider the actual shape of the ligand (which is different in each conformation) and to allow finding such configurations
+that approach $\qrand$, the distance between newly generated configurations and $\qrand$ is measured as the smallest
+distance between an atom of the ligand placed at $q$ and the 3D position of $\qrand$ ($\da(q,\qrand)$ on line~\ref{alg::expand:a} in
+        Alg.~\ref{alg::expand}).
+
 The 3D position of $q$ is generated using Normal distribution $N(0,\sigma^2)$ around $\qnear$, where $\sigma^2$ is the predefined
@@
 
-todo atom metric
-
 \begin{algorithm}[h]
@@
    configuration $\qnear$ to be expanded,
+   random configuration $\qrand$
 }
@@
 \ForEach{$l \in \L$}{
-%    $newnodes = $ false\;
     \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
@@
             $q.s = s$\;
-            \If{isCollisionFree($q$) {\bf and} \\ $(\qnew = \emptyset$ {\bf or} $\da(q, \qnear) < \da(\qnew,\qnear))$} {
+            \If{isCollisionFree($q$)}{
+                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qnear) < \da(\qnew,\qnear)$}{ \nllabel{alg::expand:a}
                     $\qnew = q$\;
@@
             } 
-        $\T$.addNode($q$)\;
-        $\T$.addEdge($\qnear,q$)\;
+        }
+        \If{$\qnew \ne \emptyset$} {
+            $\T$.addNode($\qnew$)\;
+            $\T$.addEdge($\qnear,\qnew$)\;
+            {\bf break;} // go to next conformation
+        }
     }
@@
 \end{algorithm}
+
+The configruation tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
+one new configuration can be added.
+The most time consuming part of the method is the collision-detection, which is called $m|\L|$ in each expansion step.
+The collision detection can be evaluated fast using hiearchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}
 

### 2017-06-13 paper.tex
images of m003
alg updated

@@
 \def\CFD{{\mathcal{C}^s_{free}}}
-\def\dt{d_{goal}}
+\def\dt{d_{tunnel}}
 \def\da{d_{atom}}
@@
 %spacing for algorithm environment. 1.0 mean normal spacing
-\def\gb{p_{c}}
+\def\gb{p_{tunnel}}
 
@@
 
+Ligands are small molecules interacting with the protein. 
+Due to the interaction, the relative positions of the ligand's atoms can change.
+The ligands can be modeled as a multi-link kinematic chain where the torsional angles phi/psi can change in a given range~\cite{songPFpath}.
+After the torsional angles are set, 3D position of each atom of the ligand is computed.
+By changing the angles, various conformations can be generated.
+However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
+
 % in cortes2005path:
@@
 
-Proteins are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
-Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
+Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
+%Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
 %We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
@@
 %The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
-A tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, where $n$ denotes the number of spheres,
+The flexiblity of a ligand is modeled using the set $\L$ of its conformations.
+The conformations can be prepared e.g., considering the potential energy using tools like Rosetta.
+We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
+
+%Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
+%To decrease the computational burden, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
+%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
+%For each conformation $c\in \L$ an energy potential is also calculated.
+%Let $\L_i$ denote the set of conformations that 
+%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
+%can be switched to the confomration $c_j \in C$.
+
+A protein tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, $i=1,\ldots,n$, where $n$ denotes the number of spheres,
 $t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
 The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.
-
-Ligands are small molecules interacting with the protein. 
-Due to the interaction, the relative positions of the ligand's atoms can change.
-The ligands can be modeled as a multi-link kinematic chain where the torsional angles phi/psi can change in a given range~\cite{songPFpath}.
-After the torsional angles are set, 3D position of each atom of the ligand is computed.
-By changing the angles, various conformations can be generated.
-However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
 
@@
 
-Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
-To decrease the computational burden, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
-%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
-%For each conformation $c\in \L$ an energy potential is also calculated.
-%Let $\L_i$ denote the set of conformations that 
-We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
-%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
-%can be switched to the confomration $c_j \in C$.
-
 The proteins are dense structures with narrow tunnels.
@@
 For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
+A discrete set of scales is used, i.e., $s \in \{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
+$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
+This scaling technique is used also in other tools like MoMaLigPath~\cite{cortes2005path}.
 
@@
 index of the conformation $l\in \L$ and the actual scale $s$.
-A discrete set of scales is used, i.e., $s \in {\smin, \smin+\sdelta, \ldots, \smax}$, where 
-$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
 All possible configurations form the configuration space $\C$. 
 A configuration is collision-free if there is no intersection between the protein atoms and ligand atoms scaled by $s$.
-The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.
+%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.
 
@@
 The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
-The tunnels are usually computed on proteins without ligands, which implies narrow void space.
+We propose a RRT-based method to compute the trajectories of the ligand.
+As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
+suffer from the well known narrow passage problem~\cite{hannaWIS}.
 
-The configuration space is searched using the RRT principle~\cite{lavalleRRT}.
+To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
+Instead, they are generated using the spheres of the tunnel.
 The main loop of the method is described in Alg.~\ref{alg::main}.
@@
 
-The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
-is guided using moving virtual goal~\cite{vonasek2009rrt}.
-The virtual goal $v$ is the index of the actual sphere of the tunnel, around which the samples
-are generated with probability $\gb$. 
+To guide thee growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
+The virtual goal $v$ is the index of a sphere, around which the random samples are generated with probability $\gb$.
+%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
+%is guided using moving virtual goal~\cite{vonasek2009rrt}.
 After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
@@
 that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
+Setting the virtual goal to this successor allows the tree to ``skip'' such parts of the tunnels that are
+not traversible or reachable by the ligand.
 The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
 the last sphere in the tunnel, i.e., when $v = N$.
+
+The random samples $\qrand$ are generated around the virtual goal.
+The translation of $\qrand$ is generated from Normal distribution centered at $t_v$ with radius $R_v$.
+The radius $R_v$ is set as the absolut distance from the tunnel centerline, e.g. $R_v=3\AA$.
+Alternatively, $R_v$ can be set to $R_v=r_v$ according the radius of the tunnel at the virtual goal.
 
@@
 $iteration = 0$\;
-\While{$iteration < \Imax$ and $v < N$}{
+\While{$iteration < \Imax$ {\bf and}  $v < N$}{
     \eIf{$rand() < \gb$}{
@@
 
-    \For{$i=N : v$}{ \nllabel{alg::main:a}
-        $d$ = nearest node in the tree to $P[i]$\;
-        \If{$d < DT$}{
-            $v = i$; // new virtual goal found\;
-            break\;
+    \For{$i= N,N-1,\ldots v+1,v$}{ \nllabel{alg::main:a}
+        $d$ = nearest node in the tree to $T[i]$\;
+        \If{$d < \dt $}{
+            $v = i+1$; // new virtual goal found\;
+            {\bf  break}\;
         }
@@
 The collision detection can be evaluated fast using hiearchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}
+
+\def\tmpa{0.16\textwidth}
+
+\begin{figure}
+\centering
+\begin{tabular}{cccccc}
+\includegraphics[width=\tmpa]{fig/m003-000} &
+\includegraphics[width=\tmpa]{fig/m003-001} &
+\includegraphics[width=\tmpa]{fig/m003-002} & 
+\includegraphics[width=\tmpa]{fig/m003-003} & 
+\includegraphics[width=\tmpa]{fig/m003-004} & 
+\includegraphics[width=\tmpa]{fig/m003-005} \\
+\includegraphics[width=\tmpa]{fig/m003-006} &
+\includegraphics[width=\tmpa]{fig/m003-007} &
+\includegraphics[width=\tmpa]{fig/m003-008} & 
+\includegraphics[width=\tmpa]{fig/m003-009} & 
+\includegraphics[width=\tmpa]{fig/m003-010} & 
+\includegraphics[width=\tmpa]{fig/m003-011} \\
+
+\end{tabular}
+\caption{todo and maybe delete}
+\end{figure}
 

### 2017-06-14 paper.tex
Visualization concept

@@
 
+\red{Adam: }
+As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
+On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
+Moreover, it is not feasible to directly visualize and explore the resulting trajectories due the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
+Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
+In this manner, we aim to further support the decision process of a chemist about a tunnel.
+In order to do that, we employ an ensemble of resulting trajectories in order to compute and visualize the following tunnel properties:
+\begin{itemize}
+  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
+  \item \textbf{Ligand Scale} --- The scale of the ligand wheh it passes a tunnel's part.
+  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
+\end{itemize}
+Moreover, we analyze the whole set of trajectories in order to convey the most important trajectories that enabled the ligand to pass or that prevented it from passing.
+In order to visualize the important trajectories we first cluster the trajectories according to their similarity.
+Then we choose a representant from each cluster a we visualize its trajectory in 3D using a polyline.
+Moreover, we map the representant's cluster size onto the width of the polyline.
+In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
+
+\begin{figure}
+\centering
+\begin{tabular}{cccccc}
+\includegraphics[width=0.45\textwidth]{fig/trajectories-all}
+\qquad & \qquad
+\includegraphics[width=0.45\textwidth]{fig/trajectories-clustered-21} \\
+\end{tabular}
+\caption{Visualization of resulting trajectories.
+Left: All trajectories colored according to whether they reached the end of the tunnel (green) or they did not (red).
+Right: Trajectories clustered according to their proximity colored by distinct colors.
+The different line widths represent the number of trajectories in a cluster.}
+\label{fig:trajectories}
+\end{figure}
+
 \section{Experimental verification}

### 2017-06-14 paper.tex
table  Please enter the commit message for your changes. Lines starting

--- 2017-06-13 paper.tex
@@
 \def\L{L}
+\def\LA{L_1}
+\def\LB{L_2}
 
@@
 not traversible or reachable by the ligand.
-The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
+The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
 the last sphere in the tunnel, i.e., when $v = N$.
@@
 \KwIn{
-    tunnel $T=(t_i, r_i), n=1,\ldots N$ represented by $N$ spheres,
+    tunnel $T=(t_i, r_i), n=1,\ldots,N$ represented by $N$ spheres,
     initial configuration $\qinit$,
@@
 
-    \For{$i= N,N-1,\ldots v+1,v$}{ \nllabel{alg::main:a}
-        $d$ = nearest node in the tree to $T[i]$\;
+    \For{$i= N-1,N-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
+        $d$ = nearest node in the tree towards sphere $t_i$\;
         \If{$d < \dt $}{
@@
 The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
-For each ligand conformation, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with maximal
+For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal
 scale.
-For a given conformation $l \in \L$, first the maximal scale $\smax$ is considered and
-$m$ samples are generated around $\qrand$ and tested for collision.
+First, the maximal scale $\smax$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
-If none of the samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
+If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
 until a free collision-free samples is found or until the mimimal reduced-scale $\smin$ is reached.
@@
 \section{Experimental verification}
+
+The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase (protein 1CQW) (Fig.~\ref{fig::tunnel}).
+The tunnel was computed using CAVER 3.0~\cite{caver3} for spherical probe of radius 0.9~\AA.
+The tunnel is represented by 760 spheres with average distance $0.05$~\AA.
+
+% from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
+The traversability of the tunnel was evaluated for 1-chlorpropan (denoted as $\LA$) and 1-chlorbutan (denoted as $\LB$).
+
+\begin{figure}
+\centering
+\caption{\label{fig::tunnel}
+    TODO image of the tunnel + 1cqw
+}
+\end{figure}
+
+\begin{table}
+\centering
+\caption{\label{tab::rrt}
+    Runtime and sucess ratio of the planner
+}
+\input table.003.tex
+\hskip 10pt
+\input table.004.tex
+\end{table}
+
+Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.
 

### 2017-06-14 paper.tex
comments, added first table

@@
   \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
+  \red{ i.e., number of trajectories ending in the tunnel part over all trajectories? }
   \item \textbf{Ligand Scale} --- The scale of the ligand wheh it passes a tunnel's part.
   \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
+\red{ i.e., the conditional probability we discussed in the mail?}
 \end{itemize}
+
+\red{can you please tell in detail how is the accesibility/liugand/throughput mapped to the tunnel surface/spheres? }
+
 Moreover, we analyze the whole set of trajectories in order to convey the most important trajectories that enabled the ligand to pass or that prevented it from passing.
 In order to visualize the important trajectories we first cluster the trajectories according to their similarity. 
+\red{can you please add some details how is the similarity counted? Maybe a short description of even equation.}
 Then we choose a representant from each cluster a we visualize its trajectory in 3D using a polyline.
+\red{please some reference to the hierarchical clustering}.
 Moreover, we map the representant's cluster size onto the width of the polyline.
@@
 \hskip 10pt
-\input table.004.tex
+%\input table.004.tex
 \end{table}

### 2017-06-14 paper.tex
Updates to tunnel properties

--- 2017-06-14 paper.tex
@@
 \def\L{L}
-\def\LA{L_1}
-\def\LB{L_2}
 
@@
 not traversible or reachable by the ligand.
-The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
+The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
 the last sphere in the tunnel, i.e., when $v = N$.
@@
 \KwIn{
-    tunnel $T=(t_i, r_i), n=1,\ldots,N$ represented by $N$ spheres,
+    tunnel $T=(t_i, r_i), n=1,\ldots N$ represented by $N$ spheres,
     initial configuration $\qinit$,
@@
 
-    \For{$i= N-1,N-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
-        $d$ = nearest node in the tree towards sphere $t_i$\;
+    \For{$i= N,N-1,\ldots v+1,v$}{ \nllabel{alg::main:a}
+        $d$ = nearest node in the tree to $T[i]$\;
         \If{$d < \dt $}{
@@
 The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
-For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal
+For each ligand conformation, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with maximal
 scale.
-First, the maximal scale $\smax$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
+For a given conformation $l \in \L$, first the maximal scale $\smax$ is considered and
+$m$ samples are generated around $\qrand$ and tested for collision.
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
-If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
+If none of the samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
 until a free collision-free samples is found or until the mimimal reduced-scale $\smin$ is reached.
@@
 On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
-Moreover, it is not feasible to directly visualize and explore the resulting trajectories due the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
+Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
 Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
 In this manner, we aim to further support the decision process of a chemist about a tunnel.
-In order to do that, we employ an ensemble of resulting trajectories in order to compute and visualize the following tunnel properties:
+In order to do that, we employ an ensemble of resulting trajectories, and we compute and visualize the following tunnel properties:
 \begin{itemize}
@@
 \end{itemize}
+During visualization, we map these properties onto a tunnel's surface so that they can be visually correlated with, e.g., the tunnel's shape.
+
 Moreover, we analyze the whole set of trajectories in order to convey the most important trajectories that enabled the ligand to pass or that prevented it from passing.
@@
 \section{Experimental verification}
-
-The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase (protein 1CQW) (Fig.~\ref{fig::tunnel}).
-The tunnel was computed using CAVER 3.0~\cite{caver3} for spherical probe of radius 0.9~\AA.
-The tunnel is represented by 760 spheres with average distance $0.05$~\AA.
-
-% from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
-The traversability of the tunnel was evaluated for 1-chlorpropan (denoted as $\LA$) and 1-chlorbutan (denoted as $\LB$).
-
-\begin{figure}
-\centering
-\caption{\label{fig::tunnel}
-    TODO image of the tunnel + 1cqw
-}
-\end{figure}
-
-\begin{table}
-\centering
-\caption{\label{tab::rrt}
-    Runtime and sucess ratio of the planner
-}
-\input table.003.tex
-\hskip 10pt
-\input table.004.tex
-\end{table}
-
-Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.
 

### 2017-06-14 paper.tex

--- 2017-06-14 paper.tex
@@
 \end{itemize}
+During visualization, we map these properties onto a tunnel's surface so that they can be visually correlated with, e.g., the tunnel's shape.
 

### 2017-06-14 paper.tex
Added figures of tunnel properties

@@
   \red{ i.e., number of trajectories ending in the tunnel part over all trajectories? }
-  \item \textbf{Ligand Scale} --- The scale of the ligand wheh it passes a tunnel's part.
+  \item \textbf{Ligand scale} --- The scale of the ligand wheh it passes a tunnel's part.
   \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
@@
 \end{itemize}
-During visualization, we map these properties onto a tunnel's surface so that they can be visually correlated with, e.g., the tunnel's shape.
+During visualization, we map these properties onto a tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
+
+\begin{figure}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[width=0.3\textwidth]{fig/accessibility}
+\quad & \quad
+\includegraphics[width=0.3\textwidth]{fig/ligand-scale}
+\quad & \quad
+\includegraphics[width=0.3\textwidth]{fig/throughput}
+\end{tabular}
+\caption{Tunnel properties mapped onto its surface.
+Left: Accessibility --- green denotes parts that were accesed by the ligand easily while red denotes parts that were accessible only in about 10\% of computations.
+Middle: Ligand scale --- the average scale of the ligand was \~50\% in orange parts, \~60\% in yellow parts, \~80\% in yellow to green parts and \~100\% in green parts.
+Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.}
+\label{fig:properties}
+\end{figure}
 
@@
 \centering
-\begin{tabular}{cccccc}
+\begin{tabular}{cc}
 \includegraphics[width=0.45\textwidth]{fig/trajectories-all}

### 2017-06-14 paper.tex
Added \tylde command to typeset tilde

@@
 \newcommand{\red}[1]{\textcolor{red}{#1}}
+\newcommand{\tylde}{\raisebox{-0.85ex}{\textasciitilde}}
 
@@
 \caption{Tunnel properties mapped onto its surface.
-Left: Accessibility --- green denotes parts that were accesed by the ligand easily while red denotes parts that were accessible only in about 10\% of computations.
-Middle: Ligand scale --- the average scale of the ligand was \~50\% in orange parts, \~60\% in yellow parts, \~80\% in yellow to green parts and \~100\% in green parts.
+Left: Accessibility --- green denotes parts that were accesed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
+Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
 Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.}

### 2017-06-14 paper.tex
alg + planning sota

--- 2017-06-14 paper.tex
@@
 \and 
-\inst{3}
-TODO DAVID \at TODO DAVID
+D. Bedn\'a\v{r} \at
+Loschmidt Laboratories, Department of Experimental Biology, Masaryk University, Brno, Czech Republic
+    International Centre for Clinical Research, St. Anne’s University Hospital Brno, Brno, Czech Republic
 }
@@
 related to 
-protein folding~\cite{al2012motion,gipson2012computational,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},analysis of loop motions~\cite{cortes2004geometric}, and even for tunnel detection~\cite{vonasek2016application,vonasek2017tunnel}.
+protein folding~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro}, analysis of loop motions~\cite{cortes2004geometric},
+ tunnel detection~\cite{vonasek2016application,vonasek2017tunnel} and
+ exit pathway computation~\cite{cortes2010simulating,guieysse2008structure}.
 %none of the existing work focuses on the trajectory generation in tunnels.
 
-Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query motion planning method.
-RRT incrementally builds a configuration tree $\T$ rooted at the initial configuration.
+Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
+incrementally builds a configuration tree $\T$ rooted at the initial configuration $\qinit$.
 In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node in the tree $\qnear \in \T$ is found.
@@
 
+A RRT-based method for computing exit pathways for a small molecule was presented in~\cite{cortes2010simulating}.
+Both protein and ligand are represented as flexible kinematic chains.
+To decrease the dimension of the related configuration space, the RRT-ML variant~\cite{cortes2007mlrrt} is employed.
+In RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
+        and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder
+the growth of the tree.
+The RRT-ML is also used in the MoMa-LigPath online service.
+
+The~\cite{cortes2010simulating} approach aims to find any exit pathway for a given ligand, which requires
+to search the whole configuration space of the ligand/protein complex.
+This might be a possible disadvantage, as the proteins are dense structures with relatively narrow void space, which has the
+properties of the narrow passage~\cite{hannaWIS}.
+Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
+Many iterations are needed to find a solution, which increases the computational time.
+
+The protein tunnels computed for spherical probes contain valuable information about the protein void space.
+Based on the biochemical properties, the chemists can decide if a given tunnel is relevant and can be possibly
+used by a ligand to exit the protein.
+The biochemically relevant tunnels can be then analyzed separately.
+
 In this paper, we propose to employ RRT to compute the trajectory for a given ligand in a given tunnel.
-To guide the configuration tree through the configuration space, the random samples are generated gradually using the spheres of the tunnels.
+We assume that the tunnel is computed using the standard tunnel detection tool and is represented using a sequence of spheres.
+To guide the configuration tree through the configuration space, the random samples are generated gradually using the spheres of the tunnel.
 As the tunnels are usually computed on protein models without ligands, they are usually very narrow and a ligand can barely traverse them.
@@
 
-The most related work to this paper is the MoMa-LigPath~\cite{cortes2005path}, that is based on the ML-RRT method~\cite{cortes2010simulating,cortes2007mlrrt}.
-In comparison to our solution, the method~\cite{cortes2005path} aims for find any path leading towards the active site without
-possibility to select a tunnel of interest.
+This brings two important advantages: the volume of the configuration space to be searched is decreased, as not the whole
+protein/ligand complex, but only its subspace need to be searched.
+Consequently, the relative volume of the corresponding free region of the configuration space is increased, which leads to easier motion planning.
+The proposed approach allows the researches to analyze a given tunnel for various ligands without necessity to compute the tunnel
+again for each ligand.
 
-\red{comparison to momaligpath~\cite{cortes2005path} + \cite{guieysse2008structure} }
-
-Ligands are small molecules interacting with the protein. 
-Due to the interaction, the relative positions of the ligand's atoms can change.
+%Ligands are small molecules interacting with the protein. 
+%Due to the interaction, the relative positions of the ligand's atoms can change.
 The ligands can be modeled as a multi-link kinematic chain where the torsional angles phi/psi can change in a given range~\cite{songPFpath}.
@@
 %The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
-The flexiblity of a ligand is modeled using the set $\L$ of its conformations.
+The flexibility of a ligand is modeled using the set $\L$ of its conformations.
 The conformations can be prepared e.g., considering the potential energy using tools like Rosetta.
@@
 The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
-We propose a RRT-based method to compute the trajectories of the ligand.
-As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
-suffer from the well known narrow passage problem~\cite{hannaWIS}.
-
-To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
-Instead, they are generated using the spheres of the tunnel.
+The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
+%We propose a RRT-based method to compute the trajectories of the ligand.
+%As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
+%suffer from the well known narrow passage problem~\cite{hannaWIS}.
+%To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
+%Instead, they are generated using the spheres of the tunnel.
 The main loop of the method is described in Alg.~\ref{alg::main}.
@@
 The translation of $\qrand$ is generated from Normal distribution centered at $t_v$ with radius $R_v$.
-The radius $R_v$ is set as the absolut distance from the tunnel centerline, e.g. $R_v=3\AA$.
+The radius $R_v$ is set as the absolute distance from the tunnel centerline, e.g. $R_v=3\AA$.
 Alternatively, $R_v$ can be set to $R_v=r_v$ according the radius of the tunnel at the virtual goal.
@@
 
-The configruation tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
+The configuration tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
 one new configuration can be added.
 The most time consuming part of the method is the collision-detection, which is called $m|\L|$ in each expansion step.
-The collision detection can be evaluated fast using hiearchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}
+The collision detection can be evaluated fast using hierarchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}
 
@@
 On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
-Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
+Moreover, it is not feasible to directly visualize and explore the resulting trajectories due the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
 Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
 In this manner, we aim to further support the decision process of a chemist about a tunnel.
-In order to do that, we employ an ensemble of resulting trajectories, and we compute and visualize the following tunnel properties:
+In order to do that, we employ an ensemble of resulting trajectories in order to compute and visualize the following tunnel properties:
 \begin{itemize}
@@
 \end{itemize}
-During visualization, we map these properties onto a tunnel's surface so that they can be visually correlated with, e.g., the tunnel's shape.
 
-\red{can you please tell in detail how is the accesibility/liugand/throughput mapped to the tunnel surface/spheres? }
+\red{can you please tell in detail how is the accesibility/ligand/throughput mapped to the tunnel surface/spheres? }
 
@@
 Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.
+
+\section{Discussion}
+
+guiding, parametes, different sampling, flexibility
 

### 2017-06-14 paper.tex

@@
 \newcommand{\red}[1]{\textcolor{red}{#1}}
+%\newcommand{\tylde}{\raisebox{-0.85ex}{\textasciitilde}} %does not work for vojta
+\newcommand{\tylde}{$\sim$}
 
@@
 \includegraphics[width=\tmpa]{fig/m003-011} \\
-
 \end{tabular}
@@
 \end{itemize}
+During visualization, we map these properties onto a tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
 
 \red{can you please tell in detail how is the accesibility/ligand/throughput mapped to the tunnel surface/spheres? }
+
+\begin{figure}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[width=0.3\textwidth]{fig/accessibility}
+\quad & \quad
+\includegraphics[width=0.3\textwidth]{fig/ligand-scale}
+\quad & \quad
+\includegraphics[width=0.3\textwidth]{fig/throughput}
+\end{tabular}
+\caption{Tunnel properties mapped onto its surface.
+Left: Accessibility --- green denotes parts that were accesed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
+Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
+Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.
+\label{fig:properties}
+}
+\end{figure}
+
+\red{can you please tell in detail how is the accesibility/liugand/throughput mapped to the tunnel surface/spheres? }
 

### 2017-06-14 paper.tex
Added protein+tunnel images
added table
abstract, intro

--- 2017-06-14 paper.tex
@@
 \newcommand{\red}[1]{\textcolor{red}{#1}}
-\newcommand{\tylde}{\raisebox{-0.85ex}{\textasciitilde}}
 
@@
 \title*{
-    Testing traversability of protein tunnels with flexible ligands using motion planning
+    Testing Traversability of Protein Tunnels with Flexible Ligands Using Motion Planning
 }
@@
     Adam Jur\v{c}\'\i k  \and 
-    David Bend\'{a}\v{r}  \and
+    David Bedn\'{a}\v{r}  \and
     Martin Saska }
@@
 \inst{3}
-TODO DAVID \at TODO DAVID
+David Bedn\'{a}\v{r} \at 
+Faculty of Science,
+Masaryk University,
+Kamenice 5, 625 00 Brno
 }
@@
 \abstract{
-\red{needs rewrite:
-Function and behavior of proteins are achieved by interactions with other molecules such as ligands.
-The interaction takes place in an active site inside the protein, which is accessible from the surface by so called tunnels.
-The knowledge about tunnels can help researches for example to better select target ligand in order to achieve a desired
-behavior of the protein.
-Widely used tools for tunnel detection are based on Voronoi diagrams and they compute the tunnels for a virtual spherical probe.
-The tunnels are then scored based on their bottleneck, length and also curvature. 
-It is however not easy to determine if a given non-spherical ligand can pass such a tunnel.
-To provide the chemists better information about the tunnels, the full shape of ligands should be considered.
-}
-Sampling-based motion planning techniques can be used to compute trajectories for the non-spherical ligands in the protein tunnels.
-In this paper, we propose a modification of Rapidly Exploring Random Tree (RRT) to find the trajectories.
-To enable the ligands to move in the tunnels, the planning is realized with a reduced-scale of the ligands. 
-The flexibility of ligands is achieved by using a predefined set of ligand conformations.
-For the purpose of further analysis, we computed trajectories are visualized ..
+Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
+Molecular tunnels play an important role in catalytic and thermodynamic properties and they connect catalytic active sites with the solvent. 
+Therefore, the geometrical and physico-chemical properties of tunnels need to be determined to understand the transport of ligand from and into the protein active site. 
+Current methods describing the ligand passing through the tunnel are based either only on geometry of the tunnel or on molecular dynamic simulations. 
+In this paper, we propose \red{TODO}
+Additionally, we introduce several approaches to visualization of detected tunnels which depict valuable information about various properties and behavior of the ligand.
+Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
+On the other hand, it is much less computationally demanding than methods based on molecular dynamic simulations. 
+The method can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
 }
 
-\section{Introduction \& motivation}
+\section{Introduction \& Motivation}
 
@@
 
-\red{this is fast copy and paste from previous paper, please rewrite, cut to at ca. 1 page}
+Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
+These interactions are highly influence by the ability to transport the small molecule, called ligand, to the protein active site.
+The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
+To get the ligand to the active site, there has to be the transportation path connecting the protein outer environment with the active site.
+This paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
 
-%Protein structures are essential components of all living organisms.
-Proper understanding of structure and function of proteins is important in many fields, including drug design, agriculture, cosmetics, etc.
-%Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. 
-Computational methods are often used to analyze the geometry of protein structure,  its shape, surface area, and inner void space. 
-Proteins interact with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.) which can lead to changes in properties of the protein or the interacting molecule.
-The chemical reaction between the protein and ligand takes places in so called active site, that is accessible 
-from the protein surface by a tunnel.
-
-The knowledge about the tunnels is important in many fields..
 For example, one task in protein engineering is to change selected properties of a protein, e.g.
@@
 
-To present the results to chemicists ...
-
-\red{Adam: }
-As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
-On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
-Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
-Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
-In this manner, we aim to further support the decision process of a chemist about a tunnel.
-In order to do that, we employ an ensemble of resulting trajectories, and we compute and visualize the following tunnel properties:
-\begin{itemize}
-  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
-  \red{ i.e., number of trajectories ending in the tunnel part over all trajectories? }
-  \item \textbf{Ligand scale} --- The scale of the ligand wheh it passes a tunnel's part.
-  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
-\red{ i.e., the conditional probability we discussed in the mail?}
-\end{itemize}
-During visualization, we map these properties onto a tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
-
-\begin{figure}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[width=0.3\textwidth]{fig/accessibility}
-\quad & \quad
-\includegraphics[width=0.3\textwidth]{fig/ligand-scale}
-\quad & \quad
-\includegraphics[width=0.3\textwidth]{fig/throughput}
-\end{tabular}
-\caption{Tunnel properties mapped onto its surface.
-Left: Accessibility --- green denotes parts that were accesed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
-Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
-Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.}
-\label{fig:properties}
-\end{figure}
-
-\red{can you please tell in detail how is the accesibility/liugand/throughput mapped to the tunnel surface/spheres? }
-
-Moreover, we analyze the whole set of trajectories in order to convey the most important trajectories that enabled the ligand to pass or that prevented it from passing.
-In order to visualize the important trajectories we first cluster the trajectories according to their similarity. 
-\red{can you please add some details how is the similarity counted? Maybe a short description of even equation.}
-Then we choose a representant from each cluster a we visualize its trajectory in 3D using a polyline.
-\red{please some reference to the hierarchical clustering}.
-Moreover, we map the representant's cluster size onto the width of the polyline.
-In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
-
-\begin{figure}
-\centering
-\begin{tabular}{cc}
-\includegraphics[width=0.45\textwidth]{fig/trajectories-all}
-\qquad & \qquad
-\includegraphics[width=0.45\textwidth]{fig/trajectories-clustered-21} \\
-\end{tabular}
-\caption{Visualization of resulting trajectories.
-Left: All trajectories colored according to whether they reached the end of the tunnel (green) or they did not (red).
-Right: Trajectories clustered according to their proximity colored by distinct colors.
-The different line widths represent the number of trajectories in a cluster.}
-\label{fig:trajectories}
-\end{figure}
+To present the results to chemists ...
 
@@
 
-The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase (protein 1CQW) (Fig.~\ref{fig::tunnel}).
-The tunnel was computed using CAVER 3.0~\cite{caver3} for spherical probe of radius 0.9~\AA.
-The tunnel is represented by 760 spheres with average distance $0.05$~\AA.
-
-% from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
-The traversability of the tunnel was evaluated for 1-chlorpropan (denoted as $\LA$) and 1-chlorbutan (denoted as $\LB$).
-
-\begin{figure}
-\centering
-\caption{\label{fig::tunnel}
-    TODO image of the tunnel + 1cqw
-}
-\end{figure}
-
-\begin{table}
-\centering
-\caption{\label{tab::rrt}
-    Runtime and sucess ratio of the planner
-}
-\input table.003.tex
-\hskip 10pt
-%\input table.004.tex
-\end{table}
-
-Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.
-
 \section{Conclusion }
+In the future, we will focus on the following extensions proposed by the domain experts.
+Only ligand flexibility is included in the current algorithm but the flexibility of amino acid side-chains would greatly improve the prediction. 
+Moreover, incorporation of some scoring function for energy evaluation would introduce the characterization of not only geometrical but also physico-chemical properties and would further improve the prediction ability of our approach.
 

### 2017-06-14 paper.tex

--- 2017-06-14 paper.tex
@@
 \and 
-D. Bedn\'a\v{r} \at
-Loschmidt Laboratories, Department of Experimental Biology, Masaryk University, Brno, Czech Republic
-    International Centre for Clinical Research, St. Anne’s University Hospital Brno, Brno, Czech Republic
+\inst{3}
+David Bedn\'{a}\v{r} \at 
+Faculty of Science,
+Masaryk University,
+Kamenice 5, 625 00 Brno
 }
@@
 \abstract{
-\red{needs rewrite:
-Function and behavior of proteins are achieved by interactions with other molecules such as ligands.
-The interaction takes place in an active site inside the protein, which is accessible from the surface by so called tunnels.
-The knowledge about tunnels can help researches for example to better select target ligand in order to achieve a desired
-behavior of the protein.
-Widely used tools for tunnel detection are based on Voronoi diagrams and they compute the tunnels for a virtual spherical probe.
-The tunnels are then scored based on their bottleneck, length and also curvature. 
-It is however not easy to determine if a given non-spherical ligand can pass such a tunnel.
-To provide the chemists better information about the tunnels, the full shape of ligands should be considered.
-}
-Sampling-based motion planning techniques can be used to compute trajectories for the non-spherical ligands in the protein tunnels.
-In this paper, we propose a modification of Rapidly Exploring Random Tree (RRT) to find the trajectories.
-To enable the ligands to move in the tunnels, the planning is realized with a reduced-scale of the ligands. 
-The flexibility of ligands is achieved by using a predefined set of ligand conformations.
-For the purpose of further analysis, we computed trajectories are visualized ..
-}
-
-\section{Introduction \& motivation}
+Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
+Molecular tunnels play an important role in catalytic and thermodynamic properties and they connect catalytic active sites with the solvent. 
+Therefore, the geometrical and physico-chemical properties of tunnels need to be determined to understand the transport of ligand from and into the protein active site. 
+Current methods describing the ligand passing through the tunnel are based either only on geometry of the tunnel or on molecular dynamic simulations. 
+In this paper, we propose \red{TODO}
+Additionally, we introduce several approaches to visualization of detected tunnels which depict valuable information about various properties and behavior of the ligand.
+Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
+On the other hand, it is much less computationally demanding than methods based on molecular dynamic simulations. 
+The method can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
+}
+
+\section{Introduction \& Motivation}
 
@@
 
-\red{this is fast copy and paste from previous paper, please rewrite, cut to at ca. 1 page}
-
-%Protein structures are essential components of all living organisms.
-Proper understanding of structure and function of proteins is important in many fields, including drug design, agriculture, cosmetics, etc.
-%Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. 
-Computational methods are often used to analyze the geometry of protein structure,  its shape, surface area, and inner void space. 
-Proteins interact with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.) which can lead to changes in properties of the protein or the interacting molecule.
-The chemical reaction between the protein and ligand takes places in so called active site, that is accessible 
-from the protein surface by a tunnel.
-
-The knowledge about the tunnels is important in many fields..
+Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
+These interactions are highly influence by the ability to transport the small molecule, called ligand, to the protein active site.
+The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
+To get the ligand to the active site, there has to be the transportation path connecting the protein outer environment with the active site.
+This paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
+
 For example, one task in protein engineering is to change selected properties of a protein, e.g.
@@
 
-To present the results to chemicists ...
+To present the results to chemists ...
 
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.45\textwidth]{fig/trajectories-all}
+\includegraphics[width=0.4\textwidth]{fig/trajectories-all}
 \qquad & \qquad
-\includegraphics[width=0.45\textwidth]{fig/trajectories-clustered-21} \\
+\includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
 \end{tabular}
@@
 \centering
+\includegraphics[width=0.5\textwidth]{fig/protein-tunnel-gray}
 \caption{\label{fig::tunnel}
-    TODO image of the tunnel + 1cqw
+    The protein 1CQW visualized using its secondary structure (gray) with its highest ranked tunnel (green) which was detected using CAVER 3.0.
 }
@@
 \hskip 10pt
-%\input table.004.tex
+\input table.004.tex
 \end{table}
@@
 \section{Conclusion }
+In the future, we will focus on the following extensions proposed by the domain experts.
+Only ligand flexibility is included in the current algorithm but the flexibility of amino acid side-chains would greatly improve the prediction. 
+Moreover, incorporation of some scoring function for energy evaluation would introduce the characterization of not only geometrical but also physico-chemical properties and would further improve the prediction ability of our approach.
 

### 2017-06-15 paper.tex
Added details to properties

@@
 \usepackage{amsmath}
+\usepackage{amsfonts}
 \usepackage[ruled,lined]{algorithm2e}
@@
 In this manner, we aim to further support the decision process of a chemist about a tunnel.
-In order to do that, we employ an ensemble of resulting trajectories, and we compute and visualize the following tunnel properties:
+In order to do that, we employ an ensemble of resulting trajectories $T$, and we compute and visualize the following tunnel properties:
 \begin{itemize}
   \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
-  \red{ i.e., number of trajectories ending in the tunnel part over all trajectories? }
-  \item \textbf{Ligand scale} --- The scale of the ligand wheh it passes a tunnel's part.
+We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \mathbb{R}$ is radius.
+Then for each trajectory $t$ we define its ending tunnel part $P_t^e$ as a tunnel part whose center is closest to the end position of $t$.
+Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
+  \item \textbf{Ligand scale} --- The scale of the ligand when it passes a tunnel's part.
+For the trajectory $t$ the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as a scale of a such sample of $t$ whose position is the closest to the center of $P_t$.
+Then, the average ligand scale in a tunnel part is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
+Similarly, minimum and maximum ligand scales can be defined.
   \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
-\red{ i.e., the conditional probability we discussed in the mail?}
+The througput of the tunnel part $P$ is defined as a ratio of trajectories that passed $P$ over all trajectories that reached $P$.
 \end{itemize}
-During visualization, we map these properties onto a tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
-
-\red{can you please tell in detail how is the accesibility/ligand/throughput mapped to the tunnel surface/spheres? }
+In the visualization, these properties are mapped onto a tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
+More precisely, a point $x$ on the tunnel surface is colored according to the property value in its nearest tunnel part $P$, i.e., such part $P$ whose center is the closest to $x$ among all tunnel parts.
 
@@
 Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
-Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.
+Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.}
 \label{fig:properties}
-}
 \end{figure}
-
-\red{can you please tell in detail how is the accesibility/liugand/throughput mapped to the tunnel surface/spheres? }
 

### 2017-06-15 paper.tex
planning sota

--- 2017-06-14 paper.tex
@@
 \and 
-\inst{3}
 David Bedn\'{a}\v{r} \at 
@@
 
+\vskip -50pt
 \abstract{
@@
 Therefore, the geometrical and physico-chemical properties of tunnels need to be determined to understand the transport of ligand from and into the protein active site. 
-Current methods describing the ligand passing through the tunnel are based either only on geometry of the tunnel or on molecular dynamic simulations. 
-In this paper, we propose \red{TODO}
-Additionally, we introduce several approaches to visualization of detected tunnels which depict valuable information about various properties and behavior of the ligand.
+Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamic simulations. 
+In this paper, we propose to analyze the ligand passing using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT).
+The tunnel is represented by a set of spheres, that are used to guide the growth of the RRT configuration tree in the related configuration space.
+Multiple trajectories are computed using the proposed planner in order to analyze expected behavior of the ligand in the tunnel.
+We introduce several approaches to visualization of detected tunnels which depict valuable information about various properties and behavior of the ligand.
 Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
@@
         the radius of the smallest sphere that can traverse the whole tunnel).
-%The active sites are usually connected with the protein surface with multiple pathways.
-%However, not all paths are biochemically relevant, so only paths with a given minimal bottleneck have to be reported (the bottleneck is
-%the radius of the smallest sphere that can traverse the whole tunnel).
-
-%Such decision is however not precise and it requires previous expertise in the domain.
-Typical ligands are however non-spherical and therefore it is difficult to estimate their traversability in tunnels
-computed for a spherical probe.
-It is therefore necessary to consider also the shape of the ligand and possibly also its flexibility.
-Due to the increase degrees of freedom (DOF) that need to be considered, the classical Voronoi-based methods cannot be used 
-to compute trajectories of the ligands.
-
-\red{TODO}
-something about screening: what is it, what's is the input and output, in which applications (research/industry) it is useful.
+Typical ligands are however non-spherical and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
+The decision based only on the spherical tunnels requires previour experties in the domain and yet it may be imprecise.
+
+To provide chemists more insight into the behavior of non-spherical ligands, it is necessary to compute the trajecotries of the ligand considering it's shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.
 
@@
 can cope with many-DOF objects of arbitrary geometry.
-The random samples for RRT are generated using the spheres of the tunnel.
+The random samples for RRT are generated using the spheres representing the tunnel in order to guide the configuration tree through the search space.
 To make the transportation easier, a scaled-down version of the ligand is used, similarly e.g. to~\cite{cortes2005path}.
@@
 
-%Recently, researchers started to focus on molecular dynamics simulations and study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
-%Recently, researchers started to focus on molecular dynamics simulations and studying the behavior of individual 
-%tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
-%Molecular dynamics is represented as a sequence of frames (molecule snapshots), and the tunnels need to be detected through these frames.
-%With the increasing size of simulations the correctness of predicting the most biochemically relevant tunnel increases as well.
-%Existing solutions based exclusively on Voronoi diagrams and clustering methods are time and memory consuming
-%which also gives the limitation for the maximum number of frames which they are able to analyze.
-%In such cases, the biochemists have to select a subset of the whole simulation and perform the analysis only for this selection.
-%Such analysis however provides only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted.
-%This opens new possibilities for alternative solutions enabling to explore the tunnels in each frame of the molecular dynamics.
-%In such cases the biochemists have to skip each n-th snapshot or select a subset of the whole simulation and perform the analysis only for this selection.
-%The properties of tunnels like bottleneck radius, their length and the surrounding residues can be used
-%to estimate suitability of the tunnel for a ligand.
-%However, such estimation still involve lot of experience.
+\red{TODO}
+something about screening: what is it, what's is the input and output, in which applications (research/industry) it is useful.
 
@@
 VD-based methods are memory less demanding, and also faster than the grid-based methods.
+\red{asi vyhodit?
 The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
@@
 The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
+}
 
@@
 The biggest disadvantage is that the shape of the ligand is not considered during the tunnel detection and it is therefore
-not easy to estimate if (and how) a ligand might traverse the tunnel.
-
-The task of trajectory computation for the non-spherical ligand can be formulated as a path search 
-in the related configuration space $\C$.
+not easy to estimate if (and how) a ligand might traverse the tunnels.
+%The protein tunnels computed for spherical probes contain valuable information about the protein void space.
+%Based on the biochemical properties, the chemists can decide if a given tunnel is relevant and can be possibly used by a ligand to exit the protein.
+%The biochemically relevant tunnels can be then analyzed separately.
+
+This requires to compute trajectorie of the non-spherical ligand in the tunnels, which can be formulated as a path planning problem in a high-dimensional configuration space $\C$.
+The configuration space is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation and possibly also other degrees of freedom responsible for conformation changes.
 The dimension of the configuration space is given my the DOF of the ligand, i.e., 6D for a rigid and 6D + $n$ for a flexible
 ligand with $n$ DOFs.
-%We propose to employ motion planning technique to determine possible trjaectories for a given ligand in the detected tunnels.
-%This can be solved by search a high-dimensional configuration space $\C$. % as both 3D translation, rotation possibly with additional DOFs caused
-%by the ligand's flexiblity need to be considered.
-%The dimension of the configuration space $\C$ is given by the degrees of freedom (6D for a solid ligand and 6D+$n$ for a flexible ligand with
-%$n$ rotational bonds).
-%The feasible configurations, where the ligand (robot) does not touch and obstacle, form the free region $\CF$.
-%The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
-%to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the beginning and the end of the
-%tunnel, respectively.
-%Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
-The idea of sampling-based motion planning is to randomly sample the configuration space.
-The random samples are classified as free or non-free using collision detection and  the free ones are stored into a roadmap.
+Sampling-based motion planning methods~\cite{Lav06} can be used to search this high-dimensional configuration space.
+The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
+The free samples are stored in a roadmap (a graph structures), in which a path can be searched using standard graph-search methods.
 A path in the roadmap then represents a motion in the workspace.
-Beyond robotics, sampling-based motion planners have been successfully applied in computational biology in studies
-related to 
-protein folding~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro}, analysis of loop motions~\cite{cortes2004geometric},
+
+Sampling-based methods can cope with robots (objects) with many DOFs. 
+Flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
+can changes~\cite{songPFpath}.
+This is useful e.g.
+in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
+Sampling-based methods have also been used for 
  tunnel detection~\cite{vonasek2016application,vonasek2017tunnel} and
@@
 %none of the existing work focuses on the trajectory generation in tunnels.
+%However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
 
@@
 incrementally builds a configuration tree $\T$ rooted at the initial configuration $\qinit$.
-In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node in the tree $\qnear \in \T$ is found.
+In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
 A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$.
@@
 The algorithm terminates if the tree approaches the goal configuration close enough.
+RRT is a suitable candidate for computing the trajectories of flexible ligands due to its ability to cope
+with many-DOF systems.
+Two main issues however need to be adressed: the increased dimension of the configuration space and the narrow passage problem.
 
@@
 
-A RRT-based method for computing exit pathways for a small molecule was presented in~\cite{cortes2010simulating}.
-Both protein and ligand are represented as flexible kinematic chains.
-To decrease the dimension of the related configuration space, the RRT-ML variant~\cite{cortes2007mlrrt} is employed.
+The kinematic chain representaion used to model ligand flexibility allows the sampling-based planners to generate all possible conformations.
+However, not all of them are feasible and it is therefore necessary to verify feasibility of the given conformation based on energy, which is time consuming.
+Moreover, the aditional degrees of freedom required to model the flexiblity increases dimension of the configuration space.
+RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
+
+A RRT-based method for computing exit pathways for a small flexible molecule was presented in~\cite{cortes2010simulating}.
+To cope with the many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} was employed.
 In RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
-        and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder
-the growth of the tree.
-The RRT-ML is also used in the MoMa-LigPath online service.
-
-The~\cite{cortes2010simulating} approach aims to find any exit pathway for a given ligand, which requires
+and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder the growth of the tree.
+The approach~\cite{cortes2010simulating} aims to find any exit pathway for a given ligand, which requires
 to search the whole configuration space of the ligand/protein complex.
@@
 properties of the narrow passage~\cite{hannaWIS}.
+A narrow passage is a region in the configuration space whose removal changes the connectivy of the free space.
+Narrow passages have smaller volume than other areas in it is therefore diffuclt to sample them dense enough using
+uniform distribution.
 Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
 Many iterations are needed to find a solution, which increases the computational time.
-
-The protein tunnels computed for spherical probes contain valuable information about the protein void space.
-Based on the biochemical properties, the chemists can decide if a given tunnel is relevant and can be possibly
-used by a ligand to exit the protein.
-The biochemically relevant tunnels can be then analyzed separately.
 
@@
 %Due to the interaction, the relative positions of the ligand's atoms can change.
-The ligands can be modeled as a multi-link kinematic chain where the torsional angles phi/psi can change in a given range~\cite{songPFpath}.
-After the torsional angles are set, 3D position of each atom of the ligand is computed.
-By changing the angles, various conformations can be generated.
-However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
 
@@
 %f-PRM~\cite{anshelevich2000deformable}, physically coreect deformations, time consuming
+
+%Recently, researchers started to focus on molecular dynamics simulations and study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
+%Recently, researchers started to focus on molecular dynamics simulations and studying the behavior of individual 
+%tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
+%Molecular dynamics is represented as a sequence of frames (molecule snapshots), and the tunnels need to be detected through these frames.
+%With the increasing size of simulations the correctness of predicting the most biochemically relevant tunnel increases as well.
+%Existing solutions based exclusively on Voronoi diagrams and clustering methods are time and memory consuming
+%which also gives the limitation for the maximum number of frames which they are able to analyze.
+%In such cases, the biochemists have to select a subset of the whole simulation and perform the analysis only for this selection.
+%Such analysis however provides only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted.
+%This opens new possibilities for alternative solutions enabling to explore the tunnels in each frame of the molecular dynamics.
+%In such cases the biochemists have to skip each n-th snapshot or select a subset of the whole simulation and perform the analysis only for this selection.
+%The properties of tunnels like bottleneck radius, their length and the surrounding residues can be used
+%to estimate suitability of the tunnel for a ligand.
+%However, such estimation still involve lot of experience.

### 2017-06-15 paper.tex
sota planning, algoritmhs

@@
 
+\def\rv{T_w}
+
 \def\Imax{I_{max}} %max number of iterations of RRT-based planners
@@
 Typical ligands are however non-spherical and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
-The decision based only on the spherical tunnels requires previour experties in the domain and yet it may be imprecise.
-
-To provide chemists more insight into the behavior of non-spherical ligands, it is necessary to compute the trajecotries of the ligand considering it's shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.
+The decision based only on the spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
+
+To provide chemists more insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering it's shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.
 
@@
 
-This requires to compute trajectorie of the non-spherical ligand in the tunnels, which can be formulated as a path planning problem in a high-dimensional configuration space $\C$.
-The configuration space is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation and possibly also other degrees of freedom responsible for conformation changes.
-The dimension of the configuration space is given my the DOF of the ligand, i.e., 6D for a rigid and 6D + $n$ for a flexible
+This requires to compute trajectories of the non-spherical ligand in the tunnels, which can be formulated as a path planning problem in a high-dimensional configuration space $\C$.
+The configuration space is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation and possibly also other degrees of freedom responsible for the conformation changes.
+The dimension of the configuration space is given by the DOF of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
 ligand with $n$ DOFs.
@@
 The free samples are stored in a roadmap (a graph structures), in which a path can be searched using standard graph-search methods.
-A path in the roadmap then represents a motion in the workspace.
+%A path in the roadmap then represents a motion in the workspace.
 
@@
 
-The kinematic chain representaion used to model ligand flexibility allows the sampling-based planners to generate all possible conformations.
+A well known issue of sampling-based planners is the narrow passage problem~\cite{hannaWIS}.
+A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space.
+Narrow passages have smaller volume than other regions and it is therefore difficult to sample them dense enough using uniform distribution, that is used in the basic sampling-based planners.
+The presence of narrow passages in the configuration space, especially if they contain part of the solution, requires many iterations
+in order to put enough samples there and it consequently increases the planning time.
+The protein tunnels have the properties of the narrow passages. 
+A possible approach to cope with the narrow passage property of the protein tunnels is to use the information
+about their 3D position, which is represented by the tunnel centerline.
+The centerline can be used to guide the growth of the RRT tree through the configuration space~\cite{vonasek2009rrt,denny2016dynamic}.
+
+The kinematic chain representation used to model ligand flexibility allows the sampling-based planners to generate all possible conformations.
 However, not all of them are feasible and it is therefore necessary to verify feasibility of the given conformation based on energy, which is time consuming.
-Moreover, the aditional degrees of freedom required to model the flexiblity increases dimension of the configuration space.
+Moreover, the additional degrees of freedom required to model the flexibility increases dimension of the configuration space.
 RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
@@
 and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder the growth of the tree.
-The approach~\cite{cortes2010simulating} aims to find any exit pathway for a given ligand, which requires
-to search the whole configuration space of the ligand/protein complex.
-This might be a possible disadvantage, as the proteins are dense structures with relatively narrow void space, which has the
-properties of the narrow passage~\cite{hannaWIS}.
-A narrow passage is a region in the configuration space whose removal changes the connectivy of the free space.
-Narrow passages have smaller volume than other areas in it is therefore diffuclt to sample them dense enough using
-uniform distribution.
+The approach~\cite{cortes2010simulating} may however suffer from the narrow passage problem, as it aims to find any exit pathway for a given ligand, which requires to search the whole configuration space of the ligand/protein complex.
 Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
@@
 
-In this paper, we propose to employ RRT to compute the trajectory for a given ligand in a given tunnel.
-We assume that the tunnel is computed using the standard tunnel detection tool and is represented using a sequence of spheres.
-To guide the configuration tree through the configuration space, the random samples are generated gradually using the spheres of the tunnel.
-As the tunnels are usually computed on protein models without ligands, they are usually very narrow and a ligand can barely traverse them.
-It can be assume that both the protein and the ligand adapt to each other and therefore.
-To enable the planner find a solution, the radii of the ligand, and possibly also the radii of the protein atoms, are down-scaled
-by a given factor.
-
+In this paper, we propose to analyze the traversability separately for each tunnel.
+This can bring several advantages for the biochemists.
+The tunnels in a protein being studied can be computed using standard tunnel detection tools for spherical probes and assessed
+according their length, curvature, bottleneck and biochemical properties (e.g. partial charge).
+These characteristics are already used in many research studies.
+Selected promising tunnels can be further analyzed for a given ligand to determine the traversability.
+It is not necessary to first run time consuming MD simulations with both protein and ligand being studied, but it is suitable
+to detected tunnels in the protein only.
+
+The proposed work employs the RRT planning principle with a three extensions: 
+a) the configuration space is not sampled uniformly, but using a guided sampling along the tunnel centerline,
+b) the radii of the ligand atoms are decreased, and 
+c) ligand flexibility of modeled using a set of predefined energy-suitable conformations.
+The first two extensions can help to cope with the narrow passage problem and the third modification
+is introduced in order to decrease the dimension of the configuration space and the sensitivity to the employed
+metric.
+
+\red{presunu jinam}
 This brings two important advantages: the volume of the configuration space to be searched is decreased, as not the whole
@@
 The flexibility of a ligand is modeled using the set $\L$ of its conformations.
-The conformations can be prepared e.g., considering the potential energy using tools like Rosetta.
+The conformations can be prepared e.g., considering the potential energy.
 We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
@@
 
-A protein tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, $i=1,\ldots,n$, where $n$ denotes the number of spheres,
+A protein tunnel is described by a sequence of collision-free spheres 
+$T=( (t_1, r_1),\ldots,(t_n,r_n) )$, where $n$ denotes the number of spheres,
 $t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
@@
 The proteins are dense structures with narrow tunnels.
-Depending on the protein, the tunnel bottlenecks can be even less then $1$~\AA, which is too narrow for ligands with more than 2 atoms.
+Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
 For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
@@
 All possible configurations form the configuration space $\C$. 
-A configuration is collision-free if there is no intersection between the protein atoms and ligand atoms scaled by $s$.
+A configuration is collision-free if none of the ligand atoms scaled by $s$ and placed at the
+position defined by $q$ collide with the protein atoms.
 %The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.
@@
 
-The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
-The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
+The traversability study proposed in this paper is achieved by computing many trajectories for the given ligand and tunnel, which
+is solved using a novel RRT-based planner.
+The main principle of the RRT planner is extended by guided sampling, scaling of the ligands atoms and using predefined set of ligand
+conformations.
+The main loop of the method is described in Alg.~\ref{alg::main}.
+%The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
 %We propose a RRT-based method to compute the trajectories of the ligand.
@@
 %Instead, they are generated using the spheres of the tunnel.
-The main loop of the method is described in Alg.~\ref{alg::main}.
 In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
@@
 
-To guide thee growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
-The virtual goal $v$ is the index of a sphere, around which the random samples are generated with probability $\gb$.
+To guide tree growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
+The virtual goal $v, 1\le v \le n$, is the index of a tunnel sphere.
+The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated
+around the virtual goal sphere $v$.
+%around which the random samples are generated with probability $\gb$.
 %The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
@@
 that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
-Setting the virtual goal to this successor allows the tree to ``skip'' such parts of the tunnels that are
-not traversible or reachable by the ligand.
+Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are
+not traversable or reachable by the ligand.
+This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
 The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
-the last sphere in the tunnel, i.e., when $v = N$.
-
-The random samples $\qrand$ are generated around the virtual goal.
-The translation of $\qrand$ is generated from Normal distribution centered at $t_v$ with radius $R_v$.
-The radius $R_v$ is set as the absolute distance from the tunnel centerline, e.g. $R_v=3\AA$.
-Alternatively, $R_v$ can be set to $R_v=r_v$ according the radius of the tunnel at the virtual goal.
+the last sphere in the tunnel, i.e., when $v = n$.
+
+To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
+from $N(t_v,\Sigma)$, where $\Sigma$ is diagonal matrix with diagonal entries equal to a parameter $\rv$, and the rotational
+part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
+The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
+metric for nearest-neighbor search.
 
@@
 \KwIn{
-    tunnel $T=(t_i, r_i), n=1,\ldots,N$ represented by $N$ spheres,
+    tunnel spheres $t_i$ and their radii $r_i$, $i=1,\ldots,n$, 
     initial configuration $\qinit$,
@@
 \hrule
-$v = 0$; // index of the virtual goal\\
+$v = 1$; // index of the virtual goal\\
 $iteration = 0$\;
-\While{$iteration < \Imax$ {\bf and}  $v < N$}{
+\While{$iteration < \Imax$ {\bf and}  $v < n$}{
     \eIf{$rand() < \gb$}{
@@
     $\qnear$ = nearest node in $\T$ towards $\qrand$\;
-
     expand($\qnear$)\;
-
-   % \For{$q \in S$}{
-    %    $\T$.addNode($q$)\;
-    %    $\T$.addEdge($\qnear, q$)\;
-    %}
-
-    \For{$i= N-1,N-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
+    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
         $d$ = nearest node in the tree towards sphere $t_i$\;
@@
 If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
-until a free collision-free samples is found or until the mimimal reduced-scale $\smin$ is reached.
+until a free collision-free samples is found or until the minimal reduced-scale $\smin$ is reached.
 
@@
 
-The 3D position of $q$ is generated using Normal distribution $N(0,\sigma^2)$ around $\qnear$, where $\sigma^2$ is the predefined
-parameter.
-The rotation part of $q$ is generated using $U(0,2\pi)$.
-The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated
-with $s$, the next lower $s$ is used.
+The random samples $q$ are generated similarly as in the case of $\qrand$ samples, only their translation
+part is generated around $\qnear$.
+The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated with $s$, the next lower $s$ is used.
 The expansion step is repeated for each conformation $l \in L$.
@@
   \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
-The througput of the tunnel part $P$ is defined as a ratio of trajectories that passed $P$ over all trajectories that reached $P$.
+The throughput of the tunnel part $P$ is defined as a ratio of trajectories that passed $P$ over all trajectories that reached $P$.
 \end{itemize}
@@
 \caption{Tunnel properties mapped onto its surface.
-Left: Accessibility --- green denotes parts that were accesed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
+Left: Accessibility --- green denotes parts that were accessed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
 Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
@@
 
-guiding, parametes, different sampling, flexibility
+guiding, parameters, different sampling, flexibility
 

### 2017-06-15 paper.tex
Added clustering details

--- 2017-06-15 paper.tex
@@
 \and 
+\inst{3}
 David Bedn\'{a}\v{r} \at 
@@
 
-\vskip -50pt
 \abstract{
@@
 Therefore, the geometrical and physico-chemical properties of tunnels need to be determined to understand the transport of ligand from and into the protein active site. 
-Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamic simulations. 
-In this paper, we propose to analyze the ligand passing using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT).
-The tunnel is represented by a set of spheres, that are used to guide the growth of the RRT configuration tree in the related configuration space.
-Multiple trajectories are computed using the proposed planner in order to analyze expected behavior of the ligand in the tunnel.
-We introduce several approaches to visualization of detected tunnels which depict valuable information about various properties and behavior of the ligand.
+Current methods describing the ligand passing through the tunnel are based either only on geometry of the tunnel or on molecular dynamic simulations. 
+In this paper, we propose \red{TODO}
+Additionally, we introduce several approaches to visualization of detected tunnels which depict valuable information about various properties and behavior of the ligand.
 Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
@@
         the radius of the smallest sphere that can traverse the whole tunnel).
-Typical ligands are however non-spherical and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
-The decision based only on the spherical tunnels requires previour experties in the domain and yet it may be imprecise.
-
-To provide chemists more insight into the behavior of non-spherical ligands, it is necessary to compute the trajecotries of the ligand considering it's shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.
+%The active sites are usually connected with the protein surface with multiple pathways.
+%However, not all paths are biochemically relevant, so only paths with a given minimal bottleneck have to be reported (the bottleneck is
+%the radius of the smallest sphere that can traverse the whole tunnel).
+
+%Such decision is however not precise and it requires previous expertise in the domain.
+Typical ligands are however non-spherical and therefore it is difficult to estimate their traversability in tunnels
+computed for a spherical probe.
+It is therefore necessary to consider also the shape of the ligand and possibly also its flexibility.
+Due to the increase degrees of freedom (DOF) that need to be considered, the classical Voronoi-based methods cannot be used 
+to compute trajectories of the ligands.
+
+\red{TODO}
+something about screening: what is it, what's is the input and output, in which applications (research/industry) it is useful.
 
@@
 can cope with many-DOF objects of arbitrary geometry.
-The random samples for RRT are generated using the spheres representing the tunnel in order to guide the configuration tree through the search space.
+The random samples for RRT are generated using the spheres of the tunnel.
 To make the transportation easier, a scaled-down version of the ligand is used, similarly e.g. to~\cite{cortes2005path}.
@@
 
-\red{TODO}
-something about screening: what is it, what's is the input and output, in which applications (research/industry) it is useful.
+%Recently, researchers started to focus on molecular dynamics simulations and study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
+%Recently, researchers started to focus on molecular dynamics simulations and studying the behavior of individual 
+%tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
+%Molecular dynamics is represented as a sequence of frames (molecule snapshots), and the tunnels need to be detected through these frames.
+%With the increasing size of simulations the correctness of predicting the most biochemically relevant tunnel increases as well.
+%Existing solutions based exclusively on Voronoi diagrams and clustering methods are time and memory consuming
+%which also gives the limitation for the maximum number of frames which they are able to analyze.
+%In such cases, the biochemists have to select a subset of the whole simulation and perform the analysis only for this selection.
+%Such analysis however provides only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted.
+%This opens new possibilities for alternative solutions enabling to explore the tunnels in each frame of the molecular dynamics.
+%In such cases the biochemists have to skip each n-th snapshot or select a subset of the whole simulation and perform the analysis only for this selection.
+%The properties of tunnels like bottleneck radius, their length and the surrounding residues can be used
+%to estimate suitability of the tunnel for a ligand.
+%However, such estimation still involve lot of experience.
 
@@
 VD-based methods are memory less demanding, and also faster than the grid-based methods.
-\red{asi vyhodit?
 The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
@@
 The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
-}
 
@@
 The biggest disadvantage is that the shape of the ligand is not considered during the tunnel detection and it is therefore
-not easy to estimate if (and how) a ligand might traverse the tunnels.
-%The protein tunnels computed for spherical probes contain valuable information about the protein void space.
-%Based on the biochemical properties, the chemists can decide if a given tunnel is relevant and can be possibly used by a ligand to exit the protein.
-%The biochemically relevant tunnels can be then analyzed separately.
-
-This requires to compute trajectorie of the non-spherical ligand in the tunnels, which can be formulated as a path planning problem in a high-dimensional configuration space $\C$.
-The configuration space is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation and possibly also other degrees of freedom responsible for conformation changes.
+not easy to estimate if (and how) a ligand might traverse the tunnel.
+
+The task of trajectory computation for the non-spherical ligand can be formulated as a path search 
+in the related configuration space $\C$.
 The dimension of the configuration space is given my the DOF of the ligand, i.e., 6D for a rigid and 6D + $n$ for a flexible
 ligand with $n$ DOFs.
-Sampling-based motion planning methods~\cite{Lav06} can be used to search this high-dimensional configuration space.
-The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
-The free samples are stored in a roadmap (a graph structures), in which a path can be searched using standard graph-search methods.
+%We propose to employ motion planning technique to determine possible trjaectories for a given ligand in the detected tunnels.
+%This can be solved by search a high-dimensional configuration space $\C$. % as both 3D translation, rotation possibly with additional DOFs caused
+%by the ligand's flexiblity need to be considered.
+%The dimension of the configuration space $\C$ is given by the degrees of freedom (6D for a solid ligand and 6D+$n$ for a flexible ligand with
+%$n$ rotational bonds).
+%The feasible configurations, where the ligand (robot) does not touch and obstacle, form the free region $\CF$.
+%The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
+%to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the beginning and the end of the
+%tunnel, respectively.
+%Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
+The idea of sampling-based motion planning is to randomly sample the configuration space.
+The random samples are classified as free or non-free using collision detection and  the free ones are stored into a roadmap.
 A path in the roadmap then represents a motion in the workspace.
-
-Sampling-based methods can cope with robots (objects) with many DOFs. 
-Flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
-can changes~\cite{songPFpath}.
-This is useful e.g.
-in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
-Sampling-based methods have also been used for 
+Beyond robotics, sampling-based motion planners have been successfully applied in computational biology in studies
+related to 
+protein folding~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro}, analysis of loop motions~\cite{cortes2004geometric},
  tunnel detection~\cite{vonasek2016application,vonasek2017tunnel} and
@@
 %none of the existing work focuses on the trajectory generation in tunnels.
-%However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
 
@@
 incrementally builds a configuration tree $\T$ rooted at the initial configuration $\qinit$.
-In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
+In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node in the tree $\qnear \in \T$ is found.
 A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$.
@@
 The algorithm terminates if the tree approaches the goal configuration close enough.
-RRT is a suitable candidate for computing the trajectories of flexible ligands due to its ability to cope
-with many-DOF systems.
-Two main issues however need to be adressed: the increased dimension of the configuration space and the narrow passage problem.
 
@@
 
-The kinematic chain representaion used to model ligand flexibility allows the sampling-based planners to generate all possible conformations.
-However, not all of them are feasible and it is therefore necessary to verify feasibility of the given conformation based on energy, which is time consuming.
-Moreover, the aditional degrees of freedom required to model the flexiblity increases dimension of the configuration space.
-RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
-
-A RRT-based method for computing exit pathways for a small flexible molecule was presented in~\cite{cortes2010simulating}.
-To cope with the many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} was employed.
+A RRT-based method for computing exit pathways for a small molecule was presented in~\cite{cortes2010simulating}.
+Both protein and ligand are represented as flexible kinematic chains.
+To decrease the dimension of the related configuration space, the RRT-ML variant~\cite{cortes2007mlrrt} is employed.
 In RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
-and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder the growth of the tree.
-The approach~\cite{cortes2010simulating} aims to find any exit pathway for a given ligand, which requires
+        and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder
+the growth of the tree.
+The RRT-ML is also used in the MoMa-LigPath online service.
+
+The~\cite{cortes2010simulating} approach aims to find any exit pathway for a given ligand, which requires
 to search the whole configuration space of the ligand/protein complex.
@@
 properties of the narrow passage~\cite{hannaWIS}.
-A narrow passage is a region in the configuration space whose removal changes the connectivy of the free space.
-Narrow passages have smaller volume than other areas in it is therefore diffuclt to sample them dense enough using
-uniform distribution.
 Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
 Many iterations are needed to find a solution, which increases the computational time.
+
+The protein tunnels computed for spherical probes contain valuable information about the protein void space.
+Based on the biochemical properties, the chemists can decide if a given tunnel is relevant and can be possibly
+used by a ligand to exit the protein.
+The biochemically relevant tunnels can be then analyzed separately.
 
@@
 %Due to the interaction, the relative positions of the ligand's atoms can change.
+The ligands can be modeled as a multi-link kinematic chain where the torsional angles phi/psi can change in a given range~\cite{songPFpath}.
+After the torsional angles are set, 3D position of each atom of the ligand is computed.
+By changing the angles, various conformations can be generated.
+However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
 
@@
 In order to visualize the important trajectories we first cluster the trajectories according to their similarity.
-\red{can you please add some details how is the similarity counted? Maybe a short description of even equation.}
-Then we choose a representant from each cluster a we visualize its trajectory in 3D using a polyline.
-\red{please some reference to the hierarchical clustering}.
+We define the similarity of two trajectories as follows.
+First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
+In order to obtain the samples, we find the average starting point $x_s$ of all trajectories and the most distant (w.r.t. $x_s$) ending point $x_e$ also among all trajectories.
+Using $x_s$ and $x_e$ we define spheres $S_1, ..., S_N$ such that $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
+Then, a vector $v$ for a trajectory $t$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $t$ which is nearest to the surface of $S_i$.
+Finally, we define a metric $d$ for trajectories $t_i$ and $t_j$ such as $d(t_i, t_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
+The clustering is then computed using the UPGMA clustering algorithm~\cite{sokal1958statistical}.
+When visualizing we choose a representant from each cluster a we depict its trajectory in 3D using a polyline.
 Moreover, we map the representant's cluster size onto the width of the polyline.
@@
 %f-PRM~\cite{anshelevich2000deformable}, physically coreect deformations, time consuming
-
-%Recently, researchers started to focus on molecular dynamics simulations and study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
-%Recently, researchers started to focus on molecular dynamics simulations and studying the behavior of individual 
-%tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
-%Molecular dynamics is represented as a sequence of frames (molecule snapshots), and the tunnels need to be detected through these frames.
-%With the increasing size of simulations the correctness of predicting the most biochemically relevant tunnel increases as well.
-%Existing solutions based exclusively on Voronoi diagrams and clustering methods are time and memory consuming
-%which also gives the limitation for the maximum number of frames which they are able to analyze.
-%In such cases, the biochemists have to select a subset of the whole simulation and perform the analysis only for this selection.
-%Such analysis however provides only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted.
-%This opens new possibilities for alternative solutions enabling to explore the tunnels in each frame of the molecular dynamics.
-%In such cases the biochemists have to skip each n-th snapshot or select a subset of the whole simulation and perform the analysis only for this selection.
-%The properties of tunnels like bottleneck radius, their length and the surrounding residues can be used
-%to estimate suitability of the tunnel for a ligand.
-%However, such estimation still involve lot of experience.

### 2017-06-15 paper.tex
images of confomrations

--- 2017-06-15 paper.tex
@@
 \usepackage{amsfonts}
+\usepackage{subfig}
 \usepackage[ruled,lined]{algorithm2e}
@@
 \institute{
-    V. Von\'{a}sek, Martin Saska \at 
+    V. Von\'{a}sek, Martin Saska:
     Faculty of Electrical Engineering,  
 Czech Technical University in Prague, 
-Technick\'a 2, 166 27, Prague 6, Czech Republic
-\email{vonasek@labe.felk.cvut.cz}  
-\and
-B. Kozl\'{\i}kov\'{a}, A. Jur\v{c}\'{\i}k \at
+Technick\'a 2, 166 27 Prague, Czech Republic
+\email{vonasek@labe.felk.cvut.cz}   \\
+%\and
+B. Kozl\'{\i}kov\'{a}, A. Jur\v{c}\'{\i}k:
 Faculty of Informatics,  
@@
 Botanick\'a 68a, 602 00 Brno, \\
-\and 
-David Bedn\'{a}\v{r} \at 
+%w\and 
+David Bedn\'{a}\v{r}:
 Faculty of Science,
 Masaryk University,
-Kamenice 5, 625 00 Brno
+Kamenice 5, 625 00 Brno\\
+The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
+Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
 }
@@
 
-\section{Preliminaries}
+%\section{Preliminaries}
+\section{Traversability in tunnels}
+
+\subsection{Preliminaries}
 
@@
 
-\section{Ligand traversability}
+\subsection{Motion planning}
+
+%\section{Ligand traversability}
 
@@
 
-The configuration tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
-one new configuration can be added.
-The most time consuming part of the method is the collision-detection, which is called $m|\L|$ in each expansion step.
-The collision detection can be evaluated fast using hierarchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}
-
-\def\tmpa{0.16\textwidth}
-
-\begin{figure}
-\centering
-\begin{tabular}{cccccc}
-\includegraphics[width=\tmpa]{fig/m003-000} &
-\includegraphics[width=\tmpa]{fig/m003-001} &
-\includegraphics[width=\tmpa]{fig/m003-002} & 
-\includegraphics[width=\tmpa]{fig/m003-003} & 
-\includegraphics[width=\tmpa]{fig/m003-004} & 
-\includegraphics[width=\tmpa]{fig/m003-005} \\
-\includegraphics[width=\tmpa]{fig/m003-006} &
-\includegraphics[width=\tmpa]{fig/m003-007} &
-\includegraphics[width=\tmpa]{fig/m003-008} & 
-\includegraphics[width=\tmpa]{fig/m003-009} & 
-\includegraphics[width=\tmpa]{fig/m003-010} & 
-\includegraphics[width=\tmpa]{fig/m003-011} \\
-\end{tabular}
-\caption{todo and maybe delete}
-\end{figure}
-
-\section{Visualization of results}
+%\begin{figure}
+%\centering
+%\begin{tabular}{cccccc}
+%\includegraphics[width=\tmpa]{fig/m003-000} &
+%\includegraphics[width=\tmpa]{fig/m003-001} &
+%\includegraphics[width=\tmpa]{fig/m003-002} & 
+%\includegraphics[width=\tmpa]{fig/m003-003} & 
+%\includegraphics[width=\tmpa]{fig/m003-004} & 
+%\includegraphics[width=\tmpa]{fig/m003-005} \\
+%\includegraphics[width=\tmpa]{fig/m003-006} &
+%\includegraphics[width=\tmpa]{fig/m003-007} &
+%\includegraphics[width=\tmpa]{fig/m003-008} & 
+%\includegraphics[width=\tmpa]{fig/m003-009} & 
+%\includegraphics[width=\tmpa]{fig/m003-010} & 
+%\includegraphics[width=\tmpa]{fig/m003-011} \\
+%\end{tabular}
+%\caption{todo and maybe delete}
+%\end{figure}
+%
+
+\subsection{Visualization of results}
 
@@
 Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted from the tree.
-
-To present the results to chemists ...
 
@@
 
-The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase (protein 1CQW) (Fig.~\ref{fig::tunnel}).
+The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase (protein 1CQW) (Fig.~\ref{fig::tunnel}a).
 The tunnel was computed using CAVER 3.0~\cite{caver3} for spherical probe of radius 0.9~\AA.
 The tunnel is represented by 760 spheres with average distance $0.05$~\AA.
-
 % from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
 The traversability of the tunnel was evaluated for 1-chlorpropan (denoted as $\LA$) and 1-chlorbutan (denoted as $\LB$).
+$\LA$ was represented by 12 conformations, and $\LB$ by a set of $114$ conformations.
+Examples of three conformations are depicted in Fig.~\ref{fig::tunnel}b.
+
+The tested ligands are bigger than the tunnel with, therefore the ligands have been scaled down, and four
+different scales were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
+For each of this scale, 50 collision-free initial states were found in the $2$\AA radius of the first sphere of the tunnel.
+For each ligand, each minimal scale $\smin$ and each start configuration, 100 trajectories were computed using the proposed planner.
+The parameters of the planner are: $\Imax=10000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA.
+
+The average runtimes and sizes of the build tree are shown in Tab.~\ref{tab::rrt}.
+For each trajectory
+
+\def\tmpa{0.13\textwidth}
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{fig/protein-tunnel-gray}
+\begin{tabular}{cc}
+\includegraphics[width=0.38\textwidth]{fig/protein-tunnel-gray} &
+\includegraphics[width=0.5\textwidth]{fig/ligAll}  \\
+(a) & (b)                        
+\end{tabular}                       
 \caption{\label{fig::tunnel}
-    The protein 1CQW visualized using its secondary structure (gray) with its highest ranked tunnel (green) which was detected using CAVER 3.0.
+    (a) The protein 1CQW visualized using its secondary structure (gray) with its highest ranked tunnel (green) which was detected using CAVER 3.0.
+    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
 }
@@
 
+The configuration tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
+one new configuration can be added.
+The most time consuming part of the method is the collision-detection, which is called $m|\L|$ in each expansion step.
+The collision detection can be evaluated fast using hierarchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}
+
 \section{Conclusion }
@@
 Moreover, incorporation of some scoring function for energy evaluation would introduce the characterization of not only geometrical but also physico-chemical properties and would further improve the prediction ability of our approach.
-
-\section{Acknowledgments}
-
-The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
-Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
 
@@
 %However, such estimation still involve lot of experience.
+
+\section{Acknowledgments}
+
+The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
+Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.

### 2017-06-15 paper.tex
scripts for making tables, experiments

@@
 Therefore, the geometrical and physico-chemical properties of tunnels need to be determined to understand the transport of ligand from and into the protein active site. 
-Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamic simulations. 
+Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamic (MD) simulations.
+Although the tunnels can be computed fast using e.g., Voronoi diagrams, they cannot provide information about ligand traversability, as it is 
+not included in the VD-based computation.
+On the other side, MD simulations are too slow for a daily routine.
 In this paper, we propose to analyze the ligand passing using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT).
-The tunnel is represented by a set of spheres, that are used to guide the growth of the RRT configuration tree in the related configuration space.
-Multiple trajectories are computed using the proposed planner in order to analyze expected behavior of the ligand in the tunnel.
-We introduce several approaches to visualization of detected tunnels which depict valuable information about various properties and behavior of the ligand.
+The tunnel is represented by a set of spheres, that are used to guide the exploration in the search space.
+Multiple trajectories are computed for the ligand being tested.
+We introduce several approaches for visualization of trajectories which depict valuable information about various properties and behavior of the ligand.
 Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
-On the other hand, it is much less computationally demanding than methods based on molecular dynamic simulations. 
+On the other hand, it is much less computationally demanding than methods based on MD simulations.
 The method can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
@@
 
-\begin{figure}[t]
+\begin{figure}[b]
 \centering
@@
 \caption{\label{fig::motiv}
-    Example of tunnel detection in haloalkane dehalogenase.
+    Tunnels in haloalkane dehalogenase with a possible trajectory of 1-chlorpropan.
 }
@@
 To get the ligand to the active site, there has to be the transportation path connecting the protein outer environment with the active site.
-This paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
+This paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
 
@@
 
-The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or 
-a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
-These methods compute the tunnels assuming a spherical ligand (probe), and the resulting tunnels are represented as a sequence of spheres.
+The tunnels can be detected using Voronoi diagrams~\cite{yaffe2008,caver3}.
+%The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3}.  %or a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
+The tunnels are computed assuming a spherical ligand (probe), and they are represented as a sequence of spheres.
 Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and bottleneck (i.e.,
@@
 metric.
-
-\red{presunu jinam}
-This brings two important advantages: the volume of the configuration space to be searched is decreased, as not the whole
-protein/ligand complex, but only its subspace need to be searched.
-Consequently, the relative volume of the corresponding free region of the configuration space is increased, which leads to easier motion planning.
-The proposed approach allows the researches to analyze a given tunnel for various ligands without necessity to compute the tunnel
-again for each ligand.
 
@@
 The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated
-around the virtual goal sphere $v$.
+around the sphere $t_v \in T$.
 %around which the random samples are generated with probability $\gb$.
@@
 %is guided using moving virtual goal~\cite{vonasek2009rrt}.
-After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
+After the tree reaches the sphere $t_v$, i.e., the distance of the tree to the center of $t_v$ is
 less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
@@
 The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
-metric for nearest-neighbor search.
+metric for the nearest-neighbor search.
 
@@
 to the nearest node towards $\qgoal$ can be found.
-Due to stochastic nature of RRT, it is not guaranteed that the tree always reached the goal configuration $\qgoal$.
-Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted from the tree.
+Due to stochastic nature of RRT, it is not guaranteed that the tree can always reach the goal configuration $\qgoal$.
+Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted from the tree, as even these non-complete 
+trajectories provide valuable information about the ligand behavior.
 
@@
 different scales were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
+No trajectories were found for $\smin > 0.6$.
 For each of this scale, 50 collision-free initial states were found in the $2$\AA radius of the first sphere of the tunnel.
@@
 The parameters of the planner are: $\Imax=10000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA.
+For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance $\rv$ or less) over the 100 trials.
 
 The average runtimes and sizes of the build tree are shown in Tab.~\ref{tab::rrt}.
-For each trajectory
+The tables also show the minimal, maximal success rate and the average success rate.
+The average success rate is compute as the average value of the success rates achived for all 50 initial configurations.
+
+The highest average success rate is achived for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
+The minimal and maximal success rates indicate how the initial position of the ligand influence the ability to pass the tunnels.
+The ligand $\LB$, which was represented by more conformations than $\LA$, has higher minimal and also average success rate, which
+indicates that the flexibility is important for the tunnel traversability.
+The runtime is significantly higher for the $\LB$ ligand, which is caused by the larger number of conformations that need to be
+examined in each expansion.
 
@@
 }
+{
+\renewcommand{\tabcolsep}{1pt}
 \input table.003.tex
-\hskip 10pt
+\hskip 5pt   
 \input table.004.tex
+}
 \end{table}
@@
 
-\section{Discussion}
-
-guiding, parameters, different sampling, flexibility
-
-The configuration tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
-one new configuration can be added.
-The most time consuming part of the method is the collision-detection, which is called $m|\L|$ in each expansion step.
-The collision detection can be evaluated fast using hierarchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}
+\section{Discussion \& implementation details}
+
+Sampling-based motion planning is a powerful tool for computing tajectories for flexible ligands.
+The proposed extensions to the RRT planner have been motivated by the necessity to cope with the narrow passage
+and the high dimensionaly of the flexible ligands.
+By scaling-down the geometry of ligand, which is also used in other related tools~\cite{cortes2005path}, 
+the relative volume of the free region of the configuration space increases, which increases probability of placing
+random samples $\qrand$ into the free region.
+To further cope with the narrow passage problem, the translation part of the random samples is generated using
+the 3D centerline of the tunnel.
+The sampling along the tunnel centerline (in the ditance defined by $\rv$) does not only lead to better growth of the tree in the tunnel, but it also limits the maximal deflection of the trajectory from the tunnel.
+Therefore, the bounding box of $\C$ can be defined around the tunnel in the distance $\rv$, which is typically
+smaller than the bounding box of the whole protein.
+This, together with the scaled-down ligand, increases relative volume of the free configuration space and help the RRT to
+reach the goal configuration.
+
+The runtime of the planner is curently given mainly by the collision-detection.
+As the ligand is assumed to travel only along the tunnel centerline (in the distance defined by $\rv$),
+only atoms located in the vicinity of the tunnel can be considered for collision detection.
+A fast collision-detection can be achieved either using fast hiearchical data structures as OBB (implemented e.g. in~\cite{ozcollide})
+    or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
 

### 2017-06-15 results/stat/compare.pl

@@ -1,0 +1,35 @@
+#!/usr/bin/env perl
+
+
+$| = 1;
+
+foreach my $prefix (qw/ fs fsexp2 /) {
+    foreach my $dir (glob "*-minScale*") {
+        print "Processing $dir\n";
+
+        my @all;
+        foreach my $file (glob "$dir/$prefix.*.stat") {
+            open(my $fi,"<$file");
+            my @lines = <$fi>;
+            close $fi;
+            chomp @lines;
+
+            foreach my $l (@lines) {
+                if ($l !~ m/^#/) {
+                    print "Orig: '$l'\n";
+                    $l =~ s/;/ /g;
+                    $l =~ s/=/ /g;
+                    print "New: '$l'\n";
+                    my @a = split(/\s+/,$l);
+                    push @all, "$a[1], $a[3]";
+                }
+            }
+
+        }
+        open(my $fo,">$dir/aa.$prefix");
+        foreach (@all) {
+            print $fo "$_\n";
+        }
+        close $fo;
+    }
+}

### 2017-06-15 results/stat/downStat.sh

@@ -1,0 +1,1 @@
+rsync -ave ssh --include=".*.stat" --exclude="*/" tarkil:~/proteins/test-fs/output/ .

### 2017-06-15 results/stat/makeTable.pl

@@ -1,0 +1,184 @@
+#!/usr/bin/env perl
+
+
+$| = 1;
+
+BEGIN { push @INC, "/home/vojta/work/rrtPlanning/scripts"; }
+use CStat;
+
+my %names;
+
+$names{"003"} = "\\LA";
+$names{"004"} = "\\LB";
+
+
+my $dtg = 1;
+my $dtg2 = 3;
+
+
+my %hh;
+
+foreach my $pref (qw/ ts fs /) {
+
+    my %h;
+
+    foreach my $f (glob "$pref.stat.*") {
+
+        my $lig, $scale;
+        if ($f =~ m/stat\.(.*)\.(\d+)/) {
+            $lig = $1 ;
+            $scale = $2 + 0;
+        } else {
+            print "Cannot parse file name $f\n";
+        }
+
+        my ($ddr, $length) = loadStatFileToHash($f);
+        print "Loaded $length from $f\n";
+
+        $h{$lig}{$scale}{"len"} = [$length , 0 ];
+
+        my ($min, $max, $mean, $dev, $med, @rest) = getArrayStatByName($ddr,"rrt_path_time");
+        print "time: $f: @a, mean=$mean : $dev\n";
+
+        $h{$lig}{$scale}{"time"} = [$mean , $dev ];
+
+        my ($min, $max, $mean, $dev, $med, @rest) = getArrayStatByName($ddr,"trajDistToGoal");
+        print "dist: $f: @a, mean=$mean : $dev\n";
+
+        $h{$lig}{$scale}{"dtg"} = [$mean , $dev ];
+
+
+        if ($pref eq "ts") {
+            my ($min, $max, $mean, $dev, $med, @rest) = getArrayStatByName($ddr,"treeSize");
+            print "dist: $f: @a, mean=$mean : $dev\n";
+
+            $h{$lig}{$scale}{"ts"} = [$mean , $dev ];
+        }
+
+
+        my @a = getArrayByName($ddr, "trajDistToGoal");
+
+        my $cnt = 0;
+        my $cnt2 = 0;
+        foreach my $d (@a) {
+            if ($d <= $dtg) {
+                $cnt++;
+            }
+            if ($d <= $dtg2) {
+                $cnt2++;
+            }
+        }
+        my $sr = 100.0*$cnt/@a;
+        my $sr2 = 100.0*$cnt2/@a;
+        print "$f: SR=$sr\n";
[…]

### 2017-06-15 results/stat/reportRunTime.pl

@@ -1,0 +1,75 @@
+#!/usr/bin/env perl
+
+
+$| = 1;
+
+   
+foreach my $pref (qw/ fs ts /) {
+
+    foreach my $dir (glob "*-minScale*") {
+
+        my %sgstat;
+
+        my $scale = -1;
+        my $ligand = "";
+        if ($dir =~ m/m(\d+)-minScale-(.+)/) {
+            $ligand = $1;
+            $scale = $2 + 0.0;
+        } else {
+            die "cannot parse name of $dir\n";
+        }
+
+        print "Processing $dir, scale=$scale\n";
+
+        my @all;
+        foreach my $file (glob "$dir/$pref.*.stat") {
+            print "Loading $file\r";
+            open(my $fi,"<$file");
+            my @lines = <$fi>;
+            close $fi;
+            chomp @lines;
+
+            my @tmp;
+            foreach my $l (@lines) {
+                if ($l !~ m/^#/) {
+                    push @tmp, $l;
+                }
+            }
+            push @all, @tmp;
+
+            if ($file =~ m/$pref.*\.sindex-(\d+)\.trial-(\d+)\./) {
+                my $sgi = $1 + 0;
+                my $trial = $2 + 0;
+                if (not exists $sgstat{$sgi}) {
+                    $sgstat{$sgi} = [];
+                }       
+                push @{ $sgstat{$sgi} }, @tmp;
+            } else {
+                print "Cannot parse of file: $file\n";
+            }
+        }
+
+        $scale = int($scale*10);
+        open(my $fo, ">$pref.stat.$ligand.$scale");
+        foreach (@all) {
+            print $fo "$_\n";
+        }
+        close $fo;
+ 
+        foreach my $sgi (keys %sgstat) {
+            open(my $fo, ">$pref.stat.$ligand.sgi-$sgi.scale-$scale");
+            my @a = @{ $sgstat{$sgi} };
+            foreach (@a) {
+                print $fo "$_\n";
+            }
+            close $fo;
+        }
+        
+
+
+    }
+
+}
+
+
+

### 2017-06-15 converttojpg.pl
script for converting stuff

@@ -1,0 +1,9 @@
+#!/usr/bin/env perl
+
+foreach my $f (glob "*.png") {
+    print "Converting $f .. ";
+    my $nf = $f;
+    $nf =~ s/png$/jpg/g;
+    print "convert $f $nf\n";
+    system "convert $f -background white -flatten -alpha off $nf\n";
+}

### 2017-06-22 paper.tex
abstract

@@
 \abstract{
-Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
-Molecular tunnels play an important role in catalytic and thermodynamic properties and they connect catalytic active sites with the solvent. 
-Therefore, the geometrical and physico-chemical properties of tunnels need to be determined to understand the transport of ligand from and into the protein active site. 
+%Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
+Molecular tunnels connect catalytic active sites with the solvent and they play an important role in catalytic and thermodynamic properties.
+The geometrical and physico-chemical properties of tunnels need to be determined in order to understand the transport of ligand from and into the protein active site. 
 Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamic (MD) simulations.
@@
 On the other side, MD simulations are too slow for a daily routine.
-In this paper, we propose to analyze the ligand passing using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT).
-The tunnel is represented by a set of spheres, that are used to guide the exploration in the search space.
-Multiple trajectories are computed for the ligand being tested.
-We introduce several approaches for visualization of trajectories which depict valuable information about various properties and behavior of the ligand.
-Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
-On the other hand, it is much less computationally demanding than methods based on MD simulations.
-The method can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
+Many degrees of freedom need to be considered in order to analyze the traversability, which motivates us to employ robotic motion planning tools.
+The ligand traversability through a tunnel can be evaluated using motion planning tools developed originally in robotics.
+In this paper, we propose a method based on the Rapidly Exploring Random Tree (RRT) planner to generate trajectories for the ligand inside
+tunnels being studied.
+We propose novel modifications to RRT in order to compute the trajectories around the tunnel and also to consider flexibility of the ligand.
+We introduce several approaches for visualization of the resulting trajectories which depict valuable information about various properties and behavior of the ligand.
+The goal of the paper is to show that motion planning can provide additional insight about tunnel properties, being
+less computationally demanding than the MD simulations.
+%Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
+%On the other hand, it is much less computationally demanding than methods based on MD simulations.
+The proposed approach can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
 }
@@
 The ordinary VD is computed on points representing centers of all atoms, without considering radii of the atoms.
-This may lead to detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
+%This may lead to detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
 To consider atoms with different radii, weights of individual points are determined by the van der Walls radii of the atoms in WVD.
@@
 VD-based methods are memory less demanding, and also faster than the grid-based methods.
-\red{asi vyhodit?
-The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
-finding correspondences between them.
-The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
-}
+%The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
+%finding correspondences between them.
+%The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.
 
@@
 The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
-The free samples are stored in a roadmap (a graph structures), in which a path can be searched using standard graph-search methods.
-%A path in the roadmap then represents a motion in the workspace.
+The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.
 
@@
 with many-DOF systems.
-Two main issues however need to be adressed: the increased dimension of the configuration space and the narrow passage problem.
+Two main issues however need to be addressed: the increased dimension of the configuration space and the narrow passage problem.
 
@@
 
-A well known issue of sampling-based planners is the narrow passage problem~\cite{hannaWIS}.
-A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space.
+%A well known issue of sampling-based planners is the narrow passage problem~\cite{hannaWIS}.
+A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space~\cite{hannaWIS}.
 Narrow passages have smaller volume than other regions and it is therefore difficult to sample them dense enough using uniform distribution, that is used in the basic sampling-based planners.
@@
 In RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
-and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder the growth of the tree.
+and translation) and it employs the other DOFs (i.e., that are responsible for conformation changes) if they hinder the growth of the tree.
 The approach~\cite{cortes2010simulating} may however suffer from the narrow passage problem, as it aims to find any exit pathway for a given ligand, which requires to search the whole configuration space of the ligand/protein complex.
@@
 $\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
-This scaling technique is used also in other tools like MoMaLigPath~\cite{cortes2005path}.
+This scaling technique is used also in other tools like MoMa-LigPath~\cite{cortes2005path}.
 
@@
 No trajectories were found for $\smin > 0.6$.
-For each of this scale, 50 collision-free initial states were found in the $2$\AA radius of the first sphere of the tunnel.
+For each of this scale, 50 collision-free initial states were found in the $2$\AA\ radius of the first sphere of the tunnel.
 For each ligand, each minimal scale $\smin$ and each start configuration, 100 trajectories were computed using the proposed planner.
@@
 
-The average runtimes and sizes of the build tree are shown in Tab.~\ref{tab::rrt}.
+The average runtimes and sizes of the built trees are shown in Tab.~\ref{tab::rrt}.
 The tables also show the minimal, maximal success rate and the average success rate.
-The average success rate is compute as the average value of the success rates achived for all 50 initial configurations.
-
-The highest average success rate is achived for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
+The average success rate is compute as the average value of the success rates achieved for all 50 initial configurations.
+
+The highest average success rate is achieved for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
 The minimal and maximal success rates indicate how the initial position of the ligand influence the ability to pass the tunnels.
@@
 \caption{\label{tab::rrt}
-    Runtime and sucess ratio of the planner
+    Runtime and success ratio of the planner
 }
@@
 
-Sampling-based motion planning is a powerful tool for computing tajectories for flexible ligands.
+Sampling-based motion planning is a powerful tool for computing trajectories for flexible ligands.
 The proposed extensions to the RRT planner have been motivated by the necessity to cope with the narrow passage
-and the high dimensionaly of the flexible ligands.
+and the high dimensional of the flexible ligands.
 By scaling-down the geometry of ligand, which is also used in other related tools~\cite{cortes2005path}, 
@@
 the 3D centerline of the tunnel.
-The sampling along the tunnel centerline (in the ditance defined by $\rv$) does not only lead to better growth of the tree in the tunnel, but it also limits the maximal deflection of the trajectory from the tunnel.
+The sampling along the tunnel centerline (in the distance defined by $\rv$) does not only lead to better growth of the tree in the tunnel, but it also limits the maximal deflection of the trajectory from the tunnel.
 Therefore, the bounding box of $\C$ can be defined around the tunnel in the distance $\rv$, which is typically
@@
 
-The runtime of the planner is curently given mainly by the collision-detection.
+The runtime of the planner is currently given mainly by the collision-detection.
 As the ligand is assumed to travel only along the tunnel centerline (in the distance defined by $\rv$),
 only atoms located in the vicinity of the tunnel can be considered for collision detection.
-A fast collision-detection can be achieved either using fast hiearchical data structures as OBB (implemented e.g. in~\cite{ozcollide})
+A fast collision-detection can be achieved either using fast hierarchical data structures as OBB (implemented e.g. in~\cite{ozcollide})
     or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.

### 2017-06-22 paper.tex
small stuff

@@
 Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
-These interactions are highly influence by the ability to transport the small molecule, called ligand, to the protein active site.
+These interactions are highly influenced by the ability to transport the small molecule, called ligand, to the protein active site.
 The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
@@
 
-For example, one task in protein engineering is to change selected properties of a protein, e.g.
-its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
-This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
-ligand from the outside environment to the active site or vice versa.
-
 The tunnels can be detected using Voronoi diagrams~\cite{yaffe2008,caver3}.
@@
 The tunnels are computed assuming a spherical ligand (probe), and they are represented as a sequence of spheres.
-Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and bottleneck (i.e.,
-        the radius of the smallest sphere that can traverse the whole tunnel).
+Bio-chemists decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and bottleneck (i.e., the radius of the smallest sphere that can traverse the whole tunnel).
+This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g.
+its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
+%The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
+%This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
+%ligand from the outside environment to the active site or vice versa.
+
 Typical ligands are however non-spherical and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
 The decision based only on the spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
-
 To provide chemists more insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering it's shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.
@@
 
-\red{TODO}
-something about screening: what is it, what's is the input and output, in which applications (research/industry) it is useful.
+%\red{TODO}
+%something about screening: what is it, what's is the input and output, in which applications (research/industry) it is useful.
 

### 2017-06-23 result-multiple/copy.pl
scripts

@@ -1,0 +1,29 @@
+#!/usr/bin/env perl
+
+my $outdir = "output-trajOnly";
+
+foreach my $td (glob "output/tun-*") {
+
+
+    print "Processing $td\n";
+
+    foreach my $tf (glob "$td/*") {
+    
+        my $od = $tf;
+        $od =~ s/^output/$outdir/;
+        system  "mkdir -p $od";
+
+        foreach my $tf2 (glob "$tf/*.traj.00.txt") {
+
+            if ($tf2 =~ m/sindex-(\d+).trial-(\d+)/) {
+                my $sindex = $1 + 0;
+                my $trial = $2 + 0;
+                if ($sindex < 50 and $trial < 20) {
+                    print "cp $tf2 $od/\n";
+                    system "cp $tf2 $od/";
+                }
+            }
+        }
+    }
+}
+

### 2017-06-23 result-multiple/down.sh

@@ -1,0 +1,1 @@
+rsync -ave ssh --include="*/" --include="*.stat" --include="*.traj.00.txt" --exclude="*" tarkil:~/proteins/test-fs-multiple/output .

### 2017-06-23 result-multiple/makeTable.pl

@@ -1,0 +1,192 @@
+#!/usr/bin/env perl
+
+
+$| = 1;
+
+BEGIN { push @INC, "/home/vojta/work/rrtPlanning/scripts"; }
+use CStat;
+use Storable;
+
+my %names;
+
+$names{"003"} = "\\LA";
+$names{"004"} = "\\LB";
+
+
+my $DTG = 1.5;
+
+
+my %h;
+
+my @cols = qw/ rrt_path_function_time rrt_path_time trajDistToGoal treeSize warmUpTime /;
+
+my $hashfile = "h.hash";
+
+
+my $processData = 0;
+
+
+if ($processData == 1) {
+
+
+    foreach my $tunnel (glob "output/tun*") {
+
+        foreach my $scaledir (glob "$tunnel/m003-*") {
+            print "Processing $scaledir\n";
+
+            my $scale = 0;
+            if ($scaledir =~ m/m003-minScale-(.*)/) {
+                $scale = $1 + 0;
+            } else {
+                die "Cannot parse name of scaledir $scaledir\n";
+            }
+
+            foreach my $f (glob "$scaledir/ts.*.stat") {
+
+                my $sindex;
+                my $trial;
+                if ($f =~ m/ts\.m003\.sindex-(\d+)\.trial-(\d+)\.stat/) {
+                    $sindex = $1 + 0;
+                    $trial = $2 + 0; 
+                } else {
+                    die "Cannot interpret file name $f\n";
+                }
+
+#print "Processing $f\r";
+
+                my ($ddr, $length) = loadStatFileToHash($f);
+                print "Loaded $length from $f\r";
+
+                foreach my $col (@cols) {
+                    my @a = @{ $ddr->{$col} };
+                    if (not exists $h{$tunnel}{$scale}{$sindex}{$col}{$trial}) {
+                        $h{$tunnel}{$scale}{$sindex}{$col}{$trial} = [];
+                    }
+
+                    push @{ $h{$tunnel}{$scale}{$sindex}{$col}{$trial} }, @a;
+                }
+            }
+        }
+    }
+
+    store \%h, $hashfile;
+
+} else  {
+    my $href = retrieve $hashfile; 
+    %h = %{ $href };
+}
+
+
+
+my %stat;
+
+#normalize runtimes
+foreach my $tunnel (keys %h) {
+    foreach my $scale (keys %{ $h{$tunnel} }) {
+
[…]

### 2017-06-23 result-multiple/reportreachability.pl

@@ -1,0 +1,99 @@
+#!/usr/bin/env perl
+
+
+my $DTG = 1.1;
+
+foreach my $tun (1..4) {
+    
+
+    my $caver = "ca-tunnels/tun.$tun.txt";
+    open(my $fi, "<$caver") or die "Cannot open caver tunnel: $caver: $!\n";
+    my @calines = <$fi>;
+    close $fi;
+    chomp @calines;
+
+    my @ca;
+    for(my $i = 0; $i < @calines; $i++) {
+        my ($x, $y, $z) = split(/\s+/, $calines[$i]);
+        push @ca, [$x, $y, $z];
+    }
+    print "Loaded " .  @ca . " tunnel points from $caver\n";
+
+
+    foreach my $scaledir (glob "output/tun-$tun/*") {
+        if ($scaledir =~ m/m003-minScale-(.*)/) {
+            my %reached;
+            my $scale = $1 + 0;
+
+            my $total = 0;
+
+            foreach my $tfile (glob "$scaledir/*.txt") {
+
+                open(my $fi, "<$tfile");
+                my @data = <$fi>;
+                close $data;
+
+                my @pts;
+                for(my $i=0;$i < @data; $i++) {
+                    my ($x, $y, $z, @rest) = split(/\s+/,$data[$i]);
+                    push @pts, [$x, $y, $z];
+                }
+                print "Loading $tfile, " . @pts . " points\n";
+
+                my $res = reportReachability(\@ca, \@pts);
+
+                for(my $i=0; $i <= $res; $i++) {
+                    $reached{$i}++;
+                }
+                $total++;
+            }
+
+            open(my $fo,">reached.tun-$tun.scale-$scale.txt");
+            print $fo "#reached reached/total\n";
+            foreach my $i (sort {$a <=> $b} keys %reached) {
+                print $fo "$i " . $reached{$i} . " " . 1.0*$reached{$i} / $total . "\n";
+            }
+            close $fo;
+
+
+
+        } else {
+            die "Cannot parse of name dir $scaledir: $!\n";
+        }
+    }
+
+    }
+
+
+sub reportReachability {
+    my ($caref, $ptsref) = @_;
+
+
+    my $res = -1;
+
+    for(my $i = 0; $i < @{ $caref }; $i++) {
+        
+        #find nearest in the trajectory. If less then DTG, this one is reached
+
[…]

### 2017-06-27 paper.tex
visualization destroeyd and not finished yet, I am too lazy today..

@@
 \def\qinit{\qstart}
-\def\qgoal{\qstart}
+\def\qgoal{q_{goal}}
 \def\qnear{q_{near}}
@@
 \def\da{d_{atom}}
+\def\R{\mathbb{R}}
 
@@
 Many degrees of freedom need to be considered in order to analyze the traversability, which motivates us to employ robotic motion planning tools.
-The ligand traversability through a tunnel can be evaluated using motion planning tools developed originally in robotics.
-In this paper, we propose a method based on the Rapidly Exploring Random Tree (RRT) planner to generate trajectories for the ligand inside
-tunnels being studied.
+In this paper, we propose to evaluate the tunnels using based on ligand trajectories computed by motion planning, namely using
+Rapidly Exploring Random Tree (RRT).
 We propose novel modifications to RRT in order to compute the trajectories around the tunnel and also to consider flexibility of the ligand.
@@
 A protein tunnel is described by a sequence of collision-free spheres 
-$T=( (t_1, r_1),\ldots,(t_n,r_n) )$, where $n$ denotes the number of spheres,
-$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
+$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
+$c \in \R^3$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
 The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.
@@
 
-\subsection{Motion planning}
+\subsection{Main idea \red{VV dodelam}}
+
+The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand in the tunnel.
+As the traversability may depends on the initial position of the ligand, especially its rotation, it is necessary to compute
+the trajectories from multiple starting configurations $\qinit$.
+The starting configuration $\qinit$ is generated by randomly sampling the vicinity of the first sphere $c_1$ of the tunnel.
+A sample $q$ is created by generating its translation part randomly around $t_1$ in the distance $R$.
+The rotation part of $q$ and conformation index are set randomly, the scale is set to $\smin$.
+If the sample $q$ is collision-free, it can be considered as a new starting configuration $\qinit$.
+
+Then, multiple trajectories are computed starting from $\qinit$ using the method proposed in the next section.
+
+\subsection{RRT for ligand trajectory computation}
 
@@
 To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
-from $N(t_v,\Sigma)$, where $\Sigma$ is diagonal matrix with diagonal entries equal to a parameter $\rv$, and the rotational
+from $N(c_v,\Sigma)$, where $\Sigma$ is diagonal matrix with diagonal entries equal to a parameter $\rv$, and the rotational
 part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
@@
 \KwIn{
-    tunnel spheres $t_i$ and their radii $r_i$, $i=1,\ldots,n$, 
+    tunnel spheres $c_i$ and their radii $r_i$, $i=1,\ldots,n$, 
     initial configuration $\qinit$,
@@
     \eIf{$rand() < \gb$}{
-        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
+        $\qrand$ = random sample around actual virtual goal $c_n \in T$\;
     }{
@@
     \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
-        $d$ = nearest node in the tree towards sphere $t_i$\;
+        $d$ = nearest node in the tree towards sphere $c_i$\;
         \If{$d < \dt $}{
@@
 
-%\begin{figure}
-%\centering
-%\begin{tabular}{cccccc}
-%\includegraphics[width=\tmpa]{fig/m003-000} &
-%\includegraphics[width=\tmpa]{fig/m003-001} &
-%\includegraphics[width=\tmpa]{fig/m003-002} & 
-%\includegraphics[width=\tmpa]{fig/m003-003} & 
-%\includegraphics[width=\tmpa]{fig/m003-004} & 
-%\includegraphics[width=\tmpa]{fig/m003-005} \\
-%\includegraphics[width=\tmpa]{fig/m003-006} &
-%\includegraphics[width=\tmpa]{fig/m003-007} &
-%\includegraphics[width=\tmpa]{fig/m003-008} & 
-%\includegraphics[width=\tmpa]{fig/m003-009} & 
-%\includegraphics[width=\tmpa]{fig/m003-010} & 
-%\includegraphics[width=\tmpa]{fig/m003-011} \\
-%\end{tabular}
-%\caption{todo and maybe delete}
-%\end{figure}
-%
-
-\subsection{Visualization of results}
-
-The result of the proposed planning method is a tree of collision-free configurations, from which a path leading from $\qinit$
-to the nearest node towards $\qgoal$ can be found.
-Due to stochastic nature of RRT, it is not guaranteed that the tree can always reach the goal configuration $\qgoal$.
-Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted from the tree, as even these non-complete 
-trajectories provide valuable information about the ligand behavior.
-
-\red{Adam: }
+\subsection{Computing trajectories}
+
+The result of the planner is the tree $\T$ of collision-free configurations.
+The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the neearest node in the tree towards $\qgoal$, is found.
+The path $P=(q_i), q_i \in \C$ is represented by a sequence of the collision-free configurations.
+
+Due to stochastic nature of RRT, the computed trajectories may differ in each run.
+Moreover, they are influenced by the initial configuration $\qinit$, especialy by the rotation of the ligand.
+To evaluate traversability of the given tunnel, multiple starting configurations are generated and for each
+of these configurations, multiple trajectories are computed.
+This results in a set of $n$ trajectories $P_i, i=1,\ldots,n$.
+These trajectories are then used to evaluate tunnel characteristics, which is described in the next section.
+
+\subsection{Traversability characteristics}
+
+\red{VV is significantly rewriting this section}
+
+There basic characteristics of the tunnel $T$ are computed from the trajectories: the accessiblity, throughput and the scaling factor.
+These characteristics are computed for each sphere $(c_i,r_i) \in T$ of the tunnel.
+This requires to determine which parts of the tunnel were accessed by the trajectories.
+
+The Accessibility 
+--- The probability of the ligand that it reaches a tunnel's part.
+We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \R$ is radius.
+Then for each trajectory $t$ we define its ending tunnel part $P_t^e$ as a tunnel part whose center is closest to the end position of $t$.
+Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
+
 As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
@@
 \begin{itemize}
-  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
-We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \mathbb{R}$ is radius.
-Then for each trajectory $t$ we define its ending tunnel part $P_t^e$ as a tunnel part whose center is closest to the end position of $t$.
-Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
+
 \item \textbf{Ligand scale} --- The scale of the ligand when it passes a tunnel's part.
@@
 \end{figure}
+
+\subsection{Visualization of the results}
 
@@
 
-\section{Acknowledgments}
-
-The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
-Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
+%\section{Acknowledgments}
+%
+%The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
+%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
+%
+%\begin{figure}
+%\centering
+%\begin{tabular}{cccccc}
+%\includegraphics[width=\tmpa]{fig/m003-000} &
+%\includegraphics[width=\tmpa]{fig/m003-001} &
+%\includegraphics[width=\tmpa]{fig/m003-002} & 
+%\includegraphics[width=\tmpa]{fig/m003-003} & 
+%\includegraphics[width=\tmpa]{fig/m003-004} & 
+%\includegraphics[width=\tmpa]{fig/m003-005} \\
+%\includegraphics[width=\tmpa]{fig/m003-006} &
+%\includegraphics[width=\tmpa]{fig/m003-007} &
+%\includegraphics[width=\tmpa]{fig/m003-008} & 
+%\includegraphics[width=\tmpa]{fig/m003-009} & 
+%\includegraphics[width=\tmpa]{fig/m003-010} & 
+%\includegraphics[width=\tmpa]{fig/m003-011} \\
+%\end{tabular}
+%\caption{todo and maybe delete}
+%\end{figure}
+%

### 2017-06-27 paper.tex
tunnel characteristics, not finished yet

@@
 
-\subsection{Main idea \red{VV dodelam}}
+\subsection{Main idea}
 
@@
 If the sample $q$ is collision-free, it can be considered as a new starting configuration $\qinit$.
-
 Then, multiple trajectories are computed starting from $\qinit$ using the method proposed in the next section.
@@
 
-%\section{Ligand traversability}
-
-The traversability study proposed in this paper is achieved by computing many trajectories for the given ligand and tunnel, which
-is solved using a novel RRT-based planner.
-The main principle of the RRT planner is extended by guided sampling, scaling of the ligands atoms and using predefined set of ligand
-conformations.
+The trajectories are computed using an extension of the RRT~\cite{lavalleRRT} planner.
+The RRT planner is extended by guided sampling, scaling of the ligands atoms and using predefined set of ligand conformations.
 The main loop of the method is described in Alg.~\ref{alg::main}.
@@
 To guide tree growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
-The virtual goal $v, 1\le v \le n$, is the index of a tunnel sphere.
+The virtual goal $v, 1\le v \le n$, is the index of sphere of the tunnel.
 The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated
 around the sphere $t_v \in T$.
-%around which the random samples are generated with probability $\gb$.
-%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
-%is guided using moving virtual goal~\cite{vonasek2009rrt}.
 After the tree reaches the sphere $t_v$, i.e., the distance of the tree to the center of $t_v$ is
@@
 
-\subsection{Computing trajectories}
-
 The result of the planner is the tree $\T$ of collision-free configurations.
 The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the neearest node in the tree towards $\qgoal$, is found.
-The path $P=(q_i), q_i \in \C$ is represented by a sequence of the collision-free configurations.
-
-Due to stochastic nature of RRT, the computed trajectories may differ in each run.
-Moreover, they are influenced by the initial configuration $\qinit$, especialy by the rotation of the ligand.
-To evaluate traversability of the given tunnel, multiple starting configurations are generated and for each
-of these configurations, multiple trajectories are computed.
-This results in a set of $n$ trajectories $P_i, i=1,\ldots,n$.
-These trajectories are then used to evaluate tunnel characteristics, which is described in the next section.
+The path $P=(q_i), q_i \in \C$ is represented as a sequence of the collision-free configurations.
+Let $d(P,i)$ denote the 3D Euclidean distance
+
+The path $P$ reaches a particular sphere $t_i$ in the tunnel if the 3D Euclidean distance between the sphere and its nearest configuration
+in the path is less than the threshold $T$.
+TODO VV dodelam
 
@@
 
-\red{VV is significantly rewriting this section}
-
-There basic characteristics of the tunnel $T$ are computed from the trajectories: the accessiblity, throughput and the scaling factor.
+The traversability of the tunnel $T$ is evaluated using all computed trajectories $P_i$. 
+Three basic characteristics of the tunnel are computed from the trajectories: the accessiblity, throughput and the scaling factor.
 These characteristics are computed for each sphere $(c_i,r_i) \in T$ of the tunnel.

### 2017-06-28 paper.tex
comments Bara

@@
 \institute{
-    V. Von\'{a}sek, Martin Saska:
+    V. Von\'{a}sek, M. Saska:
     Faculty of Electrical Engineering,  
 Czech Technical University in Prague, 
-Technick\'a 2, 166 27 Prague, Czech Republic
-\email{vonasek@labe.felk.cvut.cz}   \\
+Technick\'a 2, 166 27 Prague, Czech Republic,
+\email{vonasek@labe.felk.cvut.cz},   \\
 %\and
@@
 Masaryk University, 
-Botanick\'a 68a, 602 00 Brno, \\
+Botanick\'a 68a, 602~00 Brno, Czech Republic,\\
 %w\and 
-David Bedn\'{a}\v{r}:
+D. Bedn\'{a}\v{r}:
 Faculty of Science,
 Masaryk University,
-Kamenice 5, 625 00 Brno\\
+Kamenice 5, 625 00 Brno, Czech Republic\\
 The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
@@
 Molecular tunnels connect catalytic active sites with the solvent and they play an important role in catalytic and thermodynamic properties.
-The geometrical and physico-chemical properties of tunnels need to be determined in order to understand the transport of ligand from and into the protein active site. 
+The geometrical and physico-chemical properties of tunnels need to be determined in order to understand the transport of a ligand from and into the protein active site. 
 Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamic (MD) simulations.
@@
 Rapidly Exploring Random Tree (RRT).
-We propose novel modifications to RRT in order to compute the trajectories around the tunnel and also to consider flexibility of the ligand.
-We introduce several approaches for visualization of the resulting trajectories which depict valuable information about various properties and behavior of the ligand.
+We propose novel modifications to RRT in order to compute the trajectories around the tunnel and also to consider the flexibility of the ligand.
+We also introduce several approaches for visualization of the resulting trajectories which depict valuable information about various properties and behavior of the ligand.
 The goal of the paper is to show that motion planning can provide additional insight about tunnel properties, being
@@
 \caption{\label{fig::motiv}
-    Tunnels in haloalkane dehalogenase with a possible trajectory of 1-chlorpropan.
+    Tunnels in haloalkane dehalogenase with a possible trajectory of 1-chlorpropan ligand.
 }
@@
 To get the ligand to the active site, there has to be the transportation path connecting the protein outer environment with the active site.
-This paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
-
-The tunnels can be detected using Voronoi diagrams~\cite{yaffe2008,caver3}.
+These paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
+
+Traditionally, tunnels can be detected using Voronoi diagrams~\cite{yaffe2008,caver3}.
 %The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3}.  %or a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
-The tunnels are computed assuming a spherical ligand (probe), and they are represented as a sequence of spheres.
-Bio-chemists decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and bottleneck (i.e., the radius of the smallest sphere that can traverse the whole tunnel).
-This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g.
-its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
+Here the tunnels are computed assuming a spherical ligand (probe) and they are represented as a sequence of spheres.
+Biochemists decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
+This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013} or its activity of the protein towards other molecules~\cite{Pavlova2009}.
 %The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
@@
 
-Typical ligands are however non-spherical and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
-The decision based only on the spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
-To provide chemists more insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering it's shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.
-
-In this paper, we propose a novel method for computing the trajectories of the flexible ligands in a tunnel represented by spheres.
+However, ligands are typically of non-spherical shape and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
+The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
+To provide biochemists with better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.
+
+In this paper, we propose a novel method for computing the trajectories of flexible ligands in a tunnel represented by spheres.
 The proposed work is based on the sampling-based Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner, which
-can cope with many-DOF objects of arbitrary geometry.
+can deal with many-DOF objects of arbitrary geometry.
 The random samples for RRT are generated using the spheres representing the tunnel in order to guide the configuration tree through the search space.
-To make the transportation easier, a scaled-down version of the ligand is used, similarly e.g. to~\cite{cortes2005path}.
+To make the transportation easier, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
 The flexibility of the ligand is modeled using a predefined set of typical conformations.
-We propose novel visualization techniques for presenting the results to bio-chemists.
-The proposed method can be seen as an extension for the classic tunnel detection tools.
+We further propose novel visualization techniques for presenting the results to biochemists.
+The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
 
@@
 
-\section{Related work}
+\section{Related Work}
 
@@
 on the presence of atoms of the protein.
-Tunnels can be then searched using standard graph-search methods like Dijkstra's algorithm.
+Tunnels can be then searched using standard graph-search methods, like Dijkstra's algorithm.
 Besides, the grid can be used to identify other relevant properties like 
@@
 Due to the high memory consumption, these methods are not suitable for tunnel detection in dynamic proteins and therefore they
-are used primarily for analysis of static molecules or individual snapshots of dynamics.
-
-Currently the most widely used approach to tunnel detection is based on Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD).
-The ordinary VD is computed on points representing centers of all atoms, without considering radii of the atoms.
+are used primarily for analysis of static molecules or individual snapshots of molecular dynamics.
+
+Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD).
+The ordinary VD is computed on points representing centers of all atoms, without considering the radii of atoms.
 %This may lead to detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
-To consider atoms with different radii, weights of individual points are determined by the van der Walls radii of the atoms in WVD.
+To consider atoms with different radii, the weights of individual points are determined by the van der Walls radii of atoms in WVD.
 An alternative solution is to compute a non-weighted VD on an extended point set, where 
-each atom is approximated by several spheres with small radius~\cite{yaffe2008,caver3}.
-VD-based methods are memory less demanding, and also faster than the grid-based methods.
+each atom is approximated by several spheres with a small radius~\cite{yaffe2008,caver3}.
+VD-based methods are memory less demanding and also faster than the grid-based methods.
 %The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
@@
 
-The detected tunnels are evaluated using basic characteristics like bottleneck, length, curvature and a list of surrounding
-residues, which are later used to estimate possibility of the interaction.
-The biggest disadvantage is that the shape of the ligand is not considered during the tunnel detection and it is therefore
+The detected tunnels are evaluated using basic characteristics, like bottleneck, length, curvature, and a list of surrounding
+residues, which are later used to estimate the interaction possibilities.
+The biggest disadvantage is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore
 not easy to estimate if (and how) a ligand might traverse the tunnels.
@@
 
-This requires to compute trajectories of the non-spherical ligand in the tunnels, which can be formulated as a path planning problem in a high-dimensional configuration space $\C$.
-The configuration space is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation and possibly also other degrees of freedom responsible for the conformation changes.
-The dimension of the configuration space is given by the DOF of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
+This requires to compute trajectories of the non-spherical ligand in tunnels, which can be formulated as a path planning problem in a high-dimensional configuration space $\C$.
+The configuration space is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation, and possibly also other degrees of freedom responsible for the conformation changes.
+The dimension of the configuration space is given by the degrees of freedom (DOF) of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
 ligand with $n$ DOFs.
@@
 Sampling-based methods can cope with robots (objects) with many DOFs. 
-Flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
-can changes~\cite{songPFpath}.
-This is useful e.g.
-in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
+The flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
+can change~\cite{songPFpath}.
+This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
 Sampling-based methods have also been used for 
@@
 The protein tunnels have the properties of the narrow passages. 
-A possible approach to cope with the narrow passage property of the protein tunnels is to use the information
+A possible approach to cope with the narrow passage property of protein tunnels is to use the information
 about their 3D position, which is represented by the tunnel centerline.
@@
 
-The kinematic chain representation used to model ligand flexibility allows the sampling-based planners to generate all possible conformations.
-However, not all of them are feasible and it is therefore necessary to verify feasibility of the given conformation based on energy, which is time consuming.
-Moreover, the additional degrees of freedom required to model the flexibility increases dimension of the configuration space.
+The kinematic chain representation used to model the ligand flexibility allows the sampling-based planners to generate all possible conformations.
+However, not all of them are feasible and it is therefore necessary to verify the feasibility of a given conformation based on energy, which is time consuming.
+Moreover, the additional degrees of freedom required to model the flexibility increase the dimension of the configuration space.
 RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
@@
 A RRT-based method for computing exit pathways for a small flexible molecule was presented in~\cite{cortes2010simulating}.
-To cope with the many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} was employed.
-In RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
+To cope with many-DOF ligands, the RRT-ML variant~\cite{cortes2007mlrrt} was employed.
+RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
 and translation) and it employs the other DOFs (i.e., that are responsible for conformation changes) if they hinder the growth of the tree.
-The approach~\cite{cortes2010simulating} may however suffer from the narrow passage problem, as it aims to find any exit pathway for a given ligand, which requires to search the whole configuration space of the ligand/protein complex.
+The approach~\cite{cortes2010simulating} may however suffer from the narrow passage problem, as it aims to find some exit pathway for a given ligand, which requires to search the whole configuration space of the ligand/protein complex.
 Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
@@
 The tunnels in a protein being studied can be computed using standard tunnel detection tools for spherical probes and assessed
-according their length, curvature, bottleneck and biochemical properties (e.g. partial charge).
+according to their length, curvature, bottleneck, and biochemical properties (e.g., partial charge).
 These characteristics are already used in many research studies.
@@
 
-The proposed work employs the RRT planning principle with a three extensions: 
-a) the configuration space is not sampled uniformly, but using a guided sampling along the tunnel centerline,
-b) the radii of the ligand atoms are decreased, and 
-c) ligand flexibility of modeled using a set of predefined energy-suitable conformations.
-The first two extensions can help to cope with the narrow passage problem and the third modification
+The proposed work employs the RRT planning principle with three extensions: 
+a) the configuration space is not sampled uniformly, but uses a guided sampling along the tunnel centerline,
+b) the radii of the ligand atoms can be decreased, and 
+c) ligand flexibility is modeled using a set of predefined energy-suitable conformations.
+The first two extensions can help to deal with the narrow passage problem and the third modification
 is introduced in order to decrease the dimension of the configuration space and the sensitivity to the employed
@@
 By reducing the scale of the ligand's atomic radii, the configuration tree can explore more areas of the configuration space.
-Such techniques has been used for motion planning of deformable objects, 
-e.g.,~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path} 
-and for motion planning of rigid objects in among 
+Such techniques have been used for motion planning of deformable objects, 
+e.g., in~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path}, 
+and for motion planning of rigid objects among 
 flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.
@@
 %\section{Preliminaries}
-\section{Traversability in tunnels}
+\section{Traversability of Tunnels}
 
@@
 The flexibility of a ligand is modeled using the set $\L$ of its conformations.
-The conformations can be prepared e.g., considering the potential energy.
+The conformations can be prepared, e.g., considering the potential energy.
 We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
@@
 A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
-by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., it's geometric center), rotation around $x$, $y$ and $z$ axes,
+by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., its geometric center), rotation around $x$, $y$, and $z$ axes,
 index of the conformation $l\in \L$ and the actual scale $s$.
@@
 A configuration is collision-free if none of the ligand atoms scaled by $s$ and placed at the
-position defined by $q$ collide with the protein atoms.
+position defined by $q$ collides with the protein atoms.
 %The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.
 
-\subsection{Main idea}
-
-The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand in the tunnel.
-As the traversability may depends on the initial position of the ligand, especially its rotation, it is necessary to compute
+\subsection{Main Idea}
+
+The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand inside the tunnel.
+As the traversability may depend on the initial position of the ligand, especially its rotation, it is necessary to compute
 the trajectories from multiple starting configurations $\qinit$.
@@
 A sample $q$ is created by generating its translation part randomly around $t_1$ in the distance $R$.
-The rotation part of $q$ and conformation index are set randomly, the scale is set to $\smin$.
+The rotation part of $q$ and the conformation index are set randomly, the scale is set to $\smin$.
 If the sample $q$ is collision-free, it can be considered as a new starting configuration $\qinit$.
-Then, multiple trajectories are computed starting from $\qinit$ using the method proposed in the next section.
-
-\subsection{RRT for ligand trajectory computation}
+Then, multiple trajectories are computed starting from $\qinit$ using the method proposed in the following section.
+
+\subsection{RRT for Ligand Trajectory Computation}
 
 The trajectories are computed using an extension of the RRT~\cite{lavalleRRT} planner.
-The RRT planner is extended by guided sampling, scaling of the ligands atoms and using predefined set of ligand conformations.
+The RRT planner is extended by guided sampling, scaling of the ligand atoms, and using a predefined set of ligand conformations.
 The main loop of the method is described in Alg.~\ref{alg::main}.
@@
 %Instead, they are generated using the spheres of the tunnel.
-In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
-The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
+In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
+The nearest-neighbor search between the $\qrand$ and the tree is performed using the weighted 6D Euclidean metric considering
 both 3D rotation and 3D translation.
 
-To guide tree growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
+To guide the growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
 The virtual goal $v, 1\le v \le n$, is the index of sphere of the tunnel.
@@
 This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
-The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
+The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
 the last sphere in the tunnel, i.e., when $v = n$.
@@
 To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
-from $N(c_v,\Sigma)$, where $\Sigma$ is diagonal matrix with diagonal entries equal to a parameter $\rv$, and the rotational
+from $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to a parameter $\rv$, and the rotational
 part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
@@
 %\setstretch{\straa}
-\caption{\label{alg::main}main loop of the RRT planner}
+\caption{\label{alg::main}Main loop of the RRT planner}
 \KwIn{
@@
 The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
-If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the searches continues
-until a free collision-free samples is found or until the minimal reduced-scale $\smin$ is reached.
+If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
+until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
 
@@
 part is generated around $\qnear$.
-The expansion begins with the largest scale $s$ and if less then $m$ collision-free samples are generated with $s$, the next lower $s$ is used.
+The expansion begins with the largest scale $s$ and if less than $m$ collision-free samples are generated with $s$, the next lower $s$ is used.
 The expansion step is repeated for each conformation $l \in L$.
@@
 %\setstretch{\straa}
-\caption{\label{alg::expand}expand}
+\caption{\label{alg::expand}Expand}
 \KwIn{
@@
    ligand conformations $\L$,
-   scale limits $\smin, \smax$ and $\sdelta$
+   scale limits $\smin, \smax$, and $\sdelta$
 }
@@
 The result of the planner is the tree $\T$ of collision-free configurations.
-The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the neearest node in the tree towards $\qgoal$, is found.
-The path $P=(q_i), q_i \in \C$ is represented as a sequence of the collision-free configurations.
-Let $d(P,i)$ denote the 3D Euclidean distance
+The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the nearest node in the tree towards $\qgoal$, is found.
+The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
+Let $d(P,i)$ denote the 3D Euclidean distance.
 
@@
 
-\subsection{Traversability characteristics}
+\subsection{Traversability Characteristics}
 
@@
 --- The probability of the ligand that it reaches a tunnel's part.
-We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \R$ is radius.
-Then for each trajectory $t$ we define its ending tunnel part $P_t^e$ as a tunnel part whose center is closest to the end position of $t$.
+We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \R$ is its radius.
+Then, for each trajectory $t$ we define its ending tunnel part $P_t^e$ as the part whose center is closest to the end position of $t$.
 Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
 
-As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
-On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
+As described above, our approach enables to estimate the probability that the ligand passes through the tunnel.
+On the other hand, the probability does not show any details about the ligand's passage, e.g., how complicated it is to pass through some part of the tunnel, which trajectory is the most used one, etc.
 Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
 Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
-In this manner, we aim to further support the decision process of a chemist about a tunnel.
-In order to do that, we employ an ensemble of resulting trajectories $T$, and we compute and visualize the following tunnel properties:
+In order to further support the tunnel exploration process performed by biochemists.
+To do that, we employ an ensemble of resulting trajectories $T$, and we compute and visualize the following tunnel properties:
 \begin{itemize}
 
-\item \textbf{Ligand scale} --- The scale of the ligand when it passes a tunnel's part.
-For the trajectory $t$ the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as a scale of a such sample of $t$ whose position is the closest to the center of $P_t$.
-Then, the average ligand scale in a tunnel part is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
-Similarly, minimum and maximum ligand scales can be defined.
-  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
+\item \textbf{Ligand scale} --- The scale of the ligand when it passes a given tunnel's part.
+For the trajectory $t$, the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as the scale of such sample of $t$ whose position is the closest one to the center of $P_t$.
+Then, the average ligand scale is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
+Similarly, the minimum and maximum ligand scales can be defined.
+  \item \textbf{Throughput} --- The probability that the ligand passes through a tunnel's part.
 The throughput of the tunnel part $P$ is defined as a ratio of trajectories that passed $P$ over all trajectories that reached $P$.
 \end{itemize}
-In the visualization, these properties are mapped onto a tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
+In the visualization, these properties are mapped onto the tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
 More precisely, a point $x$ on the tunnel surface is colored according to the property value in its nearest tunnel part $P$, i.e., such part $P$ whose center is the closest to $x$ among all tunnel parts.
@@
 
-\subsection{Visualization of the results}
+\subsection{Visualization of Results}
 
@@
 In order to visualize the important trajectories we first cluster the trajectories according to their similarity.
-We define the similarity of two trajectories as follows.
+We define the similarity between two trajectories as follows.
 First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
@@
 Using $x_s$ and $x_e$ we define spheres $S_1, ..., S_N$ such that $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
-Then, a vector $v$ for a trajectory $t$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $t$ which is nearest to the surface of $S_i$.
+Then, a vector $v$ for a trajectory $t$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $t$ which is the nearest one to the surface of $S_i$.
 Finally, we define a metric $d$ for trajectories $t_i$ and $t_j$ such as $d(t_i, t_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
 The clustering is then computed using the UPGMA clustering algorithm~\cite{sokal1958statistical}.
-When visualizing we choose a representant from each cluster a we depict its trajectory in 3D using a polyline.
-Moreover, we map the representant's cluster size onto the width of the polyline.
+When visualizing, we choose a representative from each cluster and depict its trajectory in 3D using a polyline.
+Moreover, we map the representative's cluster size onto the width of the polyline.
 In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
@@
 \caption{Visualization of resulting trajectories.
-Left: All trajectories colored according to whether they reached the end of the tunnel (green) or they did not (red).
+Left: All trajectories colored according to whether they reached the end of the tunnel (green) or not (red).
 Right: Trajectories clustered according to their proximity colored by distinct colors.
@@
 
-\section{Experimental verification}
-
-The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase (protein 1CQW) (Fig.~\ref{fig::tunnel}a).
+\section{Experimental Verification}
+
+The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase protein (PDB ID 1CQW) (Fig.~\ref{fig::tunnel}a).
 The tunnel was computed using CAVER 3.0~\cite{caver3} for spherical probe of radius 0.9~\AA.
@@
 
-The tested ligands are bigger than the tunnel with, therefore the ligands have been scaled down, and four
+The tested ligands are bigger than the tunnel with, therefore the ligands were scaled down, and four
 different scales were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
 No trajectories were found for $\smin > 0.6$.
-For each of this scale, 50 collision-free initial states were found in the $2$\AA\ radius of the first sphere of the tunnel.
+For each scale, 50 collision-free initial states were found in the $2$\AA\ radius from the first sphere of the tunnel.
 For each ligand, each minimal scale $\smin$ and each start configuration, 100 trajectories were computed using the proposed planner.
-The parameters of the planner are: $\Imax=10000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA.
-For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance $\rv$ or less) over the 100 trials.
+The parameters of the planner are: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA.
+For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance $\rv$ or less) over 100 trials.
 
 The average runtimes and sizes of the built trees are shown in Tab.~\ref{tab::rrt}.
-The tables also show the minimal, maximal success rate and the average success rate.
-The average success rate is compute as the average value of the success rates achieved for all 50 initial configurations.
+The tables also show the minimal, maximal success rate, and the average success rate.
+The average success rate is computed as the average value of the success rates achieved for all 50 initial configurations.
 
@@
 \caption{\label{fig::tunnel}
-    (a) The protein 1CQW visualized using its secondary structure (gray) with its highest ranked tunnel (green) which was detected using CAVER 3.0.
+    (a) The protein 1CQW visualized using the cartoon representation (gray) with its highest ranked tunnel (green) which was detected using CAVER 3.0.
     (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
@@
 
-\section{Discussion \& implementation details}
+\section{Discussion \& Implementation Details}
 
@@
 The proposed extensions to the RRT planner have been motivated by the necessity to cope with the narrow passage
-and the high dimensional of the flexible ligands.
+and the high dimensionality of the flexible ligands.
 By scaling-down the geometry of ligand, which is also used in other related tools~\cite{cortes2005path}, 
-the relative volume of the free region of the configuration space increases, which increases probability of placing
+the relative volume of the free region of the configuration space increases, which increases the probability of placing
 random samples $\qrand$ into the free region.
@@
 smaller than the bounding box of the whole protein.
-This, together with the scaled-down ligand, increases relative volume of the free configuration space and help the RRT to
+This, together with the scaled-down ligand, increases the relative volume of the free configuration space and helps the RRT to
 reach the goal configuration.
@@
 only atoms located in the vicinity of the tunnel can be considered for collision detection.
-A fast collision-detection can be achieved either using fast hierarchical data structures as OBB (implemented e.g. in~\cite{ozcollide})
+Fast collision-detection can be achieved either using fast hierarchical data structures as OBB (implemented, e.g., in~\cite{ozcollide})
     or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.

### 2017-06-28 paper.tex
tunnel properties + other comments

@@
 \def\R{\mathbb{R}}
+\def\QI{Q_{init}}
 
@@
 \def\SS{\mathbf{S}}
-
-\def\RRTTD{RRT$_{\mathrm{td}}$}
-\def\RRTN{RRT$_{\mathrm{n}}$}
-\def\ths{d_\mathrm{surf}}
 
@@
 
-%\section{Preliminaries}
 \section{Traversability of Tunnels}
@@
 Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
-%Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
-%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
-%   can be detected as atoms located at $\alpha$-shape of the molecule.
-%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
 The flexibility of a ligand is modeled using the set $\L$ of its conformations.
@@
 
-\subsection{Main Idea}
-
-The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand inside the tunnel.
-As the traversability may depend on the initial position of the ligand, especially its rotation, it is necessary to compute
-the trajectories from multiple starting configurations $\qinit$.
-The starting configuration $\qinit$ is generated by randomly sampling the vicinity of the first sphere $c_1$ of the tunnel.
-A sample $q$ is created by generating its translation part randomly around $t_1$ in the distance $R$.
+\subsection{Computing initial configuration}
+
+The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand inside the tunnel, which
+requires to generate a set $\QI$ of collision-free initial configurations.
+A sample $q$ is created by generating its translation part randomly around $c_1$ in the distance $R$.
 The rotation part of $q$ and the conformation index are set randomly, the scale is set to $\smin$.
-If the sample $q$ is collision-free, it can be considered as a new starting configuration $\qinit$.
-Then, multiple trajectories are computed starting from $\qinit$ using the method proposed in the following section.
-
-\subsection{RRT for Ligand Trajectory Computation}
+If the sample $q$ is collision-free, it can be considered as a new starting configuration, so $q$ is added to $\QI$.
+For each starting configuration $\qinit \in \QI$, we compute multiple trajectories.
+Each trajectory is computed using a variant of RRT, which is proposed in the following section.
+
+\subsection{Computing single trajectory of the ligand}
 
@@
 The main loop of the method is described in Alg.~\ref{alg::main}.
-%The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
-%We propose a RRT-based method to compute the trajectories of the ligand.
-%As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
-%suffer from the well known narrow passage problem~\cite{hannaWIS}.
-%To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
-%Instead, they are generated using the spheres of the tunnel.
 In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
@@
 
-The result of the planner is the tree $\T$ of collision-free configurations.
-The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the nearest node in the tree towards $\qgoal$, is found.
+The result of each planning trial is the tree $\T$ of collision-free configurations.
+The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the nearest node in the tree towards $\qgoal$, is found in the tree.
 The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
-Let $d(P,i)$ denote the 3D Euclidean distance.
-
-The path $P$ reaches a particular sphere $t_i$ in the tunnel if the 3D Euclidean distance between the sphere and its nearest configuration
-in the path is less than the threshold $T$.
-TODO VV dodelam
 
@@
 
-The traversability of the tunnel $T$ is evaluated using all computed trajectories $P_i$. 
-Three basic characteristics of the tunnel are computed from the trajectories: the accessiblity, throughput and the scaling factor.
-These characteristics are computed for each sphere $(c_i,r_i) \in T$ of the tunnel.
-This requires to determine which parts of the tunnel were accessed by the trajectories.
-
-The Accessibility 
---- The probability of the ligand that it reaches a tunnel's part.
-We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \R$ is its radius.
-Then, for each trajectory $t$ we define its ending tunnel part $P_t^e$ as the part whose center is closest to the end position of $t$.
-Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
-
-As described above, our approach enables to estimate the probability that the ligand passes through the tunnel.
-On the other hand, the probability does not show any details about the ligand's passage, e.g., how complicated it is to pass through some part of the tunnel, which trajectory is the most used one, etc.
-Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
-Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
-In order to further support the tunnel exploration process performed by biochemists.
-To do that, we employ an ensemble of resulting trajectories $T$, and we compute and visualize the following tunnel properties:
-\begin{itemize}
-
-\item \textbf{Ligand scale} --- The scale of the ligand when it passes a given tunnel's part.
-For the trajectory $t$, the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as the scale of such sample of $t$ whose position is the closest one to the center of $P_t$.
-Then, the average ligand scale is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
-Similarly, the minimum and maximum ligand scales can be defined.
-  \item \textbf{Throughput} --- The probability that the ligand passes through a tunnel's part.
-The throughput of the tunnel part $P$ is defined as a ratio of trajectories that passed $P$ over all trajectories that reached $P$.
-\end{itemize}
-In the visualization, these properties are mapped onto the tunnel's surface (see Fig.~\ref{fig:properties}) so that they can be visually correlated with, e.g., the tunnel's shape.
-More precisely, a point $x$ on the tunnel surface is colored according to the property value in its nearest tunnel part $P$, i.e., such part $P$ whose center is the closest to $x$ among all tunnel parts.
+The trajectories are computed for all initial configurations $\qinit \in \QI$ and for each of the starting
+configuration, multiple trajecotries are obtained.
+For the purpose of this paper, we compute the tunnel characteristics using all computed trajectories.
+
+A trajectory $P$ reaches the tunnel sphere $c_i \in t$ if the 3D Euclidean distances of the 
+nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
+Let $N(i)$ denote the number of trajectories, that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
+Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput and the scaling factor.
+
+The accessibility $A(i)$ of the sphere $i$ is the probability of reaching the sphere by the computed trajectories,
+i.e., $A(i)=N(i) / N$, where $N$ is the total number of all trajectories.
+The accessibility  shows how probable is to pass the tunnel upto the sphere $i$.
+The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
+the bottleneck have low accessibility.
+
+To better measure how probable is to pass a tunnel part, we should take into account only trajectories, that even have chance
+to visit it, e.g. the trajectories that reached the preceding tunnel part.
+The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
+i.e, $T(i) = N(i+1) / N(i)$. 
+The throughput is not computed for the last sphere ($i=N$).
+The throughput shows a local accessibility of the tunnel parts.
+
+%The probability of the ligand that it reaches a tunnel's part.
+%We employ the tunnel representation from CAVER 3.0 and we define a tunnel part as a sphere $P = (x, r)$ where $x \in \mathbb{R}^3$ is sphere center and $r \in \R$ is its radius.
+%Then, for each trajectory $t$ we define its ending tunnel part $P_t^e$ as the part whose center is closest to the end position of $t$.
+%Finally, the accessibility of $P$ is defined as a ratio of trajectories that passed $P$ or ended in $P$ over all trajectories.
+%On the other hand, the probability does not show any details about the ligand's passage, e.g., how complicated it is to pass through some part of the tunnel, which trajectory is the most used one, etc.
+%Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
+
+The proposed planner is allowed to scale down radii of the ligand up to the scale $\smin$. 
+It can be expected, that narrow parts of the tunnels are more often passed with a more scaled down ligand, than
+the wider parts.
+Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
+The ligand scale is computed as the average value of scales of nearest configurations of the trajectories towards tunnel sphere
+$i$.
+
+%--- The scale of the ligand when it passes a given tunnel's part.
+%For the trajectory $t$, the ligand scale $s(P_t)$ in the tunnel part $P_t$ is defined as the scale of such sample of $t$ whose position is the closest one to the center of $P_t$.
+%Then, the average ligand scale is defined as $s(P) = \frac{1}{|T|} \sum_{t \in T} s(P_t)$.
+%Similarly, the minimum and maximum ligand scales can be defined.
+
+\subsection{Visualization of Results}
+
+%In order to fursupport the tunnel exploration process performed by biochemists.
+The computed properties of the tunnel are presented visually by mapping the accessibility, throughput and scale to the sphere representation
+of the tunnel (Fig.~\ref{fig:properties}) using a color map.
+Alternatively, the surface representation can be used to visualize the tunnel.
+In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $t_i$ is the closest to the point among all tunnel parts.
 
@@
 \begin{tabular}{ccc}
-\includegraphics[width=0.3\textwidth]{fig/accessibility}
-\quad & \quad
-\includegraphics[width=0.3\textwidth]{fig/ligand-scale}
-\quad & \quad
-\includegraphics[width=0.3\textwidth]{fig/throughput}
+\includegraphics[width=0.3\textwidth]{fig/accessibility} &
+\includegraphics[width=0.3\textwidth]{fig/throughput} &
+\includegraphics[width=0.3\textwidth]{fig/ligand-scale} \\
+  (a) & (b) & (c) \\                     
 \end{tabular}
 \caption{Tunnel properties mapped onto its surface.
-Left: Accessibility --- green denotes parts that were accessed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
-Middle: Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
-Right: Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.}
+(a) Accessibility --- green denotes parts that were accessed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
+(b) Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.
+(c) Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
 \label{fig:properties}
+}
 \end{figure}
 
-\subsection{Visualization of Results}
-
-Moreover, we analyze the whole set of trajectories in order to convey the most important trajectories that enabled the ligand to pass or that prevented it from passing.
-In order to visualize the important trajectories we first cluster the trajectories according to their similarity.
-We define the similarity between two trajectories as follows.
-First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
-In order to obtain the samples, we find the average starting point $x_s$ of all trajectories and the most distant (w.r.t. $x_s$) ending point $x_e$ also among all trajectories.
-Using $x_s$ and $x_e$ we define spheres $S_1, ..., S_N$ such that $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
-Then, a vector $v$ for a trajectory $t$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $t$ which is the nearest one to the surface of $S_i$.
-Finally, we define a metric $d$ for trajectories $t_i$ and $t_j$ such as $d(t_i, t_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
+The mapping using colors helps to identify difficult part of the tunnels, but it is also necessary to show the raw trajectories.
+This is necessary for example if the trajectories detour from a part of the tunnel.
+The na\"{i}ve visualization of all trajectories is possible, but too slow if the number of trajectories is in the order of thousands.
+
+In order to visualize the most typical trajectories, we cluster the trajectories according to their similarity.
+The number of configurations in the trajectories usually differ due to the stochastic nature of RRT and it also depends on the
+initial configurations.
+
+To compare two trajectories (which are sequence of configuration in the configuration space), they are first converted
+to a vector $v=(x_1,x_2,\ldots,x_M)$, where $x_i \in \R^3$ is a 3D
+point representing the trajectory and $M$ is the parameter determining how precisely we want to represent the trajectories.
+To obtain the vector $v$ for a given trajectory $P$, first a set of spheres 
+ $S_1, ..., S_N$, where $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
+%First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
+%In order to obtain the samples, we find the average starting point $x_s$ of all trajectories and the most distant (w.r.t. $x_s$) ending point $x_e$ also among all trajectories.
+%Using $x_s$ and $x_e$ we define spheres $S_1, ..., S_N$ such that $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
+Then, a vector $v$ for a trajectory $P$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $P$ which is the nearest one to the surface of $S_i$.
+Finally, we define a metric $d$ for trajectories $P_i$ and $P_j$ as $d(P_i, P_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
 The clustering is then computed using the UPGMA clustering algorithm~\cite{sokal1958statistical}.
@@
 \includegraphics[width=0.4\textwidth]{fig/trajectories-all}
-\qquad & \qquad
 \includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
+(a) & (b) \\                       
 \end{tabular}
 \caption{Visualization of resulting trajectories.
-Left: All trajectories colored according to whether they reached the end of the tunnel (green) or not (red).
-Right: Trajectories clustered according to their proximity colored by distinct colors.
-The different line widths represent the number of trajectories in a cluster.}
+(a) All trajectories colored according to whether they reached the end of the tunnel (green) or not (red).
+(b) Trajectories clustered according to their proximity colored by distinct colors.
+The different line widths represent the number of trajectories in a cluster.
 \label{fig:trajectories}
+}
 \end{figure}

### 2017-06-28 paper.tex
intro additional material

--- 2017-06-28 paper.tex
@@
 
-However, ligands are typically of non-spherical shape and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
+Whereas the tunnel computation is already well established research field, the simulation of ligand transportation through the detected tunnels is rather new field.
+Ligands are typically of non-spherical shape and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
 The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
 To provide biochemists with better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.
+
+The flexibility of the ligand is modeled using a predefined set of typical conformations.
+The conformation changes of protein amino acids and ligand are represented by so called rotamers, stored in different rotamer libraries (e.g., Dunbrack library~\cite{dunbrack}).
+Rotamers differ in the angles between their atoms.
+Possible rotamer conformations correspond to energetically and geometrically favorable positions.
 
@@
 To make the transportation easier, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
-The flexibility of the ligand is modeled using a predefined set of typical conformations.
+This solution is common also in different types of calculations, e.g.~\cite{kellogg}. 
+It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
+This strategy helps to overcome the limitations of the conformational discretation introduced by the rotamer approximation.
+
+Our method helps the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
+In order to predict the success of traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
+The advantage of our method is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources.
+
 We further propose novel visualization techniques for presenting the results to biochemists.

### 2017-06-28 paper.tex
corrections

@@
 
-\subsection{Computing initial configuration}
+\subsection{Computing Initial Configuration}
 
@@
 
-\subsection{Computing single trajectory of the ligand}
+\subsection{Computing Single Trajectory of the Ligand}
 
@@
 The trajectories are computed for all initial configurations $\qinit \in \QI$ and for each of the starting
-configuration, multiple trajecotries are obtained.
+configuration, multiple trajectories are obtained.
 For the purpose of this paper, we compute the tunnel characteristics using all computed trajectories.
 
-A trajectory $P$ reaches the tunnel sphere $c_i \in t$ if the 3D Euclidean distances of the 
+Trajectory $P$ reaches the tunnel sphere $c_i \in t$, if the 3D Euclidean distance of the 
 nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
-Let $N(i)$ denote the number of trajectories, that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
-Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput and the scaling factor.
+Let $N(i)$ denote the number of trajectories that reached $i-$th sphere of the tunnel, $i=1,\ldots,n$.
+Three basic characteristics of the tunnel are computed from the trajectories: accessibility, throughput, and the scaling factor.
 
 The accessibility $A(i)$ of the sphere $i$ is the probability of reaching the sphere by the computed trajectories,
-i.e., $A(i)=N(i) / N$, where $N$ is the total number of all trajectories.
-The accessibility  shows how probable is to pass the tunnel upto the sphere $i$.
+i.e., $A(i)=N(i) / N$, where $N$ is the total number of trajectories.
+The accessibility  shows how probable is to pass the tunnel up to the sphere $i$.
 The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
@@
 
-To better measure how probable is to pass a tunnel part, we should take into account only trajectories, that even have chance
-to visit it, e.g. the trajectories that reached the preceding tunnel part.
+To better measure the probability of passing a given tunnel part, we should take into account only trajectories that even have chance
+to visit this part, e.g., the trajectories that reached the preceding tunnel part.
 The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
@@
 
-The proposed planner is allowed to scale down radii of the ligand up to the scale $\smin$. 
-It can be expected, that narrow parts of the tunnels are more often passed with a more scaled down ligand, than
+The proposed planner is allowed to scale down the radii of ligand spheres up to the allowed scale $\smin$. 
+It can be expected, that narrower parts of the tunnel are more often passed with a more scaled-down ligand than
 the wider parts.
 Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
-The ligand scale is computed as the average value of scales of nearest configurations of the trajectories towards tunnel sphere
+The ligand scale is computed as the average value of scales of the nearest configurations of the trajectories towards tunnel sphere
 $i$.
@@
 %In order to fursupport the tunnel exploration process performed by biochemists.
-The computed properties of the tunnel are presented visually by mapping the accessibility, throughput and scale to the sphere representation
+The computed properties of the tunnel are presented visually by mapping the accessibility, throughput, and scale to the sphere representation
 of the tunnel (Fig.~\ref{fig:properties}) using a color map.
@@
 
-The mapping using colors helps to identify difficult part of the tunnels, but it is also necessary to show the raw trajectories.
+The mapping using colors helps to identify difficult parts of the tunnel, but it is also necessary to show the original trajectories.
 This is necessary for example if the trajectories detour from a part of the tunnel.
-The na\"{i}ve visualization of all trajectories is possible, but too slow if the number of trajectories is in the order of thousands.
+The na\"{i}ve visualization of all trajectories is possible, but too slow if we are dealing with thousands of trajectories.
 
@@
 
-To compare two trajectories (which are sequence of configuration in the configuration space), they are first converted
+To compare two trajectories (which are sequences of configurations in the configuration space), we first convert them
 to a vector $v=(x_1,x_2,\ldots,x_M)$, where $x_i \in \R^3$ is a 3D

### 2017-06-29 paper.tex
algoritmus

@@
 \def\QI{Q_{init}}
+\def\RI{R_{init}}
 
@@
 \SetKw{return}{return}
-
-\def\VV{V_{vor}}
-\def\VVA{V_{vor}^{act}}
 
@@
 Rapidly Exploring Random Tree (RRT).
-We propose novel modifications to RRT in order to compute the trajectories around the tunnel and also to consider the flexibility of the ligand.
+We propose several modifications to RRT in order to compute the trajectories around the tunnel and also to consider the flexibility of the ligand.
 We also introduce several approaches for visualization of the resulting trajectories which depict valuable information about various properties and behavior of the ligand.
@@
 
-The proteins are dense structures with narrow tunnels.
+The proteins are dense structures and the tunnels are typically narrow.
 Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
@@
 $\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
-This scaling technique is used also in other tools like MoMa-LigPath~\cite{cortes2005path}.
+The scaling-down  technique is used also in other tools like MoMa-LigPath~\cite{cortes2005path}.
 
@@
 requires to generate a set $\QI$ of collision-free initial configurations.
-A sample $q$ is created by generating its translation part randomly around $c_1$ in the distance $R$.
-The rotation part of $q$ and the conformation index are set randomly, the scale is set to $\smin$.
+In this paper we assume that the ligand has to travel from the begining to the end of the tunnel.
+As the tunnel are computed for a spherical probe, the ligand may not fit exactly to the first sphere of the tunnel.
+The initial configurations have to be searched around the first sphere of the tunnel.
+To find a new initial configuration, a random sample $q$ is generated around $c_i$ in the distance $\RI$ (translation and rotation
+ parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
 If the sample $q$ is collision-free, it can be considered as a new starting configuration, so $q$ is added to $\QI$.
-For each starting configuration $\qinit \in \QI$, we compute multiple trajectories.
-Each trajectory is computed using a variant of RRT, which is proposed in the following section.
+For each starting configuration $\qinit \in \QI$,  $K$ trajectories are created.
+Each trajectory is computed using a modified RRT, which is introduced in the following section.
 
@@
 
-The trajectories are computed using an extension of the RRT~\cite{lavalleRRT} planner.
-The RRT planner is extended by guided sampling, scaling of the ligand atoms, and using a predefined set of ligand conformations.
-The main loop of the method is described in Alg.~\ref{alg::main}.
+The task of the trajectory computation is to find a trajectory for the given ligand in the given tunnel, which can be solved
+using RRT~\cite{lavalleRRT}.
+The original RRT is exnteded to cope with specific requirmeents needed for traversability of ligands.
+First, the trajectory has to be found around the given tunnel, but deviation from the tunnel centerline are allowed.
+We assume, that the distance of the ligand from the tunnel centerline is at most $\rv$.
+Due to this requirement, sampling process of RRT has to be adapted in order to follow the tunnel, and to prevent
+construction of trajectories in the rest of the protein.
+The other modifications aim to cope with various scales of the ligand and multiple conformations,
+
+The main loop of the proposed method is described in Alg.~\ref{alg::main}.
 In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
@@
 The virtual goal $v, 1\le v \le n$, is the index of sphere of the tunnel.
-The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated
-around the sphere $t_v \in T$.
+The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated around the sphere $t_v \in T$.
 After the tree reaches the sphere $t_v$, i.e., the distance of the tree to the center of $t_v$ is
@@
 that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
-Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are
-not traversable or reachable by the ligand.
+Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
 This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
@@
 The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
-For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal
-scale.
+For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
 First, the maximal scale $\smax$ is tested and $m$ random samples are generated around $\qrand$ and tested for collision.
@@
 until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
+The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
+part is generated around $\qnear$.
 
@@
 that approach $\qrand$, the distance between newly generated configurations and $\qrand$ is measured as the smallest
-distance between an atom of the ligand placed at $q$ and the 3D position of $\qrand$ ($\da(q,\qrand)$ on line~\ref{alg::expand:a} in
+3D distance between an atom of the ligand placed at $q$ and the 3D position of $\qrand$ ($\da(q,\qrand)$ on line~\ref{alg::expand:a} in
 Alg.~\ref{alg::expand}).
-
-The random samples $q$ are generated similarly as in the case of $\qrand$ samples, only their translation
-part is generated around $\qnear$.
-The expansion begins with the largest scale $s$ and if less than $m$ collision-free samples are generated with $s$, the next lower $s$ is used.
-The expansion step is repeated for each conformation $l \in L$.
+By computing the distance between the nearest atoms, the shape of the ligand is actually considered.
+As the expansion routine attemts to find such sample that minimizes this distance to $\qrand$, it supports retraction
+of the ligand towards $\qrand$.
 
@@
 
-%Approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing ordinary Voronoi diagram.
-%Computing WVD this way is more efficient and numerically stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
-%Vertices of this WVD are filtered out if their distance to an atom is less than probe radius $\probe$.
-%Let $\VV$ contain the remaining Voronoi vertices, and let $r_v \in \VV$ denote the distance to the nearest atom.
-%The remaining Voronoi vertices $\VV$ of the WVD and their distances $r_v, v\in \VV$ to the nearest atom are then used in the following method.
-
-%A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples in the difficult areas.
-%For Probabilistic Roadmaps~\cite{kavrakiPRM}, this can be achieved simply by generating more samples in the difficult areas, e.g. along
-%medial axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
-%This requires to precompute the medial axis, which can be time consuming.
-%Another approach is to shift the random samples towards the medial axis~\cite{amatoOBPRM} or into a direction of estimated
-%medial axis~\cite{hollemanMAPRM}.
-%Generating random samples in difficult regions however does not ensure construction of a better tree in the case of RRT-based methods. 
-%In RRT, the configuration tree is expanded in the direction of random samples, but the obstacles may prevent to reach them closely~\cite{vonasekphd}.
-%Increased probability of sampling in a given region, e.g., in a narrow passage, 
-%brings advantage only if the tree is close to the region and if it can reach it.
-%it is necessary to maintain the sampling regions according the progress of the tree.
-%For example, random samples can be generated along a given workspace path~\cite{vonasek2009rrt}.
-%After the tree reaches a path point, the sampling around the point is disabled and moved to the next point of 
-%the path.
-%Similar approach is used in~\cite{denny2014marrt}, where the sampling regions are maintained with a grapth describing topology 
-%of the workspace.
-%OB-RRT~\cite{amatoOBRRT} defines several types of tree expansions
-%steps based on information about obstacles (e.g. using vectors of obstacles, tangential vectors and information about medial axis).
-
-%~\cite{amatoOBPRM,hollemanMAPRM} shift random samples towards medial axis of the environment, which helps to sample
-%narrow passages more densely. 
-%For example, the random samples generated in the configuration space can be shifted
-%towards the medial axis of the environment~\cite{amatoOBPRM,hollemanMAPRM} or even generated around the medial 
-%axis~\cite{wilmarthMAPRM,foskey01hybrid,guibas1999probabilistic,hoff2000interactive,yang2004adapting}.
-%The paper~\cite{amatoOBPRM} suggests to sample uniformly in the configuration space and shift the samples towards the medial axis.
-%Another approach is to generate the samples randomly in \CS\ and shift them closer to the medial axis~\cite{amatoOBPRM}.
-%The medial axis can be computed exactly using GVD.
-%While the medial axis can be easily computed for 2D or 3D workspaces using GVD, its computation become complicated in many dimensions.
-%In such case, the samples can be generated uniformly from the whole \CS\ and moved in an estimated direction towards the medial axis~\cite{hollemanMAPRM}.
-
 %sampling-based for 

### 2017-06-29 paper.tex
images

@@
 The virtual goal $v, 1\le v \le n$, is the index of sphere of the tunnel.
-The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated around the sphere $t_v \in T$.
-After the tree reaches the sphere $t_v$, i.e., the distance of the tree to the center of $t_v$ is
-less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
+The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated around the sphere $c_v \in T$.
+After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
+less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
 that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
@@
     tunnel spheres $c_i$ and their radii $r_i$, $i=1,\ldots,n$, 
-    initial configuration $\qinit$,
-    goal configuration $\qgoal$,
+    initial configuration $\qinit$
+%    goal configuration $\qgoal$,
 }
@@
 %\setstretch{\straa}
-\caption{\label{alg::expand}Expand}
+\caption{\label{alg::expand}expand}
 \KwIn{
@@
 }
-%\return $S$\;
 \end{algorithm}
@@
 The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
+The scale of ligand may vary in each part of the trajectory and also the conformations.
+The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.
 
@@
 
-The trajectories are computed for all initial configurations $\qinit \in \QI$ and for each of the starting
-configuration, multiple trajectories are obtained.
-For the purpose of this paper, we compute the tunnel characteristics using all computed trajectories.
-
-Trajectory $P$ reaches the tunnel sphere $c_i \in t$, if the 3D Euclidean distance of the 
+For each initial configuration $\qinit \in \QI$, $K$ trajectories are computed, which results in the set of $K |\QI|$ trajectories.
+All these trajectories are used to compute the following properties of the tunnel $T$.
+A trajectory $P$ reaches the tunnel sphere $c_i \in T$, if the 3D Euclidean distance of the 
 nearest configuration $q \in P$ towards $c_i$ is less than $r_i$ (radius of the sphere $c_i$).
@@
 The accessibility $A(i)$ of the sphere $i$ is the probability of reaching the sphere by the computed trajectories,
-i.e., $A(i)=N(i) / N$, where $N$ is the total number of trajectories.
+i.e., $A(i)=N(i) / N$, where $N=K|\QI|$ is the total number of trajectories.
 The accessibility  shows how probable is to pass the tunnel up to the sphere $i$.
@@
 The computed properties of the tunnel are presented visually by mapping the accessibility, throughput, and scale to the sphere representation
-of the tunnel (Fig.~\ref{fig:properties}) using a color map.
+of the tunnel using a color map.
 Alternatively, the surface representation can be used to visualize the tunnel.
 In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $t_i$ is the closest to the point among all tunnel parts.
+Examples of collow mapping are shown in Fig.~\ref{fig:properties}.
 
@@
 
-The mapping using colors helps to identify difficult parts of the tunnel, but it is also necessary to show the original trajectories.
-This is necessary for example if the trajectories detour from a part of the tunnel.
+Besides the color mapping, it is also necessary to show the computed trajectories.
+This is useful for example to find out in which part of the tunnel the trajectories detour.
 The na\"{i}ve visualization of all trajectories is possible, but too slow if we are dealing with thousands of trajectories.
@@
 initial configurations.
-
-To compare two trajectories (which are sequences of configurations in the configuration space), we first convert them
+% by VV ==================================================================
+%For the purpose of clusting, the trajectories have to be first represented by vectors of same length.
+%\def\cstart{\overline{c_{start}}}
+%\def\cend{\overline{c_{end}}}
+%Let $\cstart \in \R^3$ and $\cend \in \R^3$ denote the average starting and ending point of all computed trajectories.
+%A line $x(t) = \cstart + t(\cend - \cstart), 0 \le t \le 1$ is created. For each point on the line, a radius
+%$r(t)$ is assigned as $r(t) =
+
+To compare two trajectories  (which are sequences of configurations in the configuration space), they are first converted
 to a vector $v=(x_1,x_2,\ldots,x_M)$, where $x_i \in \R^3$ is a 3D
 point representing the trajectory and $M$ is the parameter determining how precisely we want to represent the trajectories.
-To obtain the vector $v$ for a given trajectory $P$, first a set of spheres 
- $S_1, ..., S_N$, where $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
+
+%by ADAM ============================================================
+%To obtain the vector $v$ for a given trajectory $P$, first a set of spheres 
+ %$S_1, ..., S_M$, where $S_i = (\cstart, r_i)$ where $r_i = i \frac{|\cstart \cend|}{M + 1}$.
  %First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
@@
 Then, a vector $v$ for a trajectory $P$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $P$ which is the nearest one to the surface of $S_i$.
+
 Finally, we define a metric $d$ for trajectories $P_i$ and $P_j$ as $d(P_i, P_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
@@
 \end{table}
+
+\begin{figure}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[width=0.32\textwidth]{fig/tunne4ac}& 
+\includegraphics[width=0.32\textwidth]{fig/tunne4bc} & 
+\includegraphics[width=0.32\textwidth]{fig/tunne4cc} \\
+(a) & (b) & (c)                        
+\end{tabular}
+\caption{\label{fig::t4}
+    todo
+}
+\end{figure}
 

### 2017-06-30 paper.tex
clustering

@@
 
+\def\cstart{\overline{c_{start}}}
+\def\cend{\overline{c_{end}}}
+
 Besides the color mapping, it is also necessary to show the computed trajectories.
 This is useful for example to find out in which part of the tunnel the trajectories detour.
-The na\"{i}ve visualization of all trajectories is possible, but too slow if we are dealing with thousands of trajectories.
-
-In order to visualize the most typical trajectories, we cluster the trajectories according to their similarity.
-The number of configurations in the trajectories usually differ due to the stochastic nature of RRT and it also depends on the
-initial configurations.
-% by VV ==================================================================
-%For the purpose of clusting, the trajectories have to be first represented by vectors of same length.
-%\def\cstart{\overline{c_{start}}}
-%\def\cend{\overline{c_{end}}}
-%Let $\cstart \in \R^3$ and $\cend \in \R^3$ denote the average starting and ending point of all computed trajectories.
-%A line $x(t) = \cstart + t(\cend - \cstart), 0 \le t \le 1$ is created. For each point on the line, a radius
-%$r(t)$ is assigned as $r(t) =
-
-To compare two trajectories  (which are sequences of configurations in the configuration space), they are first converted
-to a vector $v=(x_1,x_2,\ldots,x_M)$, where $x_i \in \R^3$ is a 3D
-point representing the trajectory and $M$ is the parameter determining how precisely we want to represent the trajectories.
-
-%by ADAM ============================================================
-%To obtain the vector $v$ for a given trajectory $P$, first a set of spheres 
- %$S_1, ..., S_M$, where $S_i = (\cstart, r_i)$ where $r_i = i \frac{|\cstart \cend|}{M + 1}$.
- %First, we represent each trajectory $t$ by a vector $v = (x_1, x_2, ..., x_N)$ of $N$ samples where $x_i \in \mathbb{R}^3$.
-%In order to obtain the samples, we find the average starting point $x_s$ of all trajectories and the most distant (w.r.t. $x_s$) ending point $x_e$ also among all trajectories.
-%Using $x_s$ and $x_e$ we define spheres $S_1, ..., S_N$ such that $S_i = (x_s, r_i)$ where $r_i = i \frac{|x_s x_e|}{N + 1}$.
-Then, a vector $v$ for a trajectory $P$ is obtained such that an $i$-th sample $x_i$ in $v$ corresponds to the position in $P$ which is the nearest one to the surface of $S_i$.
-
-Finally, we define a metric $d$ for trajectories $P_i$ and $P_j$ as $d(P_i, P_j) = \frac{1}{N} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
-The clustering is then computed using the UPGMA clustering algorithm~\cite{sokal1958statistical}.
-When visualizing, we choose a representative from each cluster and depict its trajectory in 3D using a polyline.
-Moreover, we map the representative's cluster size onto the width of the polyline.
+The simple visualization of all trajectories could however be too slow for an interactive work.
+Therefore, the trajectories are first clustered and then only the clusters are visualized.
+Due to different lengths of the trajectories, they are first converted to a normalized form.
+Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ is the 3D Euclidean distance
+of the most distant configuration from $\cstart$ from all trajectories.
+A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d \over M}$, where $i=0,\ldots, M-1$.
+The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
+where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
+For the purpose of clustering, the distance between two normalized trajectories $v_i$ and $v_j$ is defined as
+$d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
+The clustering is then computed using the UPGMA algorithm~\cite{sokal1958statistical}.
+
+The trajectories are then visualized using representative of each cluster.
+The number of trajectories in each cluster is represented by the width of the polyline.
 In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
@@
 
-The proposed method has been used to analyze a tunnel in Haloalkane dehalogenase protein (PDB ID 1CQW) (Fig.~\ref{fig::tunnel}a).
+The proposed approach has been used to analyze tunnels in Haloalkane dehalogenase protein (PDB ID 1CQW) (Fig.~\ref{fig::tunnel}a).
 The tunnel was computed using CAVER 3.0~\cite{caver3} for spherical probe of radius 0.9~\AA.
@@
 % from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
-The traversability of the tunnel was evaluated for 1-chlorpropan (denoted as $\LA$) and 1-chlorbutan (denoted as $\LB$).
+The traversability of the tunnel was evaluated for 1-chlorpropan (denoted as $\LA$) with 11 atoms and 
+and 1-chlorbutan (denoted as $\LB$) with 14 atoms.
 $\LA$ was represented by 12 conformations, and $\LB$ by a set of $114$ conformations.
@@
 
-The tested ligands are bigger than the tunnel with, therefore the ligands were scaled down, and four
-different scales were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
+Both tested ligands have more than 10 atoms and therefore they cannot fit into the tunnel of with 0.9~\AA.
+To find collision-free trajectories, the radii of ligands have to be scaled down. 
+Four different scaling-down factors were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
 No trajectories were found for $\smin > 0.6$.
-For each scale, 50 collision-free initial states were found in the $2$\AA\ radius from the first sphere of the tunnel.
-For each ligand, each minimal scale $\smin$ and each start configuration, 100 trajectories were computed using the proposed planner.
-The parameters of the planner are: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA.
+For each scale, 50 collision-free initial configurations were found in the $2$\AA\ radius from the first sphere of the tunnel.
+For each ligand, each minimal scale $\smin$ and each initial configuration, 100 trajectories were computed using the proposed planner.
+The parameters of the planner were: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA.
 For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance $\rv$ or less) over 100 trials.

### 2017-06-30 paper.tex
intro

@@
 The geometrical and physico-chemical properties of tunnels need to be determined in order to understand the transport of a ligand from and into the protein active site. 
-Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamic (MD) simulations.
-Although the tunnels can be computed fast using e.g., Voronoi diagrams, they cannot provide information about ligand traversability, as it is 
-not included in the VD-based computation.
-On the other side, MD simulations are too slow for a daily routine.
-Many degrees of freedom need to be considered in order to analyze the traversability, which motivates us to employ robotic motion planning tools.
-In this paper, we propose to evaluate the tunnels using based on ligand trajectories computed by motion planning, namely using
+Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamics (MD) simulations.
+The tunnels are usually computed using Voronoi diagrams and therefore they shown only possible movements of a single atom.
+%Traversability of ligands with more atoms 
+%It is not easy to o estimate traversability of the tunnel for a ligand with more atoms.
+%Although the tunnels can be computed fast using e.g., Voronoi diagrams, they cannot provide information about ligand traversability, as it is 
+%not included in the VD-based computation.
+Ligand movements inside protein can be computed using MD simulations, but they are too slow for a daily routine.
+The tunnels computed for a single atom can however be used as a hint for ligand transport from or to the active site.
+The task of ligand transportation through a tunnel considering its shape and also conformation changes can be formulated
+ as a motion planning problem.
+We propose to evaluate the tunnels based on ligand trajectories computed by motion planning, namely using
 Rapidly Exploring Random Tree (RRT).
-We propose several modifications to RRT in order to compute the trajectories around the tunnel and also to consider the flexibility of the ligand.
-We also introduce several approaches for visualization of the resulting trajectories which depict valuable information about various properties and behavior of the ligand.
+We present modifications to RRT in order to compute the trajectories around a given tunnel and also to consider the flexibility of the ligand.
+Besides, methods for visualization of the trajectories and they properties are presented.
+%We also introduce several approaches for visualization of the resulting trajectories which depict valuable information about various properties and behavior of the ligand.
 The goal of the paper is to show that motion planning can provide additional insight about tunnel properties, being
@@
 The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
-To get the ligand to the active site, there has to be the transportation path connecting the protein outer environment with the active site.
+To get the ligand to the active site, there has to be a transportation path connecting the protein outer environment with the active site.
 These paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).
 
-Traditionally, tunnels can be detected using Voronoi diagrams~\cite{yaffe2008,caver3}.
+Traditionally, tunnels are detected using Voronoi diagrams~\cite{yaffe2008,caver3} assuming
 %The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3}.  %or a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
-Here the tunnels are computed assuming a spherical ligand (probe) and they are represented as a sequence of spheres.
-Biochemists decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
+ a spherical ligand (probe) and they are represented as a sequence of spheres.
+Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
 This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013} or its activity of the protein towards other molecules~\cite{Pavlova2009}.
@@
 The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
-To provide biochemists with better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also conformation changes, which can be formulated as a path planning problem in the a high-dimensional configuration space.
-
+To provide biochemists a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also conformation changes.
+%which can be formulated as a path planning problem in the a high-dimensional configuration space.
+
+In this paper, we propose a novel method for computing the trajectories of flexible ligands in a tunnel represented by spheres.
+The proposed work is based on the sampling-based Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner, which
+can deal with many-DOF objects of arbitrary geometry.
+The proposed RRT-based planner generates the random samples around a virtual sphere moving through the tunnel, which
+guides the search in the configuration space.
+To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
 The flexibility of the ligand is modeled using a predefined set of typical conformations.
@@
 Possible rotamer conformations correspond to energetically and geometrically favorable positions.
-
-In this paper, we propose a novel method for computing the trajectories of flexible ligands in a tunnel represented by spheres.
-The proposed work is based on the sampling-based Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner, which
-can deal with many-DOF objects of arbitrary geometry.
-The random samples for RRT are generated using the spheres representing the tunnel in order to guide the configuration tree through the search space.
-To make the transportation easier, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
 This solution is common also in different types of calculations, e.g.~\cite{kellogg}. 
@@
 
-Our method helps the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
+The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
+Besides trajectory computation, we also propose methods for their visualization.
+Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
 In order to predict the success of traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
-The advantage of our method is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources.
-
-We further propose novel visualization techniques for presenting the results to biochemists.
-The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
+The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources.
+%We further propose visualization techniques for presenting the results to biochemists.
 
@@
 The obvious disadvantage of the grid-based methods is the high memory demand and their dependency on the grid resolution.
-Due to the high memory consumption, these methods are not suitable for tunnel detection in dynamic proteins and therefore they
-are used primarily for analysis of static molecules or individual snapshots of molecular dynamics.
+%Due to the high memory consumption, these methods are not suitable for tunnel detection in dynamic proteins and therefore they
+%are used primarily for analysis of static molecules or individual snapshots of molecular dynamics.
 
@@
 
-The detected tunnels are evaluated using basic characteristics, like bottleneck, length, curvature, and a list of surrounding
+The tunnels detected by the above mentioned approaches are evaluated using basic characteristics like bottleneck, length, curvature, and a list of surrounding
 residues, which are later used to estimate the interaction possibilities.
-The biggest disadvantage is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore
-not easy to estimate if (and how) a ligand might traverse the tunnels.
+The biggest disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a ligand might traverse the tunnels.
 %The protein tunnels computed for spherical probes contain valuable information about the protein void space.
@@
 
-This requires to compute trajectories of the non-spherical ligand in tunnels, which can be formulated as a path planning problem in a high-dimensional configuration space $\C$.
-The configuration space is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation, and possibly also other degrees of freedom responsible for the conformation changes.
+In order to determine if a ligand can pass the tunnel, it is necessary to compute the trajectory for the ligand.
+This can be considered as a motion planning problem in a high-dimensional configuration space.
+The configuration space $\C$ is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation, and possibly also other degrees of freedom responsible for the conformation changes.
 The dimension of the configuration space is given by the degrees of freedom (DOF) of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
 ligand with $n$ DOFs.
-Sampling-based motion planning methods~\cite{Lav06} can be used to search this high-dimensional configuration space.
+Sampling-based motion planning methods can be used to search this high-dimensional configuration space~\cite{Lav06}.
 The idea of sampling-based motion planning is to randomly sample $\C$ and classify the samples as free or non-free using collision detection.
@@
 
-Sampling-based methods can cope with robots (objects) with many DOFs. 
+Sampling-based motion planning methods are suitable for computing the trajectories of the ligand as they 
+can cope with many-DOF robots (objects) of arbitrary shapes.
 The flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
@@
 The algorithm terminates if the tree approaches the goal configuration close enough.
-RRT is a suitable candidate for computing the trajectories of flexible ligands due to its ability to cope
-with many-DOF systems.
-Two main issues however need to be addressed: the increased dimension of the configuration space and the narrow passage problem.
+%RRT is a suitable candidate for computing the trajectories of flexible ligands due to its ability to cope with many-DOF systems.
 
@@
 %A well known issue of sampling-based planners is the narrow passage problem~\cite{hannaWIS}.
+Two main issues need to be addressed: the increased dimension of the configuration space and the narrow passage problem.
 A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space~\cite{hannaWIS}.
@@
 
-By reducing the scale of the ligand's atomic radii, the configuration tree can explore more areas of the configuration space.
+By reducing the scale of the atomic radii, the configuration tree can explore more areas of the configuration space.
 Such techniques have been used for motion planning of deformable objects, 

### 2017-06-30 paper.tex
alg

@@
 %Due to the interaction, the relative positions of the ligand's atoms can change.
-
 % in cortes2005path:
@@
 % several orders of magnitude. Furthermore, our results also  indicate that the geometric stage can provide highly valuable information to biologists.
-
-%The growth of the configuration tree into some regions of the configuration space can however be slowed down due to the narrow passages.
-%A narrow passage is a small region of the configuration space containing part of the solution.
-%As the samples are generated from uniform distribution in RRT, probability of placing the samples into the narrow passages is low and therefore,
-%   the probability of expanding the tree through a narrow passage is small~\cite{hannaWIS}.
-%
 %This technique is similar to generation of samples near the medial exist of the environment
@@
 %or by guiding the growth of the tree around a general path in the workspace~\cite{vonasek2009rrt,denny2014marrt}.
-
 By reducing the scale of the atomic radii, the configuration tree can explore more areas of the configuration space.
@@
 flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.
+The scaling-down technique is used also in other tools like MoMa-LigPath~\cite{cortes2005path}.
 
@@
 The flexibility of a ligand is modeled using the set $\L$ of its conformations.
-The conformations can be prepared, e.g., considering the potential energy.
-We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
+The conformations can be prepared, e.g., considering the potential energy or downloaded from a library~\cite{dunbrack}.
+%We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
 
@@
 $T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
-$c \in \R^3$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
+$c_i \in \R^3$ is their 3D position and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
 The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.
@@
 $\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
-The scaling-down  technique is used also in other tools like MoMa-LigPath~\cite{cortes2005path}.
 
@@
 using RRT~\cite{lavalleRRT}.
-The original RRT is exnteded to cope with specific requirmeents needed for traversability of ligands.
+The original RRT is exnteded to cope with specific requirments needed for the traversability of ligands.
 First, the trajectory has to be found around the given tunnel, but deviation from the tunnel centerline are allowed.
-We assume, that the distance of the ligand from the tunnel centerline is at most $\rv$.
+%The deviation of the trajectory from the tunnel centerline is influenced by the parameter $\rv$.
+%We assume, that the distance of the ligand from the tunnel centerline is at most $\rv$.
 Due to this requirement, sampling process of RRT has to be adapted in order to follow the tunnel, and to prevent
@@
 metric for the nearest-neighbor search.
+The parameter $\rv$ influences the distribution of samples around the tunnel centerline. 
+By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
+of $\rv$ cause  exploration of paths around the tunnel.
+We propose to set this parameter to the average tunnel width.
 
@@
     or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
+
+We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
 

### 2017-06-30 paper.tex
conclusion

@@
 
-\section{Discussion \& Implementation Details}
-
-Sampling-based motion planning is a powerful tool for computing trajectories for flexible ligands.
-The proposed extensions to the RRT planner have been motivated by the necessity to cope with the narrow passage
-and the high dimensionality of the flexible ligands.
-By scaling-down the geometry of ligand, which is also used in other related tools~\cite{cortes2005path}, 
-the relative volume of the free region of the configuration space increases, which increases the probability of placing
-random samples $\qrand$ into the free region.
-To further cope with the narrow passage problem, the translation part of the random samples is generated using
-the 3D centerline of the tunnel.
-The sampling along the tunnel centerline (in the distance defined by $\rv$) does not only lead to better growth of the tree in the tunnel, but it also limits the maximal deflection of the trajectory from the tunnel.
-Therefore, the bounding box of $\C$ can be defined around the tunnel in the distance $\rv$, which is typically
-smaller than the bounding box of the whole protein.
-This, together with the scaled-down ligand, increases the relative volume of the free configuration space and helps the RRT to
-reach the goal configuration.
+\section{Discussion}
+
+Trajectories computed for a ligand moving in a tunnel can reveal such properties that cannot be computed
+only based on the geometry of the tunnels.
+For example, possible detours from the tunnel can be identified, which suggest that even a tunnel with a small
+bottleneck can be traversed.
+The computed trajectories however cannot be considered as `real', as real ligands move inside proteins under influnce
+of many forces.
+
+This work assumes that the tunnels are computed on a static protein with the ligands, which brings several limitations.
+First of all, proteins are dynamic structures and consequently also the tunnels are dynamic: their move, merge or even disappear due
+to motions of protein atoms.
+Currentt works in protein engineering utilizing tunnels usually employ static tunnels only, or 
+various tunnels are detected separately in each frame of molecular dynamics and the tunnels are aggregated to a single one.
+One of the reason is the difficulty of computing dynamic tunnels with both grid-based and VD-based methods.
+
+The second limitation is caused by the utilization of narrow tunnels.
+The protein atoms tend to fill the internal void space therefore the tunnels identified inside proteins without ligand tend
+to be very narrow.
+To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary and it is used also in other related tool, e.g.~\cite{cortes2005path}.
+Trajectories computed for scaled-down ligands cannot be considered as real trajectories by still they show easy and difficult parts of
+the tunnels.
+Real ligands influence protein atoms and vice versa, which enables the ligand to pass even spaces that seem to be narrow.
+
+Due to the above reasons, the computed trajectories can be considered as too optimistic
+(e.g. because they are computed on such a tunnel that is wide, but will be closed due to molecular dynamics),
+or too pesimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can in fact be
+opened in molecular dynamics).
+
+Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
+than the simple bottleneck radius, which is used nowadays.
+To better utilize the results of motion planning, it is necessary to observe behavior of ligands in many simulations with various
+proteins.
+
+To the best knowledge of the authors, this is the first work attempt to analyze traversability of ligands in protein tunnels.
+Most related works, e.g. those based on motion planning, are usually focused on protein folding, or finding arbitrary pathways
+from proteins~\cite{cortes2010simulating,guieysse2008structure}.
+We approached
 
@@
     or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
-
 We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

### 2017-06-30 paper.tex
some corrections, now slightly over page limit, sorry!!!

--- 2017-06-30 paper.tex
@@
 
-Whereas the tunnel computation is already well established research field, the simulation of ligand transportation through the detected tunnels is  rather new field.
-Ligands are typically of non-spherical shape and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
+Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
+Ligands are typically of non-spherical shape,therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
 The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
-To provide biochemists a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also conformation changes.
+To provide biochemists with a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.
 %which can be formulated as a path planning problem in the a high-dimensional configuration space.
@@
 guides the search in the configuration space.
-To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
 The flexibility of the ligand is modeled using a predefined set of typical conformations.
@@
 Possible rotamer conformations correspond to energetically and geometrically favorable positions.
+
+To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
 This solution is common also in different types of calculations, e.g.~\cite{kellogg}. 
@@
 Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
-In order to predict the success of traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
-The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources.
+In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
+The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources required for the virtual screening.
 %We further propose visualization techniques for presenting the results to biochemists.
@@
 i.e., $A(i)=N(i) / N$, where $N=K|\QI|$ is the total number of trajectories.
-The accessibility  shows how probable is to pass the tunnel up to the sphere $i$.
+The accessibility  shows how probable it is to pass the tunnel up to the sphere $i$.
 The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
@@
 The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
-i.e, $T(i) = N(i+1) / N(i)$. 
+i.e., $T(i) = N(i+1) / N(i)$. 
 The throughput is not computed for the last sphere ($i=N$).
-The throughput shows a local accessibility of the tunnel parts.
+The throughput shows a local accessibility of tunnel parts.
 
@@
 The proposed planner is allowed to scale down the radii of ligand spheres up to the allowed scale $\smin$. 
-It can be expected, that narrower parts of the tunnel are more often passed with a more scaled-down ligand than
+It can be expected that narrower parts of the tunnel are more often passed with a more scaled-down ligand than
 the wider parts.
 Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
-The ligand scale is computed as the average value of scales of the nearest configurations of the trajectories towards tunnel sphere
-$i$.
+The ligand scale is computed as the average value of scales of the nearest configurations of the trajectories towards tunnel sphere~$i$.
 
@@
 In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $t_i$ is the closest to the point among all tunnel parts.
-Examples of collow mapping are shown in Fig.~\ref{fig:properties}.
+Examples of color mapping are shown in Fig.~\ref{fig:properties}.
 
@@
 Besides the color mapping, it is also necessary to show the computed trajectories.
-This is useful for example to find out in which part of the tunnel the trajectories detour.
-The simple visualization of all trajectories could however be too slow for an interactive work.
+This is useful, for example, to find out in which part of the tunnel the trajectories detour.
+Simple visualization of all trajectories could however be too slow for an interactive manipulation.
 Therefore, the trajectories are first clustered and then only the clusters are visualized.
 Due to different lengths of the trajectories, they are first converted to a normalized form.
-Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ is the 3D Euclidean distance
+Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
 of the most distant configuration from $\cstart$ from all trajectories.
@@
 
-The trajectories are then visualized using representative of each cluster.
+The trajectories are then visualized using a representative of each cluster.
 The number of trajectories in each cluster is represented by the width of the polyline.
@@
 (b) Trajectories clustered according to their proximity colored by distinct colors.
-The different line widths represent the number of trajectories in a cluster.
+%The different line widths represent the number of trajectories in a cluster.
 \label{fig:trajectories}

### 2017-06-30 paper.tex
text polishing

@@
 
-\def\rv{T_w}
+\def\rv{R_{tunnel}}
 
@@
 \def\smax{s_{max}}
-\def\sdelta{{\Delta}_s}
+\def\sdelta{s_{\Delta}}
 
@@
 The tunnels are usually computed using Voronoi diagrams and therefore they shown only possible movements of a single atom.
-%Traversability of ligands with more atoms 
-%It is not easy to o estimate traversability of the tunnel for a ligand with more atoms.
-%Although the tunnels can be computed fast using e.g., Voronoi diagrams, they cannot provide information about ligand traversability, as it is 
-%not included in the VD-based computation.
 Ligand movements inside protein can be computed using MD simulations, but they are too slow for a daily routine.
@@
 Besides, methods for visualization of the trajectories and they properties are presented.
-%We also introduce several approaches for visualization of the resulting trajectories which depict valuable information about various properties and behavior of the ligand.
 The goal of the paper is to show that motion planning can provide additional insight about tunnel properties, being
 less computationally demanding than the MD simulations.
-%Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
-%On the other hand, it is much less computationally demanding than methods based on MD simulations.
 The proposed approach can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
@@
 To provide biochemists with a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.
-%which can be formulated as a path planning problem in the a high-dimensional configuration space.
-
-In this paper, we propose a novel method for computing the trajectories of flexible ligands in a tunnel represented by spheres.
-The proposed work is based on the sampling-based Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner, which
-can deal with many-DOF objects of arbitrary geometry.
-The proposed RRT-based planner generates the random samples around a virtual sphere moving through the tunnel, which
+
+In this paper, we propose a novel method for computing the trajectories of flexible ligands in a tunnel.
+The proposed work is based on the sampling-based Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner.
+The main idea of the planner is to generate the random samples around a virtual sphere moving through the tunnel, which
 guides the search in the configuration space.
+This allows us to focus the sampling of the configuration space around the tunnel.
 The flexibility of the ligand is modeled using a predefined set of typical conformations.
-The conformation changes of protein amino acids and ligand are represented by so called rotamers, stored in different rotamer libraries (e.g., Dunbrack library~\cite{dunbrack}).
-Rotamers differ in the angles between their atoms.
-Possible rotamer conformations correspond to energetically and geometrically favorable positions.
-
 To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
-This solution is common also in different types of calculations, e.g.~\cite{kellogg}. 
 It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
@@
 %We further propose visualization techniques for presenting the results to biochemists.
-
-%\red{TODO}
-%something about screening: what is it, what's is the input and output, in which applications (research/industry) it is useful.
 
@@
 
-In order to determine if a ligand can pass the tunnel, it is necessary to compute the trajectory for the ligand.
-This can be considered as a motion planning problem in a high-dimensional configuration space.
+In order to determine if a ligand can pass the tunnel, it is necessary to compute a trajectory considering the shape of the ligand.
+This can be formulated as a motion planning problem in a high-dimensional configuration space.
 The configuration space $\C$ is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation, and possibly also other degrees of freedom responsible for the conformation changes.
@@
 This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
-Sampling-based methods have also been used for 
- tunnel detection~\cite{vonasek2016application,vonasek2017tunnel} and
- exit pathway computation~\cite{cortes2010simulating,guieysse2008structure}.
+%Sampling-based methods have also been used for 
+% tunnel detection~\cite{vonasek2016application,vonasek2017tunnel} and
+% exit pathway computation~\cite{cortes2010simulating,guieysse2008structure}.
 %none of the existing work focuses on the trajectory generation in tunnels.
@@
 In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
-A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$.
+A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
 If $\qnew$ is collision-free, it is added to the tree.
@@
 
+%Two main issues need to be addressed: the increased dimension of the configuration space and the narrow passage problem.
 %A well known issue of sampling-based planners is the narrow passage problem~\cite{hannaWIS}.
-Two main issues need to be addressed: the increased dimension of the configuration space and the narrow passage problem.
+
+Motion planning for flexible ligands in protein tunnels brings two main issues: 
+the ligand flexiblity increases the dimension of the configuration space, and the necessity to plan in protein tunnels
+leads to the narrow passage problem.
 A narrow passage is a region in the configuration space whose removal changes the connectivity of the free space~\cite{hannaWIS}.
@@
 in order to put enough samples there and it consequently increases the planning time.
-The protein tunnels have the properties of the narrow passages. 
-A possible approach to cope with the narrow passage property of protein tunnels is to use the information
-about their 3D position, which is represented by the tunnel centerline.
-The centerline can be used to guide the growth of the RRT tree through the configuration space~\cite{vonasek2009rrt,denny2016dynamic}.
-
-The kinematic chain representation used to model the ligand flexibility allows the sampling-based planners to generate all possible conformations.
-However, not all of them are feasible and it is therefore necessary to verify the feasibility of a given conformation based on energy, which is time consuming.
-Moreover, the additional degrees of freedom required to model the flexibility increase the dimension of the configuration space.
-RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
+Typical tunnels have bottlenecks smaller than $1.0$~\AA, so they can already be considered as narrow passages for ligands with more than two atoms.
+%The protein tunnels have the properties of the narrow passages, as they are very narrow
+To cope with the narrow passages, they have to be sampled more dense.
+For the family of RRT planners, it is usefull to change the distribution of random samples according to the growth of the tree~\cite{vonasek2009rrt,vonasekphd,denny2016dynamic}.
+%about their 3D position, which is represented by the tunnel centerline.
+%The centerline can be used to guide the growth of the RRT tree through the configuration space~\cite{vonasek2009rrt,denny2016dynamic}.
+
+%The additional degrees of freedom required to model the flexibility increase the dimension of the configuration space.
+The kinematic chain representation used to model the ligand flexibility can generate all possible conformations but it also increases the dimension of the configuration space.
+Not all of conformations are however feasible and it is therefore necessary to verify the feasibility of a given conformation based on energy, which is time consuming.
+%RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
+An alterantive solution is to employ a library of known conformations.
+The conformation changes of protein amino acids and ligand are represented by so called rotamers, stored in different rotamer libraries (e.g., Dunbrack library~\cite{dunbrack}).
+Rotamers differ in the angles between their atoms.
+Possible rotamer conformations correspond to energetically and geometrically favorable positions.
+Using a library of predefined conformations is used also in other tasks, e.g.~\cite{kellogg}.
+%This solution is common also in different types of calculations, e.g.~\cite{kellogg}.
 
@@
 b) the radii of the ligand atoms can be decreased, and 
-c) ligand flexibility is modeled using a set of predefined energy-suitable conformations.
-The first two extensions can help to deal with the narrow passage problem and the third modification
-is introduced in order to decrease the dimension of the configuration space and the sensitivity to the employed
-metric.
+c) ligand flexibility is modeled using a library of predefined conformations.
+The first two extensions can help to deal with the narrow passage problem.
+By reducing the scale of the atomic radii, the configuration tree can explore more areas of the configuration space.
+The scaling-down technique have been used e.g. for motion planning of deformable objects~\cite{gayle2005path,alterovitz2008motion}, and
+they are also used in the most related MoMa-LigPath tool~\cite{cortes2005path}.
+%is used also in other related tools like like MoMa-LigPath~\cite{cortes2005path}
+%Such techniques have been used for motion planning of deformable objects, 
+%e.g., in~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path}, 
+%e.g., in~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,gayle2005path}, 
+%and for motion planning of rigid objects among 
+%flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.
+
+%and the third modification
+%is introduced in order to decrease the dimension of the configuration space.
 
@@
 %or by guiding the growth of the tree around a general path in the workspace~\cite{vonasek2009rrt,denny2014marrt}.
-By reducing the scale of the atomic radii, the configuration tree can explore more areas of the configuration space.
-Such techniques have been used for motion planning of deformable objects, 
-e.g., in~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path}, 
-and for motion planning of rigid objects among 
-flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.
-The scaling-down technique is used also in other tools like MoMa-LigPath~\cite{cortes2005path}.
 
@@
 The flexibility of a ligand is modeled using the set $\L$ of its conformations.
-The conformations can be prepared, e.g., considering the potential energy or downloaded from a library~\cite{dunbrack}.
+The conformations $L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.
+
+%or downloaded from a library~\cite{dunbrack}.
 %We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
@@
 
-The proteins are dense structures and the tunnels are typically narrow.
-Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
-For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
+%The proteins are dense structures and the tunnels are typically narrow.
+%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
+To enable motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
 A discrete set of scales is used, i.e., $s \in \{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
@@
 If the sample $q$ is collision-free, it can be considered as a new starting configuration, so $q$ is added to $\QI$.
+Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
 For each starting configuration $\qinit \in \QI$,  $K$ trajectories are created.
@@
 using RRT~\cite{lavalleRRT}.
-The original RRT is exnteded to cope with specific requirments needed for the traversability of ligands.
-First, the trajectory has to be found around the given tunnel, but deviation from the tunnel centerline are allowed.
+Here, the original RRT is exnteded to cope with the specific requirments needed for the traversability of ligands.
+First, the trajectory has to be found around the given tunnel, but deflection from the tunnel centerline is allowed.
 %The deviation of the trajectory from the tunnel centerline is influenced by the parameter $\rv$.
@@
 construction of trajectories in the rest of the protein.
-The other modifications aim to cope with various scales of the ligand and multiple conformations,
+%The other modifications aim to cope with various scales of the ligand and multiple conformations,
 
@@
 To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
-from $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to a parameter $\rv$, and the rotational
+from $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
 part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
@@
 \KwIn{
-    tunnel spheres $c_i$ and their radii $r_i$, $i=1,\ldots,n$, 
+    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
     initial configuration $\qinit$
@@
     \eIf{$rand() < \gb$}{
-        $\qrand$ = random sample around actual virtual goal $c_n \in T$\;
+        $\qrand$ = random sample around actual virtual goal $c_v \in T$\;
     }{
@@
     $\qnear$ = nearest node in $\T$ towards $\qrand$\;
-    expand($\qnear$)\;
+    expand($\qnear,\qrand$)\;
     \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
@@
 
-The core of the proposed method is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
+The core of the proposed planner is the expansion procedure (Alg.~\ref{alg::expand}) which generates new collision-free nodes around $\qnear\in\T$.
 For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
@@
    ligand conformations $\L$,
-   scale limits $\smin, \smax$, and $\sdelta$
+   scale limits $\smin, \smax$, and $\sdelta$,
+   tree $\T$
 }
@@
 \ForEach{$l \in \L$}{
-    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
+    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
         $\qnew = \emptyset$; // empty configuration\\
@@
             \If{isCollisionFree($q$)}{
-                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qnear) < \da(\qnew,\qnear)$}{ \nllabel{alg::expand:a}
+                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
                     $\qnew = q$\;
@@
 The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
-The scale of ligand may vary in each part of the trajectory and also the conformations.
-The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.
+The path is found in the tree even if the tree does not approach $\qgoal$ close enough.
+Considering also these non-feasible solutions is necessary to evaluate difficult areas of the tunnels, e.g. bottlenecks.
+The utilization of all computed paths for evaluation of tunnel difficulty is described in the next section.
 
@@
 We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
+
+parmas:
+$\RI$
+$\rv$
+$\dt$
+
+The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.
 

### 2017-06-30 paper.tex
polishing

@@
 
-The result of each planning trial is the tree $\T$ of collision-free configurations.
-The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the nearest node in the tree towards $\qgoal$, is found in the tree.
+The result of each planning trial is the tree $\T$ of collision-free configurations in which a path
+between $\qinit$ (root of the tree) and $q'$ is found, where
+ $q'$ is the nearest node in the tree towards $\qgoal$ (using 3D Euclidean metric). 
 The path $P=(q_i), q_i \in \C$ is represented as a sequence of collision-free configurations.
@@
 The accessibility shows how probable it is to pass the tunnel up to the sphere $i$.
+Obviously, the most important is $A(n)$ of the last sphere of the tunnel, which can be considered as the overal difficulty of the tunnel.
 The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
@@
 
-To better measure the probability of passing a given tunnel part, we should take into account only trajectories that even have chance
-to visit this part, e.g., the trajectories that reached the preceding tunnel part.
+%To better measure the probability of passing a given tunnel part, we should take into account only trajectories that even have chance
+%to visit this part, e.g., the trajectories that reached the preceding tunnel part.
 The throughput $T(i)$ is the ratio of trajectories that passed sphere $i$ (i.e., visited sphere $i+1$) and reached the sphere $i$,
@@
 The throughput is not computed for the last sphere ($i=N$).
-The throughput shows a local accessibility of tunnel parts.
+The throughput shows a local accessibility of tunnel parts and it can be used to detect places where most of the trajectories ends.
 
@@
 Ligand scale $L(i)$ at the tunnel sphere $i$ is the average scale of ligand that reaches the sphere $i$.
-The ligand scale is computed as the average value of scales of the nearest configurations of the trajectories towards tunnel sphere~$i$.
+%The ligand scale is computed as the average value of scales of the nearest configurations of the trajectories towards tunnel sphere~$i$.
 
@@
 %In order to fursupport the tunnel exploration process performed by biochemists.
-The computed properties of the tunnel are presented visually by mapping the accessibility, throughput, and scale to the sphere representation
-of the tunnel using a color map.
+The above defined characteristics can be shown even as a graph, or better, presented visually by mapping then using collors to the tunnel spheres.
  Alternatively, the surface representation can be used to visualize the tunnel.
-In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $t_i$ is the closest to the point among all tunnel parts.
+In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $c_i$ is the closest to the point among all tunnel parts.
 Examples of color mapping are shown in Fig.~\ref{fig:properties}.
 
-\begin{figure}
+\begin{figure}[h]
 \centering
@@
 \end{tabular}
-\caption{Tunnel properties mapped onto its surface.
-(a) Accessibility --- green denotes parts that were accessed by the ligand easily while red denotes parts that were accessible only in \tylde 10\% of computations.
-(b) Throughput --- green denotes parts that are highly probable to be passed by the ligand while red denotes those that were hard to pass.
+\caption{Tunnel properties mapped onto its surface. The tunnel begins at the top left corner.
+(a) Accessibility --- green denotes the parts that were accessed by the ligand easily ($A(i) > 80$~\%), 
+    while red denotes parts that were accessible  in $<10$~\% of cases.
+(b) Throughput --- green denotes parts that are highly probable to be passed by the ligand ($T(i) > 80$~\%), 
+    while red denotes those that were hard to pass ($T(i) < 20$~\%).
 (c) Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
@@
 This is useful, for example, to find out in which part of the tunnel the trajectories detour.
-Simple visualization of all trajectories could however be too slow for an interactive manipulation.
+Simple visualization of all trajectories could however be too slow for an interactive work.
 Therefore, the trajectories are first clustered and then only the clusters are visualized.
@@
 where $x_i$ is the 3D position of the nearest configuration $q \in P$ to the surface of the $i$-th sphere with the radius $r'_i$.
-For the purpose of clustering, the distance between two normalized trajectories $v_i$ and $v_j$ is defined as
+The distance between two normalized trajectories $v_i$ and $v_j$ is defined as
 $d(v_i, v_j) = \frac{1}{M} \sum_{1 \leq k \leq N} |x_k^i x_k^j|$.
-The clustering is then computed using the UPGMA algorithm~\cite{sokal1958statistical}.
+This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
 
@@
 \begin{tabular}{cc}
-\includegraphics[width=0.4\textwidth]{fig/trajectories-all}
+\includegraphics[width=0.4\textwidth]{fig/trajectories-all} &
 \includegraphics[width=0.4\textwidth]{fig/trajectories-clustered-21} \\
@@
 \end{tabular}
-\caption{Visualization of resulting trajectories.
-(a) All trajectories colored according to whether they reached the end of the tunnel (green) or not (red).
+\caption{Visualization of the resulting trajectories. The tunnel begins in the top left corner.
+(a) All trajectories ($\sim5000$) colored according to whether they reached the end of the tunnel (green) or not (red). 
 (b) Trajectories clustered according to their proximity colored by distinct colors.
@@
 The proposed approach has been used to analyze tunnels in Haloalkane dehalogenase protein (PDB ID 1CQW) (Fig.~\ref{fig::tunnel}a).
-The tunnel was computed using CAVER 3.0~\cite{caver3} for spherical probe of radius 0.9~\AA.
+The tunnel was computed using CAVER 3.0~\cite{caver3} for the spherical probe of radius 0.9~\AA.
 The tunnel is represented by 760 spheres with average distance $0.05$~\AA.
@@
 $\LA$ was represented by 12 conformations, and $\LB$ by a set of $114$ conformations.
-Examples of three conformations are depicted in Fig.~\ref{fig::tunnel}b.
+Examples of three conformations of $\LA$ and $\LB$ are depicted in Fig.~\ref{fig::tunnel}b.
 
@@
 No trajectories were found for $\smin > 0.6$.
-For each scale, 50 collision-free initial configurations were found in the $2$\AA\ radius from the first sphere of the tunnel.
+For each scale, 50 collision-free initial configurations were found in the $\RI=2$\AA\ radius around the first sphere of the tunnel.
 For each ligand, each minimal scale $\smin$ and each initial configuration, 100 trajectories were computed using the proposed planner.
-The parameters of the planner were: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA.
+The parameters of the planner were: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
 For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance $\rv$ or less) over 100 trials.
@@
 examined in each expansion.
+
+%The tree can be expanded by up to $|\L|$ conformations in each iterations, which gives theoretical
+%size of the tree $110\cdot 10^3$ for $\LA$ and $1140\cdot 10^3$ for $\LB$.
+%The constructed tree are however significantly smaller
 

### 2017-06-30 paper.tex
images from all tunnels, some text

@@
 \begin{tabular}{cc}
-\includegraphics[width=0.38\textwidth]{fig/protein-tunnel-gray} &
+%\includegraphics[width=0.38\textwidth]{fig/protein-tunnel-gray} &
+\includegraphics[width=0.38\textwidth]{fig/t05proteintunnels} &
 \includegraphics[width=0.5\textwidth]{fig/ligAll}  \\
@@
 \caption{\label{fig::tunnel}
-    (a) The protein 1CQW visualized using the cartoon representation (gray) with its highest ranked tunnel (green) which was detected using CAVER 3.0.
+    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
+        three tunnels (red, green, blue) detected by CAVER 3.0.
+%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
     (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
@@
 Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.
+
+\begin{figure}
+\centering
+{
+\renewcommand{\tabcolsep}{-5pt}
+\begin{tabular}{ccc}
+\includegraphics[width=0.38\textwidth]{fig/t05proteinBottle2T} & 
+\includegraphics[width=0.38\textwidth]{fig/t04goodT} & 
+\includegraphics[width=0.38\textwidth]{fig/t05goodT} \\
+\includegraphics[width=0.38\textwidth]{fig/t05thpT} &
+\includegraphics[width=0.38\textwidth]{fig/t04badT} &
+\includegraphics[width=0.38\textwidth]{fig/t05badT} \\ 
+(a) & (b) & (c)                       
+\end{tabular}                       
+\caption{\label{fig::tunnel}
+    (a) Classic bottleneck for spherical probe (top) and visualization of throughput (bottom) for the ligand $\LA$.
+    (b) Successfull (green, top) trajectories that reached end points of the tunnels,
+        and unsuccessful ones (red, bottom) for $\smin=0.4$.
+    (c) Successfull and unsuccessful trajectories for $\smin=0.5$.
+}
+}
+\end{figure}
 

### 2017-06-30 paper.tex
images flipped, added detailed images + text

@@
 The proposed approach has been used to analyze tunnels in Haloalkane dehalogenase protein (PDB ID 1CQW) (Fig.~\ref{fig::tunnel}a).
-The tunnel was computed using CAVER 3.0~\cite{caver3} for the spherical probe of radius 0.9~\AA.
-The tunnel is represented by 760 spheres with average distance $0.05$~\AA.
+Three tunnels were detected using CAVER 3.0~\cite{caver3} for the spherical probe of radius 0.9~\AA.
+The traversability of the first tunnel (represented by $n=76$ spheres), was evaluated for 1-chlorpropan (denoted as $\LA$) with 11 atoms and 
 % from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
-The traversability of the tunnel was evaluated for 1-chlorpropan (denoted as $\LA$) with 11 atoms and 
 and 1-chlorbutan (denoted as $\LB$) with 14 atoms.
@@
 %\includegraphics[width=0.38\textwidth]{fig/protein-tunnel-gray} &
-\includegraphics[width=0.38\textwidth]{fig/t05proteintunnels} &
-\includegraphics[width=0.5\textwidth]{fig/ligAll}  \\
+\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
+\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
 (a) & (b)                        
@@
         three tunnels (red, green, blue) detected by CAVER 3.0.
+        The first tunnel is depicted red.
 %        its highest ranked tunnel (green) which was detected using CAVER 3.0.
@@
 
-\begin{figure}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[width=0.32\textwidth]{fig/tunne4ac}& 
-\includegraphics[width=0.32\textwidth]{fig/tunne4bc} & 
-\includegraphics[width=0.32\textwidth]{fig/tunne4cc} \\
-(a) & (b) & (c)                        
-\end{tabular}
-\caption{\label{fig::t4}
-    todo
-}
-\end{figure}
-
-Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.
+Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
+to distance $\dt=2$~\AA) and they are considered unsuccessful otherwise.
+The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
+also in other places.
+The comparison of classic bottleneck and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
+The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
+The successful trajectories for $\smin=0.4$ reveals that the end of the tunnel No. 3 can be approached by two different pathways.
+The detail of this pathway is depicted in Fig.~\ref{fig::detail}.
+Despite the low bottleneck of this tunnel, the ligand may reach its end using an alternative path way. 
+This shows the strenght of the analysis based on the trajectories and not only on the geometric bottlenecks.
 
@@
 \end{tabular}                       
-\caption{\label{fig::tunnel}
+\caption{\label{fig::tunnel2}
     (a) Classic bottleneck for spherical probe (top) and visualization of throughput (bottom) for the ligand $\LA$.
@@
 
+\begin{figure}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[width=0.32\textwidth]{fig/tunne4ac}& 
+\includegraphics[width=0.32\textwidth]{fig/tunne4bc} & 
+\includegraphics[width=0.32\textwidth]{fig/tunne4cc} \\
+(a) & (b) & (c)                        
+\end{tabular}
+\caption{\label{fig::detail}
+    Detailed view to alternative pathway around the third tunnel in 1CQW.
+    The successful trajectories are in green (a) the unsuccessful in red (b).
+    (c) Shows visualization of protein atoms around the tunnel.
+}
+\end{figure}
+
 \section{Discussion}
@@
 For example, possible detours from the tunnel can be identified, which suggest that even a tunnel with a small
-bottleneck can be traversed.
+bottleneck can be traversed (Fig.~\ref{fig::detail}).
+
+The achieved results have to be
+
 The computed trajectories however cannot be considered as `real', as real ligands move inside proteins under influnce
@@
 The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.
+
+Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.
 

### 2017-06-30 paper.tex
conslusion, spelling, images

@@
 \usepackage[bookmarks=false]{hyperref}
+\usepackage{setspace}
 \pdfminorversion=4
@@
 The proposed work is based on the sampling-based Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner.
-The main idea of the planner is to generate the random samples around a virtual sphere moving through the tunnel, which
+The main idea of the proposed planner is to generate the random samples around a virtual sphere moving through the tunnel, which
 guides the search in the configuration space.
@@
 It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
-This strategy helps to overcome the limitations of the conformational discretation introduced by the rotamer approximation.
+%This strategy helps to overcome the limitations of the conformational discretation introduced by the rotamer approximation.
 
 The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
-Besides trajectory computation, we also propose methods for their visualization.
+%Besides trajectory computation, visualization of the results in important. %we also propose methods for their visualization.
 Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
+Therefore, methods for visualization of the results  are discussed.
 In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
@@
 can change~\cite{songPFpath}.
-This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
+This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid} or analysis of loop motions~\cite{cortes2004geometric}.
+%This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
 %Sampling-based methods have also been used for 
@@
 Motion planning for flexible ligands in protein tunnels brings two main issues: 
-the ligand flexiblity increases the dimension of the configuration space, and the necessity to plan in protein tunnels
+the ligand flexibility increases the dimension of the configuration space, and the necessity to plan in protein tunnels
 leads to the narrow passage problem.
@@
 To cope with the narrow passages, they have to be sampled more dense.
-For the family of RRT planners, it is usefull to change the distribution of random samples according to the growth of the tree~\cite{vonasek2009rrt,vonasekphd,denny2016dynamic}.
+For the family of RRT planners, it is useful to change the distribution of random samples according to the growth of the tree~\cite{vonasek2009rrt,vonasekphd,denny2016dynamic}.
 %about their 3D position, which is represented by the tunnel centerline.
@@
 %RRT-based planners are sensitive to the employed metric, that should consider also the flexibility-related DOFs.
-An alterantive solution is to employ a library of known conformations.
+An alternative solution is to employ a library of known conformations.
 The conformation changes of protein amino acids and ligand are represented by so called rotamers, stored in different rotamer libraries (e.g., Dunbrack library~\cite{dunbrack}).
@@
 requires to generate a set $\QI$ of collision-free initial configurations.
-In this paper we assume that the ligand has to travel from the begining to the end of the tunnel.
+In this paper we assume that the ligand has to travel from the beginning to the end of the tunnel.
 As the tunnel are computed for a spherical probe, the ligand may not fit exactly to the first sphere of the tunnel.
@@
 
-The task of the trajectory computation is to find a trajectory for the given ligand in the given tunnel, which can be solved
-using RRT~\cite{lavalleRRT}.
-Here, the original RRT is exnteded to cope with the specific requirments needed for the traversability of ligands.
+The task of the trajectory computation is to find a trajectory for the given ligand in the given tunnel. 
+% which can be solved using RRT~\cite{lavalleRRT}.
+Here, the original RRT is extended to cope with the specific requirements needed for the traversability of ligands.
 First, the trajectory has to be found around the given tunnel, but deflection from the tunnel centerline is allowed.
@@
 \begin{algorithm}[h]
-%\setstretch{\straa}
+{\small
+\setstretch{0.88}
 \caption{\label{alg::main}Main loop of the RRT planner}
@@
 \return $\T$\;
+}
 \end{algorithm}
@@
 By computing the distance between the nearest atoms, the shape of the ligand is actually considered.
-As the expansion routine attemts to find such sample that minimizes this distance to $\qrand$, it supports retraction
+As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction
 of the ligand towards $\qrand$.
@@
 \begin{algorithm}[h]
+{\small
+\setstretch{0.88}
 %\setstretch{\straa}
@@
 }
+}
 \end{algorithm}
@@
 The accessibility shows how probable it is to pass the tunnel up to the sphere $i$.
-Obviously, the most important is $A(n)$ of the last sphere of the tunnel, which can be considered as the overal difficulty of the tunnel.
+Obviously, the most important is $A(n)$ of the last sphere of the tunnel, which can be considered as the overall difficulty of the tunnel.
 The  ligand passage may however be strongly affected by the first bottleneck, so the parts of the tunnel located behind
@@
 %In order to fursupport the tunnel exploration process performed by biochemists.
-The above defined characteristics can be shown even as a graph, or better, presented visually by mapping then using collors to the tunnel spheres.
+The above defined characteristics can be shown even as a graph, or better, presented visually by mapping then using colors to the tunnel spheres.
  Alternatively, the surface representation can be used to visualize the tunnel.
@@
 \caption{\label{tab::rrt}
-    Runtime and success ratio of the planner
+    Runtime and success ratio for ligand $\LA$ (left) and $\LB$ (right).
 }
 {
+\small
 \renewcommand{\tabcolsep}{1pt}
 \input table.003.tex
-\hskip 5pt   
+\hskip 4pt   
 \input table.004.tex
@@
 Despite the low bottleneck of this tunnel, the ligand may reach its end using an alternative path way. 
-This shows the strenght of the analysis based on the trajectories and not only on the geometric bottlenecks.
+This shows the strength of the analysis based on the trajectories and not only on the geometric bottlenecks.
 
@@
     (a) Classic bottleneck for spherical probe (top) and visualization of throughput (bottom) for the ligand $\LA$.
-    (b) Successfull (green, top) trajectories that reached end points of the tunnels,
+    (b) Successful (green, top) trajectories that reached end points of the tunnels,
         and unsuccessful ones (red, bottom) for $\smin=0.4$.
-    (c) Successfull and unsuccessful trajectories for $\smin=0.5$.
+    (c) Successful and unsuccessful trajectories for $\smin=0.5$.
 }
@@
 
-Trajectories computed for a ligand moving in a tunnel can reveal such properties that cannot be computed
-only based on the geometry of the tunnels.
+%Trajectories computed for a ligand moving in a tunnel can reveal such properties that cannot be computed
+%only based on the geometry of the tunnels.
+
+The computed trajectories cannot be considered as `real', they should rather be considered as a hint for the biochemists.
+One of the reason is that the trajectories are computed inside a static protein and static tunnels.
+Real proteins are dynamic structures and consequently also the tunnels are dynamic: their move, merge or even disappear due to motions of protein atoms.
+Contrary, usually static tunnels are used in protein engineering, or 
+multiple tunnels are detected separately in each frame of molecular dynamics and the tunnels are then aggregated to a single one.
+Despite this simplification, researches claim that static tunnels help in their research.
+
+The second limitation is caused by the utilization of scaled-down ligands.
+The protein atoms tend to fill the internal void space therefore the tunnels identified inside proteins without ligand tend
+to be very narrow.
+To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary.
+It is used also in other related tool, e.g.~\cite{cortes2005path}.
+By allowing more aggressive scaling-down, more easier a trajectory can be found.
+The tunnels therefore should be always evaluated considering various minimal scales.
+
+Due to the above reasons, the computed trajectories can be considered as too optimistic
+(e.g. because they are computed on such a tunnel that is wide, but will be closed due to molecular dynamics),
+or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can in fact be
+opened in molecular dynamics).
+
+Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
+than the simple bottleneck radius, which is used nowadays.
 For example, possible detours from the tunnel can be identified, which suggest that even a tunnel with a small
@@
 
-The achieved results have to be
-
-The computed trajectories however cannot be considered as `real', as real ligands move inside proteins under influnce
-of many forces.
-
-This work assumes that the tunnels are computed on a static protein with the ligands, which brings several limitations.
-First of all, proteins are dynamic structures and consequently also the tunnels are dynamic: their move, merge or even disappear due
-to motions of protein atoms.
-Currentt works in protein engineering utilizing tunnels usually employ static tunnels only, or 
-various tunnels are detected separately in each frame of molecular dynamics and the tunnels are aggregated to a single one.
-One of the reason is the difficulty of computing dynamic tunnels with both grid-based and VD-based methods.
-
-The second limitation is caused by the utilization of narrow tunnels.
-The protein atoms tend to fill the internal void space therefore the tunnels identified inside proteins without ligand tend
-to be very narrow.
-To enable at least some motion of the ligands inside such narrow tunnels, the scaling-down of the ligand atoms is necessary and it is used also in other related tool, e.g.~\cite{cortes2005path}.
-Trajectories computed for scaled-down ligands cannot be considered as real trajectories by still they show easy and difficult parts of
-the tunnels.
-Real ligands influence protein atoms and vice versa, which enables the ligand to pass even spaces that seem to be narrow.
-
-Due to the above reasons, the computed trajectories can be considered as too optimistic
-(e.g. because they are computed on such a tunnel that is wide, but will be closed due to molecular dynamics),
-or too pesimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can in fact be
-opened in molecular dynamics).
-
-Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
-than the simple bottleneck radius, which is used nowadays.
-To better utilize the results of motion planning, it is necessary to observe behavior of ligands in many simulations with various
-proteins.
-
-To the best knowledge of the authors, this is the first work attempt to analyze traversability of ligands in protein tunnels.
-Most related works, e.g. those based on motion planning, are usually focused on protein folding, or finding arbitrary pathways
-from proteins~\cite{cortes2010simulating,guieysse2008structure}.
-We approached
-
-The runtime of the planner is currently given mainly by the collision-detection.
-As the ligand is assumed to travel only along the tunnel centerline (in the distance defined by $\rv$),
-only atoms located in the vicinity of the tunnel can be considered for collision detection.
-Fast collision-detection can be achieved either using fast hierarchical data structures as OBB (implemented, e.g., in~\cite{ozcollide})
-    or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
-We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
-
-parmas:
-$\RI$
-$\rv$
-$\dt$
-
-The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.
+%To better utilize the results of motion planning, it is necessary to observe behavior of ligands in many simulations with various proteins.
+%To the best knowledge of the authors, this is the first work attempt to analyze traversability of ligands in protein tunnels.
+%Most related works, e.g. those based on motion planning, are usually focused on protein folding, or finding arbitrary pathways
+%from proteins~\cite{cortes2010simulating,guieysse2008structure}.
+
+The proposed planner is very simple and it can be improved in many ways: by using a less aggressive scaling-down technique (in order
+to obtain better trajectories) or by using only a subset of conformations and a better expansion (in order to speed it up).
+Having a better planner (in the term of speed or even success rate) however does not mean that the resulting trajectories will help the biochemists more.
+
+For a daily and practical utilization of here proposed idea, it is 
+necessary to compare trajectories from existing MD simulations with the trajectories obtain using motion planning and find
+suitable parameters and thresholds (e.g. about the throughput) that will be used for making decisions.
 
@@
 
+%The runtime of the planner is currently given mainly by the collision-detection.
+%As the ligand is assumed to travel only along the tunnel centerline (in the distance defined by $\rv$),
+%only atoms located in the vicinity of the tunnel can be considered for collision detection.
+%Fast collision-detection can be achieved either using fast hierarchical data structures as OBB (implemented, e.g., in~\cite{ozcollide})
+%    or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
+%We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
+
+%parmas:
+%$\RI$
+%$\rv$
+%$\dt$
+%The presented expansion procedure favors the samples with maximal scale (they are tested first), the trajectory contains samples with higher scales at the easy part of the tunnels (e.i., wide areas), and the smaller scales are used only if the ligand has to fit into a narrow space.
+
 \section{Conclusion }
-In the future, we will focus on the following extensions proposed by the domain experts.
-Only ligand flexibility is included in the current algorithm but the flexibility of amino acid side-chains would greatly improve the prediction. 
-Moreover, incorporation of some scoring function for energy evaluation would introduce the characterization of not only geometrical but also physico-chemical properties and would further improve the prediction ability of our approach.
+%In the future, we will focus on the following extensions proposed by the domain experts.
+%Only ligand flexibility is included in the current algorithm but the flexibility of amino acid side-chains would greatly improve the prediction. 
+%Moreover, incorporation of some scoring function for energy evaluation would introduce the characterization of not only geometrical but also physico-chemical properties and would further improve the prediction ability of our approach.
+
+Protein tunnels are transporting pathways for ligands towards active sites and nowadays, they traversability is evaluated
+only considering one-atom ligands (probes).
+In this paper, we propose to analyze traversability in tunnels using motion planning, that can consider both shape
+of the protein and ligand.
+We proposed modifications to RRT method to compute trajectories for a ligand represented by a library of common conformations.
+The corresponding configuration space is sampled using guided sampling, where a virtual sphere moves along the tunnel and attracts growth
+of the tree towards it.
+The enable trajectory computation in proteins, ligands atomic radii are scaled down.
+Besides classic property like bottleneck radius, the computed trajectories allows us to evaluate traversability of tunnels using
+accessibility, throughput and also by visualizing typical pathways found by the planner.
 

### 2017-06-30 paper.tex
conclusion, visualization

@@
 
-The accessibility $A(i)$ of the sphere $i$ is the probability of reaching the sphere by the computed trajectories,
-i.e., $A(i)=N(i) / N$, where $N=K|\QI|$ is the total number of trajectories.
+The accessibility $A(i)=N(i)/N$ of the sphere $i$ is the probability of reaching the sphere $i$, where $N=K|\QI|$ is the total number of trajectories.
 The accessibility shows how probable it is to pass the tunnel up to the sphere $i$.
@@
 %In order to fursupport the tunnel exploration process performed by biochemists.
-The above defined characteristics can be shown even as a graph, or better, presented visually by mapping then using colors to the tunnel spheres.
+The above defined characteristics can be shown even as a graph, or better, presented visually by mapping them using colors to the tunnel spheres.
  Alternatively, the surface representation can be used to visualize the tunnel.
 In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $c_i$ is the closest to the point among all tunnel parts.
-Examples of color mapping are shown in Fig.~\ref{fig:properties}.
+
+The examples of the color mapping are shown in Fig.~\ref{fig:properties}.
+The accessibility (Fig.~\ref{fig:properties}a) shows that more than the half of the tunnel is not accessible (red part of the tunnel).
+The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversible.
 
@@
 Besides the color mapping, it is also necessary to show the computed trajectories.
-This is useful, for example, to find out in which part of the tunnel the trajectories detour.
+%This is useful, for example, to find out in which part of the tunnel the trajectories detour.
 Simple visualization of all trajectories could however be too slow for an interactive work.
@@
 Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
-of the most distant configuration from $\cstart$ from all trajectories.
-A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d \over M}$, where $i=0,\ldots, M-1$.
+of the most distant configuration from $\cstart$ amongst all trajectories.
+A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
 The trajectory $P=(q_1,\ldots,q_n)$ of length $n$ is represented by the normalized vector $v=(x_1,\ldots,x_M)$ of length $M$,
@@
 This distance is used in the UPGMA clustering technique~\cite{sokal1958statistical}.
-
-The trajectories are then visualized using a representative of each cluster.
-The number of trajectories in each cluster is represented by the width of the polyline.
-In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
+The trajectories can be visualized using a representative of each cluster.
+The number of trajectories in each cluster is represented by the width of the polyline (Fig.~\ref{fig:trajectories}).
+%In this manner, we are able to convey the information about all different trajectories together (see Fig.~\ref{fig:trajectories} right).
 
@@
 \end{tabular}
-\caption{Visualization of the resulting trajectories. The tunnel begins in the top left corner.
+\caption{Visualization of the trajectories. The tunnel begins in the top left corner.
 (a) All trajectories ($\sim5000$) colored according to whether they reached the end of the tunnel (green) or not (red). 
-(b) Trajectories clustered according to their proximity colored by distinct colors.
+(b) Visualization using clusters of trajectories. 
+%Trajectories clustered according to their proximity colored by distinct colors.
 %The different line widths represent the number of trajectories in a cluster.
@@
 
-The proposed approach has been used to analyze tunnels in Haloalkane dehalogenase protein (PDB ID 1CQW) (Fig.~\ref{fig::tunnel}a).
-Three tunnels were detected using CAVER 3.0~\cite{caver3} for the spherical probe of radius 0.9~\AA.
-The traversability of the first tunnel (represented by $n=76$ spheres), was evaluated for 1-chlorpropan (denoted as $\LA$) with 11 atoms and 
+The tunnels in Haloalkane dehalogenase protein (PDB ID 1CQW) has been analyzed using the proposed approach.
+Three tunnels were detected using CAVER 3.0~\cite{caver3} for the spherical probe of radius 0.9~\AA (Fig.~\ref{fig::tunnel}a).
+The traversability was evaluated for 1-chlorpropan (denoted as $\LA$) with 11 atoms and 
 % from David: m003 je "1-chlorpropan" a m004 "1-chlorbutan"
@@
 Examples of three conformations of $\LA$ and $\LB$ are depicted in Fig.~\ref{fig::tunnel}b.
-
-Both tested ligands have more than 10 atoms and therefore they cannot fit into the tunnel of with 0.9~\AA.
-To find collision-free trajectories, the radii of ligands have to be scaled down. 
+Further information about the experiments can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.
+
+Both tested ligands have more than 10 atoms and therefore they cannot fit into the tunnel of with 0.9~\AA, so
+the radii of ligands were scaled down. 
 Four different scaling-down factors were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
@@
 The parameters of the planner were: $\Imax=10,000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA, $\rv=2$~\AA.
-For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance $\rv$ or less) over 100 trials.
-
-The average runtimes and sizes of the built trees are shown in Tab.~\ref{tab::rrt}.
-The tables also show the minimal, maximal success rate, and the average success rate.
-The average success rate is computed as the average value of the success rates achieved for all 50 initial configurations.
-
-The highest average success rate is achieved for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
-The minimal and maximal success rates indicate how the initial position of the ligand influence the ability to pass the tunnels.
-The ligand $\LB$, which was represented by more conformations than $\LA$, has higher minimal and also average success rate, which
-indicates that the flexibility is important for the tunnel traversability.
+For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance 
+ $\rv=2$~\AA or less) over 100 trials.
+
+The average runtimes, success rates and sizes of the built trees for the first tunnel are shown in Tab.~\ref{tab::rrt}.
+The highest average success rate wasv achieved for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
+The difference between the minimal and maximal success rates indicates how the initial configurations of the ligand influence the ability to pass the tunnels.
+For example, the most difficult initial configuration of $\LA$ and $\smin=0.3$ lead to success rate 78~\% (the easiest 
+        initial configuration lead to success rate $100$~\%), but for the scale $\smin=0.4$, the worst success rate is
+only $42$~\% and the best 96~\%.
+The traversability of larger ligands ($\smin=0.4$ vs. $\smin=0.3$) is therefore more sensitive to the initial configuration.
+This shows the importance of testing the traversability from multiple initial configurations.
 The runtime is significantly higher for the $\LB$ ligand, which is caused by the larger number of conformations that need to be
@@
 Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
-to distance $\dt=2$~\AA) and they are considered unsuccessful otherwise.
+to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
 The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
 also in other places.
-The comparison of classic bottleneck and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
+The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
 The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
@@
 The detail of this pathway is depicted in Fig.~\ref{fig::detail}.
-Despite the low bottleneck of this tunnel, the ligand may reach its end using an alternative path way. 
+Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative path way. 
 This shows the strength of the analysis based on the trajectories and not only on the geometric bottlenecks.
@@
 One of the reason is that the trajectories are computed inside a static protein and static tunnels.
-Real proteins are dynamic structures and consequently also the tunnels are dynamic: their move, merge or even disappear due to motions of protein atoms.
-Contrary, usually static tunnels are used in protein engineering, or 
-multiple tunnels are detected separately in each frame of molecular dynamics and the tunnels are then aggregated to a single one.
-Despite this simplification, researches claim that static tunnels help in their research.
+Real proteins are dynamic structures and consequently also the tunnels are dynamic: they move, merge or even disappear due to motions of protein atoms.
+Contrary, usually static tunnels are used in protein engineering.
+%multiple tunnels are detected separately in each frame of molecular dynamics and the tunnels are then aggregated to a single one.
+Despite such a simplification, researches often used information about the static tunnels  to estimate the possibility of chemical reactions
+at the active sites.
 
@@
 It is used also in other related tool, e.g.~\cite{cortes2005path}.
-By allowing more aggressive scaling-down, more easier a trajectory can be found.
-The tunnels therefore should be always evaluated considering various minimal scales.
-
-Due to the above reasons, the computed trajectories can be considered as too optimistic
-(e.g. because they are computed on such a tunnel that is wide, but will be closed due to molecular dynamics),
-or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can in fact be
-opened in molecular dynamics).
-
+%By allowing more aggressive scaling-down, more easier a trajectory can be found.
+%The tunnels therefore should be always evaluated considering various minimal scales.
+
+Due to these reasons, the computed trajectories can be considered either as too optimistic
+(e.g. because they are computed on a wide tunnel that can be in fact closed due to molecular dynamics),
+or too pessimistic (e.g. the trajectory is not found because of narrow passage around the tunnel, which can be 
+opened due to the molecular dynamics).
 Despite these limitations, testing the tunnel traversability using motion planning technique can provide chemists more information
@@
 to obtain better trajectories) or by using only a subset of conformations and a better expansion (in order to speed it up).
-Having a better planner (in the term of speed or even success rate) however does not mean that the resulting trajectories will help the biochemists more.
-
-For a daily and practical utilization of here proposed idea, it is 
-necessary to compare trajectories from existing MD simulations with the trajectories obtain using motion planning and find
-suitable parameters and thresholds (e.g. about the throughput) that will be used for making decisions.
-
-Further information can be found at {\url{http://mrs.felk.cvut.cz/isrr2017}}.
+Having a better motion planner (in the term of the speed or even success rate) however does not automatically mean that the resulting trajectories will help the biochemists more.
+The presented work shows possible advantages of the analysis based on ligand trajectories.
+By comparing the planned trajectories with trajectories observed in MD simulations, thresholds for the success rate or minimal tunnel throughput can be set, which will help biochemists to make decisions.
 
@@
 
-Protein tunnels are transporting pathways for ligands towards active sites and nowadays, they traversability is evaluated
-only considering one-atom ligands (probes).
-In this paper, we propose to analyze traversability in tunnels using motion planning, that can consider both shape
+Protein tunnels are transporting pathways for ligands towards active sites.
+In this paper, we propose to analyze the traversability in the tunnels using motion planning, that can consider both shape
 of the protein and ligand.
-We proposed modifications to RRT method to compute trajectories for a ligand represented by a library of common conformations.
-The corresponding configuration space is sampled using guided sampling, where a virtual sphere moves along the tunnel and attracts growth
+We proposed modifications to the RRT method to compute trajectories for a ligand represented by a library of common conformations.
+The corresponding configuration space is sampled using the guided sampling, where a virtual sphere moves along the tunnel and attracts growth
 of the tree towards it.
-The enable trajectory computation in proteins, ligands atomic radii are scaled down.
-Besides classic property like bottleneck radius, the computed trajectories allows us to evaluate traversability of tunnels using
+To enable trajectory computation in narrow tunnels, atomic radii of the ligand are scaled down.
+Computed trajectories allows us to evaluate the traversability of the tunnels using
 accessibility, throughput and also by visualizing typical pathways found by the planner.
+These properties can help to better understand importance of the tunnels.
+
+%which can be useful in the virtual screening
 

### 2017-06-30 paper.tex
overal text, spelling

@@
 Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamics (MD) simulations.
-The tunnels are usually computed using Voronoi diagrams and therefore they shown only possible movements of a single atom.
+The tunnels are usually computed using Voronoi diagrams and therefore they can show possible movements only for a single atom.
 Ligand movements inside protein can be computed using MD simulations, but they are too slow for a daily routine.
@@
 We present modifications to RRT in order to compute the trajectories around a given tunnel and also to consider the flexibility of the ligand.
-Besides, methods for visualization of the trajectories and they properties are presented.
+Besides, methods for visualization of the trajectories and their properties are presented.
 The goal of the paper is to show that motion planning can provide additional insight about tunnel properties, being
@@
 Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
-Ligands are typically of non-spherical shape,therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
+Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
 The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
-To provide biochemists with a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.
-
-In this paper, we propose a novel method for computing the trajectories of flexible ligands in a tunnel.
-The proposed work is based on the sampling-based Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner.
+To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.
+
+We propose to analyze the traversability of the ligands in the tunnels using motion planning.
+The principle of Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner is used and further modified.
 The main idea of the proposed planner is to generate the random samples around a virtual sphere moving through the tunnel, which
@@
 Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
-Therefore, methods for visualization of the results  are discussed.
 In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
@@
 %We further propose visualization techniques for presenting the results to biochemists.
+%Therefore, methods for visualization of the results  are discussed.
 
@@
 
-Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD).
+Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD). %~\cite{caver3,yaffe2008}.
 The ordinary VD is computed on points representing centers of all atoms, without considering the radii of atoms.
@@
 The conformation changes of protein amino acids and ligand are represented by so called rotamers, stored in different rotamer libraries (e.g., Dunbrack library~\cite{dunbrack}).
-Rotamers differ in the angles between their atoms.
+%Rotamers differ in the angles between their atoms.
 Possible rotamer conformations correspond to energetically and geometrically favorable positions.
-Using a library of predefined conformations is used also in other tasks, e.g.~\cite{kellogg}.
 %This solution is common also in different types of calculations, e.g.~\cite{kellogg}.
@@
 These characteristics are already used in many research studies.
-Selected promising tunnels can be further analyzed for a given ligand to determine the traversability.
-It is not necessary to first run time consuming MD simulations with both protein and ligand being studied, but it is suitable
-to detected tunnels in the protein only.
+Selected tunnels can further be analyzed with a given ligand to determine the traversability.
+According to the found trajectories, researches may decided of the given ligand can reach the active site, which helps to 
+organize subsequent MD simulations or other experiments.
+
+%It is not necessary to first run time consuming MD simulations with both protein and ligand being studied.
+%but it is suitable to detected tunnels in the protein only.
 
@@
 c) ligand flexibility is modeled using a library of predefined conformations.
-The first two extensions can help to deal with the narrow passage problem.
-By reducing the scale of the atomic radii, the configuration tree can explore more areas of the configuration space.
+The first two extensions are designed to cope with the narrow passage problem.
+By reducing the scale of the atomic radii, the ligand can move more without collisions, which widens the narrow passages and increases
+the probability of placing samples into them.
 The scaling-down technique have been used e.g. for motion planning of deformable objects~\cite{gayle2005path,alterovitz2008motion}, and
 they are also used in the most related MoMa-LigPath tool~\cite{cortes2005path}.
+By using the predefined set of conformations, the number of dimensions of the configuration space does not depend on the DOF of the ligand.
+Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
+%Using a library of predefined conformations has been used also in other tasks, e.g.~\cite{kellogg}.
+
 %is used also in other related tools like like MoMa-LigPath~\cite{cortes2005path}
@@
 
-A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
+A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
 by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., its geometric center), rotation around $x$, $y$, and $z$ axes,
@@
 In this paper we assume that the ligand has to travel from the beginning to the end of the tunnel.
-As the tunnel are computed for a spherical probe, the ligand may not fit exactly to the first sphere of the tunnel.
+As the tunnels are computed for a spherical probe, the ligand may not fit exactly to the first sphere of the tunnel.
 The initial configurations have to be searched around the first sphere of the tunnel.
@@
 
-The task of the trajectory computation is to find a trajectory for the given ligand in the given tunnel. 
+The task of the trajectory computation is to find a trajectory for a ligand in the given tunnel. 
 % which can be solved using RRT~\cite{lavalleRRT}.
@@
 In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
-The nearest-neighbor search between the $\qrand$ and the tree is performed using the weighted 6D Euclidean metric considering
+The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric considering
 both 3D rotation and 3D translation.
@@
 To guide the growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
-The virtual goal $v, 1\le v \le n$, is the index of sphere of the tunnel.
-The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated around the sphere $c_v \in T$.
+The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
+The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
+%with probability $1-\gb$ from $\C$, otherwise they are generated around the sphere $c_v \in T$.
 After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
@@
 case of the non-spherical ligands.
-To consider the actual shape of the ligand (which is different in each conformation) and to allow finding such configurations
-that approach $\qrand$, the distance between newly generated configurations and $\qrand$ is measured as the smallest
+To consider the actual shape of the ligand (which is different in each conformation) and to support finding such configurations
+that maximally approach $\qrand$, the distance between newly generated configurations and $\qrand$ is measured as the smallest
 3D distance between an atom of the ligand placed at $q$ and the 3D position of $\qrand$ ($\da(q,\qrand)$ on line~\ref{alg::expand:a} in
@@
 By computing the distance between the nearest atoms, the shape of the ligand is actually considered.
-As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction
-of the ligand towards $\qrand$.
+This metric supports retraction of the ligand towards $\qrand$.
+%As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
 

### 2017-06-30 paper.tex
jdu spat.

@@
 The accessibility (Fig.~\ref{fig:properties}a) shows that more than the half of the tunnel is not accessible (red part of the tunnel).
-The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversible.
+The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversable.
 
@@
 Let $\cstart$ represent the average starting position of all trajectories and let $d_{max}$ be the 3D Euclidean distance
-of the most distant configuration from $\cstart$ amongst all trajectories.
+of the most distant configuration from $\cstart$ among all trajectories.
 A set of $M$ spheres centered at $\cstart$ are created with radii $r'_i=i {d_{max} \over M}$, where $i=0,\ldots, M-1$.
@@
 The average runtimes, success rates and sizes of the built trees for the first tunnel are shown in Tab.~\ref{tab::rrt}.
-The highest average success rate wasv achieved for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
+The highest average success rate was achieved for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
 The difference between the minimal and maximal success rates indicates how the initial configurations of the ligand influence the ability to pass the tunnels.
@@
 examined in each expansion.
+The average values of the runtimes, which are in the order of minutes, are negligible in the comparison of runtime of MD simulations.
 
@@
 The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
-The successful trajectories for $\smin=0.4$ reveals that the end of the tunnel No. 3 can be approached by two different pathways.
-The detail of this pathway is depicted in Fig.~\ref{fig::detail}.
-Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative path way. 
-This shows the strength of the analysis based on the trajectories and not only on the geometric bottlenecks.
+The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
+The detail is depicted in Fig.~\ref{fig::detail}.
+Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.
 
@@
 \centering
+{
+\renewcommand{\tabcolsep}{0pt}
 \begin{tabular}{ccc}
-\includegraphics[width=0.32\textwidth]{fig/tunne4ac}& 
-\includegraphics[width=0.32\textwidth]{fig/tunne4bc} & 
-\includegraphics[width=0.32\textwidth]{fig/tunne4cc} \\
+\includegraphics[width=0.325\textwidth]{fig/tunne4ac}& 
+\includegraphics[width=0.325\textwidth]{fig/tunne4bc} & 
+\includegraphics[width=0.325\textwidth]{fig/tunne4cc} \\
 (a) & (b) & (c)                        
 \end{tabular}
+}
 \caption{\label{fig::detail}
@@
 The second limitation is caused by the utilization of scaled-down ligands.
-The protein atoms tend to fill the internal void space therefore the tunnels identified inside proteins without ligand tend
+The protein atoms fill the internal void space therefore the tunnels identified inside proteins without ligand tend
 to be very narrow.
@@
 than the simple bottleneck radius, which is used nowadays.
-For example, possible detours from the tunnel can be identified, which suggest that even a tunnel with a small
+For example, possible detours from the tunnel can be identified, which indicates that even a tunnel with a small
 bottleneck can be traversed (Fig.~\ref{fig::detail}).
@@
 
-Protein tunnels are transporting pathways for ligands towards active sites.
+The protein tunnels are transporting pathways for ligands to the  active sites.
 In this paper, we propose to analyze the traversability in the tunnels using motion planning, that can consider both shape
